### 2022-08-05 description.tex
First draft

\section{Description}

While statistical alignment is typically performed using pairwise hidden Markov
models, finite-state transducers (FSTs) \green{present themselves as a better
alternative}.
% too brief?
FSTs \green{also} have the ability to rigorously model molecular sequence
evolution in addition to well established methods for combining them in
different ways \parencite{bradley2007transducers}.
A powerful and versatile algorithm is composition, which consists of sending the
output of one FST as the input of a second FST.
The FST model implemented in COATi is designed by composing smaller FSTs, each
representing a specific process.

% Evolution FST
Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
\ref{fig:evolution-fst}), based on existing transducers (e.g.
\cite{holmes2001evolutionary}).
The Evolution FST is formed by composing a substitution FST that encodes a 64x64
codon model and an indel FST that models insertions and deletions, including
frameshifts.
The innovation of the Evolution FST with respect to other transducers is the
combination of a codon substitution model with gaps that can occur at any
position of any length.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). Absorption and
    emission of symbols occurs between states. (a) The substitution FST
    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
    The indel FST allows for insertions (I to U) and deletions (D to V).
    Contiguous insertions and deletions are always arranged for insertions to
    precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

% FST implementation
The alignment FST is the result from composing both sequences with the Evolution
FST.
Any path through the alignment FST represents a pairwise alignment, while the
shortest path corresponds to the best alignment.
% A path through the FST that \green{results} from composing both sequences with
% the Evolution FST represents a pairwise alignment.
% The shortest path through the FST corresponds with the best alignment.
However, composing large FSTs is expensive and can be prohibitive \green{(specify
I talk about time)}.
Despite the existence of efficient C++ FST libraries (e.g. openFST
\cite{allauzen2007openfst}), runtime is still limiting for sequence longer than
a few thousand nucleotides.
\green{To solve this issue, the search for an optimal path (alignment)} can be
reformulated as a dynamic programming problem.

### 2022-08-05 description2.tex
First draft

--- 2022-08-05 description.tex
@@
 
-While statistical alignment is typically performed using pairwise hidden Markov
-models, finite-state transducers (FSTs) \green{present themselves as a better
-alternative}.
-% too brief?
-FSTs \green{also} have the ability to rigorously model molecular sequence
-evolution in addition to well established methods for combining them in
-different ways \parencite{bradley2007transducers}.
+% Summarize pair-HMMs. Brief, comment typically used but move onto FST right away
+Statistical alignment is typically performed using pairwise hidden Markov
+models (pair-HMMs).
+% This computational machines consist of two output tapes and a set of states that
+% emit symbols onto one or both tapes.
+Pair-HMMs have the ability to rigorously model molecular sequence evolution and
+can calculate the probability that two sequences are related, represented
+$P(X, Y)$ \parencite{yoon_2009_hmm}.
+However, a limitation of pair-HMMs is the ability to only model evolution of two
+related sequences from an unknown ancestor.
+Finite-state transducers (FSTs) have the ability to calculate the probability
+that a sequence $Y$ evolved from sequence $X$, represented $P(Y | X)$.
+FSTs share similar computational benefits as pair-HMMs in addition to well
+established algorithms for combining them in different ways
+\parencite{bradley2007transducers}.
 A powerful and versatile algorithm is composition, which consists of sending the
-output of one FST as the input of a second FST.
+output one FST as the input of a second FST.
 The FST model implemented in COATi is designed by composing smaller FSTs, each
@@
 \cite{holmes2001evolutionary}).
-The Evolution FST is formed by composing a substitution FST that encodes a 64x64
-codon model and an indel FST that models insertions and deletions, including
-frameshifts.
+This FST is formed by composing a substitution FST that encodes a 64x64 codon
+model and an indel FST that models insertions and deletions, including frameshifts.
 The innovation of the Evolution FST with respect to other transducers is the
@@
 % FST implementation
+% A path through the FST that \green{results} from composing both sequences with
+% the Evolution FST represent a pairwise alignment.
+% To find the most \green{probable/likely} alignment we must find the shortest
+% path through.
 The alignment FST is the result from composing both sequences with the Evolution
@@
 shortest path corresponds to the best alignment.
-% A path through the FST that \green{results} from composing both sequences with
-% the Evolution FST represents a pairwise alignment.
-% The shortest path through the FST corresponds with the best alignment.
-However, composing large FSTs is expensive and can be prohibitive \green{(specify
-I talk about time)}.
+However, composing large FSTs is an expensive operation and can be prohibitive.
 Despite the existence of efficient C++ FST libraries (e.g. openFST
-\cite{allauzen2007openfst}), runtime is still limiting for sequence longer than
-a few thousand nucleotides.
-\green{To solve this issue, the search for an optimal path (alignment)} can be
+\cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
+than a few thousand nucleotides each.
+To solve this issue, the search for an optimal path (alignment) can be
 reformulated as a dynamic programming problem.
+% Maintaining (or approximate??) the statistical framework, COATi implements a
+% Gotoh-inspired algorithm thus reducing the problem to O(nm) runtime, n and m 
+% length of sequences \green{citation?}
+
+% Alignment itself?
+
+% Evolution FST + Q matrix
+% Marginal model? - Probably not
+% DP - mention in a sentence or two.
+% 1 figure: FST model? - results table?

### 2022-08-05 figures/fig-evolution-fst.tex
First draft

% !TEX program=lualatex
\RequirePackage{luatex85}
\documentclass[tikz,border=10,10pt]{standalone}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{unicode-math}
\usepackage{fontspec}
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{xcolor}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning}
\tikzset{
  font={\fontsize{9pt}{11}\selectfont}
}

\setmainfont{TeX Gyre Heros}
\setmathfont{Latin Modern Math}

\definecolor{colorR}{RGB}{228,26,28}    % RED
\definecolor{colorB}{RGB}{55,126,184}   % BLUE
\definecolor{colorG}{RGB}{77,175,74}    % GREEN
\definecolor{colorP}{RGB}{152,78,163}   % PURPLE
\definecolor{colorO}{RGB}{255,127,0}    % ORANGE
\definecolor{colorY}{RGB}{255,255,51}   % YELLOW
\definecolor{colorBn}{RGB}{166,86,40}   % BROWN
\definecolor{colorPk}{RGB}{247,129,191} % PINK
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

\tikzstyle{line}=[draw, -stealth', very thick]
\tikzstyle{block}=[circle,fill=colorB!50,on grid]
\tikzstyle{lab}=[]
\tikzstyle{w}=[lab,midway]
\tikzstyle{e}=[lab,midway,auto=false,fill=colorP!50,font=\scriptsize]

\begin{document}
% \begin{tikzpicture}[node distance=25mm, auto]
\begin{tikzpicture}[node distance=25mm, auto,
	dottedline/.style = {ultra thick, loosely dotted,shorten >=1mm, shorten <=1mm}]
%%% MG94 marginal substitution FST
\node (titlea) {\textbf{a) Substitution}};

\node[block,below=2cm of titlea,fill=colorG!50] (start_sub) {start};
\node[block,right=15mm of start_sub] (S_sub) {S};
\node[block,right=60mm of S_sub] (M_sub) {M};
\node[block,fill=colorR!50,right=15mm of M_sub] (end_sub) {end};

\draw[line] (start_sub) -- (S_sub);
\draw[line] (S_sub) to[bend right=22.5] (M_sub);
% \draw[line] (S_sub) -- (M_sub);
\draw[line] (M_sub) -- (end_sub);

% \draw[line] (M_sub) to[bend right=85] node[e,pos=0.5] {X\textrightarrow{}Y} (S_sub);
% \draw[line] (M_sub) to[bend right=55] node[e,pos=0.5] {AAA1\textrightarrow{}A} (S_sub);
\draw[line] (M_sub) to[bend right=28.5] (S_sub);
\draw[line] (M_sub) to[bend right=16.5] (S_sub);
\draw[line] (M_sub) to[bend right=25.5] (S_sub);
\draw[line] (M_sub) to[bend right=19.5] (S_sub);
\draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5] {XXX\textrightarrow{}YYY} (S_sub);
% \draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5] {AAA1\textrightarrow{}C} (S_sub);
% \draw[line] (M_sub) to[bend right=8] node[e,pos=0.5] {AAA1\textrightarrow{}G} (S_sub);

% \draw[line] (M_sub) to[bend right=-8] node[e,pos=0.5] {AAA1\textrightarrow{}T} (S_sub);
% \draw[line] (M_sub) to[bend right=-22.5] node[e,pos=0.5] {AAA2\textrightarrow{}A} (S_sub);
% \draw[line] (M_sub) to[bend right=-65] node[e,pos=0.5] {TTT3\textrightarrow{}T} (S_sub);

% \draw[dottedline] (5,-3) -- node[below] {} + (0, -9mm);

%%% Legend

\node[below=2cm of start_sub,text width=50mm,anchor=north west] (sequences) {
\begin{tabular}{r@{\,: }l}
\multicolumn{2}{l}{\textbf{Sequences}}\\
    \hline
	X & input nucleotides\\
	Y & intermediate nucleotides\\
	Z & output nucleotides\\
	\O & nothing/empty sequence\\
\end{tabular}
};

\node[above right=-1.4cm and 5mm of sequences,text width=50mm] {
\begin{tabular}{r@{\,: }l}
\multicolumn{2}{l}{\textbf{Parameters}}\\
\hline
	$g$ & gap open weight\\
	$e$ & gap extension weight\\
\end{tabular}
};

%%% Indel FST
\node[right=10cm of titlea] (titleb) {\textbf{b) Insertion-Deletion}};

\node[block,right=12cm of start_sub,fill=colorG!50] (start1) {start};
\node[block,below=15mm of start1] (S1) {S};
\node[block,below=of S1] (I1) {I};
\node[block,right=of I1] (U1) {U};
\node[block,above=of U1] (W1) {W};
\node[block,above=20mm of W1] (D1) {D};
\node[block,right=of D1] (V1) {V};
\node[block,right=of W1] (M1) {M};
\node[block,fill=colorR!50,right=15mm of M1] (end1) {end};

\draw (D1) +(0mm,-20mm) coordinate (C1);
\draw (D1) +(50mm,-10mm) coordinate (C2);

\draw[line] (start1) -- (S1);
\draw[line] (S1) -- node[w] {$g$} (I1);
\draw[line] (S1) -- node[w] {$1-g$} (W1);
\draw[line] (U1) to[out=150,in=30]  node[w,above] {$e$} (I1);
\draw[line] (U1) -- node[w,pos=0.1,right] {$1-e$} (W1);
\draw[line] (W1) -- node[w] {$g$} (D1);
\draw[line] (W1) -- node[w] {$1-g$} (M1);
\draw[line] (V1) to[out=150,in=30] node[w,above] {$e$} (D1);
\draw[line] (D1) to[out=60,in=90] (end1 |- D1.south) to (end1);
\draw[line] (V1) -- node[w,swap] {$1-e$} (M1);
\draw[line] (M1) --  (end1);

\draw[line] (I1) to[bend right=20] (U1);
\draw[line] (I1) to[bend right=25] (U1);
\draw[line] (I1) to[bend right=35] (U1);
\draw[line] (I1) to[bend right=40] (U1);
\draw[line] (I1) to[bend right=30] node[e] {\O\textrightarrow{}Z} (U1);

\draw[line] (D1) to[bend right=20] (V1);
\draw[line] (D1) to[bend right=25] (V1);
\draw[line] (D1) to[bend right=35] (V1);
\draw[line] (D1) to[bend right=40] (V1);
\draw[line] (D1) to[bend right=30] node[e] {Y\textrightarrow{}\O} (V1);

\draw[line] (M1) to[bend left=39] (S1);
\draw[line] (M1) to[bend left=42] (S1);
\draw[line] (M1) to[bend left=48] (S1);
\draw[line] (M1) to[bend left=51] (S1);

\draw[line] (M1) to[bend left=45] node[e,pos=0.25] {YYY\textrightarrow{}ZZZ} (S1);

% \node[below left=7.5mm and 1mm of I1.east,text width=50mm,anchor=north west] {
% \begin{tabular}{r@{\,: }l}
% \multicolumn{2}{l}{}\\
% %\hline
% 	X & input nucleotides\\
% 	Y & intermediate nucleotides\\
% 	Z & output nucleotides\\
% 	\O & nothing/empty sequence\\
%     \hline
% 	$g$ & gap open weight\\
% 	$e$ & gap extension weight\\
% \end{tabular}
% };

\end{tikzpicture}
\end{document}

### 2022-08-05 introduction.tex
First draft

\section{Introduction}

% sequence alignment is important and heavily used
Sequence alignment is a fundamental task in bioinformatics and a cornerstone
step in comparative and functional genomic studies
\parencite{sequence_alignment_rosenberg_2009}.
% often seen as an ad hoc problem
Modern sequence analysis began with the heuristic homology algorithms of
Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and has
progressed to arrive at current aligners.
However, the alignment of molecular sequences is, in practice, often seen as a
tool and the alignment inference as an ad hoc problem
\parencite{morrison_MSA_2018}.

% despite available aligners, there is room for improvement
% summarize sentence into ~ "alignment is performed in the amino acid space"
A common strategy is to perform alignment inference in the amino acid space
\parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information,
underperforms compared to alignment at the codon level, and fails in the
presence of artifacts.
% to correct this common practice is to discard lots of data
Given that errors and artifacts are common in molecular data, this requires
costly curation practices that discard large amounts of information.
% errors in genomic data that lead to erroneous downstream analyses
When uncorrected, errors in genomic datasets can lead to erroneous results in
functional and comparative genomic studies \parencite{estimates_schneider_2009}.

% to address this, we present COATi
Here we present COATi, short for COdon-Aware Alignment Transducer, a statistical
pairwise aligner that incorporates codon substitution models and is
robust to artifacts present in genomic data.

### 2022-08-05 pairwise_appnote.tex
First draft

\include{header.tex}

% document begins here
\begin{document}
% \vspace*{0.35in}

% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: }}
% codon-aware pairwise aligner + artifacts
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Reed A. Cartwright\textsuperscript{1,3}
\\
\bigskip
1 Biodesign Institute, Arizona State University
\\
2 Ira A. Schools of Engineering, Arizona State University
\\
3 School of Life Sciences, Arizona State University
\\
\bigskip
* corresponding@author.mail

\end{flushleft}

% \abstract{Abstract}
\begin{abstract}
\noindent \textbf{Summary:} COATi is a statistical codon-aware pairwise aligner
that supports complex insertion-deletion models and is able to handle artifacts
present in genomic data.\\  % abstract summary can be longer.
\textbf{Availability:} The source code, along with documentation, for COATi is
freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
implemented in C++.\\
\textbf{Supplementary information:} %\green{TODO}.
\end{abstract}

% now start line numbers
\linenumbers

\input{introduction}

\input{description2}

\input{results}

\section{Conclusions/Discussion}

COATi is an FST-based application that can find the optimal alignment between a
pair of sequences in the presence of artifacts using a statistical model.
% \green{(continue)}

\section*{Acknowledgments}

The author(s) would like to thank \green{person1}, \green{person2}, and
\green{number} reviewers for their helpful suggestions.

\subsection*{Funding}
This research was founded by an NSF-IIBR grant (\green{grant number}).\\

\noindent \textit{Conflict of interest:} none declared.

% \section{References}
% \printbibliography

\nolinenumbers

\end{document}

### 2022-08-05 results.tex
First draft

\section{Results}
% prelim data comparison?

Using 4000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011}, MACSE v2.06
\parencite{ranwez_macse_2011}, MAFFT v7.407 \parencite{mafft_katoh_2002}, and
PRANK v.170427 \parencite{prank_loytynoja_2014} together with COATi.

After downloading, 1660 alignments contained gaps identified by at least one
aligner.
We randomly introduced gap patterns extracted from all five methods into the
2340 initially ungapped sequence pairs to generate the true alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics.
% by estimating $k_s$ and $k_a$ statistics.

% % Software comparison table
% \begin{table}[!ht]
% \centering
%     %\frame{
%     \input{figures/table-comp}%}
% 	\caption{Accuracy of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE,
%             on 2340 simulated sequence pairs. Perfect alignments have
%             ($d_{seq}=0$), best alignments have lowest $d_{seq}$, and imperfect
%             alignments have $d_{seq}>0$ when at least one aligner found a
%             perfect alignment. Best values are highlighted in blue.}
% 	\label{table:comp}
% \end{table}

COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments,
and had a higher positive (97.3\%) selection accuracy.
MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
average alignment error an order of magnitude larger than COATi and the accuracy
retrieving positive selection of 81.5\% and 85.8\% respectively.
PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
obtained similar results with an average alignment error two orders of magnitude
larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
The accuracy retrieving negative selection was similar across all five
aligners.

% (Table \ref{table:comp}).

### 2022-08-08 description.tex
Work on description, add d_seq to results

@@
 The Evolution FST is formed by composing a substitution FST that encodes a 64x64
-codon model and an indel FST that models insertions and deletions, including
-frameshifts.
+codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models
+insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
 The innovation of the Evolution FST with respect to other transducers is the
-combination of a codon substitution model with gaps that can occur at any
-position of any length.
+combination of a codon substitution model that allows stop codons with gaps that
+can occur at any position of any length.
 
@@
     \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
-    \caption{The evolution FST is assembled by composing a substitution FST and
+    \caption{The Evolution FST is assembled by composing a substitution FST and
     an indel FST. Each node represents a state in an FST while arcs display
@@
 a few thousand nucleotides.
-\green{To solve this issue, the search for an optimal path (alignment)} can be
+\green{To solve this issue, the search for an optimal path (alignment)} is
 reformulated as a dynamic programming problem.

### 2022-08-08 description2.tex
Work on description, add d_seq to results

@@
 \cite{holmes2001evolutionary}).
-This FST is formed by composing a substitution FST that encodes a 64x64 codon
-model and an indel FST that models insertions and deletions, including frameshifts.
+The Evolution FST is formed by composing a substitution FST that encodes a 64x64
+codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models
+insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
 The innovation of the Evolution FST with respect to other transducers is the
-combination of a codon substitution model with gaps that can occur at any
-position of any length.
+combination of a codon substitution model that allows stop codons with gaps that
+can occur at any position of any length.
 
@@
 shortest path corresponds to the best alignment.
-However, composing large FSTs is an expensive operation and can be prohibitive.
-Despite the existence of efficient C++ FST libraries (e.g. openFST
-\cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
-than a few thousand nucleotides each.
-To solve this issue, the search for an optimal path (alignment) can be
-reformulated as a dynamic programming problem.
-% Maintaining (or approximate??) the statistical framework, COATi implements a
-% Gotoh-inspired algorithm thus reducing the problem to O(nm) runtime, n and m 
-% length of sequences \green{citation?}
+All FST operations including creation of models, composition, search for the
+shortest path, and other optimization algorithms are performed using the C++
+openFST library \parencite{allauzen2007openfst}.
 
-% Alignment itself?
+% However, composing large FSTs is an expensive operation and can be prohibitive.
+% Despite the existence of efficient C++ FST libraries (e.g. openFST
+% \cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
+% than a few thousand nucleotides each.
+% To solve this issue, the search for an optimal path (alignment) is
+% reformulated as a dynamic programming problem.
 
@@
 % Marginal model? - Probably not
-% DP - mention in a sentence or two.
+% DP - probably not
 % 1 figure: FST model? - results table?

### 2022-08-08 pairwise_appnote.tex
Work on description, add d_seq to results

@@
 present in genomic data.\\  % abstract summary can be longer.
-\textbf{Availability:} The source code, along with documentation, for COATi is
+\textbf{Availability:} The source code for COATi, along with documentation, is
 freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
@@
 
-The author(s) would like to thank \green{person1}, \green{person2}, and
-\green{number} reviewers for their helpful suggestions.
+% The authors would like to thank \green{person1}, \green{person2}, and
+% \green{number} reviewers for their helpful suggestions.
 

### 2022-08-08 results.tex
Work on description, add d_seq to results

@@
 
-COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
-p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
+COATi was significantly more accurate, $d_{seq}$ value of $6$x$10^{-4}$  than
+other aligners; all p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
 one-tailed Wilcoxon signed rank test.
 In addition, COATi produced more perfect alignments, less imperfect alignments,
-and had a higher positive (97.3\%) selection accuracy.
+and had a higher positive selection accuracy (97.3\%).
 MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
-average alignment error an order of magnitude larger than COATi and the accuracy
-retrieving positive selection of 81.5\% and 85.8\% respectively.
+average alignment error an order of magnitude larger than COATi ($6$x$10^{-3}$)
+and the accuracy retrieving positive selection of 81.5\% and 85.8\% respectively.
 PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
 obtained similar results with an average alignment error two orders of magnitude
-larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
+larger than COATi ($0.01$) and a 87.3\% and 69.1\% accuracy retrieving positive
+selection.
 The accuracy retrieving negative selection was similar across all five
-aligners.
+aligners (98.6\% $\pm$ 1.25\%).
 

### 2022-08-15 description2.tex
Fix typo
Add lab comments

@@
 $P(X, Y)$ \parencite{yoon_2009_hmm}.
-However, a limitation of pair-HMMs is the ability to only model evolution of two
-related sequences from an unknown ancestor.
+However, a limitation of pair-HMMs is the ability to only model the evolution of
+two related sequences from an unknown ancestor.
 Finite-state transducers (FSTs) have the ability to calculate the probability
 that a sequence $Y$ evolved from sequence $X$, represented $P(Y | X)$.
-FSTs share similar computational benefits as pair-HMMs in addition to well
-established algorithms for combining them in different ways
+FSTs have similar benefits to pair-HMMs in addition to
+well-established algorithms for combining FSTs in different ways
 \parencite{bradley2007transducers}.
 A powerful and versatile algorithm is composition, which consists of sending the
-output one FST as the input of a second FST.
+output of one FST into the input of a second FST.
 The FST model implemented in COATi is designed by composing smaller FSTs, each
@@
     an indel FST. Each node represents a state in an FST while arcs display
-    possible transitions between states (and their weights). Absorption and
-    emission of symbols occurs between states. (a) The substitution FST
-    encodes a 64x64 codon substitution model with 64 arcs from M to S. (b)
-    The indel FST allows for insertions (I to U) and deletions (D to V).
-    Contiguous insertions and deletions are always arranged for insertions to
-    precede deletions to limit equivalent alignments.}
+    possible transitions between states (and their weights). (a) The
+    substitution FST encodes a 64x64 codon substitution model with 64 arcs from
+    M to S. (b) The indel FST allows for insertions (I to U) and deletions
+    (D to V). Contiguous insertions and deletions are always arranged for
+    insertions to precede deletions to limit equivalent alignments.}
     \label{fig:evolution-fst}
@@
 shortest path corresponds to the best alignment.
-All FST operations including creation of models, composition, search for the
+All FST operations including model development, composition, search for the
 shortest path, and other optimization algorithms are performed using the C++

### 2022-08-15 introduction.tex
Add lab comments

@@
 Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and has
-progressed to arrive at current aligners.
+progressed to arrive at current aligners such as Clustal$\Omega$
+\parencite{clustal_omega_sievers_2011}, MAFFT \parencite{mafft_katoh_2002}, or
+PRANK \parencite{prank_loytynoja_2014}.
 However, the alignment of molecular sequences is, in practice, often seen as a
@@
 underperforms compared to alignment at the codon level, and fails in the
-presence of artifacts.
+presence of artifacts such as frameshifts and early stop codons.
 % to correct this common practice is to discard lots of data
@@
 % errors in genomic data that lead to erroneous downstream analyses
-When uncorrected, errors in genomic datasets can lead to erroneous results in
+Errors in genomic datasets can lead to erroneous results in
 functional and comparative genomic studies \parencite{estimates_schneider_2009}.

### 2022-08-15 pairwise_appnote.tex
Fix typo
Add title
Add lab comments

@@
 \begin{flushleft}
-{\Large\textbf{COATi: }}
+{\Large\textbf{COATi: statistical pairwise aligner robust to artifacts}}
 % codon-aware pairwise aligner + artifacts
@@
 
-\section{Conclusions/Discussion}
+\section{Discussion}
 
-COATi is an FST-based application that can find the optimal alignment between a
-pair of sequences in the presence of artifacts using a statistical model.
-% \green{(continue)}
+COATi is an FST-based application that can calculate the optimal alignment
+between a pair of sequences in the presence of artifacts using a statistical
+model.
+It will allow researchers to analyze more data with higher accuracy and
+facilitate the study of important biological processes that shape genomic data.
 
@@
 \subsection*{Funding}
-This research was founded by an NSF-IIBR grant (\green{grant number}).\\
+This research was funded by an NSF-IIBR grant (\green{grant number}).\\
 
@@
 % \section{References}
-% \printbibliography
+\printbibliography
 

### 2022-08-15 results.tex
Add lab comments

--- 2022-08-05 results.tex
@@
 We used the data set to evaluate the accuracy of popular cutting edge aligners
-Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011}, MACSE v2.06
-\parencite{ranwez_macse_2011}, MAFFT v7.407 \parencite{mafft_katoh_2002}, and
-PRANK v.170427 \parencite{prank_loytynoja_2014} together with COATi.
+Clustal$\Omega$ v1.2.4, %\parencite{clustal_omega_sievers_2011},
+MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407, and
+%\parencite{mafft_katoh_2002}
+PRANK v.170427 % \parencite{prank_loytynoja_2014}
+together with COATi.
 
@@
 
-COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
-p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
-one-tailed Wilcoxon signed rank test.
-In addition, COATi produced more perfect alignments, less imperfect alignments,
-and had a higher positive (97.3\%) selection accuracy.
-MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
-average alignment error an order of magnitude larger than COATi and the accuracy
-retrieving positive selection of 81.5\% and 85.8\% respectively.
-PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
-obtained similar results with an average alignment error two orders of magnitude
-larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
-The accuracy retrieving negative selection was similar across all five
-aligners.
+COATi was significantly (p < $1.714$x$10^{-8}$) more accurate at inferring
+simulated alignments compared to other aligners according to the one-tailed
+Wilcoxon signed ranked test, with an average alignment error ($d_{seq}$) value
+of $6$x$10^{-4}$.
+In addition, COATi produced more perfect alignments ($d_{seq}=0$), less imperfect
+alignments, and more accurately retrieved events of positive selection (97.3\%).
+Compared to COATi, MACSE (allows frameshifts) and MAFFT (using DNA) had an
+average alignment error an order of magnitude larger than COATi ($6$x$10^{-3}$)
+and a lower accuracy retrieving events of positive selection at a rate of 81.5\%
+and 85.8\% respectively.
+PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) had
+an average alignment error two orders of magnitude larger than COATi ($0.01$)
+and a 87.3\% and 69.1\% accuracy retrieving events of positive selection,
+respectively.
+The accuracy retrieving events of negative selection was similar across all five
+aligners (98.6\% $\pm$ 1.25\%).
 

### 2022-08-15 introduction.tex
Reorder introduction content

@@
 presence of artifacts such as frameshifts and early stop codons.
+% errors in genomic data that lead to erroneous downstream analyses
+Errors in genomic datasets can lead to erroneous results in
+functional and comparative genomic studies \parencite{estimates_schneider_2009}.
 % to correct this common practice is to discard lots of data
@@
 costly curation practices that discard large amounts of information.
-% errors in genomic data that lead to erroneous downstream analyses
-Errors in genomic datasets can lead to erroneous results in
-functional and comparative genomic studies \parencite{estimates_schneider_2009}.
-
 % to address this, we present COATi
-Here we present COATi, short for COdon-Aware Alignment Transducer, a statistical
-pairwise aligner that incorporates codon substitution models and is
-robust to artifacts present in genomic data.
+To address this, we present COATi, short for COdon-Aware Alignment Transducer,
+a statistical pairwise aligner that incorporates codon substitution models and
+is robust to artifacts present in genomic data.

### 2022-11-15 description.tex
Add marginal, update results

@@
 
-While statistical alignment is typically performed using pairwise hidden Markov
-models, finite-state transducers (FSTs) \green{present themselves as a better
-alternative}.
-% too brief?
-FSTs \green{also} have the ability to rigorously model molecular sequence
-evolution in addition to well established methods for combining them in
-different ways \parencite{bradley2007transducers}.
-A powerful and versatile algorithm is composition, which consists of sending the
-output of one FST as the input of a second FST.
+% Summarize pair-HMMs. Brief, comment typically used but move onto FST right away
+Statistical alignment is typically performed using pairwise hidden Markov models
+(pair-HMMs).
+Pair-HMMs are computational machines that have the ability to rigorously model
+molecular sequence evolution and can calculate the probability that two
+sequences are related, represented $P(X, Y)$ \parencite{yoon_2009_hmm}.
+However, a limitation of pair-HMMs is the ability to only model the evolution
+of two related sequences from an unknown ancestor.
+
+Finite-state transducers (FSTs) have the same benefits as pair-HMMS with the
+additional ability to generate a descendant sequence given an ancestral one.
+Properly weighted, an FST can calculate the probability that a sequence $Y$
+(descendant) evolved from sequence $X$ (ancestor), represented $P(Y | X)$.
+Furthermore, the existence of well-established algorithms for combining FSTs in
+different ways \parencite{bradley2007transducers} allows the design of complex
+models by combining simpler FSTs.
+A powerful and versatile algorithm for comparative sequence analysis is
+composition, which consists of sending the output of one FST into the input of a
+second FST.
 The FST model implemented in COATi is designed by composing smaller FSTs, each
@@
 
-% Evolution FST
 Pairwise alignment in COATi is implemented via the Evolution FST (Fig.
@@
 insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
+The substitution models available are Muse and Gaut \parencite{muse_gaut_1994}
+(MG94) and the Empirical Codon Model \parencite{kosiol_ECM_2007} (ECM).
 The innovation of the Evolution FST with respect to other transducers is the
@@
 shortest path corresponds to the best alignment.
-% A path through the FST that \green{results} from composing both sequences with
-% the Evolution FST represents a pairwise alignment.
-% The shortest path through the FST corresponds with the best alignment.
-However, composing large FSTs is expensive and can be prohibitive \green{(specify
-I talk about time)}.
-Despite the existence of efficient C++ FST libraries (e.g. openFST
-\cite{allauzen2007openfst}), runtime is still limiting for sequence longer than
-a few thousand nucleotides.
-\green{To solve this issue, the search for an optimal path (alignment)} is
-reformulated as a dynamic programming problem.
+All FST operations including model development, composition, search for the
+shortest path, and other optimization algorithms are performed using the C++
+openFST library \parencite{allauzen2007openfst}.
+
+% However, composing large FSTs is an expensive operation and can be prohibitive.
+% Despite the existence of efficient C++ FST libraries (e.g. openFST
+% \cite{allauzen2007openfst}), runtime is still limiting for sequence pairs longer
+% than a few thousand nucleotides each.
+% To solve this issue, the search for an optimal path (alignment) is
+% reformulated as a dynamic programming problem.
+
+COATi also features a marginal substitution model with substitution probability
+matrix
+
+\[P'_{ijp} = \mathlarger{\mathlarger{\sum}}_{cod} \thickspace \begin{cases}
+P(i \thinspace | \thinspace cod) \enspace \text{if} \enspace cod_p = j\\
+0 \enspace \text{otherwise}
+\end{cases} \]
+
+Where $P'_{ijp}$ represents the probability that codon $i$ from the ancestor
+sequence changes to nucleotide $j$ of the descendant sequence at position $p$
+$\in \{0, 1, 2\}$ of the reading frame.
+$P$ is the substitution probability matrix MG94 or ECM.
+This model emphasizes the position where the substitution in a codon occurs and
+helps restrict the effects of low quality data in the descendant sequence.
+In combination with the indel FST alignment with the marginal model can be
+implemented using dynamic programming, which results in a significant speed up
+with similar accuracy.
+
+% Evolution FST + Q matrix
+% Marginal model? - Probably not
+% DP - probably not
+% 1 figure: FST model? - results table?

### 2022-11-15 introduction.tex
Add marginal, update results

--- 2022-08-15 introduction.tex
@@
 % sequence alignment is important and heavily used
-Sequence alignment is a fundamental task in bioinformatics and a cornerstone
-step in comparative and functional genomic studies
+Sequence alignment is a cornerstone step in bioinformatics
 \parencite{sequence_alignment_rosenberg_2009}.
-% often seen as an ad hoc problem
-Modern sequence analysis began with the heuristic homology algorithms of
-Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and has
-progressed to arrive at current aligners such as Clustal$\Omega$
-\parencite{clustal_omega_sievers_2011}, MAFFT \parencite{mafft_katoh_2002}, or
-PRANK \parencite{prank_loytynoja_2014}.
-However, the alignment of molecular sequences is, in practice, often seen as a
-tool and the alignment inference as an ad hoc problem
-\parencite{morrison_MSA_2018}.
-
+% errors in genomic data that lead to erroneous downstream analyses
+Uncorrected errors in sequence alignment can lead to erroneous results in
+functional and comparative genomic studies \parencite{estimates_schneider_2009}.
+% to correct this common practice is to discard lots of data
+Given that errors and artifacts are common in molecular data, this requires
+costly curation practices that discard large amounts of information.
 % despite available aligners, there is room for improvement
 % summarize sentence into ~ "alignment is performed in the amino acid space"
-A common strategy is to perform alignment inference in the amino acid space
-\parencite{bininda2005transalign,abascal2010translatorx}.
+In addition, a common strategy is to perform alignment inference in the amino
+acid space \parencite{bininda2005transalign,abascal2010translatorx}.
 While this approach is an improvement over DNA models, it discards information,
@@
 presence of artifacts such as frameshifts and early stop codons.
-% to correct this common practice is to discard lots of data
-Given that errors and artifacts are common in molecular data, this requires
-costly curation practices that discard large amounts of information.
-% errors in genomic data that lead to erroneous downstream analyses
-Errors in genomic datasets can lead to erroneous results in
-functional and comparative genomic studies \parencite{estimates_schneider_2009}.
+Although some aligners incorporate codon substitution models, they do not
+support frameshifts or lack as statistical model.
 
 % to address this, we present COATi
-Here we present COATi, short for COdon-Aware Alignment Transducer, a statistical
-pairwise aligner that incorporates codon substitution models and is
-robust to artifacts present in genomic data.
+To address this problem, we present COATi, short for COdon-Aware Alignment
+Transducer, a statistical pairwise aligner that incorporates codon substitution
+models and is robust to artifacts present in genomic data.

### 2022-11-15 methods.tex
Add marginal, update results

\section{Methods}

Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011},
MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407
\parencite{mafft_katoh_2002}, and PRANK v.170427 \parencite{prank_loytynoja_2014}
together with COATi fst and marginal models.

After downloading, sequences longer than 6000 nucleotides were filtered out
(2232), and 8369 alignments contained gaps identified by at least one
aligner.
We randomly introduced gap patterns extracted from all five methods into the
5399 initially ungapped sequence pairs to generate the true alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\parencite{ka_ks_li_1993}.

### 2022-11-15 results.tex
Add marginal, update results

--- 2022-08-05 results.tex
@@
 
-Using 4000 human genes and their gorilla homologous pairs from the ENSEMBL
-database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
-alignments with empirical gap patterns.
-We used the data set to evaluate the accuracy of popular cutting edge aligners
-Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011}, MACSE v2.06
-\parencite{ranwez_macse_2011}, MAFFT v7.407 \parencite{mafft_katoh_2002}, and
-PRANK v.170427 \parencite{prank_loytynoja_2014} together with COATi.
-
-After downloading, 1660 alignments contained gaps identified by at least one
-aligner.
-We randomly introduced gap patterns extracted from all five methods into the
-2340 initially ungapped sequence pairs to generate the true alignments.
-Alignment accuracy was measured using the distance metric $d_{seq}$
-\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
-alignments.
-In addition, accuracy of positive and negative selection was calculated
-using the $F_1$ score by estimating $k_s$ and $k_a$ statistics.
 % by estimating $k_s$ and $k_a$ statistics.
@@
 
-COATi was significantly more accurate (lower $d_{seq}$) than other aligners; all
-p-values were equal (MAFFT) or less than $1.714$x$10^{-8}$ according to the
-one-tailed Wilcoxon signed rank test.
-In addition, COATi produced more perfect alignments, less imperfect alignments,
-and had a higher positive (97.3\%) selection accuracy.
-MACSE (allows frameshifts) and MAFFT (using DNA) performed similarly with an
-average alignment error an order of magnitude larger than COATi and the accuracy
-retrieving positive selection of 81.5\% and 85.8\% respectively.
-PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) also
-obtained similar results with an average alignment error two orders of magnitude
-larger than COATi and a 87.3\% and 69.1\% accuracy retrieving positive selection.
-The accuracy retrieving negative selection was similar across all five
-aligners.
+Both models in COATi were significantly (p < $2.2$x$10^{-16}$) more accurate at
+inferring simulated alignments compared to other aligners according to the
+one-tailed Wilcoxon signed ranked test, with an average alignment error
+($d_{seq}$) value of $1$x$10^{-3}$.
+In addition, COATi produced more perfect alignments ($d_{seq}=0$), less imperfect
+alignments, and more accurately retrieved events of positive selection
+(fst model 91.9\%, marginal model 90.8\%) (Supplementary Table 1).
+Compared to COATi, MACSE (allows frameshifts) and MAFFT (using DNA) had an
+average alignment error an order of magnitude larger than COATi and a lower
+accuracy retrieving events of positive selection at a rate of 81.5\%
+and 85.8\% respectively.
+PRANK (using codons) and Clustal$\Omega$ (using amino acid translations) had
+an average alignment error two orders of magnitude larger than COATi and a
+87.3\% and 69.1\% accuracy retrieving events of positive selection, respectively.
+The accuracy retrieving events of negative selection was similar across all five
+aligners (97.7\% $\pm$ 1.67\%).
 

### 2022-11-15 suppl.tex
Add marginal, update results

\include{header.tex}

\begin{document}

{\large\textbf{Supplementary material}}

\begin{table}[!ht]
\centering
 \frame{\input{figures/table-comp}}
 \caption{Accuracy of COATi fst, COATi marginal, PRANK, MAFFT, Clustal$\Omega$,
         and MACSE, on 5399 simulated sequence pairs. Perfect alignments have
         ($d_{seq}=0$), best alignments have lowest $d_{seq}$, and imperfect
         alignments have $d_{seq}>0$ when at least one aligner found a
         perfect alignment.}
 \label{table:comp}
\end{table}

\end{document}

### 2022-12-07 abstract.tex
Letter format draft

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses including phylogenetic inference, ancestral sequence reconstruction, and gene annotation. Errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomics.
Common artifacts in unpublished reference genomes include abiological frameshifts and early stop codons.
While for model organisms this is eventually fixed, for many species this is not the case and current curation efforts require discarding large amounts of data.
To address this need, we present COATi, a statistical codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
COATi will allow users to reduce the amount of discarded data while generating more accurate sequence alignments.
\end{abstract}

### 2022-12-07 alignpair_letter.tex
Letter format draft

--- 2022-08-15 pairwise_appnote.tex
@@
 \begin{flushleft}
-{\Large\textbf{COATi: statistical pairwise aligner robust to artifacts}}
-% codon-aware pairwise aligner + artifacts
+{\Large\textbf{COATi: statistical pairwise alignment of protein coding sequences}}
 \newline
@@
 Juan J. Garcia Mesa\textsuperscript{1,2},
+Ziqi Zhu\textsuperscript{1,3},
 Reed A. Cartwright\textsuperscript{1,3}
@@
 
-% \abstract{Abstract}
-\begin{abstract}
-\noindent \textbf{Summary:} COATi is a statistical codon-aware pairwise aligner
-that supports complex insertion-deletion models and is able to handle artifacts
-present in genomic data.\\  % abstract summary can be longer.
-\textbf{Availability:} The source code for COATi, along with documentation, is
-freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
-implemented in C++.\\
-\textbf{Supplementary information:} %\green{TODO}.
-\end{abstract}
+\input{abstract}
 
@@
 
-\input{description2}
+\input{materials_methods}
 
-\input{results}
+\input{results_discussion}
 
-\section{Discussion}
+\section*{Availability}
+The source code for COATi, along with documentation, is
+freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
+implemented in C++.
+Code to replicate the analysis can be found on GitHug: \url{https://github.com/jgarciamesa/TODO}.
 
-COATi is an FST-based application that can calculate the optimal alignment
-between a pair of sequences in the presence of artifacts using a statistical
-model.
-It will allow researchers to analyze more data with higher accuracy and
-facilitate the study of important biological processes that shape genomic data.
+% \section*{Supplementary information}
+% \green{TODO}.
 

### 2022-12-07 figures/fig-aln.tex
Letter format draft

% !TEX program=lualatex
\RequirePackage{luatex85}
\documentclass[11pt,tikz]{standalone}

\usepackage{luatextra} % also loads fixltx2e, fontspec, xunicode
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{unicode-math}
\usepackage{xcolor}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,matrix,fit}

\setmainfont{TeX Gyre Heros}
\setmathfont{Latin Modern Math}
\setmonofont{Source Code Pro}

\definecolor{colorR}{RGB}{228,26,28}    % RED
\definecolor{colorB}{RGB}{55,126,184}   % BLUE
\definecolor{colorG}{RGB}{77,175,74}    % GREEN
\definecolor{colorP}{RGB}{152,78,163}   % PURPLE
\definecolor{colorO}{RGB}{255,127,0}    % ORANGE
\definecolor{colorY}{RGB}{255,255,51}   % YELLOW
\definecolor{colorBn}{RGB}{166,86,40}   % BROWN
\definecolor{colorPk}{RGB}{247,129,191} % PINK
\definecolor{colorGy}{RGB}{153,153,153} % GRAY
\definecolor{asublue}{RGB}{0,163,224}

\tikzstyle{aln}=[matrix of nodes, nodes in empty cells]
\tikzstyle{best}=[fill=asublue!20,rounded corners]
\tikzstyle{lab}=[font=\normalsize]
\tikzstyle{indel}=[thick, rounded corners]
\tikzstyle{phase0}=[draw=colorGy]
\tikzstyle{phase1}=[draw=colorP]
\tikzstyle{phase2}=[draw=colorO]

\newcommand{\e}[1]{\textbf{\textcolor{colorR}{#1}}}

\begin{document}
\begin{tikzpicture}[node distance=5mm,font=\footnotesize]

\node (titleBiology) {\textbf{a) Biology}};
% biology
\matrix (biology)[aln,best,below right=-1mm and -15mm of titleBiology] {
   & Ser & His & Lys & Gly & Arg &            &            & Asp & Ala & \\
A: & TCC & CAT & AAG & GGG & CGG & T\verb|--| & \verb|-|CG & GAC & GCC & \verb|---|\\
D: & TCC & CA\verb|-| & \verb|--|G & GGG & CGG & TCC & CAG & GAC & GCC & ACG\\
   & Ser &            &            & Gly & Arg & Ser & Gln & Asp & Ala & Thr\\
};
\node (bio17) [fit=(biology-1-7)(biology-1-8)]{Ser};
\node [fit=(biology-4-3)(biology-4-4)]{Gln};
\draw [indel,phase1] (biology-2-7.north west) rectangle (biology-3-8.south east);
\draw [indel,phase0] (biology-2-11.north west) rectangle (biology-3-11.south east);
\draw [indel,phase2] (biology-2-3.north west) rectangle (biology-3-4.south east);

\node (titlePrank) [below right=2.6cm and -1.8cm of titleBiology] {\textbf{b) Prank (codon)}};
% prank codon --gaprate=0.01
\matrix (prank) [aln,below right=-1mm and -2.5cm of titlePrank] {
   & Ser & His     & Lys & Gly  & Arg & Ser &           & Asp & Ala & \\
A: & TCC & CAT     & AAG & GGG  & CGG & TCG &\verb|---| & GAC & GCC & \verb|---|\\
D: & TCC & CA\e{G} & \verb|---| & GGG & CGG & TC\e{C} & CAG & GAC & GCC & ACG\\
   & Ser & Gln     &            & Gly     & Arg & Ser   & Gln  & Asp & Ala & Thr\\
};
\draw [indel,phase0] (prank-2-8.north west) rectangle (prank-3-8.south east);
\draw [indel,phase0] (prank-2-11.north west) rectangle (prank-3-11.south east);
\draw [indel,phase0] (prank-2-4.north west) rectangle (prank-3-4.south east);

\node (titleMafftClustalo) [below right=2.4cm and -2.75cm of titlePrank] {\textbf{c) MAFFT \& Clustal$\pmb{\Omega}$ \& MACSE}};
% mafft dna & clustalo amino acids
\matrix(mafft-clustalo) [aln,below right=-1mm and -4.75cm of titleMafftClustalo] {
   & Ser & His     & Lys     & Gly     & Arg     & Ser & Asp & Ala & \\
A: & TCC & CAT & AAG & GGG     & CGG     & TCG & GAC & GCC & \verb|---|\\
D: & TCC & CA\e{G}     & \e{GG}G     & \e{C}GG & \e{TCC} & \e{CA}G & GAC & GCC & ACG\\
   & Ser & Gln     & Gly     & Arg     & Ser     & Gln & Asp & Ala & Thr\\
};
\draw [indel,phase0] (mafft-clustalo-2-10.north west) rectangle (mafft-clustalo-3-10.south east);

\node (titleCoati) [below right=2.4cm and -5.05cm of titleMafftClustalo] {\textbf{d) COATi}};
% coati mar-mg
\matrix (coati) [aln,best,below right=-1mm and -1.35cm of titleCoati] {
   & Ser & His & Lys & Gly & Arg &            &            & Asp & Ala & \\
A: & TCC & CAT & AAG & GGG & CGG & T\verb|--| & \verb|-|CG & GAC & GCC & \verb|---|\\
D: & TCC & CA\verb|-| & \verb|--|G & GGG & CGG & TCC & CAG & GAC & GCC & ACG\\
   & Ser &            &            & Gly & Arg & Ser & Gln & Asp & Ala & Thr\\
};
\node (cti17) [fit=(coati-1-7)(coati-1-8)]{Ser};
\node [fit=(coati-4-3)(coati-4-4)]{Gln};
\draw [indel,phase1] (coati-2-7.north west) rectangle (coati-3-8.south east);
\draw [indel,phase0] (coati-2-11.north west) rectangle (coati-3-11.south east);
\draw [indel,phase2] (coati-2-3.north west) rectangle (coati-3-4.south east);

%%%%%

% \draw (a.north west) node[lab,anchor=center] (dlab) {\textbf{a)}};
% \draw (dlab |- c.north west) node[lab,anchor=center] {\textbf{b)}};

\node[] () [below = 0.2em] at (coati.south) {};

\end{tikzpicture}
\end{document}

### 2022-12-07 figures/fig-evolution-fst.tex
Add marginal, update results
Letter format draft

@@
 \draw[line] (W1) -- node[w] {$1-g$} (M1);
+\draw[line] (M1) -- (end1);
 \draw[line] (V1) to[out=150,in=30] node[w,above] {$e$} (D1);
-\draw[line] (D1) to[out=60,in=90] (end1 |- D1.south) to (end1);
+\draw[line] (D1) to[out=60,in=90] node[w] {$1-e$} (end1 |- D1.south) to (end1);
 \draw[line] (V1) -- node[w,swap] {$1-e$} (M1);
-\draw[line] (M1) --  (end1);
 
@@
 
-\draw[line] (M1) to[bend left=45] node[e,pos=0.25] {YYY\textrightarrow{}ZZZ} (S1);
+\draw[line] (M1) to[bend left=45] node[e,pos=0.25] {Y\textrightarrow{}Z} (S1);
 

### 2022-12-07 introduction.tex
Letter format draft

@@
-\section{Introduction}
+\section*{Introduction}
 
-% sequence alignment is important and heavily used
-Sequence alignment is a cornerstone step in bioinformatics
-\parencite{sequence_alignment_rosenberg_2009}.
-% errors in genomic data that lead to erroneous downstream analyses
-Uncorrected errors in sequence alignment can lead to erroneous results in
-functional and comparative genomic studies \parencite{estimates_schneider_2009}.
-% to correct this common practice is to discard lots of data
-Given that errors and artifacts are common in molecular data, this requires
-costly curation practices that discard large amounts of information.
-% despite available aligners, there is room for improvement
-% summarize sentence into ~ "alignment is performed in the amino acid space"
-In addition, a common strategy is to perform alignment inference in the amino
-acid space \parencite{bininda2005transalign,abascal2010translatorx}.
-While this approach is an improvement over DNA models, it discards information,
-underperforms compared to alignment at the codon level, and fails in the
-presence of artifacts such as frameshifts and early stop codons.
-Although some aligners incorporate codon substitution models, they do not
-support frameshifts or lack as statistical model.
+Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \parencite{sequence_alignment_rosenberg_2009}.
+Modern sequence analysis began with the heuristic homology algorithms of Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}.
 
-% to address this, we present COATi
-To address this problem, we present COATi, short for COdon-Aware Alignment
-Transducer, a statistical pairwise aligner that incorporates codon substitution
-models and is robust to artifacts present in genomic data.
+A common strategy is to perform alignment inference in the amino acid space \parencite{bininda2005transalign,abascal2010translatorx}.
+While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
+Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
+In addition, while modeling indels to appear within codons is rare, this is often the case (Zhu \& Cartwright, personal communication, 2019).
+Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).
+
+\begin{figure}[h!]
+% \begin{framed}
+    \begin{minipage}[c]{0.65\textwidth}
+        \includegraphics[scale=1]{fig-aln.pdf}
+    \end{minipage}\hfill
+    \begin{minipage}[c]{0.35\textwidth}
+        \caption{
+        \textbf{Standard algorithms produce suboptimal alignments.}
+        (a) shows a possible alignment of an ancestor (A) sequence and a descendant (D) sequence.
+        (b), (c), and (d) are the results of different aligners.
+        Nucleotide mismatches are highlighted in red, phase zero, one, and two indels are shown in gray, purple, and orange, respectively.
+        Phase one indel (purple) is also of type I (synonymous) while phase two indel (orange) is of type II (nonsynonymous). 
+        COATi is able to retrieve the biological alignment (highlighted in blue).
+        }
+    \label{fig:aln}
+    \end{minipage}
+% \end{framed}
+\end{figure}
+
+Uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \parencite{estimates_schneider_2009}.
+Current methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
+To address this problem, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates codon substitution models and is robust to artifacts present in modern genomic data.

### 2022-12-07 materials_methods.tex
Letter format draft

\section*{Materials and Methods}  % temporary section name

Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \parencite{bradley2007transducers}.
Pair-HMMs are computational machines with two output tapes that contain a finite number of states typically labeled match, insert, and delete that emit symbols (nucleotides or amino acids) to one or both tapes.
Each tape represents a sequence and a path through a pair-HMM is a possible pairwise alignment.
Conceptually, these machines generate two sequences ($X$ and $Y$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(X, Y)$ \parencite{yoon_2009_hmm}.

A limitation of pair-HMMs is the ability to only model the evolution of two related sequences from an unknown ancestor.
Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature to generate a descendant sequence given an ancestral one.
FSTs consume symbols from an input tape and emit symbols to an output tape.
Properly weighted, an FST can calculate the probability that a descendant sequence $Y$ evolved from an ancestor sequence $X$, represented by $P(Y | X)$.
Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \parencite{bradley2007transducers}.
A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
The model implemented in COATi is designed by composing smaller FSTs, each representing a specific process.

Genomes for model organisms are often of high quality after being refined over many iterations and having their coding sequences meticulously curated.
On the contrary, non-model organisms typically have lower-quality genomes that have been only partially curated.
Low-quality genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \parencite{jackman2018tigmint}.
FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.

COATi implements the pairwise alignment of a low-quality sequence (descendant)  against a high-quality sequence (ancestor) as a path through the Evolution FST (Fig. \ref{fig:evolution-fst}), based on existing transducers (e.g. \cite{holmes2001evolutionary}).
This FST is the result of composing a substitution FST that encodes a codon model (Fig. \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig. \ref{fig:evolution-fst}-b).
A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.

Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
Any path through this FST represents a pairwise alignment, while the shortest path corresponds to the best alignment.
All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \parencite{allauzen2007openfst}.

\begin{figure}[h!]
\begin{framed}
\centering
    \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
    \caption{The Evolution FST is assembled by composing a substitution FST and
    an indel FST. Each node represents a state in an FST while arcs display
    possible transitions between states (and their weights). (a) The
    substitution FST encodes a 61x64 codon substitution model with $61\cdot 64$ arcs
    from M to S. (b) The indel FST allows for insertions (I to U) and deletions
    (D to V). Contiguous insertions and deletions are always arranged for
    insertions to precede deletions to limit equivalent alignments.}
    \label{fig:evolution-fst}
\end{framed}
\end{figure}

Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
COATi implements the Muse and Gaut (1994) codon model (codon-triplet-mg) and the Empirical Codon Model \parencite{kosiol_ECM_2007} (codon-triplet-ecm).
It also lets the user provide a codon substitution matrix.
The default FST model (codon-triplet-mg) does not allow substitutions from stop codons, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.

COATi also features a marginal substitution model with probability matrix

\[P'_{ijp} = \mathlarger{\mathlarger{\sum}}_{cod} \thickspace \begin{cases}
P(i \thinspace | \thinspace cod) \enspace \text{if} \enspace cod_p = j\\
0 \enspace \text{otherwise}
\end{cases} \]

Where $P'_{ijp}$ represents the probability that codon $i$ from the ancestor
sequence changes to nucleotide $j$ of the descendant sequence at position $p$
$\in \{0, 1, 2\}$ of the reading frame.
$P$ is the substitution probability matrix Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

### 2022-12-07 results_discussion.tex
Letter format draft

\section*{Results and Discussion}
Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
alignments with empirical gap patterns.
We used the data set to evaluate the accuracy of popular cutting edge aligners
Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011},
MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407
\parencite{mafft_katoh_2002}, and PRANK v.170427 \parencite{prank_loytynoja_2014}
together with COATi.

After downloading, sequences longer than 6000 nucleotides were filtered out
(2232), and 8369 alignments contained gaps identified by at least one
aligner.
We randomly introduced gap patterns extracted from all five methods into the
5399 initially ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\parencite{ka_ks_li_1993}.

% Software comparison table
\begin{table}[!ht]
\centering
 \frame{\input{figures/table-comp}}
 \caption{Accuracy of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE, on 5399
         simulated sequence pairs. Perfect alignments have ($d_{seq}=0$), best
         alignments have lowest $d_{seq}$, and imperfect alignments have
         $d_{seq}>0$ when at least one aligner found a perfect alignment.}
 \label{table:comp}
\end{table}

COATi was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.2\cdot10-16$ according to the one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
It obtained better results compared to a wide variety of alignment strategies.
Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.

Despite human and gorilla sequences having a relatively short evolutionary distance, COATi showed a biologically significant improvement over other methods, with an average alignment error nine-fold smaller than the next best method.
COATi is an FST-based application that can calculate the optimal alignment
between a pair of sequences in the presence of artifacts using a statistical
model.
It will allow researchers to analyze more data with higher accuracy and
facilitate the study of important biological processes that shape genomic data.

### 2023-01-24 abstract.tex
First round of edits.

@@
 \noindent
-Sequence alignment is an essential method in bioinformatics and the basis of many analyses including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
-Errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomics.
-Common artifacts in unpublished reference genomes include abiological frameshifts and early stop codons.
-While for model organisms this is eventually fixed, for many species this is not the case and current curation efforts require discarding large amounts of data.
-To address this need, we present COATi, a statistical codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
+Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
+Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
+For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences annotated in reference genomes.
+While this is eventually fixed in the reference genomes of model organisms, many genomes contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
+To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
 COATi will allow users to reduce the amount of discarded data while generating more accurate sequence alignments.

### 2023-01-24 introduction.tex
First round of edits.

@@
 Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \parencite{sequence_alignment_rosenberg_2009}.
-Modern sequence analysis began with the heuristic homology algorithms of Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}.
-
-A common strategy is to perform alignment inference in the amino acid space \parencite{bininda2005transalign,abascal2010translatorx}.
+While sophisticated advances have been made, the challenge of alignment inference has not been fully solved \parencite{art_morrison_2015}.
+%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}.
+%
+The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \parencite[e.g.][]{bininda2005transalign,abascal2010translatorx}.
 While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
 Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
-In addition, while modeling indels to appear within codons is rare, this is often the case (Zhu \& Cartwright, personal communication, 2019).
+In addition, while modeling indels to appear within codons is rare, this is often the case (\TODO{cite Ziqi's dissertation here and the mouse genome paper}).
 Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).

### 2023-02-04 introduction.tex
Additional comments and changes by Reed

@@
 
+% Figures are 86 mm or 178 mm wide
 \begin{figure}[h!]
-% \begin{framed}
-    \begin{minipage}[c]{0.65\textwidth}
+    \centering%
     \includegraphics[scale=1]{fig-aln.pdf}
-    \end{minipage}\hfill
-    \begin{minipage}[c]{0.35\textwidth}
+    \par
     \caption{
-        \textbf{Standard algorithms produce suboptimal alignments.}
-        (a) shows a possible alignment of an ancestor (A) sequence and a descendant (D) sequence.
+        Standard algorithms produce suboptimal alignments.
+        (a) shows a possible alignment of an ancestor sequence (A) and a descendant sequence (D).
         (b), (c), and (d) are the results of different aligners.
-        Nucleotide mismatches are highlighted in red, phase zero, one, and two indels are shown in gray, purple, and orange, respectively.
-        Phase one indel (purple) is also of type I (synonymous) while phase two indel (orange) is of type II (nonsynonymous). 
-        COATi is able to retrieve the biological alignment (highlighted in blue).
+        Nucleotide mismatches are highlighted in red. Phase 0, phase 1, and phase 2 indels are shown in gray, purple, and orange, respectively.
+        Additionally, the orange indel is type II (an amino-acid indel plus an amino-acid change) while the purple indel is type I (an amino-acid indel only).
+        COATi is the only aligner able to retrieve the biological alignment in this example.
         }
     \label{fig:aln}
-    \end{minipage}
-% \end{framed}
 \end{figure}

### 2023-02-04 materials_methods.tex
Additional comments and changes by Reed

@@
 
-Genomes for model organisms are often of high quality after being refined over many iterations and having their coding sequences meticulously curated.
-On the contrary, non-model organisms typically have lower-quality genomes that have been only partially curated.
-Low-quality genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \parencite{jackman2018tigmint}.
+Genome quality impacts conclusions drawn from comparative genomic studies. \TODO{Juan, add citations.}
+Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
+In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
+These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \parencite{jackman2018tigmint}.
 FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.
 
-COATi implements the pairwise alignment of a low-quality sequence (descendant)  against a high-quality sequence (ancestor) as a path through the Evolution FST (Fig.
+COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig.
 \ref{fig:evolution-fst}), based on existing transducers (e.g.
 \cite{holmes2001evolutionary}).
+Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
 This FST is the result of composing a substitution FST that encodes a codon model (Fig.
@@
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \parencite{allauzen2007openfst}.
+However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
+This additional state space increase the computational complexity of the alignment algorithm.
 
 \begin{figure}[h!]
-\begin{framed}
 \centering
 \includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
-    \caption{The Evolution FST is assembled by composing a substitution FST and
-    an indel FST. Each node represents a state in an FST while arcs display
-    possible transitions between states (and their weights). (a) The
-    substitution FST encodes a 61x64 codon substitution model with $61\cdot 64$ arcs
-    from M to S. (b) The indel FST allows for insertions (I to U) and deletions
-    (D to V). Contiguous insertions and deletions are always arranged for
-    insertions to precede deletions to limit equivalent alignments.}
+\par
+\caption{The Evolution FST is assembled by composing a substitution FST and an indel FST.
+Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
+(a) The substitution FST encodes a $61 \times 64 $ codon substitution model with 3904 arcs from M to S.
+These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
+The weight of each arc is a conditional probability derived from a codon substitution model.
+(b) The indel FST allows for insertions (I to U) and deletions (D to V).
+Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
+Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
 \label{fig:evolution-fst}
-\end{framed}
 \end{figure}
@@
 It also lets the user provide a codon substitution matrix.
-The default FST model (codon-triplet-mg) does not allow substitutions from stop codons, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
+The default FST model (codon-triplet-mg) does not allow substitutions from stop codons \TODO{Juan is this still correct?
+Do we need to mention the extra error rate.}, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
 
-COATi also features a marginal substitution model with probability matrix
+To reduce the runtime complexity of COATi, we have also developed an approximation of the Evolution FST that can be implemented with standard dynamic programming techniques.
+This approximation uses a marginal substitution model where the output nucleotides are independent of one another and only depend on the input codon and position.
+This produces a $\left(61 \times 3 \right) \times 4$ substitution model and eliminates the need to track dependencies between output nucleotides.
 
-\[P'_{ijp} = \mathlarger{\mathlarger{\sum}}_{cod} \thickspace \begin{cases}
-P(i \thinspace | \thinspace cod) \enspace \text{if} \enspace cod_p = j\\
-0 \enspace \text{otherwise}
-\end{cases} \]
-
-Where $P'_{ijp}$ represents the probability that codon $i$ from the ancestor
-sequence changes to nucleotide $j$ of the descendant sequence at position $p$
-$\in \{0, 1, 2\}$ of the reading frame.
-$P$ is the substitution probability matrix Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
+A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame positions.
+%
+Specifically, let
+$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
+%
+\[
+P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
+\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
+\]
+%
+represent the marginal transition probabilities, where
+$p \in \{0, 1, 2\}$ is the position of the descendent nucleotide relative to the ancestral reading frame.
+COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
 These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.

### 2023-02-04 results_discussion.tex
Additional comments and changes by Reed

@@
 Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
-database \parencite{ensembl_hubbard_2002} we simulated a data set of pairwise
+database \parencite{ensembl_hubbard_2002}, we simulated a data set of pairwise
 alignments with empirical gap patterns.
@@
 
-After downloading, sequences longer than 6000 nucleotides were filtered out
-(2232), and 8369 alignments contained gaps identified by at least one
-aligner.
-We randomly introduced gap patterns extracted from all five methods into the
-5399 initially ungapped sequence pairs to generate the benchmark alignments.
+After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 8369 sequence pairs that contained gaps identified by at least one aligner, and 5399 ungapped sequence pairs.
+We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
 Alignment accuracy was measured using the distance metric $d_{seq}$
@@
 \centering
- \frame{\input{figures/table-comp}}
- \caption{Accuracy of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE, on 5399
-         simulated sequence pairs. Perfect alignments have ($d_{seq}=0$), best
-         alignments have lowest $d_{seq}$, and imperfect alignments have
-         $d_{seq}>0$ when at least one aligner found a perfect alignment.}
+\input{figures/table-comp}
+ \caption{COATi generates better alignments than other alignment algorithms.
+Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 5399 empirically simulated sequence pairs.
+Perfect alignments have $d_{seq}=0$, best alignments have the lowest $d_{seq}$, and imperfect alignments have $d_{seq}>0$ when at least one aligner found a perfect alignment.}
  \label{table:comp}
@@
 
-COATi was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.2\cdot10-16$ according to the one-tailed Wilcoxon signed rank test.
+COATi was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.2 \cdot 10^{-16}$ according to the one-tailed Wilcoxon signed rank test.
 In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).

### 2023-02-04 alignpair_letter.tex
Add testing github repo
Additional comments and changes by Reed

@@
 Ziqi Zhu\textsuperscript{1,3},
-Reed A. Cartwright\textsuperscript{1,3}
+Reed A. Cartwright\textsuperscript{1,3,*}
 \\
@@
 \bigskip
-* corresponding@author.mail
+* cartwright@asu.edu
 
@@
 implemented in C++.
-Code to replicate the analysis can be found on GitHug: \url{https://github.com/jgarciamesa/TODO}.
+Code to replicate the analysis can be found on GitHug: \url{https://github.com/jgarciamesa/coati-testing}.
 
@@
 
+\TODO{This}
+
 % The authors would like to thank \green{person1}, \green{person2}, and
@@
 \subsection*{Funding}
-This research was funded by an NSF-IIBR grant (\green{grant number}).\\
+This research was funded by NSF award DBI-1929850.\\
 
 \noindent \textit{Conflict of interest:} none declared.
+
+\TODO{Juan, check all references for consistency.}
 

### 2023-02-08 alignpair_letter.tex
Change references to MBE required style

@@
 implemented in C++.
-Code to replicate the analysis can be found on GitHug: \url{https://github.com/jgarciamesa/coati-testing}.
+Code to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing}.
 
 % \section*{Supplementary information}
-% \green{TODO}.
 
@@
 
-\TODO{This}
-
-% The authors would like to thank \green{person1}, \green{person2}, and
-% \green{number} reviewers for their helpful suggestions.
-
-\subsection*{Funding}
 This research was funded by NSF award DBI-1929850.\\
@@
 
-\TODO{Juan, check all references for consistency.}
-
 % \section{References}
-\printbibliography
+\bibliographystyle{mbe}%
+\setlength{\bibhang}{0pt}
+\bibliography{alignpair_letter.bib}
 

### 2023-02-08 introduction.tex
Change references to MBE required style

@@
 
-Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \parencite{sequence_alignment_rosenberg_2009}.
-While sophisticated advances have been made, the challenge of alignment inference has not been fully solved \parencite{art_morrison_2015}.
-%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}.
+Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citembe{sequence_alignment_rosenberg_2009}.
+While sophisticated advances have been made, the challenge of alignment inference has not been fully solved \citembe{art_morrison_2015}.
+%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}.
 %
-The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \parencite[e.g.][]{bininda2005transalign,abascal2010translatorx}.
+The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space (e.g.
+\citetwo{bininda2005transalign}{abascal2010translatorx}).
 While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
 Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
-In addition, while modeling indels to appear within codons is rare, this is often the case (\TODO{cite Ziqi's dissertation here and the mouse genome paper}).
+In addition, while modeling indels to appear within codons is rare, this is often the case \citembetwo{zhu2022profiling}{taylor2004occurrence}.
 Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).
@@
 
-Uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \parencite{estimates_schneider_2009}.
+Uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembe{estimates_schneider_2009}.
 Current methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.

### 2023-02-08 materials_methods.tex
Change references to MBE required style

@@
 
-Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \parencite{bradley2007transducers}.
+Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citembe{bradley2007transducers}.
 Pair-HMMs are computational machines with two output tapes that contain a finite number of states typically labeled match, insert, and delete that emit symbols (nucleotides or amino acids) to one or both tapes.
 Each tape represents a sequence and a path through a pair-HMM is a possible pairwise alignment.
-Conceptually, these machines generate two sequences ($X$ and $Y$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(X, Y)$ \parencite{yoon_2009_hmm}.
+Conceptually, these machines generate two sequences ($X$ and $Y$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(X, Y)$ \citembe{yoon_2009_hmm}.
 
@@
 Properly weighted, an FST can calculate the probability that a descendant sequence $Y$ evolved from an ancestor sequence $X$, represented by $P(Y | X)$.
-Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \parencite{bradley2007transducers}.
+Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
 A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
@@
 In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
-These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \parencite{jackman2018tigmint}.
+These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citembe{jackman2018tigmint}.
 FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.
@@
 Any path through this FST represents a pairwise alignment, while the shortest path corresponds to the best alignment.
-All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \parencite{allauzen2007openfst}.
+All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citembe{allauzen2007openfst}.
 However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
@@
 Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
-COATi implements the Muse and Gaut (1994) codon model (codon-triplet-mg) and the Empirical Codon Model \parencite{kosiol_ECM_2007} (codon-triplet-ecm).
+COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citembe{kosiol_ECM_2007} (codon-triplet-ecm).
 It also lets the user provide a codon substitution matrix.

### 2023-02-08 results_discussion.tex
Change references to MBE required style

@@
 Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
-database \parencite{ensembl_hubbard_2002}, we simulated a data set of pairwise
+database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
 alignments with empirical gap patterns.
 We used the data set to evaluate the accuracy of popular cutting edge aligners
-Clustal$\Omega$ v1.2.4 \parencite{clustal_omega_sievers_2011},
-MACSE v2.06 \parencite{ranwez_macse_2011}, MAFFT v7.407
-\parencite{mafft_katoh_2002}, and PRANK v.170427 \parencite{prank_loytynoja_2014}
+Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
+MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.407
+\citembe{mafft_katoh_2002}, and PRANK v.170427 \citembe{prank_loytynoja_2014}
 together with COATi.
@@
 Alignment accuracy was measured using the distance metric $d_{seq}$
-\parencite{metrics_blackburne_whelan_2011} between simulated and inferred
+\citembe{metrics_blackburne_whelan_2011} between simulated and inferred
 alignments.
@@
 using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-\parencite{ka_ks_li_1993}.
+\citembe{ka_ks_li_1993}.
 

### 2023-02-08 materials_methods.tex
Add missing citations

@@
 
-Genome quality impacts conclusions drawn from comparative genomic studies. \TODO{Juan, add citations.}
+Genome quality impacts conclusions drawn from comparative genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
 Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.

### 2023-02-08 materials_methods.tex
Explain indicator func, early stop codons

@@
 It also lets the user provide a codon substitution matrix.
-The default FST model (codon-triplet-mg) does not allow substitutions from stop codons \TODO{Juan is this still correct?
-Do we need to mention the extra error rate.}, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
+The default FST model (codon-triplet-mg) does not allow early stop codons on the ancestor sequence, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
 
@@
 represent the marginal transition probabilities, where
-$p \in \{0, 1, 2\}$ is the position of the descendent nucleotide relative to the ancestral reading frame.
+$p \in \{0, 1, 2\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function defined
+%
+$I(y) := \{ 1 \text{ if } y = Y_p\text{, } 0 \text{ otherwise}\}$.
+%
 COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.

### 2023-03-10 abstract.tex
Make changes and add notes.

@@
 Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
-For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences annotated in reference genomes.
-While this is eventually fixed in the reference genomes of model organisms, many genomes contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
+For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
+While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
 To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
-COATi will allow users to reduce the amount of discarded data while generating more accurate sequence alignments.
+COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments.
 \end{abstract}

### 2023-03-10 alignpair_letter.tex
Make changes and add notes.

@@
 \bigskip
-1 Biodesign Institute, Arizona State University
+1 The Biodesign Institute, Arizona State University, Tempe, Arizona, USA
 \\
-2 Ira A. Schools of Engineering, Arizona State University
+2 Ira A.\ Fulton Schools of Engineering, Arizona State University, Tempe, Arizona, USA
 \\
-3 School of Life Sciences, Arizona State University
+3 School of Life Sciences, Arizona State University, Tempe, Arizona, USA
 \\

### 2023-03-10 introduction.tex
Add missing citations
Make changes and add notes.

@@
 Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citembe{sequence_alignment_rosenberg_2009}.
-While sophisticated advances have been made, the challenge of alignment inference has not been fully solved \citembe{art_morrison_2015}.
+While sophisticated advancements have been made, the challenge of alignment inference has not been fully solved \citembe{art_morrison_2015}.
 %Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}.
@@
 Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
-In addition, while modeling indels to appear within codons is rare, this is often the case \citembetwo{zhu2022profiling}{taylor2004occurrence}.
-Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).
+In addition, existing aligners typically force gaps to occur between codons, whereas in natural sequences, only about 40\% of indels occur between codons \citembetwo{taylor2004occurrence}{zhu2022profiling}.
+This mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).
 
@@
 
-Uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembe{estimates_schneider_2009}.
-Current methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
+Genome quality impacts conclusions drawn from comparative genomic studies, and uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
+Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
+In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
+These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citembe{jackman2018tigmint}.
+%
+When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
+current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
 To address this problem, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates codon substitution models and is robust to artifacts present in modern genomic data.

### 2023-03-10 materials_methods.tex
Make changes and add notes.

@@
 Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citembe{bradley2007transducers}.
-Pair-HMMs are computational machines with two output tapes that contain a finite number of states typically labeled match, insert, and delete that emit symbols (nucleotides or amino acids) to one or both tapes.
+Pair-HMMs are computational machines with two output tapes.
+Pair-HMMs contain a finite number of states---typically labeled match, insert, and delete---that emit symbols (nucleotides or amino acids) to one or both tapes.
 Each tape represents a sequence and a path through a pair-HMM is a possible pairwise alignment.
-Conceptually, these machines generate two sequences ($X$ and $Y$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(X, Y)$ \citembe{yoon_2009_hmm}.
+Conceptually, these machines generate two sequences ($\text{X}$ and $\text{Y}$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(\text{X}, \text{Y})$ \citembe{yoon_2009_hmm}.
 
-A limitation of pair-HMMs is the ability to only model the evolution of two related sequences from an unknown ancestor.
-Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature to generate a descendant sequence given an ancestral one.
+A limitation of pair-HMMs is that they only model the evolution of two related sequences from an unknown ancestor.
+Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that can can model the generation a descendant sequence given an ancestral one.
 FSTs consume symbols from an input tape and emit symbols to an output tape.
-Properly weighted, an FST can calculate the probability that a descendant sequence $Y$ evolved from an ancestor sequence $X$, represented by $P(Y | X)$.
+Properly weighted, an FST can calculate the probability that a descendant sequence, $\text{Y}$, evolved from an ancestral sequence, $\text{X}$, represented by $P(\text{Y} | \text{X})$.
 Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
 A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
-The model implemented in COATi is designed by composing smaller FSTs, each representing a specific process.
+COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
 
-Genome quality impacts conclusions drawn from comparative genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
-Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
-In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
-These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citembe{jackman2018tigmint}.
-FSTs and their powerful methods provide a well-suited framework to statistically align a sequence from a non-model organism against a sequence from a model organism.
-
-COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig.
-\ref{fig:evolution-fst}), based on existing transducers (e.g.
-\cite{holmes2001evolutionary}).
+COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig.\ \ref{fig:evolution-fst}) \citep[c.f.][]{holmes2001evolutionary}.
 Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
-This FST is the result of composing a substitution FST that encodes a codon model (Fig.
-\ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.
-\ref{fig:evolution-fst}-b).
+This FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
 A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.
@@
 Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
-Any path through this FST represents a pairwise alignment, while the shortest path corresponds to the best alignment.
+Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment.
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citembe{allauzen2007openfst}.
 However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
-This additional state space increase the computational complexity of the alignment algorithm.
+This additional state space increases the computational complexity of the alignment algorithm.
 
@@
 It also lets the user provide a codon substitution matrix.
-The default FST model (codon-triplet-mg) does not allow early stop codons on the ancestor sequence, although it supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
+The default FST model (codon-triplet-mg) does not allow early stop codons in the ancestor sequence; although, it does supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
 
@@
 
-A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame positions.
+A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position.
 %
@@
 %
-$I(y) := \{ 1 \text{ if } y = Y_p\text{, } 0 \text{ otherwise}\}$.
+$I(e) = \{ 1 \text{ if $e$ is true and } 0 \text{ otherwise}\}$.
 %
 COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
-These models emphasize the position where the substitution in a codon occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
+These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
 In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.

### 2023-03-10 results_discussion.tex
Make changes and add notes.

@@
 \section*{Results and Discussion}
-Using 16000 human genes and their gorilla homologous pairs from the ENSEMBL
+Using 16000 human genes and their gorilla orthologs from the ENSEMBL
 database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
 alignments with empirical gap patterns.
-We used the data set to evaluate the accuracy of popular cutting edge aligners
+We used the data set to evaluate the accuracy of popular aligners
 Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
@@
 \citembe{mafft_katoh_2002}, and PRANK v.170427 \citembe{prank_loytynoja_2014}
-together with COATi.
+together with COATi. \TODO{which version of coati are we using. Should we show both? Or show both in a supplement?}
 
@@
 using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-\citembe{ka_ks_li_1993}.
+\citembe{ka_ks_li_1993}. \TODO{briefly explain F1 score}
 
@@
 facilitate the study of important biological processes that shape genomic data.
+
+\TODO{Add one short paragraph about future work. E.g. including a three-mer based indel model, and weighting indel phases differently.}

### 2023-04-28 abstract.tex
Update results and fix typos

@@
 For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
-While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
+While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain these artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
 To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.

### 2023-04-28 materials_methods.tex
Update results and fix typos

@@
-\section*{Materials and Methods}  % temporary section name
+\section*{Materials and Methods}
 
@@
 A limitation of pair-HMMs is that they only model the evolution of two related sequences from an unknown ancestor.
-Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that can can model the generation a descendant sequence given an ancestral one.
+Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that can model the generation of a descendant sequence given an ancestral one.
 FSTs consume symbols from an input tape and emit symbols to an output tape.
@@
 Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
-(a) The substitution FST encodes a $61 \times 64 $ codon substitution model with 3904 arcs from M to S.
+(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from M to S.
 These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
@@
 Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
+On top of the indel FST we add a base calling error FST (Figure S1) to model sequencing errors.
 Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
@@
 It also lets the user provide a codon substitution matrix.
-The default FST model (codon-triplet-mg) does not allow early stop codons in the ancestor sequence; although, it does supports mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
+The default FST model (codon-triplet-mg) does not allow early stop codons in the ancestor sequence; although, it does support mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
 

### 2023-04-28 results_discussion.tex
Update results and fix typos

@@
 Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
-MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.407
-\citembe{mafft_katoh_2002}, and PRANK v.170427 \citembe{prank_loytynoja_2014}
-together with COATi. \TODO{which version of coati are we using. Should we show both? Or show both in a supplement?}
+MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
+\citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}
+together with COATi.
 
@@
 using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-\citembe{ka_ks_li_1993}. \TODO{briefly explain F1 score}
+\citembe{ka_ks_li_1993}.
+The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
 
@@
 
-COATi was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.2 \cdot 10^{-16}$ according to the one-tailed Wilcoxon signed rank test.
+COATi, using the codon-triplet-ecm model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $4.0 \cdot 10^{-06}$ according to the one-tailed Wilcoxon signed rank test.
 In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
@@
 
-\TODO{Add one short paragraph about future work. E.g. including a three-mer based indel model, and weighting indel phases differently.}
+Future works include extending the indel FST to combine a three-mer gap model with a frameshift parameter.
+We also plan on comparing the marginal and triplet models to evaluate the implications of the marginalization.
+Additionally, it would be valuable to weigh indel phases differently.

### 2023-04-28 figures/fig-base-calling-error.tex
Add base calling error figure

--- 2022-08-05 figures/fig-evolution-fst.tex
@@
 \tikzstyle{e}=[lab,midway,auto=false,fill=colorP!50,font=\scriptsize]
+\tikzstyle{p}=[lab,midway,auto=false,fill=colorO!50,font=\scriptsize]
 
 \begin{document}
-% \begin{tikzpicture}[node distance=25mm, auto]
 \begin{tikzpicture}[node distance=25mm, auto,
 	dottedline/.style = {ultra thick, loosely dotted,shorten >=1mm, shorten <=1mm}]
-%%% MG94 marginal substitution FST
-\node (titlea) {\textbf{a) Substitution}};
+\node (titlea) {\textbf{Base Calling Error FST}};
 
@@
 \node[block,right=15mm of start_sub] (S_sub) {S};
+\node[block,above right=10mm and 30mm of S_sub] (ij_sub) {};
+\node[block,right=30mm of S_sub] (ii_sub) {};
+\node[block,below right=10mm and 30mm of S_sub] (in_sub) {};
 \node[block,right=60mm of S_sub] (M_sub) {M};
@@
 \draw[line] (start_sub) -- (S_sub);
-\draw[line] (S_sub) to[bend right=22.5] (M_sub);
-% \draw[line] (S_sub) -- (M_sub);
 \draw[line] (M_sub) -- (end_sub);
 
-% \draw[line] (M_sub) to[bend right=85] node[e,pos=0.5] {X\textrightarrow{}Y} (S_sub);
-% \draw[line] (M_sub) to[bend right=55] node[e,pos=0.5] {AAA1\textrightarrow{}A} (S_sub);
-\draw[line] (M_sub) to[bend right=28.5] (S_sub);
-\draw[line] (M_sub) to[bend right=16.5] (S_sub);
-\draw[line] (M_sub) to[bend right=25.5] (S_sub);
-\draw[line] (M_sub) to[bend right=19.5] (S_sub);
-\draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5] {XXX\textrightarrow{}YYY} (S_sub);
-% \draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5] {AAA1\textrightarrow{}C} (S_sub);
-% \draw[line] (M_sub) to[bend right=8] node[e,pos=0.5] {AAA1\textrightarrow{}G} (S_sub);
+\draw[line] (M_sub) to[bend right=20] node[e] {i\textrightarrow{}j} (ij_sub);
+\draw[line] (ij_sub) to[bend right=20] node[p] {$\varepsilon$} (S_sub);
+\draw[line] (M_sub) to node[e] {i\textrightarrow{}i} (ii_sub);
+\draw[line] (ii_sub) to node[p] {$1 - 3 \cdot \varepsilon$} (S_sub);
+\draw[line] (M_sub) to[bend right=-20] node[e] {i\textrightarrow{}N} (in_sub);
+\draw[line] (in_sub) to[bend right=-20] (S_sub);
 
-% \draw[line] (M_sub) to[bend right=-8] node[e,pos=0.5] {AAA1\textrightarrow{}T} (S_sub);
-% \draw[line] (M_sub) to[bend right=-22.5] node[e,pos=0.5] {AAA2\textrightarrow{}A} (S_sub);
-% \draw[line] (M_sub) to[bend right=-65] node[e,pos=0.5] {TTT3\textrightarrow{}T} (S_sub);
-
-% \draw[dottedline] (5,-3) -- node[below] {} + (0, -9mm);
+\draw[line] (S_sub) to[bend left=-60] (M_sub);
 
@@
     \hline
-	X & input nucleotides\\
-	Y & intermediate nucleotides\\
-	Z & output nucleotides\\
-	\O & nothing/empty sequence\\
+	i & input nucleotides\\
+	j & intermediate nucleotides\\
+	N & input ambiguous nucleotide N\\
 \end{tabular}
@@
 
-\node[above right=-1.4cm and 5mm of sequences,text width=50mm] {
+\node[above right=-1.03cm and 5mm of sequences,text width=50mm] {
 \begin{tabular}{r@{\,: }l}
@@
 \hline
-	$g$ & gap open weight\\
-	$e$ & gap extension weight\\
+	$\varepsilon$ & base calling error weight\\
 \end{tabular}
@@
 
-%%% Indel FST
-\node[right=10cm of titlea] (titleb) {\textbf{b) Insertion-Deletion}};
-
-\node[block,right=12cm of start_sub,fill=colorG!50] (start1) {start};
-\node[block,below=15mm of start1] (S1) {S};
-\node[block,below=of S1] (I1) {I};
-\node[block,right=of I1] (U1) {U};
-\node[block,above=of U1] (W1) {W};
-\node[block,above=20mm of W1] (D1) {D};
-\node[block,right=of D1] (V1) {V};
-\node[block,right=of W1] (M1) {M};
-\node[block,fill=colorR!50,right=15mm of M1] (end1) {end};
-
-\draw (D1) +(0mm,-20mm) coordinate (C1);
-\draw (D1) +(50mm,-10mm) coordinate (C2);
-
-\draw[line] (start1) -- (S1);
-\draw[line] (S1) -- node[w] {$g$} (I1);
-\draw[line] (S1) -- node[w] {$1-g$} (W1);
-\draw[line] (U1) to[out=150,in=30]  node[w,above] {$e$} (I1);
-\draw[line] (U1) -- node[w,pos=0.1,right] {$1-e$} (W1);
-\draw[line] (W1) -- node[w] {$g$} (D1);
-\draw[line] (W1) -- node[w] {$1-g$} (M1);
-\draw[line] (V1) to[out=150,in=30] node[w,above] {$e$} (D1);
-\draw[line] (D1) to[out=60,in=90] (end1 |- D1.south) to (end1);
-\draw[line] (V1) -- node[w,swap] {$1-e$} (M1);
-\draw[line] (M1) --  (end1);
-
-\draw[line] (I1) to[bend right=20] (U1);
-\draw[line] (I1) to[bend right=25] (U1);
-\draw[line] (I1) to[bend right=35] (U1);
-\draw[line] (I1) to[bend right=40] (U1);
-\draw[line] (I1) to[bend right=30] node[e] {\O\textrightarrow{}Z} (U1);
-
-\draw[line] (D1) to[bend right=20] (V1);
-\draw[line] (D1) to[bend right=25] (V1);
-\draw[line] (D1) to[bend right=35] (V1);
-\draw[line] (D1) to[bend right=40] (V1);
-\draw[line] (D1) to[bend right=30] node[e] {Y\textrightarrow{}\O} (V1);
-
-\draw[line] (M1) to[bend left=39] (S1);
-\draw[line] (M1) to[bend left=42] (S1);
-\draw[line] (M1) to[bend left=48] (S1);
-\draw[line] (M1) to[bend left=51] (S1);
-
-\draw[line] (M1) to[bend left=45] node[e,pos=0.25] {YYY\textrightarrow{}ZZZ} (S1);
-
-% \node[below left=7.5mm and 1mm of I1.east,text width=50mm,anchor=north west] {
-% \begin{tabular}{r@{\,: }l}
-% \multicolumn{2}{l}{}\\
-% %\hline
-% 	X & input nucleotides\\
-% 	Y & intermediate nucleotides\\
-% 	Z & output nucleotides\\
-% 	\O & nothing/empty sequence\\
-%     \hline
-% 	$g$ & gap open weight\\
-% 	$e$ & gap extension weight\\
-% \end{tabular}
-% };
-
 \end{tikzpicture}

### 2023-05-04 figures/fig-base-calling-error.tex
Merge suppl materials

@@
 	dottedline/.style = {ultra thick, loosely dotted,shorten >=1mm, shorten <=1mm}]
-\node (titlea) {\textbf{Base Calling Error FST}};
+%\node (titlea) {\textbf{Base Calling Error FST}};
 
-\node[block,below=2cm of titlea,fill=colorG!50] (start_sub) {start};
+\node[block,fill=colorG!50] (start_sub) {start};
 \node[block,right=15mm of start_sub] (S_sub) {S};
@@
 \draw[line] (M_sub) to[bend right=20] node[e] {i\textrightarrow{}j} (ij_sub);
-\draw[line] (ij_sub) to[bend right=20] node[p] {$\varepsilon$} (S_sub);
+\draw[line] (ij_sub) to[bend right=20] node[w] {$\varepsilon$} (S_sub);
 \draw[line] (M_sub) to node[e] {i\textrightarrow{}i} (ii_sub);
-\draw[line] (ii_sub) to node[p] {$1 - 3 \cdot \varepsilon$} (S_sub);
+\draw[line] (ii_sub) to node[w] {$1 - 3 \cdot \varepsilon$} (S_sub);
 \draw[line] (M_sub) to[bend right=-20] node[e] {i\textrightarrow{}N} (in_sub);
@@
 
-\node[below=2cm of start_sub,text width=50mm,anchor=north west] (sequences) {
+\node[below right=2cm and -1.75cm of start_sub,text width=50mm,anchor=north west] (sequences) {
 \begin{tabular}{r@{\,: }l}
@@
     \hline
-	i & input nucleotides\\
-	j & intermediate nucleotides\\
-	N & input ambiguous nucleotide N\\
+	$i \rightarrow i$ & matching input to intermediate base\\
+	$i \rightarrow j$ & mismatching input to intermediate base\\
+	$i \rightarrow N$ & input to intermediate ambiguous base N\\
 \end{tabular}
@@
 
-\node[above right=-1.03cm and 5mm of sequences,text width=50mm] {
+\node[above right=-1.03cm and 25mm of sequences,text width=50mm] {
 \begin{tabular}{r@{\,: }l}

### 2023-05-04 results_discussion.tex
Merge suppl materials

@@
 
-After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 8369 sequence pairs that contained gaps identified by at least one aligner, and 5399 ungapped sequence pairs.
+After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 6048 sequence pairs that contained gaps identified by at least one aligner, and 7719 ungapped sequence pairs.
 We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
@@
  \caption{COATi generates better alignments than other alignment algorithms.
-Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 5399 empirically simulated sequence pairs.
+Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7719 empirically simulated sequence pairs.
 Perfect alignments have $d_{seq}=0$, best alignments have the lowest $d_{seq}$, and imperfect alignments have $d_{seq}>0$ when at least one aligner found a perfect alignment.}

### 2023-05-04 supplementary_materials.Rmd
Merge suppl materials

---
title: "Supplementary Materials"
#author: "Juan J. García Mesa"
#date: "`r Sys.Date()`"
geometry: margin=1.8cm
output:
    pdf_document:
        fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE}
library(knitr)

summary_table = function(results_file, coati_model) {
    # read results_summary data
    results = read.csv(results_file)

    # extract dseq values for each aligner
    coati = results$dseq[which(results$aligner == coati_model)]
    clustalomega = results$dseq[which(results$aligner == "clustalo")]
    macse = results$dseq[which(results$aligner == "macse")]
    mafft = results$dseq[which(results$aligner == "mafft")]
    prank = results$dseq[which(results$aligner == "prank")]

    dseq_summary = data.frame(coati = mean(coati),
                      prank = mean(prank),
                      mafft = mean(mafft),
                      clustalo = mean(clustalomega),
                      macse = mean(macse), row.names = "$d_{seq}$")

    ############################################################################
    # one-side wilcoxon rank test
    wtp = wilcox.test(coati, prank)
    wtmf = wilcox.test(coati, mafft)
    wtc = wilcox.test(coati, clustalomega)
    wtmc = wilcox.test(coati, macse)
    wilcoxon = data.frame(coati = NA, prank = wtp$p.value, mafft = wtmf$p.value,
                          clustalo = wtc$p.value, macse = wtmc$p.value,
                          row.names = "Wilcoxon rank test")
    stats_table = rbind(dseq_summary, wilcoxon)

    ############################################################################
    # perfect, best, and imperfect alignments
    source("supplementary_materials/scripts/number_alignments.R")
    alns = num_alns(results_file)

    # fix row and column names and add to stats table
    row.names(alns)[which(row.names(alns) == coati_model)] = "coati"
    colnames(alns) = c("Perfect alignments", "Best alignments", "Imperfect alignments")
    stats_table = rbind(stats_table, t(alns))

    ############################################################################
    # selection (ka ks)
    source("supplementary_materials/scripts/kaks.R")

    aligners = c(coati_model, "prank", "mafft", "clustalo", "macse")
    selection = data.frame("F1.pos" = double(),
                           "F1.neg" = double())

    # calculate F1 score for positive and negative selection for each aligner
    for(model in aligners) {
        model_rows = results[which(results$aligner == model), ]
        pos = ps_accuracy(model_rows$ref_omega, model_rows$aln_omega)
        neg = ns_accuracy(model_rows$ref_omega, model_rows$aln_omega)

        selection[nrow(selection) + 1, ] = c(pos, neg)
    }

    # fix row and column names and add to stats table
    colnames(selection) = c("F1-score pos selection", "F1-score neg selection")
    rownames(selection) = c("coati", "prank", "mafft", "clustalo", "macse")
    stats_table = rbind(stats_table, t(selection))
    colnames(stats_table) = c(coati_model, "PRANK", "MAFFT", "ClustalOmega", "MACSE")

    # display stats table
    stats_table
}

```

```{r echo = FALSE, results = 'asis'}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")

results = "supplementary_data/tri-mg/results/results_summary.csv"
print(kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7719 simulated sequence pairs. Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))

results = "supplementary_data/tri-ecm/results/results_summary.csv"
print(kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7678 simulated sequence pairs. Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))

results = "supplementary_data/mar-mg/results/results_summary.csv"
print(kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7661 simulated sequence pairs. Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))

results = "supplementary_data/mar-ecm/results/results_summary.csv"
print(kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7710 simulated sequence pairs. Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))
```

\newpage

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = "\\label{fig:dseq-tri-mg} Comparison of $d_{seq}$ values between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE."}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
plot_dseq_main("supplementary_data/tri-mg/results/results_summary.csv", "tri-mg")
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap =  "\\label{fig:dseq-tri-ecm} Comparison of $d_{seq}$ values between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE."}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
plot_dseq_main("supplementary_data/tri-ecm/results/results_summary.csv", "tri-ecm")
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap =  "\\label{fig:dseq-mar-mg} Comparison of $d_{seq}$ values between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE."}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
plot_dseq_main("supplementary_data/mar-mg/results/results_summary.csv", "mar-mg")
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap =  "\\label{fig:dseq-mar-ecm} Comparison of $d_{seq}$ values between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE."}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
plot_dseq_main("supplementary_data/mar-ecm/results/results_summary.csv", "mar-ecm")
```

\newpage

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.cap="\\label{fig:figs} Base calling error FST. Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
```

### 2023-05-04 supplementary_materials/scripts/kaks.R

@@ -1,0 +1,111 @@
+suppressMessages(library(seqinr))
+library(stringr)
+
+#################### code modified from Reed A Cartwright ######################
+ka = function(a) { if(class(a) != "alignment") {return(NA)}
+	ka = kaks(a)$ka[1]
+	ka
+}
+
+ks = function(a) {
+	if(class(a) != "alignment") {return(NA)}
+	ks = kaks(a)$ks[1]
+	ks
+}
+
+omega = function(a) { #dN/dS
+    if(class(a) != "alignment") {return(NA)}
+	k = kaks(a)
+	ka = k$ka[1]
+	ks = k$ks[1]
+	if(ka < 0 || ks < 0) {
+		return(0)
+	} else if(ka == 0 && ks == 0) {
+		return(0)
+	} else if(ks == 0) {
+		return(10)
+	} else {
+		return(ka/ks)
+	}
+}
+
+ps_accuracy = function(ref,w) {
+	tp = length(intersect(which(w>1),which(ref>1)))
+	fp = length(intersect(which(w>1),which(ref<1)))
+	fn = length(intersect(which(w<1),which(ref>1)))
+	return(2*tp/(2*tp+fp+fn))
+}
+
+ns_accuracy = function(ref,w) {
+	tp = length(intersect(which(w<1),which(ref<1)))
+	fp = length(intersect(which(w<1),which(ref>1)))
+	fn = length(intersect(which(w>1),which(ref<1)))
+	return(2*tp/(2*tp+fp+fn))
+}
+
+process_aln = function(d,fasta) {
+	ret = list()
+
+	for(f in fasta) {
+		aln = try(read.alignment(str_c(d,"/",f),"fasta"),silent=TRUE)
+		if(class(aln) == "try-error") {
+			ret[[f]] = list()
+		} else {
+			seq = aln$seq %>% str_split(pattern="")
+			ung = (seq[[1]] %in% c("a","c","g","t","n"))
+			seq = lapply(seq,subset,ung) # subset = "["
+			aln$seq = lapply(seq,str_c,collapse="")
+			ret[[f]] = aln
+		}
+	}
+	ret
+}
+
+################################################################################
+
+k_metric = function(ARGS) {
+    results = data.frame("ka RMSE" = double(),
+                         "ks RMSE" = double(),
+                         "F1-score pos selection" = double(),
+                         "F1-score neg selection" = double())
+	for(i in ARGS){
+		# read alignments
[…]

### 2023-05-04 supplementary_materials/scripts/number_alignments.R

@@ -1,0 +1,46 @@
+suppressMessages(suppressWarnings(library(dplyr)))
+
+
+num_alns = function(filename) {
+    results_summary = read.csv(file = filename, header = TRUE, stringsAsFactors = FALSE)
+
+    # extract dseq values for all aligners per each reference alignment
+    split_tibble <- function(tibble, col = 'col') tibble %>% split(., .[, col])
+    split_results = split_tibble(results_summary, 'ref_name')
+    dseq = t(sapply(split_results, function(x){ return(x$dseq)}))
+
+    col_s = ncol(dseq)
+    row_s = nrow(dseq)
+
+    results = data.frame("perfect" = integer(col_s),
+                         "best" = integer(col_s),
+                         "imperfect" = integer(col_s))
+
+    # perfect alignments (dseq = 0)
+    if(col_s == 2) {
+        results$perfect = length(which(dseq[,2] == 0))
+    } else{
+        results$perfect = apply(dseq[,],2,function(x){length(which(x == 0))})
+    }
+
+    # best alignments (lower d_seq)
+    for(r in 1:(row_s)) {
+        for(i in which(min(dseq[r, ])==dseq[r, ])){
+            results$best[i] = results$best[i] + 1
+        }
+    }
+
+    # imperfect alignments (dseq != 0 when another methods has dseq = 0)
+    for(r in 1:(row_s)) {
+        if(min(dseq[r, ]) == 0){
+            for(i in which(dseq[r, ]>0)) {
+            results$imperfect[i] = results$imperfect[i] + 1
+            }
+        }
+    }
+
+    rownames(results) = split_results[[1]]$aligner
+    return(results)
+}
+
+

### 2023-05-04 supplementary_materials/scripts/plot_dseq.R

@@ -1,0 +1,67 @@
+library(ggplot2)
+library(ggpubr)
+
+# x-y plot of dseq
+plot_dseq_main = function(results_file, coati_model) {
+    # read results
+    results = read.csv(results_file)
+    
+    # extract dseq values for each aligner
+    dseq = data.frame(COATi = results$dseq[which(results$aligner == coati_model)],
+                      ClustalOmega = results$dseq[which(results$aligner == "clustalo")],
+                      MACSE = results$dseq[which(results$aligner == "macse")],
+                      MAFFT = results$dseq[which(results$aligner == "mafft")],
+                      PRANK = results$dseq[which(results$aligner == "prank")])
+    
+    
+    # Hex plot
+    #ggplot(dseq, aes(x = coati, y = mafft)) +
+        #geom_hex(bins = 50) +
+        #scale_fill_viridis_c() +
+        #geom_abline(slope = 1, intercept = 0, color = "red") +
+        #xlim(-0.01, max(dseq[, c("coati", "mafft")])) +
+        #ylim(-0.01, max(dseq[, c("coati", "mafft")])) +
+        #coord_fixed(ratio = 1/1)
+    
+    # Scatter plot
+    
+    prank = ggplot(dseq, aes(x = COATi, y = PRANK)) +
+        geom_point(alpha = 0.1, color = "#33618B") +
+        geom_abline(slope = 1, intercept = 0, color  = "black") +
+        xlim(0, max(dseq[, c("COATi", "PRANK")])) +
+        ylim(0, max(dseq[, c("COATi", "PRANK")])) +
+        coord_fixed(ratio = 1/1)
+    
+    mafft = ggplot(dseq, aes(x = COATi, y = MAFFT)) +
+        geom_point(alpha = 0.1, color = "#33618B") +
+        geom_abline(slope = 1, intercept = 0, color  = "black") +
+        xlim(0, max(dseq[, c("COATi", "MAFFT")])) +
+        ylim(0, max(dseq[, c("COATi", "MAFFT")])) +
+        coord_fixed(ratio = 1/1)
+    
+    macse = ggplot(dseq, aes(x = COATi, y = MACSE)) +
+        geom_point(alpha = 0.1, color = "#33618B") +
+        geom_abline(slope = 1, intercept = 0, color  = "black") +
+        xlim(0, max(dseq[, c("COATi", "MACSE")])) +
[…]

### 2023-05-09 alignpair_letter.tex
Edit document to improve clarity and accuracy.

@@
-\include{header.tex}
-
+\input{header.tex}
 % document begins here
@@
 \section*{Availability}
-The source code for COATi, along with documentation, is
-freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
-implemented in C++.
-Code to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing}.
+The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++.
+Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing}.
 
@@
 
-This research was funded by NSF award DBI-1929850.\\
+This research was funded by NSF award DBI-1929850.
 

### 2023-05-09 introduction.tex
Edit document to improve clarity and accuracy.

@@
 %
-The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space (e.g.
-\citetwo{bininda2005transalign}{abascal2010translatorx}).
+The alignment of protein-coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space (e.g.\ \citetwo{bininda2005transalign}{abascal2010translatorx}).
 While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
@@
         Standard algorithms produce suboptimal alignments.
-        (a) shows a possible alignment of an ancestor sequence (A) and a descendant sequence (D).
+        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
         (b), (c), and (d) are the results of different aligners.

### 2023-05-09 materials_methods.tex
Merge suppl materials
Edit document to improve clarity and accuracy.

@@
 Pair-HMMs contain a finite number of states---typically labeled match, insert, and delete---that emit symbols (nucleotides or amino acids) to one or both tapes.
-Each tape represents a sequence and a path through a pair-HMM is a possible pairwise alignment.
+Each tape represents a sequence, and a path through a pair-HMM is a possible pairwise alignment.
 Conceptually, these machines generate two sequences ($\text{X}$ and $\text{Y}$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(\text{X}, \text{Y})$ \citembe{yoon_2009_hmm}.
@@
 This FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
-A key innovation of this FST with respect to others is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.
+A key innovation of this FST, with respect to others, is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.
 
@@
 Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
-On top of the indel FST we add a base calling error FST (Figure S1) to model sequencing errors.
+On top of the indel FST, we add a base-calling-error FST (Supplemental Materials Figure 5) to model sequencing errors.
 Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
@@
 %
-COATi contains marginal models for both Muse and Gaut or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
+COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
 These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.

### 2023-05-09 results_discussion.tex
Edit document to improve clarity and accuracy.

@@
 alignments with empirical gap patterns.
-We used the data set to evaluate the accuracy of popular aligners
+We used the data set to evaluate the accuracy of COATi and a suite of popular aligners spanning multiple alignment methods:
 Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
 MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
-\citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}
-together with COATi.
+\citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}.
 
@@
 We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
-Alignment accuracy was measured using the distance metric $d_{seq}$
-\citembe{metrics_blackburne_whelan_2011} between simulated and inferred
+Alignment accuracy was measured using the distance metric, $d_{seq}$
+\citembe{metrics_blackburne_whelan_2011}, between simulated and inferred
 alignments.
@@
 using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-\citembe{ka_ks_li_1993}.
+\citembe{ka_ks_li_1993} \TODO{reference section in supplemental methods about this. How do ks and ka define positive and negative selection?}.
 The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
@@
 
-COATi, using the codon-triplet-ecm model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $4.0 \cdot 10^{-06}$ according to the one-tailed Wilcoxon signed rank test.
-In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
-It obtained better results compared to a wide variety of alignment strategies.
-Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
-MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
-PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.
+COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
+It was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $4.0 \cdot 10^{-06}$ according to the one-tailed, paired Wilcoxon signed-rank tests.
+\TODO{reference figures in supplement.}
+%
+In addition, COATi produced more best alignments than any other method and was most accurate in identifying positive and negative selection (Table \ref{table:comp}).
+It found the third highest number of perfect alignments and the third lowest number of imperfect alignments.
+Manual inspection of these results indicate that the differences between COATi, MAFFT, and PRANK in generating perfect alignments were minor and due to sequence pairs with multiple optimal alignments.
+
+Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
+MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
+PRANK, using a codon model, had a similar average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
 
@@
 model.
-It will allow researchers to analyze more data with higher accuracy and
-facilitate the study of important biological processes that shape genomic data.
+Using COATI will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
 
-Future works include extending the indel FST to combine a three-mer gap model with a frameshift parameter.
+Future work include extending the indel FST to combine a 3-mer gap model with a frameshift parameter and weighing each indel phase differently to reflect known selection on indel phases \citembe{zhu2022profiling}.
 We also plan on comparing the marginal and triplet models to evaluate the implications of the marginalization.
-Additionally, it would be valuable to weigh indel phases differently.

### 2023-05-12 results_discussion.tex
Add pseudocounts for plotting dseq

--- 2023-05-04 results_discussion.tex
@@
 \input{figures/table-comp}
+ \vspace{1mm}
+ \footnotesize{* PRANK produced 50 empty alignments, calculations are based on 7869 alignments.}
  \caption{COATi generates better alignments than other alignment algorithms.
 Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7719 empirically simulated sequence pairs.
-Perfect alignments have $d_{seq}=0$, best alignments have the lowest $d_{seq}$, and imperfect alignments have $d_{seq}>0$ when at least one aligner found a perfect alignment.}
+Perfect alignments have $d_{seq}=0$, best alignments have the lowest $d_{seq}$, and imperfect alignments have $d_{seq}>0$ when at least one aligner found a perfect alignment.
+Alignments with equal score to the true alignment are considered perfect.}
  \label{table:comp}
@@
 
-COATi, using the codon-triplet-ecm model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $4.0 \cdot 10^{-06}$ according to the one-tailed Wilcoxon signed rank test.
+COATi, using the codon-triplet-ecm model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.06 \cdot 10^{-79}$ according to the one-tailed Wilcoxon signed rank test.
 In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).

### 2023-05-12 supplementary_materials/scripts/number_alignments.R
Consider alternative alignments

@@ -8,6 +8,8 @@ num_alns = function(filename) {
     split_tibble <- function(tibble, col = 'col') tibble %>% split(., .[, col])
     split_results = split_tibble(results_summary, 'ref_name')
     dseq = t(sapply(split_results, function(x){ return(x$dseq)}))
+    score = t(sapply(split_results, function(x){ return(x$score)}))
+    ref_score = t(sapply(split_results, function(x){ return(x$ref_score)}))
 
     col_s = ncol(dseq)
     row_s = nrow(dseq)
@@ -17,22 +19,28 @@ num_alns = function(filename) {
                          "imperfect" = integer(col_s))
 
     # perfect alignments (dseq = 0)
-    if(col_s == 2) {
-        results$perfect = length(which(dseq[,2] == 0))
-    } else{
-        results$perfect = apply(dseq[,],2,function(x){length(which(x == 0))})
+    # alignments with same score as the reference alignment are considered dseq = 0
+    for(i in 1:row_s) {
+        alt = which(score[i, ] == ref_score[i])
+        if(length(alt) > 0) {
+            dseq[i, alt] = 0
+        }
+        z = which(dseq[i, ] == 0)
+        if(length(z) > 0){
+            results$perfect[z] = results$perfect[z] + 1
+        }
     }
 
     # best alignments (lower d_seq)
     for(r in 1:(row_s)) {
-        for(i in which(min(dseq[r, ])==dseq[r, ])){
+        for(i in which(min(dseq[r, ], na.rm = TRUE) == dseq[r, ])){
             results$best[i] = results$best[i] + 1
         }
     }
 
     # imperfect alignments (dseq != 0 when another methods has dseq = 0)
     for(r in 1:(row_s)) {
-        if(min(dseq[r, ]) == 0){
+        if(min(dseq[r, ], na.rm = TRUE) == 0){
             for(i in which(dseq[r, ]>0)) {
             results$imperfect[i] = results$imperfect[i] + 1
             }

### 2023-05-12 supplementary_materials.Rmd
Update comp table and supplementary materials

@@
 
-    dseq_summary = data.frame(coati = mean(coati),
-                      prank = mean(prank),
-                      mafft = mean(mafft),
-                      clustalo = mean(clustalomega),
-                      macse = mean(macse), row.names = "$d_{seq}$")
-
-    ############################################################################
-    # one-side wilcoxon rank test
-    wtp = wilcox.test(coati, prank)
-    wtmf = wilcox.test(coati, mafft)
-    wtc = wilcox.test(coati, clustalomega)
-    wtmc = wilcox.test(coati, macse)
-    wilcoxon = data.frame(coati = NA, prank = wtp$p.value, mafft = wtmf$p.value,
-                          clustalo = wtc$p.value, macse = wtmc$p.value,
-                          row.names = "Wilcoxon rank test")
-    stats_table = rbind(dseq_summary, wilcoxon)
+    dseq_summary = data.frame(coati = mean(coati, na.rm = TRUE),
+                      prank = mean(prank, na.rm = TRUE),
+                      mafft = mean(mafft, na.rm = TRUE),
+                      clustalo = mean(clustalomega, na.rm = TRUE),
+                      macse = mean(macse, na.rm = TRUE),
+                      row.names = "$d_{seq}$")
 
@@
     colnames(alns) = c("Perfect alignments", "Best alignments", "Imperfect alignments")
-    stats_table = rbind(stats_table, t(alns))
+    stats_table = rbind(dseq_summary, t(alns))
 
@@
     stats_table = rbind(stats_table, t(selection))
-    colnames(stats_table) = c(coati_model, "PRANK", "MAFFT", "ClustalOmega", "MACSE")
+    colnames(stats_table) = c(coati_model, "PRANK*", "MAFFT", "ClustalOmega", "MACSE")
+    stats_table = stats_table[, c(coati_model, "MAFFT", "PRANK*", "MACSE", "ClustalOmega")]
 
@@
 
+wilcoxon_test = function(results_file, coati_model) {
+    # read results_summary data
+    results = read.csv(results_file)
+
+    # extract dseq values for each aligner
+    coati = results$dseq[which(results$aligner == coati_model)]
+    clustalomega = results$dseq[which(results$aligner == "clustalo")]
+    macse = results$dseq[which(results$aligner == "macse")]
+    mafft = results$dseq[which(results$aligner == "mafft")]
+    prank = results$dseq[which(results$aligner == "prank")]
+
+    # one-side paired wilcoxon rank test
+    wtp = wilcox.test(x = coati, y = prank, paired = TRUE, alternative = "less")
+    wtmf = wilcox.test(x = coati, y = mafft, paired = TRUE, alternative = "less")
+    wtc = wilcox.test(x = coati, y = clustalomega, paired = TRUE, alternative = "less")
+    wtmc = wilcox.test(x = coati, y = macse, paired = TRUE, alternative = "less")
+    w = c(wtp$p.value, wtmf$p.value, wtc$p.value, wtmc$p.value)
+    w
+}
 ```
@@
 ```{r echo = FALSE, results = 'asis'}
+library(kableExtra)
 library(knitr)
 source("supplementary_materials/scripts/plot_dseq.R")
-
 results = "supplementary_data/tri-mg/results/results_summary.csv"
-print(kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7719 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))
-
+k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7719 simulated sequence pairs.
+Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment.
+"), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 50 empty alignments, calculations are based on 7669 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+#
 results = "supplementary_data/tri-ecm/results/results_summary.csv"
-print(kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7678 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))
-
+k = kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7678 simulated sequence pairs.
+Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 41 empty alignments, calculations are based on 7637 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+#
 results = "supplementary_data/mar-mg/results/results_summary.csv"
-print(kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7661 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))
-
+k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7661 simulated sequence pairs.
+Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment.
+"), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 50 empty alignments, calculations are based on 7611 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+#
 results = "supplementary_data/mar-ecm/results/results_summary.csv"
-print(kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7710 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq}$>0 when at least one aligner found a perfect alignment.")))
+k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7710 simulated sequence pairs.
+Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 35 empty alignments, calculations are based on 7675 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
 ```
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = "\\label{fig:dseq-tri-mg} Comparison of $d_{seq}$ values between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE."}
+```{r echo = FALSE}
+results = "supplementary_data/tri-mg/results/results_summary.csv"
+pval_tri_mg = max(wilcoxon_test(results, "tri-mg"))
+results = "supplementary_data/tri-ecm/results/results_summary.csv"
+pval_tri_ecm = max(wilcoxon_test(results, "tri-ecm"))
+results = "supplementary_data/mar-mg/results/results_summary.csv"
+pval_mar_mg = max(wilcoxon_test(results, "mar-mg"))
+results = "supplementary_data/mar-ecm/results/results_summary.csv"
+pval_mar_ecm = max(wilcoxon_test(results, "mar-ecm"))
+```
+
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
+COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
 library(knitr)
 source("supplementary_materials/scripts/plot_dseq.R")
-plot_dseq_main("supplementary_data/tri-mg/results/results_summary.csv", "tri-mg")
+suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg"))
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap =  "\\label{fig:dseq-tri-ecm} Comparison of $d_{seq}$ values between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE."}
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
+COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
 library(knitr)
 source("supplementary_materials/scripts/plot_dseq.R")
-plot_dseq_main("supplementary_data/tri-ecm/results/results_summary.csv", "tri-ecm")
+suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm"))
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap =  "\\label{fig:dseq-mar-mg} Comparison of $d_{seq}$ values between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE."}
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
+COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
 library(knitr)
 source("supplementary_materials/scripts/plot_dseq.R")
-plot_dseq_main("supplementary_data/mar-mg/results/results_summary.csv", "mar-mg")
+suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg"))
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap =  "\\label{fig:dseq-mar-ecm} Comparison of $d_{seq}$ values between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE."}
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
+COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
 library(knitr)
 source("supplementary_materials/scripts/plot_dseq.R")
-plot_dseq_main("supplementary_data/mar-ecm/results/results_summary.csv", "mar-ecm")
+suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm"))
 ```

### 2023-05-12 results_discussion.tex
Fix F1 score description

@@
 \citembe{ka_ks_li_1993}.
-The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
+The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 0 representing a perfect result.
 

### 2023-05-15 results_discussion.tex
Add suppl method kaks paragraph

@@
 using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-\citembe{ka_ks_li_1993}.
+\citembe{ka_ks_li_1993} (Supplementary Methods).
 The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 0 representing a perfect result.
@@
 
-COATi, using the codon-triplet-ecm model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.06 \cdot 10^{-79}$ according to the one-tailed Wilcoxon signed rank test.
+COATi, using the codon-triplet-mg model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.06 \cdot 10^{-79}$ according to the one-tailed Wilcoxon signed rank test (Supplementary Materials Figure 1).
 In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).

### 2023-05-15 supplementary_materials.Rmd
Add suppl method kaks paragraph

@@
 ---
-title: "Supplementary Materials"
-#author: "Juan J. García Mesa"
-#date: "`r Sys.Date()`"
-geometry: margin=1.8cm
+title: "Supplementary Materials for \" COATi: statistical pairwise alignment of protein coding sequences\""
+author: Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright
 output:
@@
         fig_caption: yes
+bibliography: alignpair_letter.bib
 ---
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
@@
 ```
+
+\newpage
+
+# Supplementary Methods
+
+Ks and Ka represent the number of substitutions per synonymous and non-synonymous sites.
+The ratio of nonsynonymous (Ka) to synonymous (Ks) nucleotide substitution rates indicates the selective pressures acting on genes.
+If the ratio is significantly greater than 1, it suggests positive selective pressure, meaning that nonsynonymous substitutions occur more frequently than synonymous substitutions.
+A ratio around 1 can indicate either neutral evolution at the protein level or a mixture of positive and negative selective pressures.
+If the ratio is less than 1, it indicates a pressure to maintain protein sequence, known as purifying selection.
+Ks and Ka are calculated using the R package seqinr v.4.2-30 [@seqinr].
+
+# References
+
+<div id="refs"></div>

### 2023-05-15 results_discussion.tex

@@
 
-COATi, using the codon-triplet-mg model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.06 \cdot 10^{-79}$ according to the one-tailed Wilcoxon signed rank test (Supplementary Materials Figure 1).
-In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
-It obtained better results compared to a wide variety of alignment strategies.
-Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
-MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
-PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.
+COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
+It was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.1 \cdot 10^{-79}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
+%
+In addition, COATi produced more best alignments than any other method and was most accurate in identifying positive and negative selection (Table \ref{table:comp}).
+It found the third highest number of perfect alignments and the third lowest number of imperfect alignments.
+Manual inspection of these results indicate that the differences between COATi, MAFFT, and PRANK in generating perfect alignments were minor and due to sequence pairs with multiple optimal alignments.
+
+Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
+MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
+PRANK, using a codon model, had a similar average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
 

### 2023-05-19 results_discussion.tex
Update number alignments and results

@@
 Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7719 empirically simulated sequence pairs.
-Perfect alignments have $d_{seq}=0$, best alignments have the lowest $d_{seq}$, and imperfect alignments have $d_{seq}>0$ when at least one aligner found a perfect alignment.
-Alignments with equal score to the true alignment are considered perfect.}
+Perfect alignments have the same score as the true alignment, best alignments have the lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
  \label{table:comp}
@@
 %
-In addition, COATi produced more best alignments than any other method and was most accurate in identifying positive and negative selection (Table \ref{table:comp}).
-It found the third highest number of perfect alignments and the third lowest number of imperfect alignments.
-Manual inspection of these results indicate that the differences between COATi, MAFFT, and PRANK in generating perfect alignments were minor and due to sequence pairs with multiple optimal alignments.
+In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
 

### 2023-05-19 supplementary_materials/scripts/number_alignments.R

@@ -18,17 +18,14 @@ num_alns = function(filename) {
                          "best" = integer(col_s),
                          "imperfect" = integer(col_s))
 
-    # perfect alignments (dseq = 0)
-    # alignments with same score as the reference alignment are considered dseq = 0
+    # perfect alignments (same score as the true alignment + dseq = 0)
+    is_perfect = matrix(data = FALSE, nrow = row_s, ncol = col_s)
+
     for(i in 1:row_s) {
-        alt = which(score[i, ] == ref_score[i])
-        if(length(alt) > 0) {
-            dseq[i, alt] = 0
-        }
-        z = which(dseq[i, ] == 0)
-        if(length(z) > 0){
-            results$perfect[z] = results$perfect[z] + 1
-        }
+        perf =  which(score[i, ] == ref_score[i])
+        is_perfect[i, perf] = TRUE
+
+        results$perfect[perf] = results$perfect[perf] + 1
     }
 
     # best alignments (lower d_seq)
@@ -38,12 +35,11 @@ num_alns = function(filename) {
         }
     }
 
-    # imperfect alignments (dseq != 0 when another methods has dseq = 0)
+    # imperfect alignments (score is diff than the true aln when at least one method found a perf aln)
     for(r in 1:(row_s)) {
-        if(min(dseq[r, ], na.rm = TRUE) == 0){
-            for(i in which(dseq[r, ]>0)) {
-            results$imperfect[i] = results$imperfect[i] + 1
-            }
+        if(any(is_perfect[r, ] == TRUE)) {
+            imp = intersect(which(is_perfect[r, ] == FALSE), which(!is.na(dseq[r, ])))
+            results$imperfect[imp] = results$imperfect[imp] + 1
         }
     }
 

### 2023-05-21 results_discussion.tex
Update results and table.

@@
 
-After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 6048 sequence pairs that contained gaps identified by at least one aligner, and 7719 ungapped sequence pairs.
+After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and then aligned the remaining pairs with all five methods.
+At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7719 sequence pairs.
 We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
-Alignment accuracy was measured using the distance metric, $d_{seq}$
-\citembe{metrics_blackburne_whelan_2011}, between simulated and inferred
-alignments.
-In addition, accuracy of positive and negative selection was calculated
-using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
+The accuracy of inferred alignments compared to the benchmarks was measured using \citeauthor{metrics_blackburne_whelan_2011}'s \citeyearpar{metrics_blackburne_whelan_2011} distance metric, $d_{seq}$.
+Additionally, because different alignments may be evolutionary equivalent, we scored alignments using the codon-marginal-mg model, and any alignment that had the same score as the benchmark alignment was considered equivalent to the benchmark alignment.
+The accuracy of identifying positive and negative selection was calculated using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
 \citembe{ka_ks_li_1993} (Supplementary Methods).
-The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 0 representing a perfect result.
+The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
 
@@
  \vspace{1mm}
- \footnotesize{* PRANK produced 50 empty alignments, calculations are based on 7869 alignments.}
+ \footnotesize{\textsuperscript{*}PRANK produced 50 empty alignments, calculations are based on 7869 alignments.}
  \caption{COATi generates better alignments than other alignment algorithms.
 Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7719 empirically simulated sequence pairs.
-Perfect alignments have the same score as the true alignment, best alignments have the lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
+Best alignments have the lowest $d_{seq}$ (including ties), perfect alignments have the same score as the true alignment, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
  \label{table:comp}
@@
 COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
-It was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.1 \cdot 10^{-79}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
+It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $2.1 \cdot 10^{-79}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
 %
-In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
+In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}).
 
@@
 MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
-PRANK, using a codon model, had a similar average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
+PRANK, using a codon model, had an verage alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
 

### 2023-05-22 results_discussion.tex
Fix prank alignments in results table

@@
  \vspace{1mm}
- \footnotesize{\textsuperscript{*}PRANK produced 50 empty alignments, calculations are based on 7869 alignments.}
+ \footnotesize{\textsuperscript{*}PRANK produced 50 empty alignments, calculations are based on 7669 alignments.}
  \caption{COATi generates better alignments than other alignment algorithms.
@@
 MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
-PRANK, using a codon model, had an verage alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
+PRANK, using a codon model, had an average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
 

### 2023-06-15 figures/fig-aln.tex
First round of edits.
Update alignments figure

@@
 \tikzstyle{aln}=[matrix of nodes, nodes in empty cells]
-\tikzstyle{best}=[fill=asublue!20,rounded corners]
+%\tikzstyle{best}=[fill=asublue!20,rounded corners]
+\tikzstyle{best}=[rounded corners]
 \tikzstyle{lab}=[font=\normalsize]
-\tikzstyle{indel}=[thick, rounded corners]
+\tikzstyle{indel}=[ultra thick, rounded corners]
 \tikzstyle{phase0}=[draw=colorGy]
@@
 A: & TCC & CAT & AAG & GGG & CGG & T\verb|--| & \verb|-|CG & GAC & GCC & \verb|---|\\
-D: & TCC & CA\verb|-| & \verb|--|G & GGG & CGG & TCC & CAG & GAC & GCC & ACG\\
-   & Ser &            &            & Gly & Arg & Ser & Gln & Asp & Ala & Thr\\
+D: & TCC & CA\verb|-| & \verb|--|G & GGG & CGG & TCC & CCG & GAC & GCC & ACG\\
+   & Ser &            &            & Gly & Arg & Ser & Pro & Asp & Ala & Thr\\
 };
@@
 
-\node (titlePrank) [below right=2.6cm and -1.8cm of titleBiology] {\textbf{b) Prank (codon)}};
+\node (titlePrankMacse) [below right=2.6cm and -1.8cm of titleBiology] {\textbf{b) PRANK (codon) and MACSE}};
 % prank codon --gaprate=0.01
-\matrix (prank) [aln,below right=-1mm and -2.5cm of titlePrank] {
+\matrix (prank-macse) [aln,below right=-1mm and -4.6cm of titlePrankMacse] {
    & Ser & His     & Lys & Gly  & Arg & Ser &           & Asp & Ala & \\
 A: & TCC & CAT     & AAG & GGG  & CGG & TCG &\verb|---| & GAC & GCC & \verb|---|\\
-D: & TCC & CA\e{G} & \verb|---| & GGG & CGG & TC\e{C} & CAG & GAC & GCC & ACG\\
-   & Ser & Gln     &            & Gly     & Arg & Ser   & Gln  & Asp & Ala & Thr\\
+D: & TCC & CA\e{G} & \verb|---| & GGG & CGG & TC\e{C} & CCG & GCC & GCC & ACG\\
+   & Ser & Gln     &            & Gly     & Arg & Ser   & Pro  & Asp & Ala & Thr\\
 };
-\draw [indel,phase0] (prank-2-8.north west) rectangle (prank-3-8.south east);
-\draw [indel,phase0] (prank-2-11.north west) rectangle (prank-3-11.south east);
-\draw [indel,phase0] (prank-2-4.north west) rectangle (prank-3-4.south east);
+\draw [indel,phase0] (prank-macse-2-8.north west) rectangle (prank-macse-3-8.south east);
+\draw [indel,phase0] (prank-macse-2-11.north west) rectangle (prank-macse-3-11.south east);
+\draw [indel,phase0] (prank-macse-2-4.north west) rectangle (prank-macse-3-4.south east);
 
-\node (titleMafftClustalo) [below right=2.4cm and -2.75cm of titlePrank] {\textbf{c) MAFFT \& Clustal$\pmb{\Omega}$ \& MACSE}};
+\node (titleMafftClustalo) [below right=2.4cm and -4.9cm of titlePrankMacse] {\textbf{c) MAFFT and Clustal$\pmb{\Omega}$}};
 % mafft dna & clustalo amino acids
-\matrix(mafft-clustalo) [aln,below right=-1mm and -4.75cm of titleMafftClustalo] {
+\matrix(mafft-clustalo) [aln,below right=-1mm and -3.5cm of titleMafftClustalo] {
    & Ser & His     & Lys     & Gly     & Arg     & Ser & Asp & Ala & \\
 A: & TCC & CAT & AAG & GGG     & CGG     & TCG & GAC & GCC & \verb|---|\\
-D: & TCC & CA\e{G}     & \e{GG}G     & \e{C}GG & \e{TCC} & \e{CA}G & GAC & GCC & ACG\\
-   & Ser & Gln     & Gly     & Arg     & Ser     & Gln & Asp & Ala & Thr\\
+D: & TCC & CA\e{G}     & \e{GG}G     & \e{C}GG & \e{TCC} & \e{C}CG & GAC & GCC & ACG\\
+   & Ser & Gln     & Gly     & Arg     & Ser     & Pro & Asp & Ala & Thr\\
 };
@@
 
-\node (titleCoati) [below right=2.4cm and -5.05cm of titleMafftClustalo] {\textbf{d) COATi}};
+\node (titleCoati) [below right=2.4cm and -3.75cm of titleMafftClustalo] {\textbf{d) COATi}};
 % coati mar-mg

### 2023-12-21 figures/fig-fst-base-calling-new.tex
Add new fst figures

% !TEX program=lualatex
\documentclass[tikz]{standalone}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
\usepackage{fontspec}
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{xcolor}
\usepackage{xfrac}

\usetikzlibrary{shapes,arrows,positioning,calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.text}
\usetikzlibrary{hobby,spath3,intersections}

\setmainfont{TeX Gyre Heros}
\setmathfont{Latin Modern Math}

% \tikzset{
%   font={\fontsize{6}{7}\selectfont}
% }

%%% Okabe Ito set
\definecolor{colorO}{RGB}{230,159,0}
\definecolor{colorSb}{RGB}{86,180,233}
\definecolor{colorBg}{RGB}{0,158,115}
\definecolor{colorY}{RGB}{240,228,66}
\definecolor{colorBu}{RGB}{0,114,178}
\definecolor{colorV}{RGB}{213,94,0}
\definecolor{colorRp}{RGB}{204,121,167}
\definecolor{colorBk}{RGB}{0,0,0}
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

% (86 mm) or double (178 mm) column width

\tikzset{
  end/.style={fill=none, text=black, inner sep=1pt},
  block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2em,node distance=4em},
  line/.style={draw, -stealth', thick},
  lab/.style={text = black, inner sep=1pt},
  title/.style={text = black, font=\bfseries}
}

\begin{document}
\fontsize{8}{9}\selectfont
\begin{tikzpicture}
%\useasboundingbox (0,0) rectangle +(178mm,178mm);

\node[end, anchor=west] at (0,0) (start) {start};
\node[end, anchor=east] at (86mm,0) (end) {end};

\node[block,right = of start] (s) {S}; 
\node[block,left = of end] (m) {M};

\draw[line] (start) to (s);
\draw[line] (m) to (end);

\coordinate (mid) at ($(s.center)!0.5!(m.center)$);

\draw[line] (m) to (s);

% \path[spath/save=sm] (s.center) circle (1em+0.2pt)
%                   (m.center) circle (1em+0.2pt);

\foreach \y [count=\yi] in {
  {A:A/$(1-e)$}, {A:C/$(e/3)$}, {A:G/$(e/3)$}, {A:T/$(e/3)$},
  {C:A/$(e/3)$}, {C:C/$(1-e)$}, {C:G/$(e/3)$}, {C:T/$(e/3)$}
} {
    \coordinate (y) at ($(mid) + (0, 15.5em) + \yi*(0, -1.75em)$);
    % \path[spath/save=arc] (s.center) to[curve through={(y)}] (m.center);

    % \tikzset{
    %   spath/split at intersections with={arc}{sm},
    %   spath/remove components={arc}{1,3}
    % }
    % \path[line,spath/use=arc];

    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=south] at (y) {\y};
}

\foreach \y [count=\yi] in {
  {G:A/$(e/3)$}, {G:C/$(e/3)$}, {G:G/$(1-e)$}, {G:T/$(e/3)$},
  {T:A/$(e/3)$}, {T:C/$(e/3)$}, {T:G/$(e/3)$}, {T:T/$(1-e)$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}

\node[lab,fill=white] at (mid) {ε:ε/$1$};

\draw (start.west) + (0, 15.5em) node[title,anchor=west] (subfig-a) {a) Error FST};

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\node[end, anchor=west] at (0,-70mm) (start) {start};
\node[end, anchor=east] at (86mm,-70mm) (end) {end};

\node[block,right = of start] (s) {S}; 
\node[block,left = of end] (m) {M};

\draw[line] (start) to (s);
\draw[line] (m) to (end);

\coordinate (mid) at ($(s.center)!0.5!(m.center)$);

\draw[line] (m) to (s);

\foreach \y [count=\yi] in {
  {A:A/$1$}, {C:C/$1$}, {G:G/$1$}, {T:T/$1$}
} {
    \coordinate (y) at ($(mid) + (0, 8.5em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=south] at (y) {\y};
}

\foreach \y [count=\yi] in {
  {A:N/$1$}, {C:N/$1$}, {G:N/$1$}, {T:N/$1$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}

\node[lab,fill=white] at (mid) {ε:ε/$1$};

\draw (start.west) + (0, 8.5em) node[title,anchor=west] (subfig-b) {b) Ambiguity FST};

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\node[end, anchor=west] at (92mm,-3.5em) (start) {start};
\node[end, anchor=east] at (178mm,-3.5em) (end) {end};

\node[block,right = of start] (s) {S}; 
\node[block,left = of end] (m) {M};

\draw[line] (start) to (s);
\draw[line] (m) to (end);

\coordinate (mid) at ($(s.center)!0.5!(m.center)$);

\draw[line] (m) to (s);

\foreach \y [count=\yi] in {
  {A:N/$1$}, {A:A/$(1-e)$}, {A:C/$(e/3)$}, {A:G/$(e/3)$}, {A:T/$(e/3)$},
  {C:N/$1$}, {C:A/$(e/3)$}, {C:C/$(1-e)$}, {C:G/$(e/3)$}, {C:T/$(e/3)$}
} {
    \coordinate (y) at ($(mid) + (0, 19em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=south] at (y) {\y};
}

\foreach \y [count=\yi] in {
  {G:N/$1$}, {G:A/$(e/3)$}, {G:C/$(e/3)$}, {G:G/$(1-e)$}, {G:T/$(e/3)$},
  {T:N/$1$}, {T:A/$(e/3)$}, {T:C/$(e/3)$}, {T:G/$(e/3)$}, {T:T/$(1-e)$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}

\node[lab,fill=white] at (mid) {ε:ε/$1$};

\draw (start.west) + (0, 19em) node[title,anchor=west] (subfig-b) {c) Error $∘$ Ambiguity FST};

\end{tikzpicture}
\end{document}

### 2023-12-21 figures/fig-indel-fst.tex
Add new fst figures

\definecolor{colorR}{RGB}{228,26,28}    % RED
\definecolor{colorB}{RGB}{55,126,184}   % BLUE
\definecolor{colorG}{RGB}{77,175,74}    % GREEN
\definecolor{colorP}{RGB}{152,78,163}   % PURPLE
\definecolor{colorO}{RGB}{255,127,0}    % ORANGE
\definecolor{colorY}{RGB}{255,255,51}   % YELLOW
\definecolor{colorBn}{RGB}{166,86,40}   % BROWN
\definecolor{colorPk}{RGB}{247,129,191} % PINK
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

\tikzstyle{line}=[draw, -stealth', very thick]
\tikzstyle{block}=[circle,fill=colorB!50,on grid]
\tikzstyle{lab}=[]
\tikzstyle{w}=[lab,midway]
\tikzstyle{e}=[lab,midway,above=0.2mm,auto=false,font=\small]

\begin{tikzpicture}[node distance=40mm, auto,
	dottedline/.style = {ultra thick, loosely dotted,shorten >=1mm, shorten <=1mm}]

%%% Indel FST
% \node[below left=3cm and -14mm of titlea.east] (titleb) {\textbf{b) Insertion-Deletion}};
\node (titleb) {};

\node[block,below right=1.5cm and -1.5cm of titleb.center,fill=colorG!50] (start1) {start};
\node[block,below=15mm of start1] (S1) {M};
\node[block,below=of S1] (I1) {U};
\node[block,right=of I1] (U1) {I};
\node[block,above=of U1] (W1) {W};
\node[block,above=20mm of W1] (D1) {V};
\node[block,right=of D1] (V1) {D};
\node[block,right=of W1] (M1) {S};
\node[block,fill=colorR!50,right=20mm of M1] (end1) {end};

\draw (D1) +(0mm,-20mm) coordinate (C1);
\draw (D1) +(50mm,-10mm) coordinate (C2);
\draw[line] (start1) -- (S1);
\draw[line] (S1) -- node[w] {$g$} (I1);
\draw[line] (S1) -- node[w] {$1-g$} (W1);
\draw[line] (U1) to[out=120,in=60]  node[w,above] {$e$} (I1);
\draw[line] (U1) -- node[w,pos=0.1,right] {$1-e$} (W1);
\draw[line] (W1) -- node[w] {$g$} (D1);
\draw[line] (W1) -- node[w] {$1-g$} (M1);
\draw[line] (M1) -- node[w] {} (end1);
\draw[line] (V1) to[out=120,in=60] node[w,above] {$e$} (D1);
\draw[line] (V1) -- node[w] {$1-e$} (M1);
%insertions
\draw[line] (I1) to[bend right=-25] node[e,pos=0.48,below=-5mm] {\O:A/$\grande{\pi}_{\text{A}}$} (U1);
\draw[line] (I1) to[bend right=0] node[e,pos=0.48,below=-5mm] {\O:C/ $\grande{\pi}_{\text{C}}$} (U1);
\draw[line] (I1) to[bend right=25] node[e,pos=0.48,below=-5mm] {\O:G/ $\grande{\pi}_{\text{G}}$} (U1);
\draw[line] (I1) to[bend right=50] node[e,pos=0.48,below=-5mm] {\O:T/ $\grande{\pi}_{\text{T}}$} (U1);
% deletion
\draw[line] (D1) to[bend right=-25] node[e,pos=0.5,below=-4.5mm] {A:\O} (V1);
\draw[line] (D1) to[bend right=0] node[e,pos=0.5,below=-4.5mm] {C:\O} (V1);
\draw[line] (D1) to[bend right=25] node[e,pos=0.5,below=-4.5mm] {G:\O} (V1);
\draw[line] (D1) to[bend right=50] node[e,pos=0.5,below=-4.5mm] {T:\O} (V1);
% A -> {A,C,G,T}
\draw[line] (M1) to[bend left=25] (S1);
\draw[line] (M1) to[bend left=35] (S1);
\draw[line] (M1) to[bend left=45] (S1);
\draw[line] (M1) to[bend left=55] (S1);
\path[] (M1) to[bend left=25] node[e,pos=0.8,below=-3mm,fill=white] {A:A} (S1);
\path[] (M1) to[bend left=35] node[e,pos=0.6,below=-3mm,fill=white] {C:C} (S1);
\path[] (M1) to[bend left=45] node[e,pos=0.4,below=-3mm,fill=white] {G:G} (S1);
\path[] (M1) to[bend left=55] node[e,pos=0.2,below=-3mm,fill=white] {T:T} (S1);
% % dots
% \node[below left=9mm and -1mm of W1,font=\Huge] {$\vdots$};
% % T -> {A,C,G,T}
% \draw[line] (M1) to[bend left=55] (S1);
% \draw[line] (M1) to[bend left=59] (S1);
% \draw[line] (M1) to[bend left=63] (S1);
% \draw[line] (M1) to[bend left=67] (S1);
% \path[] (M1) to[bend left=55] node[e,pos=0.8,below=-3mm,fill=white] {T:A} (S1);
% \path[] (M1) to[bend left=59] node[e,pos=0.6,below=-3mm,fill=white] {T:C} (S1);
% \path[] (M1) to[bend left=63] node[e,pos=0.4,below=-3mm,fill=white] {T:G} (S1);
% \path[] (M1) to[bend left=67] node[e,pos=0.2,below=-3mm,fill=white] {T:T} (S1);

%%%% Base calling error FST
% \node[right=6cm of titleb.east] (titlec) {\textbf{Base Calling Error}};

% \node[block,below=1.5cm of titlec,fill=colorG!50] (start_bce) {start};
% \node[block,right=15mm of start_bce] (S_bce) {M};
% \node[block,right=60mm of S_bce] (M_bce) {S};
% \node[block,fill=colorR!50,right=15mm of M_bce] (end_bce) {end};

% \draw[line] (start_bce) -- (S_bce);
% \draw[line] (M_bce) -- (end_bce);

% % \draw[line] (M_bce) to[bend right=25] node[w,below=1mm,pos=0.66] {$\varepsilon$} (S_bce);
% \draw[line] (M_bce) to[bend right=25] node[e] {y$_\text{i}$:z$_\text{j}$/$\grande{\varepsilon}$} (S_bce);
% % \draw[line] (M_bce) to node[w,below=0.2mm,pos=0.66] {$1 - 3 \cdot \varepsilon$} (S_bce);
% \draw[line] (M_bce) to node[e] {y$_\text{i}$:z$_\text{j}$/$1-3\cdot \grande{\varepsilon}$} (S_bce);
% \draw[line] (M_bce) to[bend right=-25] (S_bce);
% \draw[line] (M_bce) to[bend right=-25] node[e] {y$_\text{i}$:N} (S_bce);

% \draw[line] (S_bce) to[bend left=-60] (M_bce);

%%% Legend

% \node[below left=2cm and 1.5cm of start_bce,text width=50mm,anchor=north west] (sequences) {
\node[below right=5.8cm and 4.5cm of titleb,text width=50mm,anchor=north west] (sequences) {
\begin{tabular}{r@{\,: }l}
\multicolumn{2}{l}{\textbf{Parameters}}\\
    \hline
	% X & input nucleotides\\
	% y & input sequence\\
	% z & output sequence\\
	\O & empty sequence\\
	$g$ & gap open weight\\
	$e$ & gap extension weight\\
	$\grande{\pi}$ & residue stationary freq.\\
        % Y$_i$:Y$_i$ & matches\\
        % Y$_i$:Y$_j$ & mismatches\\
 % 	$i \rightarrow i$ & matching input to intermediate base\\
	% $i \rightarrow j$ & mismatching input to intermediate base\\
	% $i \rightarrow N$ & input to intermediate ambiguous base N\\
\end{tabular}
};

\end{tikzpicture}

### 2023-12-21 figures/fig-subst-fst-new.tex
Add new fst figures

% !TEX program=lualatex
\documentclass[tikz]{standalone}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
\usepackage{fontspec}
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{xcolor}

\usetikzlibrary{shapes,arrows,positioning}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.text}
\usetikzlibrary{hobby,spath3,intersections}

\setmainfont{TeX Gyre Heros}
\setmathfont{Latin Modern Math}

% \tikzset{
%   font={\fontsize{6}{7}\selectfont}
% }

%%% Okabe Ito set
\definecolor{colorO}{RGB}{230,159,0}
\definecolor{colorSb}{RGB}{86,180,233}
\definecolor{colorBg}{RGB}{0,158,115}
\definecolor{colorY}{RGB}{240,228,66}
\definecolor{colorBu}{RGB}{0,114,178}
\definecolor{colorV}{RGB}{213,94,0}
\definecolor{colorRp}{RGB}{204,121,167}
\definecolor{colorBk}{RGB}{0,0,0}
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

\tikzstyle{line}=[draw, -stealth', very thick]
\tikzstyle{lab}=[]
\tikzstyle{w}=[lab,midway]
\tikzstyle{e}=[lab,midway,above=0.2mm,auto=false,font=\scriptsize]

% (86 mm) or double (178 mm) column width

\begin{document}
\fontsize{6}{7}\selectfont
\begin{tikzpicture}[
  block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2.1em,node distance=4em},
  start/.style={block,fill=none,text=black, inner sep=1pt},
  end/.style={block,fill=none,text=black, inner sep=1pt},
  line/.style={draw, -stealth', very thick},
  label/.style={text = black}
]

\node[start] (start) {start};
\node[block,right = 5em of start] (s) {S};
\node[block, right = 7em of s] (u0) {};
\node[block, right = 21em of s] (v0) {};
\node[block,right = 28em of s] (m) {M};
\node[end,right = 5em of m] (e) {end};

\node[block,below = of u0] (u1) {};
\node[block,below = of u1] (u2) {};
\node[block,below = of u2] (u3) {};
\node[on grid, below = 4.25em of u3] (u4) {};

\node[block,below = of v0] (v1) {};
\node[block,below = of v1] (v2) {};
\node[block,below = of v2] (v3) {};
\node[on grid, below = 4.25em of v3] (v4) {};

\draw[line] (start) to (s); 
\draw[line] (m) to (e); 
\draw[line] (m) to[bend right=30] (s);

%\draw[red] (start.west) ++(0,15mm) rectangle +(86mm,-86mm);

%%%%%%%%

\path[spath/save = curve0] (s.center) to[curve through={(u0.center) .. (v0.center)}] (m.center);

\path[spath/save = c0]
    (s.center) circle (1.1em)
    (u0.center) circle (1.1em)
    (v0.center) circle (1.1em)
    (m.center) circle (1.1em);

\tikzset{
    spath/.cd,
    split at intersections with={curve0}{c0},
    get components of={curve0}\cpts
}

\draw[line,spath/restore=\getComponentOf{\cpts}{2}];
\draw[line,spath/restore=\getComponentOf{\cpts}{4}];
\draw[line,spath/restore=\getComponentOf{\cpts}{6}];

\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{2}}{0.5}] {A:A};

\node[anchor = north] at (spath cs:{\getComponentOf{\cpts}{4}} 0.5) {A:A/P(AAA|AAA)};
\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{6}}{0.5}]  {A:A};

%%%%%%%%

\path[spath/save = curve1] (s.center) to[curve through={(u1.center) .. (v1.center)}] (m.center);

\path[spath/save = c1]
    (s.center) circle (1.1em)
    (u1.center) circle (1.1em)
    (v1.center) circle (1.1em)
    (m.center) circle (1.1em);

\tikzset{
    spath/.cd,
    split at intersections with={curve1}{c1},
    get components of={curve1}\cpts
}

\draw[line,spath/restore=\getComponentOf{\cpts}{2}];
\draw[line,spath/restore=\getComponentOf{\cpts}{4}];
\draw[line,spath/restore=\getComponentOf{\cpts}{6}];

\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{2}}{0.58}] {A:A};

\node[label, anchor = north] at (spath cs:{\getComponentOf{\cpts}{4}} 0.5) {A:A/P(AAC|AAA)};
\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{6}}{0.42}]  {A:C};

%%%%%%%%

\path[spath/save = curve2] (s.center) to[curve through={(u2.center) .. (v2.center)}] (m.center);

\path[spath/save = c2]
    (s.center) circle (1.1em)
    (u2.center) circle (1.1em)
    (v2.center) circle (1.1em)
    (m.center) circle (1.1em);

\tikzset{
    spath/.cd,
    split at intersections with={curve2}{c2},
    get components of={curve2}\cpts
}

\draw[line,spath/restore=\getComponentOf{\cpts}{2}];
\draw[line,spath/restore=\getComponentOf{\cpts}{4}];
\draw[line,spath/restore=\getComponentOf{\cpts}{6}];

\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{2}}{0.67}] {A:A};

\node[label, anchor = north] at (spath cs:{\getComponentOf{\cpts}{4}} 0.5) {A:A/P(AAG|AAA)};
\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{6}}{0.33}]  {A:G};

%%%%%%%%

\path[spath/save = curve3] (s.center) to[curve through={(u3.center) .. (v3.center)}] (m.center);

\path[spath/save = c3]
    (s.center) circle (1.1em)
    (u3.center) circle (1.1em)
    (v3.center) circle (1.1em)
    (m.center) circle (1.1em);

\tikzset{
    spath/.cd,
    split at intersections with={curve3}{c3},
    get components of={curve3}\cpts
}

\draw[line,spath/restore=\getComponentOf{\cpts}{2}];
\draw[line,spath/restore=\getComponentOf{\cpts}{4}];
\draw[line,spath/restore=\getComponentOf{\cpts}{6}];

\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{2}}{0.75}] {A:A}; 
\node[label, anchor = north] at (spath cs:{\getComponentOf{\cpts}{4}} 0.5) {A:A/P(AAT|AAA)};

\node[label, below, transform shape, spath/transform to={\getComponentOf{\cpts}{6}}{0.25}] {A:T};

%%%%%%%%

\path[spath/save = curve4] (s.center) to[curve through={(u4.center) .. (v4.center)}] (m.center);

\path[spath/save = c4]
    (s.center) circle (1.1em)
    (m.center) circle (1.1em);

\tikzset{
    spath/.cd,
    split at intersections with={curve4}{c4},
    get components of={curve4}\cpts
}

\draw[line,draw=darkgray,dotted,spath/use=\getComponentOf{\cpts}{2}];

\draw[draw=white,line width=1ex,
spath/split at keep middle={\getComponentOf{\cpts}{2}}{0.37}{0.6},
spath/use=\getComponentOf{\cpts}{2}];

\path[spath/use=\getComponentOf{\cpts}{2},
  postaction={decorate,decoration={text along path,
  text align=center,text color=darkgray,raise=-0.5ex,
  text={|\itshape|3717 additional arcs}}}];

\end{tikzpicture}
\end{document}

### 2023-12-21 figures/fig-subst-fst.tex
Add new fst figures

\definecolor{colorR}{RGB}{228,26,28}    % RED
\definecolor{colorB}{RGB}{55,126,184}   % BLUE
\definecolor{colorG}{RGB}{77,175,74}    % GREEN
\definecolor{colorP}{RGB}{152,78,163}   % PURPLE
\definecolor{colorO}{RGB}{255,127,0}    % ORANGE
\definecolor{colorY}{RGB}{255,255,51}   % YELLOW
\definecolor{colorBn}{RGB}{166,86,40}   % BROWN
\definecolor{colorPk}{RGB}{247,129,191} % PINK
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

\tikzstyle{line}=[draw, -stealth', very thick]
\tikzstyle{block}=[circle,fill=colorB!50,on grid]
\tikzstyle{lab}=[]
\tikzstyle{w}=[lab,midway]
\tikzstyle{e}=[lab,midway,above=0.2mm,auto=false,font=\scriptsize]

% \begin{document}
% \begin{tikzpicture}[node distance=25mm, auto]
\begin{tikzpicture}[node distance=25mm, auto,
	dottedline/.style = {ultra thick, loosely dotted,shorten >=1mm, shorten <=1mm}]
%%% MG94 marginal substitution FST
% \node (titlea) {};

\node[block,fill=colorG!50] (start_sub) {start};
\node[block,right=15mm of start_sub] (S_sub) {M};
\node[block,right=75mm of S_sub] (M_sub) {S};
\node[block,fill=colorR!50,right=15mm of M_sub] (end_sub) {end};

% AAA->AAA
\node[block,above right=12mm and 20mm of S_sub,minimum size=15pt] (s2) {};
\node[block,above right=3mm and 20mm of s2,minimum size=15pt] (s1) {};
% % AAA->AAC
\node[block,above right=4mm and 20mm of S_sub,minimum size=15pt] (s4) {};
\node[block,below right=-1mm and 20mm of s4,minimum size=15pt] (s3) {};
% TTT->TTT
\node[block,below right=6mm and 20mm of S_sub,minimum size=15pt] (s122) {};
\node[block,below right=3mm and 20mm of s122,minimum size=15pt] (s121) {};

\draw[line] (start_sub) -- (S_sub);
\draw[line] (S_sub) to[bend right=40] (M_sub);
\draw[line] (M_sub) -- (end_sub);

% AAA->AAA
% \draw[line] (M_sub) to[bend right=15] node[w,pos=0.7,above right=3mm and 1mm,font=\scriptsize,rotate=-20] {$P(AAA|AAA)$} (s1);
\draw[line] (M_sub) to[bend right=15] node[e,pos=0.4,rotate=-20] {A:A/$P(AAA|AAA)$} (s1);
\draw[line] (s1) to[bend right=15] node[e,pos=0.5] {A:A} (s2);
\draw[line] (s2) to[bend right=15] node[e,pos=0.5] {A:A} (S_sub);

% % AAA->AAC
\draw[line] (M_sub) to[bend right=5] node[e,pos=0.5,rotate=-8,below=-5mm] {A:A/$P(AAC|AAA)$} (s3);
\draw[line] (s3) to[bend right=8] node[e,pos=0.5] {A:A} (s4);
\draw[line] (s4) to[bend right=8] node[e,pos=0.5] {A:C} (S_sub);

\node[below=9mm of s1,font=\Huge] {$\vdots$};
\node[below=5mm of s2,font=\Huge] {$\vdots$};

% TTT->TTT
% \draw[line] (M_sub) to[bend right=-5] node[w,pos=0.5,above right=1mm and -11mm, rotate=5,font=\scriptsize] {$P(TTT|TTT)$} (s121);
\draw[line] (M_sub) to[bend right=-5] node[e,pos=0.5,rotate=9] {T:T/$P(TTT|TTT)$} (s121);
\draw[line] (s121) to[bend right=-5] node[e,pos=0.5] {T:T} (s122);
\draw[line] (s122) to[bend right=-5] node[e,pos=0.4] {T:T} (S_sub);
\end{tikzpicture}

### 2023-12-21 figures/fig-transcription.tex
Add new fst figures

\definecolor{colorR}{RGB}{228,26,28}    % RED
\definecolor{colorB}{RGB}{55,126,184}   % BLUE
\definecolor{colorG}{RGB}{77,175,74}    % GREEN
\definecolor{colorP}{RGB}{152,78,163}   % PURPLE
\definecolor{colorO}{RGB}{255,127,0}    % ORANGE
\definecolor{colorY}{RGB}{255,255,51}   % YELLOW
\definecolor{colorBn}{RGB}{166,86,40}   % BROWN
\definecolor{colorPk}{RGB}{247,129,191} % PINK
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

\tikzstyle{line}=[draw, -stealth', very thick]
\tikzstyle{block}=[circle,fill=colorB!50,on grid]
\tikzstyle{lab}=[]
\tikzstyle{w}=[lab,midway]
\tikzstyle{e}=[lab,midway,below=-4mm,auto=false,font=\scriptsize]

% \begin{document}
% \begin{tikzpicture}[node distance=25mm, auto]
\begin{tikzpicture}[node distance=25mm, auto,
	dottedline/.style = {ultra thick, loosely dotted,shorten >=1mm, shorten <=1mm}]

% complimenting FST
\node (titlea) {\textbf{a) DNA Complement}};

\node[block,below left=1cm and -10mm of titlea.west,fill=colorG!50] (start1) {start};
\node[block,right=15mm of start1,minimum size=20pt] (cod0) {};
\node[block,right=20mm of cod0,minimum size=20pt] (cod1) {};
\node[block,right=20mm of cod1,minimum size=20pt] (cod2) {};
\node[block,right=20mm of cod2,minimum size=20pt] (cod3) {};
\node[block,fill=colorR!50,right=15mm of cod3] (end1) {end};

\draw[line] (start1) -- (cod0);
\draw[line] (cod3) -- (end1);

\draw[line] (cod0) to[bend right=-60] node[e] {A:T} (cod1);
\draw[line] (cod0) to[bend right=-25] node[e] {C:G} (cod1);
\draw[line] (cod0) to[bend right=25] node[e,below=-4.5mm] {G:C} (cod1);
\draw[line] (cod0) to[bend right=60] node[e] {T:A} (cod1);

\draw[line] (cod1) to[bend right=-60] node[e] {A:T} (cod2);
\draw[line] (cod1) to[bend right=-25] node[e] {C:G} (cod2);
\draw[line] (cod1) to[bend right=25] node[e,below=-4.5mm] {G:C} (cod2);
\draw[line] (cod1) to[bend right=60] node[e] {T:A} (cod2);

\draw[line] (cod2) to[bend right=-60] node[e] {A:T} (cod3);
\draw[line] (cod2) to[bend right=-25] node[e] {C:G} (cod3);
\draw[line] (cod2) to[bend right=25] node[e,below=-4.5mm] {G:C} (cod3);
\draw[line] (cod2) to[bend right=60] node[e] {T:A} (cod3);

% T->U
\node[right=6cm of titlea] (titleb) {\textbf{b) Replacing thymine with uracil}};

\node[block,below right=1cm and 12mm of titleb.west,fill=colorG!50] (start2) {start};
\node[block,right=20mm of start2,minimum size=30pt] (cod4) {};
\node[block,fill=colorR!50,right=20mm of cod4] (end2) {end};

\draw[line] (start2) -- (cod4);
\draw[line] (cod4) -- (end2);

\draw[line] (cod4) to[out=165,in=105,min distance=2mm,looseness=5] node[e,below left=0.5mm and -6mm] {A:A} (cod4);
\draw[line] (cod4) to[out=85,in=15,min distance=2mm,looseness=5] node[e,below left=0.5mm and -1mm] {C:C} (cod4);
\draw[line] (cod4) to[out=-15,in=285,min distance=2mm,looseness=5] node[e,above left=0.5mm and -1mm] {G:G} (cod4);
\draw[line] (cod4) to[out=255,in=195,min distance=2mm,looseness=5] node[e,above left=0.5mm and -6mm] {T:U} (cod4);

% transcription FST
\node[below right=3cm and 25mm of titlea.west] (titlec) {\textbf{c) DNA Transcription}};

\node[block,below right=1cm and 1cm of titlec.west,fill=colorG!50] (start3) {start};
\node[block,right=15mm of start3,minimum size=20pt] (cod5) {};
\node[block,right=20mm of cod5,minimum size=20pt] (cod6) {};
\node[block,right=20mm of cod6,minimum size=20pt] (cod7) {};
\node[block,right=20mm of cod7,minimum size=20pt] (cod8) {};
\node[block,fill=colorR!50,right=15mm of cod8] (end3) {end};

\draw[line] (start3) -- (cod5);
\draw[line] (cod8) -- (end3);

\draw[line] (cod5) to[bend right=-60] node[e] {A:U} (cod6);
\draw[line] (cod5) to[bend right=-25] node[e] {C:G} (cod6);
\draw[line] (cod5) to[bend right=25] node[e,below=-4.5mm] {G:C} (cod6);
\draw[line] (cod5) to[bend right=60] node[e] {T:A} (cod6);

\draw[line] (cod6) to[bend right=-60] node[e] {A:U} (cod7);
\draw[line] (cod6) to[bend right=-25] node[e] {C:G} (cod7);
\draw[line] (cod6) to[bend right=25] node[e,below=-4.5mm] {G:C} (cod7);
\draw[line] (cod6) to[bend right=60] node[e] {T:A} (cod7);

\draw[line] (cod7) to[bend right=-60] node[e] {A:U} (cod8);
\draw[line] (cod7) to[bend right=-25] node[e] {C:G} (cod8);
\draw[line] (cod7) to[bend right=25] node[e,below=-4.5mm] {G:C} (cod8);
\draw[line] (cod7) to[bend right=60] node[e] {T:A} (cod8);

\end{tikzpicture}

### 2023-12-21 introduction.tex
Add new fst figures

@@
 The alignment of protein-coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space (e.g.\ \citetwo{bininda2005transalign}{abascal2010translatorx}).
-While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
-Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
-In addition, existing aligners typically force gaps to occur between codons, whereas in natural sequences, only about 40\% of indels occur between codons \citembetwo{taylor2004occurrence}{zhu2022profiling}.
+While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts, such as frameshifts and early stop codons.
+While some aligners can utilize codon substitution models, they are often not robust against coding-sequence artifacts. Additionally these aligners 
+force gaps to occur between codons, whereas in natural sequences, only about 42\% of indels occur between codons \citembetwo{taylor2004occurrence}{zhu2022profiling}.
 This mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).
@@
         (b), (c), and (d) are the results of different aligners.
-        Nucleotide mismatches are highlighted in red. Phase 0, phase 1, and phase 2 indels are shown in gray, purple, and orange, respectively.
+        Nucleotide mismatches are highlighted in red. Phase 1, phase 2, and phase 3 indels are shown in purple, orange, and gray respectively.
         Additionally, the orange indel is type II (an amino-acid indel plus an amino-acid change) while the purple indel is type I (an amino-acid indel only).
@@
 
-Genome quality impacts conclusions drawn from comparative genomic studies, and uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
-Genomes for model organisms often get refined over many iterations and achieve high quality with meticulously curated protein coding sequences.
-In contrast, genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
+Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citembethree{estimates_schneider_2009}{effect_fletcher_2010}{hubisz2011error}.
+While genomes for model organisms often get refined over many iterations and contain meticulously curated protein-coding sequences, 
+genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
 These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citembe{jackman2018tigmint}.
@@
 current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
-To address this problem, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates codon substitution models and is robust to artifacts present in modern genomic data.
+
+To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets.

### 2023-12-21 materials_methods.tex
Add new fst figures

@@
 \section*{Materials and Methods}
+
+\subsection*{Statistical Alignment}
 
@@
 A limitation of pair-HMMs is that they only model the evolution of two related sequences from an unknown ancestor.
-Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that can model the generation of a descendant sequence given an ancestral one.
+Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that they can model the generation of a descendant sequence given an ancestral one.
 FSTs consume symbols from an input tape and emit symbols to an output tape.
 Properly weighted, an FST can calculate the probability that a descendant sequence, $\text{Y}$, evolved from an ancestral sequence, $\text{X}$, represented by $P(\text{Y} | \text{X})$.
-Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
-A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
+% Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
+% A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
+% COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
+
+Furthermore, there are well-established algorithms for combining FSTs in different ways, including concatenation, composition, intersection, union, or reversal, allowing the design of complex models by combining simpler FSTs \citep{bradley2007transducers,silvestre2021machine}.
+Specifically, composition is a powerful and versatile algorithm for comparative sequence analysis, consisting of sending the output of one FST into the input of a second FST.
+Composition allows combining two or more FSTs to create a new, more complex transducer.
+Figure \ref{fig:transcription} illustrates how DNA transcription for a codon can be achieved by composing a complementing FST with a transducer that replaces thymines with uracil, where the three nucleotides are read and complemented in \ref{fig:transcription}-a, which are then used as the input of \ref{fig:transcription}-b, resulting in the transcription of the codon (Fig.~\ref{fig:transcription}-c).
 COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
+
+\begin{figure}[!ht]
+    \centering
+    %\includegraphics[width=\textwidth]{figures/fig-error-fst-new.pdf}
+    \subimport{figures}{fig-fst-base-calling-new.tex}
+    \caption[DNA Transcription via FST Composition]{DNA transcription via FST composition.
+The composition of (a) a codon complimenting FST and (b) an FST that replaces T with U generates (c) an FST that implements codon transcription ($a~\circ~b = c$).}
+    \label{fig:transcription}
+\end{figure}
 
@@
 Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
-This FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
+The assumption is that the reference sequence is in-phase, which is used to help preserve the reading frame and safeguard against possible frameshifts in the ``descendant''.
+Therefore, the high-quality sequence must not contain incomplete codons (the number of nucleotides is multiple of three) and be free of any ambiguous nucleotides or early stop codons.
+In contrast, the potentially lower-quality sequence has no such restrictions and can be of any length, contain early stop codons---treated as artifacts---, and include ambiguous codons.
+
+The Evolution FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
 A key innovation of this FST, with respect to others, is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.
+
+\begin{figure}[h!]
+\centering
+\setlength{\columnsep}{-3cm}
+\begin{multicols}{2}
+    \hspace*{-2.8cm}\resizebox{0.55\textwidth}{!}{\input{figures/fig-subst-fst}}
+    \columnbreak
+    \resizebox{0.55\textwidth}{!}{\input{figures/fig-indel-fst}}
+\end{multicols}
+\par
+\caption{The Evolution FST is assembled by composing a substitution FST and an indel FST.
+Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
+(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M.
+These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
+The weight of each arc is a conditional probability derived from a codon substitution model.
+(b) The indel FST allows for insertions (U to I) and deletions (V to D).
+Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
+On top of the indel FST, we add a base-calling-error FST (Supplemental Materials Figure 1) to model sequencing errors.
+Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
+\label{fig:evolution-fst}
+\end{figure}
 
@@
 Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment.
+When equally optimal alignments exist, ties are broken according to the FST shortest-path algorithm.
+An example of an FST-based alignment can be found in Supplementary Materials Figure 2.
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citembe{allauzen2007openfst}.
@@
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=\textwidth]{fig-evolution-fst.pdf}
-\par
-\caption{The Evolution FST is assembled by composing a substitution FST and an indel FST.
-Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
-(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from M to S.
-These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
-The weight of each arc is a conditional probability derived from a codon substitution model.
-(b) The indel FST allows for insertions (I to U) and deletions (D to V).
-Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
-On top of the indel FST, we add a base-calling-error FST (Supplemental Materials Figure 5) to model sequencing errors.
-Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
-\label{fig:evolution-fst}
-\end{figure}
+\section*{Codon Substitution Models}
 
@@
 Specifically, let
-$P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)$ represent transition probabilities from a standard codon model, and
+\[
+P_\text{cod}\left( Y_1 = y_1, Y_2 = y_2, Y_3 = y_3 |
+                   X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
+\]
+represent transition probabilities from a standard codon model, and
 %
 \[
-P_\text{mar}\left(Y_p = y \middle| X_0 \cdot X_1 \cdot X_2\right) =
-\sum_{Y_0 \cdot Y_1 \cdot Y_2} I(Y_p = y) P_\text{cod}\left(Y_0 \cdot Y_1 \cdot Y_2 \middle| X_0 \cdot X_1 \cdot X_2\right)
+P_\text{mar}\left(Y_p = y \middle| X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
+=
+\sum_{y_1, y_2, y_3} I(y_p = y)
+P_\text{cod}\left( y_1, y_2, y_3 |
+                   x_1, x_2, x_3 \right)
 \]
@@
 represent the marginal transition probabilities, where
-$p \in \{0, 1, 2\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function defined
-%
-$I(e) = \{ 1 \text{ if $e$ is true and } 0 \text{ otherwise}\}$.
-%
+$p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function.
 COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
@@
 In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.
+
+\subsection*{Empirical Simulation Algorithm}
+
+I downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
+database \citep{ensembl_hubbard_2002}.
+After downloading, I removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with all five methods.
+At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7761 sequence pairs.
+Then, I randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
+
+The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length.
+The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome.
+Assigning one of the sequences as the reference, to distinguish between insertions and deletions, CIGAR strings can also summarize pairwise alignments by grouping the number of contiguous matches or mismatches `M', deletions `D', and insertions `I'.
+The resulting pattern combines these letters preceded by the number of characters for each section as they appear in the alignment.
+This pattern is introduced by replacing nucleotides with gaps as indicated by deletions on one sequence and randomly introducing residues where the CIGAR strings indicated insertions.
+
+% Several safety checks are in place to ensure the algorithm runs correctly and the result is accurate.
+The assertions are divided into checking lengths and maintaining the reading frame of each section.
+The simulation can fail if the length of the sequences is different or if, without counting insertions, the length of the pattern to be inserted is longer.
+In addition, maintaining the phases of each section in the CIGAR string is important to avoid introducing errors such as frameshifts.
+
+I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
+% After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
+I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
+Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
+MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
+\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
+
+\subsection*{Metrics}
+
+To quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools, I used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}.
+This metric accounts for indels and is more informative than conventional distance scores like sum-of-pairs or total columns.
+Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the probability that a randomly selected residue will be aligned to a different location against a sequence that does not contain such residue.
+
+The computation of $d_{seq}$ involves characterizing the gaps present in the alignment.
+Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$.
+The distance between two alignments $A$ and $B$ is the average across all characters of the symmetric difference (or Hamming distance), represented as `$\triangle$' between homology sets over the length of such sets:
+
+\begin{equation}
+d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
+\end{equation}
+
+\noindent where $c$ is the sum of the sequence lengths.
+
+In addition, I compared the number of perfectly and imperfectly retrieved alignments for each aligner.
+Perfect alignments are defined as those with a distance of zero to the reference alignment ($d_{seq} = 0$), indicating 100\% similarity.
+Notably, a set of sequences can have more than one optimal alignment under the same evolutionary model (same score), despite algorithms typically producing a single result.
+Consequently, to account for evolutionary equivalent alignments, I scored all alignments using the marginal model and also considered perfect those with scores identical to the benchmark.
+Furthermore, I counted the number of alignments with the lowest distance $d_{seq}$ to the true alignment, including ties, reported as best alignments.
+Moreover, I computed the count of imperfect alignments, where an alignment is considered imperfect when its distance to the reference alignment is greater than zero ($d_{seq} > 0$) and another method successfully produced an alignment with 100\% similarity.
+This analysis exposes instances where all aligners fall short of achieving a perfect result in addition to a direct comparison.
+
+% \cyan{The first sentence intro is a good idea, but rewrite!}
+% Sequencing artifacts and errors made in alignment can lead to inaccurate results in genomics. This is especially true in studies of positive selection and positively selected genes.
+
+To evaluate how well the aligners were able to identify positive and negative selection, I estimated $k_s$ and $k_a$ statistics.
+$k_s$ and $k_a$ are, respectively, the number of substitutions per synonymous site (no changes at the amino acid level) and per non-synonymous site (introduces changes at the amino acid level) between two protein-coding genes.
+They are also denoted as $d_s$ and $d_n$ in the literature.
+I used the R package seqinr \citep{seqinr} to estimate these metrics, which follows the popular method put forth by Li \citep{ka_ks_li_1993}.
+First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate.
+A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous.
+Second, the nucleotide changes between the two sequences are counted and divided as transitional (A$\leftrightarrow$G, C$\leftrightarrow$T) and transversional (\{A, G\}$\leftrightarrow$\{C, T\}).
+Third, the Kimura two-parameter distance \citep{kimura1980simple} is used to estimate the number of transitions and transversions per site type (nondegenerate, twofold degenerate, and fourfold degenerate), which is used as a correction factor for multiple hits.
+Finally, $k_s$ is the estimate of the average transitional rate at twofold and fourfold degenerate sites, and $k_a$ is the estimate of the average transversional rate at nondegenerate and twofold sites.
+In the results, these metrics are reported as the F$_1$ score, which is the harmonic mean of precision (true positives over total positives) and recall (true positives over true positives and false negatives).
+This score ranges between 0 and 1, with a score of 1 representing a perfect result.
+
+In addition, we compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura 2-parameter model \citep{kimura1980simple}.
+The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions.
+The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions.
+The resulting distance provides a quantitative measure of the evolutionary divergence between the sequences.
+These calculations were performed using the R software package ape \citep{paradis2019ape}.

### 2023-12-21 results_discussion.tex
Add new fst figures

@@
 \section*{Results and Discussion}
-Using 16000 human genes and their gorilla orthologs from the ENSEMBL
-database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
-alignments with empirical gap patterns.
-We used the data set to evaluate the accuracy of COATi and a suite of popular aligners spanning multiple alignment methods:
-Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
-MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
-\citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}.
+% Using 16000 human genes and their gorilla orthologs from the ENSEMBL
+% database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
+% alignments with empirical gap patterns.
+% We used the data set to evaluate the accuracy of COATi and a suite of popular aligners spanning multiple alignment methods:
+% Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
+% MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
+% \citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}.
 
-After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and then aligned the remaining pairs with all five methods.
-At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7719 sequence pairs.
-We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
-The accuracy of inferred alignments compared to the benchmarks was measured using \citeauthor{metrics_blackburne_whelan_2011}'s \citeyearpar{metrics_blackburne_whelan_2011} distance metric, $d_{seq}$.
-Additionally, because different alignments may be evolutionary equivalent, we scored alignments using the codon-marginal-mg model, and any alignment that had the same score as the benchmark alignment was considered equivalent to the benchmark alignment.
-The accuracy of identifying positive and negative selection was calculated using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-\citembe{ka_ks_li_1993} (Supplementary Methods).
-The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
+% After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and then aligned the remaining pairs with all five methods.
+% At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7719 sequence pairs.
+% We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
+% The accuracy of inferred alignments compared to the benchmarks was measured using \citeauthor{metrics_blackburne_whelan_2011}'s \citeyearpar{metrics_blackburne_whelan_2011} distance metric, $d_{seq}$.
+% Additionally, because different alignments may be evolutionary equivalent, we scored alignments using the codon-marginal-mg model, and any alignment that had the same score as the benchmark alignment was considered equivalent to the benchmark alignment.
+% The accuracy of identifying positive and negative selection was calculated using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
+% \citembe{ka_ks_li_1993} (Supplementary Methods).
+% The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
+
+COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
+It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
+%
+In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}).
+Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi is substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).
+
+Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
+MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
+PRANK, using a codon model, had an average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
 
@@
  \vspace{1mm}
- \footnotesize{\textsuperscript{*}PRANK produced 50 empty alignments, calculations are based on 7669 alignments.}
+ \footnotesize{\textsuperscript{*}PRANK produced 42 empty alignments, calculations are based on 7719 alignments.}
  \caption{COATi generates better alignments than other alignment algorithms.
-Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7719 empirically simulated sequence pairs.
+Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7761 empirically simulated sequence pairs.
 Best alignments have the lowest $d_{seq}$ (including ties), perfect alignments have the same score as the true alignment, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
@@
 
-COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
-It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $2.1 \cdot 10^{-79}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
-%
-In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}).
-
-Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
-MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
-PRANK, using a codon model, had an average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
+To test how well COATi performs when the roles of reference and low-quality sequence are reverted, I aligned the 7761 simulated alignments using gorilla as the reference.
+Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons in the gorilla sequence on the remaining alignments.
+While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence.
+Despite this limitation, I analyzed the 4003 alignments and compared the results with all methods, including COATi using the human sequence as the reference.
+The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the ancestor in comparison to the reverse (Supplementary Materials Tab.
+6).
+However, the results for COATi continue to be a significant improvement over other aligners.
 
 Despite human and gorilla sequences having a relatively short evolutionary distance, COATi showed a biologically significant improvement over other methods, with an average alignment error nine-fold smaller than the next best method.
-COATi is an FST-based application that can calculate the optimal alignment
-between a pair of sequences in the presence of artifacts using a statistical
-model.
-Using COATI will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
+COATi is an FST-based application that can calculate the optimal alignment between a pair of sequences in the presence of artifacts using a statistical model.
+Using COATi will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
 

### 2023-12-21 supplementary_materials.Rmd
Add new fst figures

@@
 ---
-title: "Supplementary Materials for \" COATi: statistical pairwise alignment of protein coding sequences\""
+title: "Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\""
 author: Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright
@@
 k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7719 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment.
-"), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
 print(footnote(k, general = "PRANK produced 50 empty alignments, calculations are based on 7669 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
@@
 k = kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7678 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
 print(footnote(k, general = "PRANK produced 41 empty alignments, calculations are based on 7637 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
@@
 results = "supplementary_data/mar-mg/results/results_summary.csv"
-k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7661 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment.
-"), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 50 empty alignments, calculations are based on 7611 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7666 simulated sequence pairs.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 47 empty alignments, calculations are based on 7619 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
 #
 results = "supplementary_data/mar-ecm/results/results_summary.csv"
-k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7710 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 35 empty alignments, calculations are based on 7675 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7717 simulated sequence pairs.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 40 empty alignments, calculations are based on 7677 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
 ```

### 2023-12-21 alignpair_letter.tex
Add new fst figures
Refactor manuscript

@@
-\input{header.tex}
+% !TEX program=lualatex
+\documentclass[12pt,letterpaper]{article}
+
+\usepackage[letterpaper,margin=1in]{geometry}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{mathtools}
+\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
+\usepackage{fontspec}
+\usepackage{microtype}
+
+\setmainfont{TeX Gyre Pagella}
+\setmathfont{Latin Modern Math}
+
+\usepackage{url}
+\usepackage{graphicx}
+
+\usepackage[compact,tiny]{titlesec}
+
+\usepackage[authoryear,round]{natbib}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows,automata,positioning,shapes,decorations.text,backgrounds,matrix}
+\usepackage{wrapfig, framed}
+\usepackage[font=small,labelfont=bf]{caption}
+\usepackage[hidelinks]{hyperref}
+\usepackage{listings}
+\usepackage{relsize}
+\usepackage[left]{lineno}
+\usepackage{colortbl}
+\usepackage{multicol}
+
+\pagenumbering{arabic}
+\newcommand*\pct{\scalebox{.9}{\%}}
+\newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\green}[1]{\textcolor{green}{#1}}
+\newcommand{\cyan}[1]{\textcolor{cyan}{#1}}
+\newcommand{\grande}[1]{\mathlarger{\mathlarger{#1}}}
+
+\hypersetup{
+    colorlinks=true,
+    linkcolor=black,
+    urlcolor=blue,
+    citecolor=black
+}
+
+\newcommand{\TODO}[1]{\begingroup\bfseries\color{red}TODO:~#1\endgroup}
+
 % document begins here
 \begin{document}
-% \vspace*{0.35in}
-
 % title goes here:
 \begin{flushleft}
-{\Large\textbf{COATi: statistical pairwise alignment of protein coding sequences}}
+{\Large\textbf{COATi: statistical pairwise alignment of protein-coding sequences}}
 \newline
@@
 Ziqi Zhu\textsuperscript{1,3},
-Reed A. Cartwright\textsuperscript{1,3,*}
-\\
+Reed A. Cartwright\textsuperscript{1,3,*}\\
 \bigskip
-1 The Biodesign Institute, Arizona State University, Tempe, Arizona, USA
-\\
-2 Ira A.\ Fulton Schools of Engineering, Arizona State University, Tempe, Arizona, USA
-\\
-3 School of Life Sciences, Arizona State University, Tempe, Arizona, USA
-\\
+1 The Biodesign Institute, Arizona State University, Tempe, Arizona, USA\\
+2 Ira A.\ Fulton Schools of Engineering, Arizona State University, Tempe, Arizona, USA\\
+3 School of Life Sciences, Arizona State University, Tempe, Arizona, USA\\
 \bigskip
@@
 
-\input{abstract}
+\begin{abstract}
+\noindent
+Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
+Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
+For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
+While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain these artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
+To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
+COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments.
+\end{abstract}
 
@@
 
-\input{introduction}
+\section*{Introduction}
 
-\input{materials_methods}
+Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \citep{sequence_alignment_rosenberg_2009}.
+While sophisticated advancements have been made, the challenge of alignment inference has not been fully solved \citep{art_morrison_2015}.
+%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}.
+%
+The alignment of protein-coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \citep[e.g.][]{bininda2005transalign,abascal2010translatorx}.
+While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts, such as frameshifts and early stop codons.
+While some aligners can utilize codon substitution models, they are often not robust against coding-sequence artifacts.
+Additionally these aligners  force gaps to occur between codons, whereas in natural sequences, only about 42\% of indels occur between codons \citep{taylor2004occurrence, zhu2022profiling}.
+This mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).
 
-\input{results_discussion}
+% Figures are 86 mm or 178 mm wide
+\begin{figure}[h!]
+    \centering%
+    \includegraphics[scale=1]{figures/fig-aln.pdf}
+    \par
+    \caption{
+        Standard algorithms produce suboptimal alignments.
+        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
+        (b), (c), and (d) are the results of different aligners.
+        Nucleotide mismatches are highlighted in red. Phase 1, phase 2, and phase 3 indels are shown in purple, orange, and gray respectively.
+        Additionally, the orange indel is type II (an amino-acid indel plus an amino-acid change) while the purple indel is type I (an amino-acid indel only).
+        COATi is the only aligner able to retrieve the biological alignment in this example.
+        }
+    \label{fig:aln}
+\end{figure}
+
+Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citep{estimates_schneider_2009, effect_fletcher_2010, hubisz2011error}.
+While genomes for model organisms often get refined over many iterations and contain meticulously curated protein-coding sequences, 
+genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
+These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citep{jackman2018tigmint}.
+%
+When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
+current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
+
+To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets.
+
+\section*{Materials and Methods}
+
+\subsection*{Statistical Alignment}
+
+Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citep{bradley2007transducers}.
+Pair-HMMs are computational machines with two output tapes.
+Pair-HMMs contain a finite number of states---typically labeled match, insert, and delete---that emit symbols (nucleotides or amino acids) to one or both tapes.
+Each tape represents a sequence, and a path through a pair-HMM is a possible pairwise alignment.
+Conceptually, these machines generate two sequences ($\text{X}$ and $\text{Y}$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(\text{X}, \text{Y})$ \citep{yoon_2009_hmm}.
+
+A limitation of pair-HMMs is that they only model the evolution of two related sequences from an unknown ancestor.
+Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that they can model the generation of a descendant sequence given an ancestral one.
+FSTs consume symbols from an input tape and emit symbols to an output tape.
+Properly weighted, an FST can calculate the probability that a descendant sequence, $\text{Y}$, evolved from an ancestral sequence, $\text{X}$, represented by $P(\text{Y} | \text{X})$.
+% Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
+% A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
+% COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
+
+Furthermore, there are well-established algorithms for combining FSTs in different ways, including concatenation, composition, intersection, union, or reversal, allowing the design of complex models by combining simpler FSTs \citep{bradley2007transducers,silvestre2021machine}.
+Specifically, composition is a powerful and versatile algorithm for comparative sequence analysis, consisting of sending the output of one FST into the input of a second FST.
+Composition allows combining two or more FSTs to create a new, more complex transducer.
+Figure \ref{fig:transcription} illustrates how DNA transcription for a codon can be achieved by composing a complementing FST with a transducer that replaces thymines with uracil, where the three nucleotides are read and complemented in \ref{fig:transcription}-a, which are then used as the input of \ref{fig:transcription}-b, resulting in the transcription of the codon (Fig.~\ref{fig:transcription}-c).
+COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale=1]{figures/fig-fst-base-calling-new.pdf}
+    \caption[DNA Transcription via FST Composition]{DNA transcription via FST composition.
+The composition of (a) a codon complimenting FST and (b) an FST that replaces T with U generates (c) an FST that implements codon transcription ($a~\circ~b = c$).}
+    \label{fig:transcription}
+\end{figure}
+
+COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig.\ \ref{fig:evolution-fst}) \citep[c.f.][]{holmes2001evolutionary}.
+Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
+The assumption is that the reference sequence is in-phase, which is used to help preserve the reading frame and safeguard against possible frameshifts in the ``descendant''.
+Therefore, the high-quality sequence must not contain incomplete codons (the number of nucleotides is multiple of three) and be free of any ambiguous nucleotides or early stop codons.
+In contrast, the potentially lower-quality sequence has no such restrictions and can be of any length, contain early stop codons---treated as artifacts---, and include ambiguous codons.
+
+The Evolution FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
+A key innovation of this FST, with respect to others, is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.
+
+\begin{figure}[h!]
+\centering
+\setlength{\columnsep}{-3cm}
+\begin{multicols}{2}
+    \hspace*{-2.8cm}\resizebox{0.55\textwidth}{!}{\input{figures/fig-subst-fst}}
+    \columnbreak
+    \resizebox{0.55\textwidth}{!}{\input{figures/fig-indel-fst}}
+\end{multicols}
+\par
+\caption{The Evolution FST is assembled by composing a substitution FST and an indel FST.
+Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
+(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M.
+These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
+The weight of each arc is a conditional probability derived from a codon substitution model.
+(b) The indel FST allows for insertions (U to I) and deletions (V to D).
+Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
+On top of the indel FST, we add a base-calling-error FST (Supplemental Materials Figure 1) to model sequencing errors.
+Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
+\label{fig:evolution-fst}
+\end{figure}
+
+Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
+Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment.
+When equally optimal alignments exist, ties are broken according to the FST shortest-path algorithm.
+An example of an FST-based alignment can be found in Supplementary Materials Figure 2.
+All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
+However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
+This additional state space increases the computational complexity of the alignment algorithm.
+
+\section*{Codon Substitution Models}
+
+Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
+COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep{kosiol_ECM_2007} (codon-triplet-ecm).
+It also lets the user provide a codon substitution matrix.
+The default FST model (codon-triplet-mg) does not allow early stop codons in the ancestor sequence; although, it does support mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
+
+To reduce the runtime complexity of COATi, we have also developed an approximation of the Evolution FST that can be implemented with standard dynamic programming techniques.
+This approximation uses a marginal substitution model where the output nucleotides are independent of one another and only depend on the input codon and position.
+This produces a $\left(61 \times 3 \right) \times 4$ substitution model and eliminates the need to track dependencies between output nucleotides.
+
+A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position.
+%
+Specifically, let
+\[
+P_\text{cod}\left( Y_1 = y_1, Y_2 = y_2, Y_3 = y_3 |
+                   X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
+\]
+represent transition probabilities from a standard codon model, and
+%
+\[
+P_\text{mar}\left(Y_p = y \middle| X_1 = x_1, X_2 = x_2, X_3 = x_3 \right)
+=
+\sum_{y_1, y_2, y_3} I(y_p = y)
+P_\text{cod}\left( y_1, y_2, y_3 |
+                   x_1, x_2, x_3 \right)
+\]
+%
+represent the marginal transition probabilities, where
+$p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function.
+COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
+These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
+In combination with the indel model, alignment using the marginal model is implemented using dynamic programming.
+
+\subsection*{Empirical Simulation Algorithm}
+
+I downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
+database \citep{ensembl_hubbard_2002}.
+After downloading, I removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with all five methods.
+At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7761 sequence pairs.
+Then, I randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
+
+The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length.
+The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome.
+Assigning one of the sequences as the reference, to distinguish between insertions and deletions, CIGAR strings can also summarize pairwise alignments by grouping the number of contiguous matches or mismatches `M', deletions `D', and insertions `I'.
+The resulting pattern combines these letters preceded by the number of characters for each section as they appear in the alignment.
+This pattern is introduced by replacing nucleotides with gaps as indicated by deletions on one sequence and randomly introducing residues where the CIGAR strings indicated insertions.
+
+% Several safety checks are in place to ensure the algorithm runs correctly and the result is accurate.
+The assertions are divided into checking lengths and maintaining the reading frame of each section.
+The simulation can fail if the length of the sequences is different or if, without counting insertions, the length of the pattern to be inserted is longer.
+In addition, maintaining the phases of each section in the CIGAR string is important to avoid introducing errors such as frameshifts.
+
+I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
+% After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
+I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
+Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
+MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
+\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
+
+\subsection*{Metrics}
+
+To quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools, I used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}.
+This metric accounts for indels and is more informative than conventional distance scores like sum-of-pairs or total columns.
+Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the probability that a randomly selected residue will be aligned to a different location against a sequence that does not contain such residue.
+
+The computation of $d_{seq}$ involves characterizing the gaps present in the alignment.
+Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$.
+The distance between two alignments $A$ and $B$ is the average across all characters of the symmetric difference (or Hamming distance), represented as `$\triangle$' between homology sets over the length of such sets:
+
+\begin{equation}
+d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
+\end{equation}
+
+\noindent where $c$ is the sum of the sequence lengths.
+
+In addition, I compared the number of perfectly and imperfectly retrieved alignments for each aligner.
+Perfect alignments are defined as those with a distance of zero to the reference alignment ($d_{seq} = 0$), indicating 100\% similarity.
+Notably, a set of sequences can have more than one optimal alignment under the same evolutionary model (same score), despite algorithms typically producing a single result.
+Consequently, to account for evolutionary equivalent alignments, I scored all alignments using the marginal model and also considered perfect those with scores identical to the benchmark.
+Furthermore, I counted the number of alignments with the lowest distance $d_{seq}$ to the true alignment, including ties, reported as best alignments.
+Moreover, I computed the count of imperfect alignments, where an alignment is considered imperfect when its distance to the reference alignment is greater than zero ($d_{seq} > 0$) and another method successfully produced an alignment with 100\% similarity.
+This analysis exposes instances where all aligners fall short of achieving a perfect result in addition to a direct comparison.
+
+% \cyan{The first sentence intro is a good idea, but rewrite!}
+% Sequencing artifacts and errors made in alignment can lead to inaccurate results in genomics. This is especially true in studies of positive selection and positively selected genes.
+
+To evaluate how well the aligners were able to identify positive and negative selection, I estimated $k_s$ and $k_a$ statistics.
+$k_s$ and $k_a$ are, respectively, the number of substitutions per synonymous site (no changes at the amino acid level) and per non-synonymous site (introduces changes at the amino acid level) between two protein-coding genes.
+They are also denoted as $d_s$ and $d_n$ in the literature.
+I used the R package seqinr \citep{seqinr} to estimate these metrics, which follows the popular method put forth by Li \citep{ka_ks_li_1993}.
+First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate.
+A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous.
+Second, the nucleotide changes between the two sequences are counted and divided as transitional (A$\leftrightarrow$G, C$\leftrightarrow$T) and transversional (\{A, G\}$\leftrightarrow$\{C, T\}).
+Third, the Kimura two-parameter distance \citep{kimura1980simple} is used to estimate the number of transitions and transversions per site type (nondegenerate, twofold degenerate, and fourfold degenerate), which is used as a correction factor for multiple hits.
+Finally, $k_s$ is the estimate of the average transitional rate at twofold and fourfold degenerate sites, and $k_a$ is the estimate of the average transversional rate at nondegenerate and twofold sites.
+In the results, these metrics are reported as the F$_1$ score, which is the harmonic mean of precision (true positives over total positives) and recall (true positives over true positives and false negatives).
+This score ranges between 0 and 1, with a score of 1 representing a perfect result.
+
+In addition, we compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura 2-parameter model \citep{kimura1980simple}.
+The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions.
+The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions.
+The resulting distance provides a quantitative measure of the evolutionary divergence between the sequences.
+These calculations were performed using the R software package ape \citep{paradis2019ape}.
+
+\section*{Results and Discussion}
+% Using 16000 human genes and their gorilla orthologs from the ENSEMBL
+% database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
+% alignments with empirical gap patterns.
+% We used the data set to evaluate the accuracy of COATi and a suite of popular aligners spanning multiple alignment methods:
+% Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
+% MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
+% \citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}.
+
+% After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and then aligned the remaining pairs with all five methods.
+% At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7719 sequence pairs.
+% We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
+% The accuracy of inferred alignments compared to the benchmarks was measured using \citeauthor{metrics_blackburne_whelan_2011}'s \citeyearpar{metrics_blackburne_whelan_2011} distance metric, $d_{seq}$.
+% Additionally, because different alignments may be evolutionary equivalent, we scored alignments using the codon-marginal-mg model, and any alignment that had the same score as the benchmark alignment was considered equivalent to the benchmark alignment.
+% The accuracy of identifying positive and negative selection was calculated using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
+% \citembe{ka_ks_li_1993} (Supplementary Methods).
+% The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
+
+COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
+It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
+%
+In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}).
+Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi is substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).
+
+Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
+MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
+PRANK, using a codon model, had an average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
+
+% Software comparison table
+\begin{table}[!ht]
+\centering
+\input{figures/table-comp}
+ \vspace{1mm}
+ \footnotesize{\textsuperscript{*}PRANK produced 42 empty alignments, calculations are based on 7719 alignments.}
+ \caption{COATi generates better alignments than other alignment algorithms.
+Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7761 empirically simulated sequence pairs.
+Best alignments have the lowest $d_{seq}$ (including ties), perfect alignments have the same score as the true alignment, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
+ \label{table:comp}
+\end{table}
+
+To test how well COATi performs when the roles of reference and low-quality sequence are reverted, I aligned the 7761 simulated alignments using gorilla as the reference.
+Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons in the gorilla sequence on the remaining alignments.
+While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence.
+Despite this limitation, I analyzed the 4003 alignments and compared the results with all methods, including COATi using the human sequence as the reference.
+The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the ancestor in comparison to the reverse (Supplementary Materials Tab.
+6).
+However, the results for COATi continue to be a significant improvement over other aligners.
+
+Despite human and gorilla sequences having a relatively short evolutionary distance, COATi showed a biologically significant improvement over other methods, with an average alignment error nine-fold smaller than the next best method.
+COATi is an FST-based application that can calculate the optimal alignment between a pair of sequences in the presence of artifacts using a statistical model.
+Using COATi will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
+
+Future work include extending the indel FST to combine a 3-mer gap model with a frameshift parameter and weighing each indel phase differently to reflect known selection on indel phases \citep{zhu2022profiling}.
+We also plan on comparing the marginal and triplet models to evaluate the implications of the marginalization.
 
@@
 
+\TODO{Reviewers, Editor, and COmmittee members}
+
 \noindent \textit{Conflict of interest:} none declared.

### 2023-12-23 alignpair_letter.tex
Add new COATi FST figure.

@@
 
-Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citep{bradley2007transducers}.
+In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{Lunter2005-hk}.
+While approaches vary, a statistical aligner for a pair of sequences $X$ and $Y$ typically finds an alignment $A$ that maximizes the joint probability $P(X, Y, A)$ or samples alignments from the posterior $P(A | X, Y)$.
+This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{bradley2007transducers}.
 Pair-HMMs are computational machines with two output tapes.
-Pair-HMMs contain a finite number of states---typically labeled match, insert, and delete---that emit symbols (nucleotides or amino acids) to one or both tapes.
 Each tape represents a sequence, and a path through a pair-HMM is a possible pairwise alignment.
-Conceptually, these machines generate two sequences ($\text{X}$ and $\text{Y}$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(\text{X}, \text{Y})$ \citep{yoon_2009_hmm}.
+Conceptually, these machines generate two sequences from an unknown ancestor and can calculate the probability that two sequences are related, i.e.\ $P(X, Y)$ \citep{yoon_2009_hmm}.
+
+While the use of pair-HMMs is ubiquitous in bioinformatics, their
 
@@
     \centering
-    \includegraphics[scale=1]{figures/fig-fst-base-calling-new.pdf}
-    \caption[DNA Transcription via FST Composition]{DNA transcription via FST composition.
-The composition of (a) a codon complimenting FST and (b) an FST that replaces T with U generates (c) an FST that implements codon transcription ($a~\circ~b = c$).}
-    \label{fig:transcription}
+    \includegraphics[width=\textwidth]{figures/fig-fst-base-calling-new.pdf}
+    \caption[FSTs more the generation of sequences]{%
+    Finite state transducers (FSTs) model the generation of an output sequences based on an input sequence.
+    %
+    (a) A graph of a probabilistic FST \citep{cotterell-etal-2014-stochastic} for base-calling errors using a Mealy-machine architecture, where parameter $u$ is the error rate.
+This graph contains two states (S and M) connected by arcs, with labels ``input symbols~:~output symbols~/~weight''.
+Arcs consume symbols from the input sequence and emit symbols to the output sequence.
+Weights describe the probability that an arc is taken given the input symbols.
+Epsilon (ε) is a special symbol denoting that no symbols were either consumed or emitted.
+    %
+    (b) An FST for matching sequences against ambiguous nucleotides (N).
+This FST is not a true probabilistic FST and cannot be used to simulate output sequences since it is missing a parameter to control how often Ns are added to the output sequence.
+    %
+    (c) An FST that results from the composition ($∘$ operation) of the Error FST with the Ambiguity FST.
+As with (b), this composed FST cannot be used to simulate output sequences; however, it does properly weight the ambiguous nucleotide N as representing any other symbol.
+    %
+    %Let $x$ be a symbol in the input of (a), $y$ be a symbol in the output of (b), and $z$ be a symbol in both the output of (a) and the input of (b).
+An arc in the composed FST (c) has an input symbol $x$, an output symbol $y$, and a weight that is the combination over $z$ of the aggregate weights of all $x$-$z$-$y$ arc pairs.
+    %
+    }\label{fig:base-calling}
 \end{figure}

### 2023-12-23 figures/fig-fst-base-calling-new.tex
Add new COATi FST figure.

@@
 \usepackage{xcolor}
-\usepackage{xfrac}
 
-\usetikzlibrary{shapes,arrows,positioning,calc}
+\usetikzlibrary{shapes,arrows.meta,positioning,calc}
 \usetikzlibrary{decorations.markings}
@@
   block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2em,node distance=4em},
-  line/.style={draw, -stealth', thick},
+  line/.style={draw, -Stealth, thick},
   lab/.style={text = black, inner sep=1pt},
@@
 
-\node[block,right = of start] (s) {S}; 
-\node[block,left = of end] (m) {M};
+\node[block,right = of start] (m) {M}; 
+\node[block,left = of end] (s) {S};
 
-\draw[line] (start) to (s);
-\draw[line] (m) to (end);
+\draw[line] (start) to (m);
+\draw[line] (s) to (end);
 
@@
 \foreach \y [count=\yi] in {
-  {A:A/$(1-e)$}, {A:C/$(e/3)$}, {A:G/$(e/3)$}, {A:T/$(e/3)$},
-  {C:A/$(e/3)$}, {C:C/$(1-e)$}, {C:G/$(e/3)$}, {C:T/$(e/3)$}
+  {A:A/$(1-u)$}, {A:C/$(u/3)$}, {A:G/$(u/3)$}, {A:T/$(u/3)$},
+  {C:A/$(u/3)$}, {C:C/$(1-u)$}, {C:G/$(u/3)$}, {C:T/$(u/3)$}
 } {
@@
 \foreach \y [count=\yi] in {
-  {G:A/$(e/3)$}, {G:C/$(e/3)$}, {G:G/$(1-e)$}, {G:T/$(e/3)$},
-  {T:A/$(e/3)$}, {T:C/$(e/3)$}, {T:G/$(e/3)$}, {T:T/$(1-e)$}
+  {G:A/$(u/3)$}, {G:C/$(u/3)$}, {G:G/$(1-u)$}, {G:T/$(u/3)$},
+  {T:A/$(u/3)$}, {T:C/$(u/3)$}, {T:G/$(u/3)$}, {T:T/$(1-u)$}
 } {
@@
 \foreach \y [count=\yi] in {
-  {A:N/$1$}, {A:A/$(1-e)$}, {A:C/$(e/3)$}, {A:G/$(e/3)$}, {A:T/$(e/3)$},
-  {C:N/$1$}, {C:A/$(e/3)$}, {C:C/$(1-e)$}, {C:G/$(e/3)$}, {C:T/$(e/3)$}
+  {A:N/$1$}, {A:A/$(1-u)$}, {A:C/$(u/3)$}, {A:G/$(u/3)$}, {A:T/$(u/3)$},
+  {C:N/$1$}, {C:A/$(u/3)$}, {C:C/$(1-u)$}, {C:G/$(u/3)$}, {C:T/$(u/3)$}
 } {
@@
 \foreach \y [count=\yi] in {
-  {G:N/$1$}, {G:A/$(e/3)$}, {G:C/$(e/3)$}, {G:G/$(1-e)$}, {G:T/$(e/3)$},
-  {T:N/$1$}, {T:A/$(e/3)$}, {T:C/$(e/3)$}, {T:G/$(e/3)$}, {T:T/$(1-e)$}
+  {G:N/$1$}, {G:A/$(u/3)$}, {G:C/$(u/3)$}, {G:G/$(1-u)$}, {G:T/$(u/3)$},
+  {T:N/$1$}, {T:A/$(u/3)$}, {T:C/$(u/3)$}, {T:G/$(u/3)$}, {T:T/$(1-u)$}
 } {

### 2023-12-23 figures/fig-fst-coati.tex
Add new COATi FST figure.

% !TEX program=lualatex
\documentclass[tikz]{standalone}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
\usepackage{fontspec}
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{xcolor}

\usetikzlibrary{shapes,arrows.meta,positioning,calc}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.text}
%\usetikzlibrary{hobby,spath3,intersections}

\setmainfont{TeX Gyre Heros}
\setmathfont{Latin Modern Math}

%%% Okabe Ito set
\definecolor{colorO}{RGB}{230,159,0}
\definecolor{colorSb}{RGB}{86,180,233}
\definecolor{colorBg}{RGB}{0,158,115}
\definecolor{colorY}{RGB}{240,228,66}
\definecolor{colorBu}{RGB}{0,114,178}
\definecolor{colorV}{RGB}{213,94,0}
\definecolor{colorRp}{RGB}{204,121,167}
\definecolor{colorBk}{RGB}{0,0,0}
\definecolor{colorGy}{RGB}{153,153,153} % GRAY

% (86 mm) or double (178 mm) column width

\tikzset{
  end/.style={fill=none, text=black, inner sep=1pt},
  block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2em,node distance=4em},
  line/.style={draw, -Stealth, thick},
  lab/.style={text = black, inner sep=1pt},
  title/.style={text = black, font=\bfseries}
}

\begin{document}
\fontsize{8}{9}\selectfont
\begin{tikzpicture}

\node[end, anchor=west] at (0,0) (start) {start};
\node[end, anchor=east] at (86mm,0) (end) {end};

\node[block,right = of start] (m) {M}; 
\node[block,left = of end] (s) {S};

\draw[line] (start) to (m);
\draw[line] (s) to (end);

\coordinate (mid) at ($(s.center)!0.5!(m.center)$);

\draw[line] (m) to (s);
\node[lab,fill=white] at (mid) {ε:ε/$1$};

\foreach \y [count=\yi] in {
   {AAA:AAT/$P(AAT|AAA)$}, {AAA:AAG/$P(AAG|AAA)$},
   {AAA:AAC/$P(AAC|AAA)$}, {AAA:AAA/$P(AAA|AAA)$}
 } {
     \coordinate (y) at ($(mid) + (0em, -0.25em) + \yi*(0, 1.75em)$);
     \node[lab,anchor=south] (lab-\yi) at (y) {\y};
}

\path[line] (s) to[out=160,in=0] (lab-1.south) to[out=180,in=20] (m);
\path[line] (s) to[out=140,in=0,looseness=1.2] (lab-2.south)
                to[out=180,in=40,looseness=1.2] (m);
\path[line] (s) to[out=120,in=0,looseness=1.2] (lab-3.south)
                to[out=180,in=60,looseness=1.2] (m);
\path[line] (s) to[out=100,in=0,looseness=1.3] (lab-4.south)
                to[out=180,in=80,looseness=1.3] (m);

%%%%%%%%%%%%%%%%%%%%

\draw (mid) + (0em, -2em) coordinate (y);

\path[line,draw=darkgray,dotted,
    postaction={-,dash={on 9.4em off 24em phase -4.8em},
      draw=white,line width=2pt},
    postaction={decorate,decoration={text along path,
      text align=center,text color=darkgray,raise=-0.5ex,
      reverse path,
      text={|\itshape|3717 additional arcs}}}
] (s)
  to[out=200,in=0,looseness=1] (y)
  to[out=180,in=-20,looseness=1] (m);

\draw (start.west) + (0, 9.25em) node[title,anchor=west] (subfig-a) {a) Codon Substitution FST};

%%%%%%%%%%

\node[block, anchor=west] at (92mm,0mm) (m) {M};
\node[end, anchor=east] at (178mm,0mm) (end) {end};

\node[end,above = 2em of m] (start) {start}; 
\node[block,left = 4.2em of end] (s) {S};

\draw (m.west) + (0, 9.25em) node[title,anchor=west] (subfig-b) {b) Indel FST};

\draw[line] (start) to (m);
\draw[line] (s) to (end);

\coordinate (mid) at ($(s.center)!0.5!(m.center)$);

\node[block] (w) at (mid) {W};

\node[block] (h) at ($(m.center)!0.9! -90:(w.center)$) {H};
\node[block] (i) at ($(w.center)!0.9!  90:(m.center)$) {I};

\node[block] (c) at ($(w.center)!0.5!  90:(s.center)$) {C};
\node[block] (d) at ($(s.center)!0.5! -90:(w.center)$) {D};

\draw[line] (m) to node[lab,above] {ε:ε/$(1-g)$} (w);
\draw[line] (m) to node[lab,left] {ε:ε/$g$} (h);

\draw[line] (w) to node[lab,above] {ε:ε/$(1-g)$} (s);
\draw[line] (w) to node[lab,left] {ε:ε/$g$} (c);

%% Insertions

\draw[line] (i) to node[lab,right,pos=0.1] {ε:ε/$(1-e)$} (w);
\draw[line] (i) to node[lab,fill=white] {ε:ε/$e$} (h);
\draw[line] (h) to[bend left=40] node[lab,above] {ε:A/$\pi_{\text{A}}$} (i);
\draw[line] (h) to[bend left=15] node[lab,above] {ε:C/$\pi_{\text{C}}$} (i);
\draw[line] (h) to[bend right=15] node[lab,below] {ε:G/$\pi_{\text{G}}$} (i);
\draw[line] (h) to[bend right=40] node[lab,below] {ε:T/$\pi_{\text{T}}$} (i);

%% Deletions

\draw[line] (d) to node[lab,right] {ε:ε/$(1-e)$} (s);
\draw[line] (d) to node[lab,fill=white] {ε:ε/$e$} (c);
\draw[line] (c) to[bend left=40] node[lab,above] {A:ε/$1$} (d);
\draw[line] (c) to[bend left=15] node[lab,above] {C:ε/$1$} (d);
\draw[line] (c) to[bend right=15] node[lab,below] {G:ε/$1$} (d);
\draw[line] (c) to[bend right=40] node[lab,below] {T:ε/$1$} (d);

%% Substitutions

\draw[white,line width=2pt] ($(w)!0.245!(i)$) -- +(0,4pt);
\draw[white,line width=2pt] ($(w)!0.385!(i)$) -- +(0,4pt);
\draw[white,line width=2pt] ($(w)!0.525!(i)$) -- +(0,4pt);
\draw[white,line width=2pt] ($(w)!0.665!(i)$) -- +(0,4pt);

\coordinate (suba) at ($(w.center)!0.30! 20:(i.center) +(2em,0.5em)$);
\coordinate (subc) at ($(w.center)!0.45! 20:(i.center) +(2em,0.5em)$);
\coordinate (subg) at ($(w.center)!0.60! 20:(i.center) +(2em,0.5em)$);
\coordinate (subt) at ($(w.center)!0.75! 20:(i.center) +(2em,0.5em)$);

\draw[line] (s) to[out=225,in=0] (suba) to[out=180,in=350] (m);
\draw[line] (s) to[out=240,in=0] (subc) to[out=180,in=335] (m);
\draw[line] (s) to[out=255,in=0] (subg) to[out=180,in=320] (m);
\draw[line] (s) to[out=270,in=0] (subt) to[out=180,in=305] (m);

\node[lab,fill=white] at (suba) {A:A/$1$};
\node[lab,fill=white] at (subc) {C:C/$1$};
\node[lab,fill=white] at (subg) {G:G/$1$};
\node[lab,fill=white] at (subt) {T:T/$1$};

%%%%%%%%%%%%%%%%%

\node[lab,anchor=west] (intape) at (0mm,-25mm)  {input};
\node[lab,anchor=east] (outtape) at (86mm,-25mm)  {output};

\draw (intape.west) + (0, 4em) node[title,anchor=west] (subfig-c) {c) COATi FST};

\draw (intape.center)
    ++(10em, 0em) coordinate (b)
    ++(5em, 0em) coordinate (c)
    ++(5em, 0em) coordinate (d)
;

\draw[line] (intape) to node[lab,anchor=base,yshift=0.5em,xshift=-2pt] {Substitution} (b);
\draw[line] (b) to node[lab,anchor=base,yshift=0.5em,xshift=-2pt] {Indel} (c);
\draw[line] (c) to node[lab,anchor=base,yshift=0.5em,xshift=-2pt] {Error} (d);
\draw[line] (d) to node[lab,anchor=base,yshift=0.5em,xshift=-2pt] {Ambiguity} (outtape);

\draw[thick,decorate,decoration={brace,amplitude=5pt,mirror,raise=1em}]
(intape) to node[lab,yshift=-2em, anchor=north] {COATi FST} (outtape);

\end{tikzpicture}
\end{document}

### 2023-12-24 alignpair_letter.tex
Continue working on the method section.

@@
 
-\subsection*{Statistical Alignment}
+\subsection*{Statistical Alignment via Finite State Tranducers}
 
 In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{Lunter2005-hk}.
-While approaches vary, a statistical aligner for a pair of sequences $X$ and $Y$ typically finds an alignment $A$ that maximizes the joint probability $P(X, Y, A)$ or samples alignments from the posterior $P(A | X, Y)$.
+An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies.
+While approaches vary, a statistical aligner for a pair of sequences, $X$ and $Y$, typically finds an alignment, $Aln$, that maximizes the joint probability $P(Aln, X, Y)$ or samples alignments from the posterior $P(Aln | X, Y)$.
 This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{bradley2007transducers}.
 Pair-HMMs are computational machines with two output tapes.
-Each tape represents a sequence, and a path through a pair-HMM is a possible pairwise alignment.
-Conceptually, these machines generate two sequences from an unknown ancestor and can calculate the probability that two sequences are related, i.e.\ $P(X, Y)$ \citep{yoon_2009_hmm}.
+Each tape represents a sequence, and a path through the pair-HMM represents an alignment of the two sequences.
+Conceptually, Pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.
 
-While the use of pair-HMMs is ubiquitous in bioinformatics, their
+While the use of pair-HMMs is ubiquitous in bioinformatics, they are limited to modeling the evolution of two related sequences from an unknown ancestor.
+As an alternative, finite-state transducers (FSTs, Fig.~\ref{fig:base-calling}) allow researchers to model the evolution of a descendant sequence from an ancestral sequence.
+FSTs are computational machines with one input tape and one output tape and provide similar benefits to pair-HMMs, while being more suitable for evolutionary models \citep{bradley2007transducers}.
+FSTs consume symbols from an input tape and emit symbols to an output tape based on the symbols consumed and the structure of the FST.
+Conceptually, FSTs generate a descendant sequence, $D$, from a known ancestor, $A$, and can calculate the conditional probability $P(Aln, D | A)$.
 
-A limitation of pair-HMMs is that they only model the evolution of two related sequences from an unknown ancestor.
-Finite-state transducers (FSTs) have similar benefits to pair-HMMs with the additional feature that they can model the generation of a descendant sequence given an ancestral one.
-FSTs consume symbols from an input tape and emit symbols to an output tape.
-Properly weighted, an FST can calculate the probability that a descendant sequence, $\text{Y}$, evolved from an ancestral sequence, $\text{X}$, represented by $P(\text{Y} | \text{X})$.
-% Furthermore, well-established algorithms for combining FSTs in different ways allow the design of complex models by combining simpler FSTs \citembe{bradley2007transducers}.
-% A powerful and versatile algorithm for comparative sequence analysis is composition, which consists of sending the output of one FST into the input of a second FST.
-% COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
-
-Furthermore, there are well-established algorithms for combining FSTs in different ways, including concatenation, composition, intersection, union, or reversal, allowing the design of complex models by combining simpler FSTs \citep{bradley2007transducers,silvestre2021machine}.
-Specifically, composition is a powerful and versatile algorithm for comparative sequence analysis, consisting of sending the output of one FST into the input of a second FST.
-Composition allows combining two or more FSTs to create a new, more complex transducer.
-Figure \ref{fig:transcription} illustrates how DNA transcription for a codon can be achieved by composing a complementing FST with a transducer that replaces thymines with uracil, where the three nucleotides are read and complemented in \ref{fig:transcription}-a, which are then used as the input of \ref{fig:transcription}-b, resulting in the transcription of the codon (Fig.~\ref{fig:transcription}-c).
-COATi uses composition to derive a statistical alignment model from the combination of smaller FSTs, each representing a specific process.
+There are well-established algorithms for combining FSTs in different ways allowing the design of complex models by combining simpler FSTs, including concatenation, composition, intersection, union, and reversal \citep{bradley2007transducers,silvestre2021machine}.
+Specifically, composition is an algorithm to combine two FSTs by sending the output of one FST into the input of another, creating a new, more complex transducer.
+Figure~\ref{fig:base-calling} illustrates how FSTs modeling sequencing errors (Fig.~\ref{fig:base-calling}a) and ambiguity (Fig.~\ref{fig:base-calling}b) can be combined via composition to produce an FST that does both (Fig.~\ref{fig:base-calling}c).
+Conceptually, composition creates an FST that generates a descendant sequence from a known ancestor via an unknown intermediate, $J$, and can calculate the conditional probability $P(D|A) = \sum_{J} P(D|J)P(J|A)$.
 
@@
 
-COATi implements the pairwise alignment of a potentially lower-quality sequence against a high-quality sequence as a path through the Evolution FST (Fig.\ \ref{fig:evolution-fst}) \citep[c.f.][]{holmes2001evolutionary}.
-Here, COATi treats the high-quality (reference) sequence as the ``ancestor'' and the potentially lower-quality sequence as the ``descendant''.
-The assumption is that the reference sequence is in-phase, which is used to help preserve the reading frame and safeguard against possible frameshifts in the ``descendant''.
-Therefore, the high-quality sequence must not contain incomplete codons (the number of nucleotides is multiple of three) and be free of any ambiguous nucleotides or early stop codons.
-In contrast, the potentially lower-quality sequence has no such restrictions and can be of any length, contain early stop codons---treated as artifacts---, and include ambiguous codons.
+\subsection*{The COATi FST}
 
-The Evolution FST is the result of composing a substitution FST that encodes a codon model (Fig.\ \ref{fig:evolution-fst}-a) and an indel FST that models insertions and deletions, including frameshifts (Fig.\ \ref{fig:evolution-fst}-b).
-A key innovation of this FST, with respect to others, is the combination of a codon substitution model with a nucleotide-based geometric indel model that allows gaps to occur at any position.
+COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}.
+We call this transducer the COATi FST.
+A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons.
+This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.
+
+Since codon substitution is the first process in COATi's model, the input sequence to the COATi FST must be compatible with a codon-substitution model, i.e.\ be a multiple of three nucleotides and not contain any ambiguous symbols or stop codons.
+The phases, reading frames, and amino-acid contexts of alignment columns is determined by the input sequence, and better alignments will be generated if the input is of high quality and free of artifacts.
+Depending on context, we may refer to the input sequence as the ``ancestral'' or ``reference'' sequence.
+In contrast, the output sequence must be compatible with the ambiguity FST, can be of any length, and can contain any nucleotides or ``N''.
+This allows COATi to align lower-quality sequences that may contain artifacts against a high-quality reference sequence.
+We may refer to the output sequence as the ``descendant'', ``non-reference'', or ``query'' sequence.
 
 \begin{figure}[h!]
-\centering
-\setlength{\columnsep}{-3cm}
-\begin{multicols}{2}
-    \hspace*{-2.8cm}\resizebox{0.55\textwidth}{!}{\input{figures/fig-subst-fst}}
-    \columnbreak
-    \resizebox{0.55\textwidth}{!}{\input{figures/fig-indel-fst}}
-\end{multicols}
-\par
-\caption{The Evolution FST is assembled by composing a substitution FST and an indel FST.
-Each node represents a state in an FST while arcs display possible transitions between states (and their weights). Unlabeled arcs have weights of 1.
+\includegraphics[width=\textwidth]{figures/fig-fst-coati.pdf}
+\caption{The COATi FST is built from simpler FSTs via composition.
 (a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M.
@@
 The weight of each arc is a conditional probability derived from a codon substitution model.
-(b) The indel FST allows for insertions (U to I) and deletions (V to D).
+See Fig.~\ref{fig:base-calling} for more details about reading this graph.
+%
+(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the ``gap opening parameter'' and $e$ is the ``gap extension'' parameter.
 Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
-On top of the indel FST, we add a base-calling-error FST (Supplemental Materials Figure 1) to model sequencing errors.
-Contiguous insertions and deletions are always arranged for insertions to precede deletions to limit equivalent alignments.}
-\label{fig:evolution-fst}
+This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions.
+%
+(c) The COATi FST is derived via composition from the codon substitution, indel, error, and ambiguity FSTs.
+}
+\label{fig:coati-fst}
 \end{figure}

### 2023-12-24 alignpair_letter.tex
Continue work on updating the methodology.

@@
 Pair-HMMs are computational machines with two output tapes.
-Each tape represents a sequence, and a path through the pair-HMM represents an alignment of the two sequences.
-Conceptually, Pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.
+Each tape represents one sequence, and a path through the pair-HMM represents an alignment of the two sequences.
+Conceptually, pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.
 
@@
 We call this transducer the COATi FST.
+Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
+COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
+It also lets the user provide a codon substitution matrix.
 A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons.
@@
 We may refer to the output sequence as the ``descendant'', ``non-reference'', or ``query'' sequence.
+The choice of which sequence is the input sequence and which sequence is the output sequence is left up to the user.
 
@@
 
-Composing both sequences with the Evolution FST results in the transducer of all possible alignments.
+In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}.
+By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences.
 Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment.
-When equally optimal alignments exist, ties are broken according to the FST shortest-path algorithm.
+If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm.
+All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
 An example of an FST-based alignment can be found in Supplementary Materials Figure 2.
-All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
-However, the Evolution FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
+
+\section*{The Marginal Model}
+
+The COATi FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
 This additional state space increases the computational complexity of the alignment algorithm.
-
-\section*{Codon Substitution Models}
-
-Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
-COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep{kosiol_ECM_2007} (codon-triplet-ecm).
-It also lets the user provide a codon substitution matrix.
-The default FST model (codon-triplet-mg) does not allow early stop codons in the ancestor sequence; although, it does support mutations to (early) stop codons under the assumption that these are artifacts common in low-quality data.
-
-To reduce the runtime complexity of COATi, we have also developed an approximation of the Evolution FST that can be implemented with standard dynamic programming techniques.
+To reduce the runtime complexity of COATi, we have also developed an approximation of the COATi FST that can be implemented with standard dynamic programming techniques.
 This approximation uses a marginal substitution model where the output nucleotides are independent of one another and only depend on the input codon and position.
@@
 A marginal substitution model is calculated from a standard substitution model by calculating the marginal probabilities that each ancestral codon produces specific descendant nucleotides at each reading frame position.
+Specifically, let
 %
-Specifically, let
 \[
@@
 \]
-represent transition probabilities from a standard codon model, and
+%
+represent transition probabilities from a codon model, and
 %
@@
 represent the marginal transition probabilities, where
-$p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I$ is an indicator function.
+$p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function.
 COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.

### 2023-12-28 alignpair_letter.tex
Check in revisions to methods.

@@
 
-The COATi FST has a large state space to keep track of codon substitution rates when codons might be interspersed with indel events.
+The COATi FST has a large state space to keep track of codon substitution rates when codons can be interspersed with indel events.
 This additional state space increases the computational complexity of the alignment algorithm.
@@
 %
-represent the marginal transition probabilities, where
-$p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function.
-COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg (default model) and codon-marginal-ecm.
+represent the marginal transition probabilities, where $p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function.
+COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg and codon-marginal-ecm.
 These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
@@
 
-I downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
+In order to compare COATi against other aligners on realistic datasets, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
+Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
 database \citep{ensembl_hubbard_2002}.
-After downloading, I removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with all five methods.
-At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7761 sequence pairs.
-Then, I randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
+After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: COATi, Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011}, MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
+\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
+Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps, and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.
 
-The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length.
+We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
+We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments.
+The phases of gaps were preserved as were the spacings of any clusters of gaps.
+Segments of matches that were 99 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
+(This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.) We deleted nucleotides from the gorilla sequences of ungapped pairs to create deletions, and we inserted random nucleotides according to their stationary frequency to the gorilla sequences to create insertions.
+The human sequences were unchanged.
+
+% The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length.
 The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome.
@@
 
-I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
-% After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
-I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
-Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
-MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
-\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
+% I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
+% % After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
+% I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
+% Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
+% MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
+% \citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
 
-\subsection*{Metrics}
+\subsection*{Alignment Accuracy Measures}
 
-To quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools, I used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}.
-This metric accounts for indels and is more informative than conventional distance scores like sum-of-pairs or total columns.
-Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the probability that a randomly selected residue will be aligned to a different location against a sequence that does not contain such residue.
+We used multiple statistics to to quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools.
 
-The computation of $d_{seq}$ involves characterizing the gaps present in the alignment.
+\paragraph{Alignment Error.}
+To quantify the error between estimated alignments and the benchmark alignments, we used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}.
+Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the fraction of nucleotides in the sequence pair that are aligned differently between estimated and benchmark alignments.
+This metric summarizes each alignment by building homology sets for each nucleotide in the sequence pair.
+Briefly, if an alignment column contains position $i$ from the first sequence and position $j$ from the second sequence, then $\{j\}$ is the homology set for position $i$ in the first sequence and $\{i\}$ is the homology set for position $j$ in the second sequence.
+If a position is aligned against a gap, then its homology set is simply $\{\text{gap}\}$.
+This treats all gaps equally as the location of the gap is not recorded.
+Finally, $d_{seq}$ is calculated  between an estimated and a benchmark alignment as the average, normalized Hamming distance between the homology sets for each nucleotide in the sequence pair.
+See \cite{metrics_blackburne_whelan_2011} for further details.
+
+% The computation of $d_{seq}$ involves characterizing the gaps present in the alignment.
 Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$.
@@
 
-\begin{equation}
-d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
-\end{equation}
+% \begin{equation}
+% d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
+% \end{equation}
 
-\noindent where $c$ is the sum of the sequence lengths.
+% \noindent where $c$ is the sum of the sequence lengths.
 
-In addition, I compared the number of perfectly and imperfectly retrieved alignments for each aligner.
-Perfect alignments are defined as those with a distance of zero to the reference alignment ($d_{seq} = 0$), indicating 100\% similarity.
-Notably, a set of sequences can have more than one optimal alignment under the same evolutionary model (same score), despite algorithms typically producing a single result.
-Consequently, to account for evolutionary equivalent alignments, I scored all alignments using the marginal model and also considered perfect those with scores identical to the benchmark.
-Furthermore, I counted the number of alignments with the lowest distance $d_{seq}$ to the true alignment, including ties, reported as best alignments.
-Moreover, I computed the count of imperfect alignments, where an alignment is considered imperfect when its distance to the reference alignment is greater than zero ($d_{seq} > 0$) and another method successfully produced an alignment with 100\% similarity.
-This analysis exposes instances where all aligners fall short of achieving a perfect result in addition to a direct comparison.
+\paragraph{Perfect, Imperfect, and Best Alignments.}
+We quantified the number of perfect, imperfect and best alignments each aligner produced.
+We define perfect alignments as alignments with a distance of zero to their benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero.
+Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from impacting our results.
+Evolutionary equivalence was determined by scoring alignments using COATi's marginal model.
+Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark was greater than zero.
+%
+Imperfect alignments are defined as alignments that are not perfect when another method successfully produces a perfect alignment for the same pair of sequences.
+%
+Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties.
+%
+Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.
 
@@
 
-To evaluate how well the aligners were able to identify positive and negative selection, I estimated $k_s$ and $k_a$ statistics.
-$k_s$ and $k_a$ are, respectively, the number of substitutions per synonymous site (no changes at the amino acid level) and per non-synonymous site (introduces changes at the amino acid level) between two protein-coding genes.
-They are also denoted as $d_s$ and $d_n$ in the literature.
-I used the R package seqinr \citep{seqinr} to estimate these metrics, which follows the popular method put forth by Li \citep{ka_ks_li_1993}.
-First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate.
+% Ks and Ka are, respectively, the number of substitutions per synonymous site and per non-synonymous site between two protein-coding genes.
+
+\paragraph{Selection.}
+Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
+To evaluate the impact of aligners on selection identification, we estimated $k_s$ and $k_a$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment.
+Briefly, $k_s$ is the number of substitutions per synonymous site and $k_a$ is the number of substitutions per non-synonymous site between two protein-coding sequences.
+We used the method of \cite{ka_ks_li_1993} to estimate these statistics as implemented in the R package seqinr \citep{seqinr}.
+%
+We considered any alignment to be showing evidence of positive selection if $k_a/k_s > 1$ and negative selection if $k_a/k_s < 1$.
+We estimated F\textsubscript{1} scores for both positive and negative selection by comparing estimated alignments to benchmark alignments.
+F\textsubscript{1} is the harmonic mean of precision and recall and allows us to measure how well aligners produce alignments that correctly identify or exclude positive selection or negative selection.
+
+% First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate.
 A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous.
@@
 
-In addition, we compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura 2-parameter model \citep{kimura1980simple}.
+\paragraph{Evolutionary Distance.}
+Alignment error also negatively impacts the estimation of evolutionary distances and phylogenetic trees.
+To quantify the impact of aligners on the estimation of evolutionary distances, we compared Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} calculated for the estimated and benchmark alignments.
+This distance corrects for multiple mutations at a site and takes into account differences in transition and transversion rates.
+It also assumes equal nucleotide frequencies and no variations in the rate of substitution across sites.
+While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
+We used the R software package ape \citep{paradis2019ape} to calculate K2P distances.
+
+% e compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura's 2-parameter distance \citep{kimura1980simple}.
 The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions.
 The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions.
-The resulting distance provides a quantitative measure of the evolutionary divergence between the sequences.
-These calculations were performed using the R software package ape \citep{paradis2019ape}.
 

### 2023-12-29 alignpair_letter.tex
COmmit completed first round of manuscript changes. I ended up touching just about everything to make it feel more like a research paper and less like an application note.

@@
 It also lets the user provide a codon substitution matrix.
-A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons.
+A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[c.f.][]{ranwez_macse_2011,ranwez_macse_2018}.
 This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.
@@
 database \citep{ensembl_hubbard_2002}.
-After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: COATi, Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011}, MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
-\citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
+After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ v1.2.4 via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE v2.06, (4) the DNA model of MAFFT v7.505, and (5) the codon model of PRANK v.150803 \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}.
 Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps, and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.
@@
 
-COATi, using the codon-triplet-mg model, obtained better results compared to a wide variety of alignment strategies.
+COATi obtained better results compared to a wide variety of alignment strategies.
 It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
-%
+Notably, the average alignment error of the second best protocol was six times larger than COATi's.
 In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}).
-Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi is substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).
+Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi was substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).
 
-Clustal$\Omega$ generated alignments via amino acid translations and obtained the highest average alignment error while having difficulties retrieving positive selection.
-MACSE used a DNA-AA hybrid model, allowing frameshifts, and obtained similar results to MAFFT using a DNA model.
-PRANK, using a codon model, had an average alignment error between MACSE/MAFFT and Clustal$\Omega$ but was unable to generate alignments for some sequence pairs.
+Aligning sequences using amino-acid translation via ClustalΩ obtained the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
+These results are not surprising because alignments in amino-acid space only permit phase-3 gaps and are sensitive to frameshift artifacts, which inflate divergence estimates and impact selection detection.
+Conversely, we obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT.
+These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts.
+If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments.
+And finally, our PRANK protocol used a codon substitution model which should have been an improvement over both amino-acid and nucleotide models; however, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
+In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.
 
@@
 
-To test how well COATi performs when the roles of reference and low-quality sequence are reverted, I aligned the 7761 simulated alignments using gorilla as the reference.
-Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons in the gorilla sequence on the remaining alignments.
+COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
+To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference.
+Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons or other artifacts in 3758 simulated gorilla sequences.
 While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence.
-Despite this limitation, I analyzed the 4003 alignments and compared the results with all methods, including COATi using the human sequence as the reference.
-The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the ancestor in comparison to the reverse (Supplementary Materials Tab.
+Despite this limitation, we analyzed the 4003 alignments and compared the results across methods, including COATi using the human sequence as the reference.
+The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the reference in comparison to the reverse (Supplementary Materials Tab.
 6).
@@
 
-Despite human and gorilla sequences having a relatively short evolutionary distance, COATi showed a biologically significant improvement over other methods, with an average alignment error nine-fold smaller than the next best method.
-COATi is an FST-based application that can calculate the optimal alignment between a pair of sequences in the presence of artifacts using a statistical model.
-Using COATi will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
+COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model.
+While COATi has multiple modes and models, here we have focused on evaluating the utility of COATi FST for estimating pairwise alignments.
+We have shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances in the presence of genomic artifacts.
+In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
 
-Future work include extending the indel FST to combine a 3-mer gap model with a frameshift parameter and weighing each indel phase differently to reflect known selection on indel phases \citep{zhu2022profiling}.
-We also plan on comparing the marginal and triplet models to evaluate the implications of the marginalization.
+COATi is under active development.
+We plan on extending it to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
+We plan on improving its multiple-sequence alignment and alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
+Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
 
@@
 This research was funded by NSF award DBI-1929850.
-
-\TODO{Reviewers, Editor, and COmmittee members}
+%
+The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Oskan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects.
+The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.
 

### 2024-01-01 response_r1.md
Begin writing response to reviewers and updating manuscript further as we respond.

---
title: Response to Reviewers
subtitle: "COATi: statistical pairwise alignment of protein-coding sequences"
author:
	- Juan J. García Mesa
	- Ziqi Zhu
	- Reed A. Cartwright
mainfont: TeX Gyre Pagella
geometry:
	- margin=1in
---

<!-- pandoc --pdf-engine=lualatex -o response_r1.pdf response_r1.md -->

We would like to thank the associate editor and two reviewers for their helpful comments and suggestions. We have addressed all of the concerns brought to our attention as described below.

## Associate Editor

 1. Comment: *The general consensus about this paper is that the question and method (codon-aware alignment) are both extremely important and suitable for MBE, but the manuscript, written essentially as a short software note, is simply too sparse to adequately present the solution, eliding over important issues and methodological details.*

    We thank the editor and reviews for the positive feedback. As requested we have broaden the scope of the manuscript from a software note to a larger research paper about our COATi software. Nearly all of the paper has been rewritten, expanded, and reconfigured to provide more details and background about our work. If certain concepts need additional exposition, we will be happy to make those adjustments.

 1. Comment: *In addition to the reviewers' comments, one more critical question strikes me. The method makes uses of a simplification procedure predicated on the concept of treating one sequence as if it is the ancestor of the other, which allows modeling the problem using FSTs rather than HMMs. This modeling choice introduces additional directionality into the model. If there are two sequences (A and B) the method arbitrarily choses one for the ancestor and one for the descendent. What is not at all clear is whether this directionality is important, i.e., is the same result generated if the model goes A-->B as it would if it went B-->A. This is a fundamentally critical question that needs to be addressed as the stability of any alignment method should not be dependent on the order of the sequences within the input file.*

    The editor is correct that our model is directional. This is not a limitation of the model but rather a feature. We have added more details about our FSTs and our reasoning for using them. In particular, we now explain that our input sequence (i.e. "the ancestor") is required to have a length that is a multiple of 3 and missing stop codons and ambiguous nucleotides (circa line 84). The output sequence (i.e. "the descendant") does not have these restrictions. This asymmetry is due to COATi needing one high-quality reference sequence to establish reading-frame contexts.

    Our primary results are presented using an empirically simulated benchmark derived from human-gorilla sequence pairs. These simulations were not intended to create gorilla sequences that were compatible with the restrictions that COATi places on input sequences. Therefore, we were only able to swap the order of ~50% of the benchmark sequence pairs to measure the impact of directionality on our results. The results show a decrease, albeit small, in accuracy across all metrics when the lower quality sequence is used as the reference. We now report these results in the main paper and supplementary materials.

    We have thought further about how to reduce the dependency the directionality of our model, but we do not have any improvements to report at this time.

## Reviewer 1

 1. Comment: *It is surprising this work has not been extended to something larger*

    We appreciate the reviewer's suggestion that our ideas may be suitable for longer sequences. We feel that COATi, with a bit of work, would be the correct tool for aligning long-read sequences generated from target genes against a reference. Our existing suite of FSTs can support both exonic and intronic regions, and via concatenation we can create a new FST that supports specific reference genes' exon and intron segments. The challenge is to figure out the best user-interface for specifying such information as well as validating the result. We choose to tackle these challenges in a future research project.

    We now comment on this in the discussion: "COATi is under active development.... We plan on ... implement[ing] new models for aligning long-read sequences of genes against reference genomes."

## Reviewer 2

1. Comment: *I would like to start this review with an apology. This review took me a bit longer than I had hoped.*

    We also apologize for taking longer than expected to submit this revision, as the first author needed to focus on defending his dissertation last semester. We are happy to report that he passed his defense without issue and is now Dr. García Mesa, joining Dr. Zhu who graduated the year before.

1. Comment: *With that said, I would like to start by saying that Garcia Mesa and colleagues have taken on an important problem that has (in my opinion) received substantially less attention than it should – the problem that sequence alignment for coding regions ignores the potential for indels to occur out of phase with codon boundaries. I find the approach that the authors propose to be exciting and potentially very useful. I also believe that MBE is an appropriate venue for this work, and I think MBE readers will be excited.*

    Thank you for the encouraging words.

 1. Comment: *However, I also feel that the manuscript is hard to read, and I think it requires substantial revision for clarity before it is acceptable.*

    We have taken the opportunity to do a substantial revision of the manuscript, including aiming for a larger research paper, instead of a brief note. He hope that it is now easier to read and answers the concerns of the reviewers. If further details are needed, we are happy to include them.

 1. Comment: *I will start with the good part of the manuscript – the problem is stated clearly in the introduction. Figure 1 is excellent, and it illustrates the fact that codon-based aligners can make the error of forcing indels to lie between codons.*

    Thank you for the encouraging words.

 1. Comment: *I am confused by the state space. Perhaps they could be enumerated in the supplementary information. Obviously, there are 61 possible codon states (assuming the standard code). It strikes me that there are 12 possible 1 nucleotide, 2 gap states (i.e., N--, -N-, and --N, where N is any nucleotide) and 48 possible 2 nucleotide, 1 gap states (i.e., NN-, N-N, and -NN). I would assume that -N- and N-N would be very rare, but I am fine with including them for the purpose of completeness. There is also the all-gap state, which brings the total up to 122 states.*

    Our FSTs do not have the same state space concept as in phylogenetic models. Consider the codon substitution FST (Fig. 3a). Its input and output sequences have state spaces of four symbols (A, C, G, and T). This FST has two internal states (M and S) and 61x61 arcs from S to M. Each of these 3721 arcs consume three nucleotides from the input sequence, emit three nucleotides to the output sequence, and represent one transition in a standard codon substitution model.

    Similarly, consider the ambiguity FST (Fig 2b). Its input sequence has four symbols (A, C, G, and T) and its output sequence has five symbols (A, C, G, T, and N). It has two internal states (M and S), and 8 arcs from S to M. Four of these arc consume a nucleotide and emit the same nucleotide. The other four arcs consume a nucleotide and emit N. This is the FST that allows COATi to support ambiguous nucleotides in the non-reference/output sequence.

    We hope the updated manuscript offers clarity about these points.

 1. Comment: *I can see how one can forbid ungapped emission of stop codons. However, I could see problematic cases like …TACAAG… (…Tyr-Lys…) where a deletion of three bases after the first T -or- after the first A would yield a stop codon. Is that forbidden? Or is it allowed in the “low quality sequence” in the pair?*

    Our Indel FST does not consider codon translations when deleting or inserting nucleotides. Indels can occur that create early stop codons. However, this will only happen if the output sequence is of low quality and contains an early stop codon. We consider this a feature of COATi as it will not fail to align output sequences that contain early stop codons.

 1. Comment: *I don’t see how they can use a codon model when indels can occur within codons. This is especially true for the Kosiol ECM. Since the ECM is purely empirical it can be viewed as a matrix of rates of change between 61 symbols. I think this should also be true for the Muse and Gaut codon model.*

    One of the advantages of using FSTs and the composition algorithm is that it handles this for us. TODO...

 1. Comment: *It might be possible to improve Fig. 2 to clarify this. It is very difficult to work through, and I am still having trouble despite staring at the figure quite a bit. I wonder if showing the example alignment for Fig. 1 might help. Perhaps something could be done in the supplement.*

    We have updated Figure 2 and added a new Figure 3 that we feel more adequately explains the finite-state transducers that we are using and the composition algorithm that allows us to combine them in a reasonable manner. We also include in the supplement .... TODO

 1. Comment: *Finally, how is the “low quality sequence” in the pair identified?*

    COATi uses the first sequence in the input file as the input/reference sequence, and the second sequence as the output/non-reference sequence. This gives users the power to place low-quality sequences as the output sequence when aligning a pair of sequences using COATi FST. We now mention this in the manuscript circa line 95.

 1. Comment: *Perhaps one of the biggest problems with the methods is that the first two paragraphs in the results cover a lot of methods. However, they cover them in an abbreviated format that is almost impossible to follow. For example, I think that the way indels are simulated are by taking alignments of empirical human-gorilla sequence pairs and then mapping the gaps from those pairs onto simulated sequences, where those sequences were simulated without gaps. Is that correct? I think the actual methods need to be spelled out much more carefully and put in the methods section.*

    We have added details about how we empirically simulated benchmark alignments (circa line 137). The revision now contains information about our procedure, as well as an example of it being applied to data. We believe that this update answers the concerns of the reviewer.

 1. Comment: *In this context, it would be good to state how Ka and Ks are calculated in the methods. The supplement has one sentence about how they are calculated and a lot about what Ka/Ks > 1 means. The material about interpretation is not a problem, but I think many MBE readers will not need it. I’d really like to see an explicit articulation of how the stats were collected in the main paper.*

    We used sequinr to calculate Ka and Ks and have expanded our discussion to expain which method that sequinr uses to calculate these stats and how that method works (briefly).

 1. Comment: *I would also like to authors to articulate something about equally optimal alignments (or slightly suboptimal alignments). The github page has information on “coati sample - align two sequences and sample alignments” and it seems to me that equally optimal alignments might be common. In fact, if one looks at the biological alignment in Fig. 1 the second indel:*

    ```
    …T-- -CG… (Ser)
    …TCC CAG… (Ser-Gln)
    ```

    *Could just as easily be:*

    ```
    …TC- --G… (Ser)
    …TCC CAG… (Ser-Gln)
    ```

    *In fact, this alternative alignment seems better. Minimally, the first alignment requires a three-base pair indel and a C<->A change whereas the second would only require a three-base pair indel.*

    Thank you for pointing out a typo in our Figure 1. It was supposed to be TCC CCG (Ser-Pro) in the bottom sequence there. (No C<->A change.) We've fixed it in the revision.

    Back to your question, `coati alignpair` breaks ties according to the implementation of the "shortest path" algorithm in the OpenFST library (when using an FST-based model like tri-mg). We currently use OpenFST to implement our FST-based algorithms. We now mention this in the manuscript (circa line 100).

    Furthermore, `coati sample` doesn't need to break ties as it randomly samples alignments from the posterior. This is useful for studies that need a better picture of alignment space than retrieving a single, best alignment. We've utilized it to study indels across the tree of life (e.g. Zhu 2022), and it will be described in greater detail in a future paper.

 1. Comment: *Finally, I would like to see the impact of alternative alignments on distances – perhaps logdet or K2P or F84 distances. It might be informative to ask how often each alignment method results in the largest distance. It strikes me that methods that “overalign” (e.g., their Fig. 1c) would tend to have the largest distances.*

    We have quantified each aligner's impact on K2P distance estimation. We report these results as the percent of benchmark alignments which aligners produce overestimated distances. COATi overestimates only about 11% of benchmarks and is better than the other methods.

 1. Comment: *It would also be nice to summarize Ka/Ks and evolutionary distances for empirical human-gorilla aligiments. It strikes me that they already have the data, having used it to get parameters for simulation.*

    TODO...

 1. Comment: *Minor (but still important) issue: The values in the supplementary information tables are reported to five decimal points. It is a judgement call whether this is overkill for the proportions, but it is very distracting for integer counts (i.e., just use “5678” rather than “5678.00000” for the number of perfect alignments)*

    TODO...

 1. Comment: *I hope the authors are given a chance to revise this. In my opinion, the approach has a lot of potential. But I also think it needs a lot of work to clarify. In fact, my confusion about how the simulations were done is such that I would really like to see methods presented more clearly before I make a final decision on validity. If they did the simulations the way I described I think their approach is good, but I am honestly not sure given what is written.*

    Thank you for the encouraging words.

### 2024-01-01 alignpair_letter.tex
Begin writing response to reviewers and updating manuscript further as we respond.

@@
 \\
-Juan J. Garcia Mesa\textsuperscript{1,2},
+Juan J. García Mesa\textsuperscript{1,2},
 Ziqi Zhu\textsuperscript{1,3},
@@
 While some aligners can utilize codon substitution models, they are often not robust against coding-sequence artifacts.
-Additionally these aligners  force gaps to occur between codons, whereas in natural sequences, only about 42\% of indels occur between codons \citep{taylor2004occurrence, zhu2022profiling}.
-This mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).
+
+Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}).
+Gaps that occur after the first position or second position of a codon are known as phase-1 and phase-2 gaps, respectively.
+Gaps that occur between codons are known as either phase-3 gaps (this study) or phase-0 gaps \citep[e.g.][]{taylor2004occurrence}.
+While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps.
+Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).
 
@@
         (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
-        (b), (c), and (d) are the results of different aligners.
-        Nucleotide mismatches are highlighted in red. Phase 1, phase 2, and phase 3 indels are shown in purple, orange, and gray respectively.
-        Additionally, the orange indel is type II (an amino-acid indel plus an amino-acid change) while the purple indel is type I (an amino-acid indel only).
-        COATi is the only aligner able to retrieve the biological alignment in this example.
+        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
+        %
+        Indels in protein-coding sequences can be classified as having one of three difference phases and being one of two different types.
+        Phases refer the the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
+        Phase-1, phase-2, and phase-3 indels are shown in purple, orange, and gray respectively.
+        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the purple indel is type-I (an amino-acid indel only).
+The difference between in-frame and frameshift indels is not displayed.
         }
@@
 
+\subsection*{Alignment Strategies}
+
+In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ v1.2.4 via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE v2.06, (4) the DNA model of MAFFT v7.505, and (5) the codon model of PRANK v.150803 \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}.
+Together, these five methods allow us to evaluate both different alignment strategies and different software implementations.
+
 \subsection*{Empirical Simulation Algorithm}
 
-In order to compare COATi against other aligners on realistic datasets, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
-Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL
-database \citep{ensembl_hubbard_2002}.
-After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with five alignment methods: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ v1.2.4 via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE v2.06, (4) the DNA model of MAFFT v7.505, and (5) the codon model of PRANK v.150803 \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}.
-Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps, and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.
-
+In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
+Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL database \citep{ensembl_hubbard_2002}.
+After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with the five alignment methods.
+Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.
 We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
@@
 Segments of matches that were 99 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
-(This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.) We deleted nucleotides from the gorilla sequences of ungapped pairs to create deletions, and we inserted random nucleotides according to their stationary frequency to the gorilla sequences to create insertions.
-The human sequences were unchanged.
+(This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.)
+
+For example, consider the gapped alignment represented by the CIGAR string
+%
+``170M 3D 10M 6I 102M'' % Can use \allowbreak{} here instead of space
+%
+applied to an ungapped alignment of human and gorilla sequences that are 300 nucleotides long.
+First, we modify the CIGAR string to insert flexible lengths for any match segment that is 99 or more nucleotides long while preserving phase.
+This results in a new CIGAR string of
+%
+``101M *M 3D 10M 6I 99M *M'',
+%
+where * represents the locations that have flexible length.
+Considering only matches and deletions, this CIGAR string has 213 fixed nucleotides, leaving 87 nucleotides in the human sequence to be allocated to the flexible locations.
+Since there are two locations of flexible length, we draw one random break point uniformly while maintaining phase, producing the final CIGAR string of
+%
+``101M 9M 3D 10M 6I 99M 78M'',
+%
+which is used to add gaps into the target, ungapped sequence pair.
+To apply deletions, we remove the corresponding nucleotides from the gorilla sequence, and to apply insertions, we add random nucleotides according to their stationary frequency to the gorilla sequence at the corresponding locations.
+The human sequence is left unchanged.
 
@@
 Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
-To evaluate the impact of aligners on selection identification, we estimated $k_s$ and $k_a$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment.
-Briefly, $k_s$ is the number of substitutions per synonymous site and $k_a$ is the number of substitutions per non-synonymous site between two protein-coding sequences.
-We used the method of \cite{ka_ks_li_1993} to estimate these statistics as implemented in the R package seqinr \citep{seqinr}.
-%
-We considered any alignment to be showing evidence of positive selection if $k_a/k_s > 1$ and negative selection if $k_a/k_s < 1$.
+To evaluate the impact of aligners on selection identification, we estimated $K_S$ and $K_A$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment.
+Briefly, $K_S$ is the number of substitutions per synonymous site and $K_A$ is the number of substitutions per non-synonymous site between two protein-coding sequences.
+We used the method developed by \cite{ka_ks_li_1993} and independently derived by \cite{Pamilo1993} to estimate these statistics as implemented in the R package seqinr \citep{seqinr}.
+
+Briefly, this method takes two aligned sequences and classifies the sites in each sequence as non-degenerate, two-fold degenerate, or four-fold degenerate based on the standard genetic code.
+A site is non-degenerate if 0/3 possible nucleotide changes to that site are synonymous, two-fold degenerate if 1/3 of the possible changes are synonymous, and four-fold degenerate if 3/3 possible changes are synonymous.
+(The rare three-fold degenerate sites are treated as two-fold degenerate in this method.) First the method calculates the average numbers of sites of each type in the sequence pair ($L_0$, $L_2$, and $L_4$).
+Next, following \cite{Li1985}, it uses Kimura's two-parameter model \citep{kimura1980simple} and the alignment to estimate the numbers of transitions ($A_i$) and transversions ($B_i$) that occurred per each $i$-th type.
+Using these statistics, \cite{ka_ks_li_1993} estimated $K_S$ and $K_A$ as
+\begin{align*}
+K_S &= \frac{L_2 A_2 + L_4 A_4}{L_2 + L_4} + B_4 &
+K_A &= A_0 + \frac{L_0 B_0 + L_2 B_2}{L_0 + L_2}
+\end{align*}
+
+We considered any alignment to be showing evidence of positive selection if $K_A/K_S > 1$ and negative selection if $K_A/K_S < 1$.
 We estimated F\textsubscript{1} scores for both positive and negative selection by comparing estimated alignments to benchmark alignments.
-F\textsubscript{1} is the harmonic mean of precision and recall and allows us to measure how well aligners produce alignments that correctly identify or exclude positive selection or negative selection.
+F\textsubscript{1} is the harmonic mean of precision and recall:
+\[
+F_1 = \frac{2 \text{TP}}{2 \text{TP} + \text{FP} + \text{FN}}
+\]
+where TP is the number of alignments that correctly predicted positive (or negative) selection, FP is the number of alignments that incorrectly predicted positive (or negative) selection, and FN is the number of alignments that incorrectly predicted the absence of positive (or negative) selection.
+F\textsubscript{1} allows us to measure how well aligners produce alignments that correctly identify the presence and absence of positive selection or negative selection.
 

### 2024-01-01 figures/fig-aln.tex
Update supplementary materials

@@
 A: & TCC & CAT     & AAG & GGG  & CGG & TCG &\verb|---| & GAC & GCC & \verb|---|\\
-D: & TCC & CA\e{G} & \verb|---| & GGG & CGG & TC\e{C} & CCG & GCC & GCC & ACG\\
+D: & TCC & CA\e{G} & \verb|---| & GGG & CGG & TC\e{C} & CCG & GAC & GCC & ACG\\
    & Ser & Gln     &            & Gly     & Arg & Ser   & Pro  & Asp & Ala & Thr\\

### 2024-01-01 supplementary_materials.tex
Update supplementary materials

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Supplementary Materials for ``COATi: statistical pairwise alignment of protein coding sequences''},
  pdfauthor={Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Supplementary Materials for ``COATi: statistical pairwise
alignment of protein coding sequences''}
\author{Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright}
\date{}

\begin{document}
\maketitle

\begin{table}[H]

\caption{\label{tab:table-results-1}Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs. Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
  & dseq & Perfect alns & Best alns & Imperfect alns & F1 pos selection & F1 neg selection\\
\hline
Triplet-MG94 & 0.00221 & 5793 & 5139 & 1048 & 0.98073 & 0.99809\\
\hline
MAFFT & 0.01471 & 5292 & 4692 & 1549 & 0.84314 & 0.98411\\
\hline
PRANK* & 0.01828 & 4725 & 4774 & 2116 & 0.86749 & 0.98698\\
\hline
MACSE & 0.01399 & 2861 & 3737 & 3980 & 0.79456 & 0.98199\\
\hline
ClustalOmega & 0.02929 & 2893 & 2615 & 3948 & 0.68691 & 0.96938\\
\hline
\multicolumn{7}{l}{\rule{0pt}{1em}\textit{*} PRANK produced 42 empty alignments, calculations are based on 7719 alignments.}\\
\end{tabular}
\end{table}
\begin{table}[H]

\caption{\label{tab:table-results-2}Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs. Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
  & dseq & Perfect alns & Best alns & Imperfect alns & F1 pos selection & F1 neg selection\\
\hline
Triplet-ECM & 0.00238 & 5689 & 5045 & 1118 & 0.97803 & 0.99779\\
\hline
MAFFT & 0.01451 & 5338 & 4677 & 1469 & 0.86048 & 0.98549\\
\hline
PRANK* & 0.01903 & 4803 & 4851 & 2004 & 0.89250 & 0.98912\\
\hline
MACSE & 0.01352 & 2903 & 3787 & 3904 & 0.82181 & 0.98359\\
\hline
ClustalOmega & 0.02801 & 2979 & 2624 & 3828 & 0.72337 & 0.97244\\
\hline
\multicolumn{7}{l}{\rule{0pt}{1em}\textit{*} PRANK produced 69 empty alignments, calculations are based on 7692 alignments.}\\
\end{tabular}
\end{table}
\begin{table}[H]

\caption{\label{tab:table-results-3}Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7755 simulated sequence pairs. Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
  & dseq & Perfect alns & Best alns & Imperfect alns & F1 pos selection & F1 neg selection\\
\hline
Marginal-MG94 & 0.00222 & 5808 & 5220 & 1075 & 0.97671 & 0.99766\\
\hline
MAFFT & 0.01505 & 5301 & 4782 & 1582 & 0.85147 & 0.98455\\
\hline
PRANK* & 0.01974 & 4856 & 5015 & 2027 & 0.89928 & 0.99000\\
\hline
MACSE & 0.01429 & 2855 & 3893 & 4028 & 0.81569 & 0.98349\\
\hline
ClustalOmega & 0.02870 & 2901 & 2610 & 3982 & 0.72399 & 0.97171\\
\hline
\multicolumn{7}{l}{\rule{0pt}{1em}\textit{*} PRANK produced 60 empty alignments, calculations are based on 7695 alignments.}\\
\end{tabular}
\end{table}
\begin{table}[H]

\caption{\label{tab:table-results-4}Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7767 simulated sequence pairs. Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
  & dseq & Perfect alns & Best alns & Imperfect alns & F1 pos selection & F1 neg selection\\
\hline
Marginal-ECM & 0.00229 & 5781 & 5135 & 1081 & 0.97052 & 0.99710\\
\hline
MAFFT & 0.01473 & 5379 & 4813 & 1483 & 0.85011 & 0.98491\\
\hline
PRANK* & 0.01953 & 4830 & 4918 & 2032 & 0.87752 & 0.98790\\
\hline
MACSE & 0.01400 & 2953 & 3893 & 3909 & 0.78977 & 0.98159\\
\hline
ClustalOmega & 0.02918 & 2892 & 2611 & 3970 & 0.67847 & 0.96785\\
\hline
\multicolumn{7}{l}{\rule{0pt}{1em}\textit{*} PRANK produced 49 empty alignments, calculations are based on 7718 alignments.}\\
\end{tabular}
\end{table}
\begin{table}[H]

\caption{\label{tab:table-results-5}Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7798 simulated sequence pairs with gorilla as the reference. Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
  & dseq & Perfect alns & Best alns & Imperfect alns & F1 pos selection & F1 neg selection\\
\hline
Triplet-MG94 & 0.00217 & 5870 & 5162 & 1030 & 0.98450 & 0.99853\\
\hline
MAFFT & 0.01445 & 5450 & 4803 & 1450 & 0.84704 & 0.98508\\
\hline
PRANK* & 0.02126 & 4942 & 5026 & 1958 & 0.89805 & 0.99042\\
\hline
MACSE & 0.01340 & 2966 & 3989 & 3934 & 0.79608 & 0.98260\\
\hline
ClustalOmega & 0.02860 & 3034 & 2711 & 3866 & 0.69147 & 0.97026\\
\hline
\multicolumn{7}{l}{\rule{0pt}{1em}\textit{*} PRANK produced 35 empty alignments, calculations are based on 7763 alignments.}\\
\end{tabular}
\end{table}
\begin{table}[H]

\caption{\label{tab:table-results-6}Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, MACSE, and codon-triplet-mg with gorila as the reference on 4003 of the 7761 simulated sequence pairs where the gorilla sequence was simulated without early stop codons, incomplete codons, or ambiguous nucleotides. Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
  & dseq & Perfect alns & Best alns & Imperfect alns & F1 pos selection & F1 neg selection\\
\hline
Triplet-MG & 0.00113 & 3501 & 2890 & 309 & 0.99278 & 0.99932\\
\hline
MAFFT & 0.00586 & 3162 & 2704 & 648 & 0.91064 & 0.99137\\
\hline
PRANK & 0.00358 & 2829 & 2673 & 981 & 0.90332 & 0.99084\\
\hline
MACSE & 0.00448 & 2552 & 2434 & 1258 & 0.87234 & 0.98857\\
\hline
ClustalOmega & 0.02099 & 1772 & 1554 & 2038 & 0.75960 & 0.97686\\
\hline
Triplet-MG-gor-ref & 0.00118 & 3463 & 2816 & 347 & 0.98993 & 0.99904\\
\hline
\end{tabular}
\end{table}

\newpage

\begin{figure}
\centering
\includegraphics{figures/fig-base-calling-error.pdf}
\caption{\label{fig:figs} Base calling error FST. Arcs from M to S
generate matches; however, here they can introduce single-nucleotide
errors, which can generate stop codon artifacts.}
\end{figure}

\begin{figure}

{\centering \includegraphics{figures/aln-example-graph} \includegraphics{figures/aln-example}

}

\caption{\label{fig:aln-example} FST of all possible alignments between `CTC' and `CTG' (top) and the best alignment, determined by the shortest path (bottom). Figures generated using the OpenFST library.}\label{fig:aln-example}
\end{figure}

\begin{figure}

{\centering \includegraphics{figures/dseq_plots_tri-mg}

}

\caption{\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\leq 1.25e-76$.}\label{fig:dseq1}
\end{figure}

\begin{figure}

{\centering \includegraphics{figures/dseq_plots_tri-ecm}

}

\caption{\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\leq 3.23e-48$.}\label{fig:dseq2}
\end{figure}

\begin{figure}

{\centering \includegraphics{figures/dseq_plots_mar-mg}

}

\caption{\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\leq 1.99e-53$.}\label{fig:dseq3}
\end{figure}

\begin{figure}

{\centering \includegraphics{figures/dseq_plots_mar-ecm}

}

\caption{\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\leq 1.44e-52$.}\label{fig:dseq4}
\end{figure}

\begin{figure}

{\centering \includegraphics{figures/dseq_plots_reverse}

}

\caption{\label{fig:dseq-reverse} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE with gorilla as the reference. COATi was significantly more accurate than other aligners; all p-values were $\leq 1.75e-64$.}\label{fig:dseq5}
\end{figure}

\newpage

\begin{figure}
\centering
\includegraphics{figures/k2p_distances.pdf}
\caption{\label{fig:k2p} K2P distances. Evolutionary distances between
reference and retrieved alignments for each method calculated using the
Kimura 2-parameter model. Notably, COATi performs best while the other
tools tend to overestimate the evolutionary distance.}
\end{figure}

\end{document}

### 2024-01-01 supplementary_materials/scripts/kaks.R

@@ -29,6 +29,34 @@ omega = function(a) { #dN/dS
 	}
 }
 
+fdr_pos = function(ref, w) {
+	tp = length(intersect(which(w>1),which(ref>1)))
+	fp = length(intersect(which(w>1),which(ref<1)))
+	fn = length(intersect(which(w<1),which(ref>1)))
+	return(fp/(tp+fp))
+}
+
+fnr_pos = function(ref, w) {
+	tp = length(intersect(which(w>1),which(ref>1)))
+	fp = length(intersect(which(w>1),which(ref<1)))
+	fn = length(intersect(which(w<1),which(ref>1)))
+	return(fn/(tp+fn))
+}
+
+fdr_neg = function(ref, w) {
+	tp = length(intersect(which(w<1),which(ref<1)))
+	fp = length(intersect(which(w<1),which(ref>1)))
+	fn = length(intersect(which(w>1),which(ref<1)))
+	return(fp/(tp+fp))
+}
+
+fnr_neg = function(ref, w) {
+	tp = length(intersect(which(w<1),which(ref<1)))
+	fp = length(intersect(which(w<1),which(ref>1)))
+	fn = length(intersect(which(w>1),which(ref<1)))
+	return(fn/(tp+fn))
+}
+
 ps_accuracy = function(ref,w) {
 	tp = length(intersect(which(w>1),which(ref>1)))
 	fp = length(intersect(which(w>1),which(ref<1)))

### 2024-01-01 supplementary_materials/scripts/plot_dseq.R

@@ -2,7 +2,7 @@ library(ggplot2)
 library(ggpubr)
 
 # x-y plot of dseq
-plot_dseq_main = function(results_file, coati_model) {
+plot_dseq_main = function(results_file, coati_model, save = FALSE) {
     # read results
     results = read.csv(results_file)
     
@@ -14,14 +14,15 @@ plot_dseq_main = function(results_file, coati_model) {
                       PRANK = results$dseq[which(results$aligner == "prank")])
     
     # Hex plot
-    
+
     prank = ggplot(dseq, aes(x = log10(COATi), y = log10(PRANK))) +
         geom_hex() +
         scale_fill_viridis_c() +
         geom_abline(slope = 1, intercept = 0, color  = "black") +
         xlim(-4.5, 0) +
         ylim(-4.5, 0) +
-        coord_fixed(ratio = 1/1)
+        coord_fixed(ratio = 1/1) +
+        theme_minimal()
     
     mafft = ggplot(dseq, aes(x = log10(COATi), y = log10(MAFFT))) +
         geom_hex() +
@@ -29,7 +30,8 @@ plot_dseq_main = function(results_file, coati_model) {
         geom_abline(slope = 1, intercept = 0, color  = "black") +
         xlim(-4.5, 0) +
         ylim(-4.5, 0) +
-        coord_fixed(ratio = 1/1)
+        coord_fixed(ratio = 1/1) +
+        theme_minimal()
     
     macse = ggplot(dseq, aes(x = log10(COATi), y = log10(MACSE))) +
         geom_hex() +
@@ -37,18 +39,32 @@ plot_dseq_main = function(results_file, coati_model) {
         geom_abline(slope = 1, intercept = 0, color  = "black") +
         xlim(-4.5, 0) +
         ylim(-4.5, 0) +
-        coord_fixed(ratio = 1/1)
-    
+        coord_fixed(ratio = 1/1) +
+        theme_minimal()
+ 
     clustalo = ggplot(dseq, aes(x = log10(COATi), y = log10(ClustalOmega))) +
         geom_hex() +
         scale_fill_viridis_c() +
         geom_abline(slope = 1, intercept = 0, color  = "black") +
         xlim(-4.5, 0) +
         ylim(-4.5, 0) +
-        coord_fixed(ratio = 1/1)
+        coord_fixed(ratio = 1/1) +
+        theme_minimal()
+
+    # dseq_plots = ggarrange(prank, mafft, clustalo, macse,
[…]

### 2024-01-01 supplementary_materials/scripts/results_gor_ref.R

@@ -1,0 +1,109 @@
+# create a results file only using the reference alignments that COATi triplet-mg
+#  was able to align with gorilla as the reference
+
+library(tidyverse)
+
+gor_ref_results = function() {
+    
+    ## Clean and save gor-ref results
+    results_raw = read.csv("supplementary_data/gor-ref/results/results_summary.csv",
+                       header = TRUE)
+    alns = results_raw |> filter(aligner == "gor-ref") |> drop_na(score) |> select(ref_name) |> unlist()
+    results = results_raw |> filter(ref_name %in% alns)
+    results_file = "supplementary_data/gor-ref/results/results_summary_clean.csv"
+    write_csv(x = results, file = results_file)
+    
+    trimg = results |> filter(aligner == "tri-mg")
+    mafft = results |> filter(aligner == "mafft")
+    prank = results |> filter(aligner == "prank")
+    macse = results |> filter(aligner == "macse")
+    clustal = results |> filter(aligner == "clustalo")
+    goref = results |> filter(aligner == "gor-ref")
+    
+    aligners = c("tri-mg", "mafft", "prank", "macse", "clustalo", "gor-ref")
+    stats_table = data.frame(dseq = c(mean(trimg$dseq, na.rm = TRUE),
+                                      mean(mafft$dseq, na.rm = TRUE),
+                                      mean(prank$dseq, na.rm = TRUE),
+                                      mean(macse$dseq, na.rm = TRUE),
+                                      mean(clustal$dseq, na.rm = TRUE),
+                                      mean(goref$dseq, na.rm = TRUE)),
+                             row.names = aligners)
+    
+    stats_table
+    
+    ###########################################################################
+    # perfect, best, and imperfect alignments
+    source("supplementary_materials/scripts/number_alignments.R")
+    alns = num_alns(results_file)
+    
+    # fix row and column names and add to stats table
+    colnames(alns) = c("Perfect alns", "Best alns", "Imperfect alns")
+    alns = alns[aligners, ]
+    stats_table = cbind(stats_table, alns)
+    
+    ############################################################################
+    # selection (ka ks)
+    source("supplementary_materials/scripts/kaks.R")
+    
+    selection = data.frame("F1.pos" = double(),
+                           "F1.neg" = double())
+    
[…]

### 2024-01-01 figures/fig-base-calling-error.tex
Update permissions.

@@
 \tikzstyle{block}=[circle,fill=colorB!50,on grid]
-\tikzstyle{lab}=[]
-\tikzstyle{w}=[lab,midway]
-\tikzstyle{e}=[lab,midway,auto=false,fill=colorP!50,font=\scriptsize]
-\tikzstyle{p}=[lab,midway,auto=false,fill=colorO!50,font=\scriptsize]
+\tikzstyle{w}=[below,pos=0.66]
+\tikzstyle{e}=[pos=0.33,auto=false,fill=colorP!50,font=\scriptsize]
 
@@
 \node[block,right=15mm of start_sub] (S_sub) {S};
-\node[block,above right=10mm and 30mm of S_sub] (ij_sub) {};
-\node[block,right=30mm of S_sub] (ii_sub) {};
-\node[block,below right=10mm and 30mm of S_sub] (in_sub) {};
 \node[block,right=60mm of S_sub] (M_sub) {M};
@@
 
-\draw[line] (M_sub) to[bend right=20] node[e] {i\textrightarrow{}j} (ij_sub);
-\draw[line] (ij_sub) to[bend right=20] node[w] {$\varepsilon$} (S_sub);
-\draw[line] (M_sub) to node[e] {i\textrightarrow{}i} (ii_sub);
-\draw[line] (ii_sub) to node[w] {$1 - 3 \cdot \varepsilon$} (S_sub);
-\draw[line] (M_sub) to[bend right=-20] node[e] {i\textrightarrow{}N} (in_sub);
-\draw[line] (in_sub) to[bend right=-20] (S_sub);
+\draw[line] (M_sub) to[bend right=25] node[w] {$\varepsilon$} (S_sub);
+\draw[line] (M_sub) to[bend right=25] node[e] {i\textrightarrow{}j} (S_sub);
+\draw[line] (M_sub) to node[w] {$1 - 3 \cdot \varepsilon$} (S_sub);
+\draw[line] (M_sub) to node[e] {i\textrightarrow{}i} (S_sub);
+\draw[line] (M_sub) to[bend right=-25] (S_sub);
+\draw[line] (M_sub) to[bend right=-25] node[e] {i\textrightarrow{}N} (S_sub);
 

### 2024-01-01 figures/fig-evolution-fst.tex
Additional comments and changes by Reed
Update permissions.

@@
 \draw[line] (M_sub) to[bend right=19.5] (S_sub);
-\draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5] {XXX\textrightarrow{}YYY} (S_sub);
+\draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5]
+{X X X\textrightarrow{}Y Y Y} (S_sub);
 % \draw[line] (M_sub) to[bend right=22.5] node[e,pos=0.5] {AAA1\textrightarrow{}C} (S_sub);
@@
 
-\node[above right=-1.4cm and 5mm of sequences,text width=50mm] {
+\node[above right=-1.80cm and 5mm of sequences,text width=50mm] {
 \begin{tabular}{r@{\,: }l}
@@
 	$e$ & gap extension weight\\
+	$\pi$ & nucleotide stationary frequency\\
 \end{tabular}
@@
 \draw[line] (I1) to[bend right=40] (U1);
-\draw[line] (I1) to[bend right=30] node[e] {\O\textrightarrow{}Z} (U1);
+\draw[line] (I1) to[bend right=30] node[e] {\O\textrightarrow{}Z} node[w,below=2mm] {$\pi_{\text{Z}}$} (U1);
 

### 2024-01-01 supplementary_materials.Rmd
Begin writing response to reviewers and updating manuscript further as we respond.
Update permissions.

--- 2023-05-15 supplementary_materials.Rmd
@@
         fig_caption: yes
+        keep_tex: true
 bibliography: alignpair_letter.bib
@@
 ```{r echo = FALSE}
-library(knitr)
+suppressPackageStartupMessages(library(knitr))
+suppressPackageStartupMessages(library(tidyverse))
 
@@
     # extract dseq values for each aligner
-    coati = results$dseq[which(results$aligner == coati_model)]
-    clustalomega = results$dseq[which(results$aligner == "clustalo")]
-    macse = results$dseq[which(results$aligner == "macse")]
-    mafft = results$dseq[which(results$aligner == "mafft")]
-    prank = results$dseq[which(results$aligner == "prank")]
+    coati = results |> filter(aligner == coati_model) |> select(dseq)
+    clustal = results |> filter(aligner == "clustalo") |> select(dseq)
+    macse = results |> filter(aligner == "macse") |> select(dseq)
+    mafft = results |> filter(aligner == "mafft") |> select(dseq)
+    prank = results |> filter(aligner == "prank") |> select(dseq)
 
-    dseq_summary = data.frame(coati = mean(coati, na.rm = TRUE),
-                      prank = mean(prank, na.rm = TRUE),
-                      mafft = mean(mafft, na.rm = TRUE),
-                      clustalo = mean(clustalomega, na.rm = TRUE),
-                      macse = mean(macse, na.rm = TRUE),
-                      row.names = "$d_{seq}$")
+    stats_table = data.frame(dseq = c(mean(coati$dseq, na.rm = TRUE),
+                          mean(mafft$dseq, na.rm = TRUE),
+                          mean(prank$dseq, na.rm = TRUE),
+                          mean(macse$dseq, na.rm = TRUE),
+                          mean(clustal$dseq, na.rm = TRUE)), row.names = c(coati_model, "mafft", "prank", "macse", "clustalo"))
 
-    ############################################################################
+    ###########################################################################
     # perfect, best, and imperfect alignments
@@
     # fix row and column names and add to stats table
-    row.names(alns)[which(row.names(alns) == coati_model)] = "coati"
-    colnames(alns) = c("Perfect alignments", "Best alignments", "Imperfect alignments")
-    stats_table = rbind(dseq_summary, t(alns))
+    colnames(alns) = c("Perfect alns", "Best alns", "Imperfect alns")
+    alns = alns[c(coati_model, "mafft", "prank", "macse", "clustalo"), ]
+    stats_table = cbind(stats_table, alns)
 
@@
 
-    aligners = c(coati_model, "prank", "mafft", "clustalo", "macse")
+    aligners = c(coati_model, "mafft", "prank", "macse", "clustalo")
     selection = data.frame("F1.pos" = double(),
@@
     # fix row and column names and add to stats table
-    colnames(selection) = c("F1-score pos selection", "F1-score neg selection")
-    rownames(selection) = c("coati", "prank", "mafft", "clustalo", "macse")
-    stats_table = rbind(stats_table, t(selection))
-    colnames(stats_table) = c(coati_model, "PRANK*", "MAFFT", "ClustalOmega", "MACSE")
-    stats_table = stats_table[, c(coati_model, "MAFFT", "PRANK*", "MACSE", "ClustalOmega")]
+    colnames(selection) = c("F1 pos selection", "F1 neg selection")
+    rownames(selection) = c("coati", "mafft", "prank", "macse", "clustalo")
+    stats_table = cbind(stats_table, selection)
+    if(coati_model == "tri-mg") {
+        coati_model = "Triplet-MG94"
+    } else if(coati_model == "tri-ecm") {
+        coati_model = "Triplet-ECM"
+    } else if(coati_model == "mar-mg") {
+        coati_model = "Marginal-MG94"
+    } else if(coati_model == "mar-ecm") {
+        coati_model = "Marginal-ECM"
+    }
+    rownames(stats_table) = c(coati_model, "MAFFT", "PRANK*", "MACSE", "ClustalOmega")
 
@@
 
-```{r echo = FALSE, results = 'asis'}
-library(kableExtra)
-library(knitr)
-source("supplementary_materials/scripts/plot_dseq.R")
+```{r setup-table-results, echo = FALSE}
+suppressMessages(library(kableExtra))
+suppressMessages(library(knitr))
+suppressMessages(library(tidyverse))
+```
+
+```{r table-results-1, echo = FALSE, results = 'asis'}
 results = "supplementary_data/tri-mg/results/results_summary.csv"
-k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7719 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment.
-"), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 50 empty alignments, calculations are based on 7669 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-#
+k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 42 empty alignments, calculations are based on 7719 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+```
+```{r table-results-2, echo = FALSE, results = 'asis'}
 results = "supplementary_data/tri-ecm/results/results_summary.csv"
-k = kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7678 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 41 empty alignments, calculations are based on 7637 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-#
+k = kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 69 empty alignments, calculations are based on 7692 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+```
+```{r table-results-3, echo = FALSE, results = 'asis'}
 results = "supplementary_data/mar-mg/results/results_summary.csv"
-k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7661 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment.
-"), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 50 empty alignments, calculations are based on 7611 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-#
+k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7755 simulated sequence pairs.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 60 empty alignments, calculations are based on 7695 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+```
+```{r table-results-4, echo = FALSE, results = 'asis'}
 results = "supplementary_data/mar-ecm/results/results_summary.csv"
-k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7710 simulated sequence pairs.
-Perfect alignments have $d_{seq}$ = 0, best alignments have lowest $d_{seq}$, and imperfect alignments have $d_{seq} > 0$ when at least one aligner found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 35 empty alignments, calculations are based on 7675 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7767 simulated sequence pairs.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 49 empty alignments, calculations are based on 7718 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+```
+```{r table-results-5, echo = FALSE, results = 'asis'}
+results = "supplementary_data/reverse/results/results_summary.csv"
+k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7798 simulated sequence pairs with gorilla as the reference.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(footnote(k, general = "PRANK produced 35 empty alignments, calculations are based on 7763 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
+```
+```{r table-results-6, echo = FALSE, results = 'asis'}
+source("supplementary_materials/scripts/results_gor_ref.R")
+k = kable(gor_ref_results(), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, MACSE, and codon-triplet-mg with gorila as the reference on 4003 of the 7761 simulated sequence pairs where the gorilla sequence was simulated without early stop codons, incomplete codons, or ambiguous nucleotides.
+Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
+print(k)
 ```
@@
 
-```{r echo = FALSE}
+```{r base-calling-error, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap="\\label{fig:figs} Base calling error FST.
+Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
+knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
+```
+
+```{r aln-example, echo = FALSE, fig.show='hold', fig.align='center', fig.cap="\\label{fig:aln-example} FST of all possible alignments between `CTC' and `CTG' (top) and the best alignment, determined by the shortest path (bottom).
+Figures generated using the OpenFST library."}
+knitr::include_graphics(path = c("figures/aln-example-graph.pdf", "figures/aln-example.pdf"))
+```
+
+```{r pvals, echo = FALSE}
 results = "supplementary_data/tri-mg/results/results_summary.csv"
@@
 pval_mar_ecm = max(wilcoxon_test(results, "mar-ecm"))
+results = "supplementary_data/reverse/results/results_summary.csv"
+pval_reverse = max(wilcoxon_test(results, "tri-mg"))
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r dseq1, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
@@
 source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg"))
+suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg", save = TRUE))
+knitr::include_graphics(path = "figures/dseq_plots_tri-mg.pdf")
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r dseq2, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
@@
 source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm"))
+suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm", save = TRUE))
+knitr::include_graphics(path = "figures/dseq_plots_tri-ecm.pdf")
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r dseq3, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
@@
 source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg"))
+suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg", save = TRUE))
+knitr::include_graphics(path = "figures/dseq_plots_mar-mg.pdf")
 ```
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
+```{r dseq4, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
 COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
@@
 source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm"))
+suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm", save = TRUE))
+knitr::include_graphics(path = "figures/dseq_plots_mar-ecm.pdf")
+```
+
+```{r dseq5, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-reverse} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE with gorilla as the reference.
+COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_reverse, digits = 3), "$.")}
+library(knitr)
+source("supplementary_materials/scripts/plot_dseq.R")
+suppressWarnings(plot_dseq_main("supplementary_data/reverse/plot_distance.csv", "reverse", save = TRUE))
+knitr::include_graphics(path = "figures/dseq_plots_reverse.pdf")
 ```
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.cap="\\label{fig:figs} Base calling error FST.
-Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
-knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
+```{r k2p-dist, echo = FALSE, fig.cap="\\label{fig:k2p} K2P distances.
+Evolutionary distances between reference and retrieved alignments for each method calculated using the Kimura 2-parameter model.
+Notably, COATi performs best while the other tools tend to overestimate the evolutionary distance."}
+knitr::include_graphics(path = "figures/k2p_distances.pdf")
 ```
 
-\newpage
+<!-- \newpage -->
 
-# Supplementary Methods
+<!-- # Supplementary Methods -->
 
-Ks and Ka represent the number of substitutions per synonymous and non-synonymous sites.
+<!-- Ks and Ka represent the number of substitutions per synonymous and non-synonymous sites.
 The ratio of nonsynonymous (Ka) to synonymous (Ks) nucleotide substitution rates indicates the selective pressures acting on genes.
@@
 If the ratio is less than 1, it indicates a pressure to maintain protein sequence, known as purifying selection.
-Ks and Ka are calculated using the R package seqinr v.4.2-30 [@seqinr].
+-->
+<!-- Ks and Ka are calculated using the R package seqinr v.4.2-30 [@seqinr]. -->
 
-# References
+<!-- # References -->
 
-<div id="refs"></div>
+<!-- <div id="refs"></div> -->

### 2024-01-05 alignpair_letter.tex
Figure and supplement work

@@
 There are well-established algorithms for combining FSTs in different ways allowing the design of complex models by combining simpler FSTs, including concatenation, composition, intersection, union, and reversal \citep{bradley2007transducers,silvestre2021machine}.
-Specifically, composition is an algorithm to combine two FSTs by sending the output of one FST into the input of another, creating a new, more complex transducer.
+Specifically, composition is an algorithm to combine two FSTs by sending the output of one FST into the input of another, creating a new, more complex transducer \citep{mohri2005weighted}.
 Figure~\ref{fig:base-calling} illustrates how FSTs modeling sequencing errors (Fig.~\ref{fig:base-calling}a) and ambiguity (Fig.~\ref{fig:base-calling}b) can be combined via composition to produce an FST that does both (Fig.~\ref{fig:base-calling}c).
@@
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
-An example of an FST-based alignment can be found in Supplementary Materials Figure 2.
+An example of an FST-based alignment can be found in Supplementary Materials Figure 1.
 

### 2024-01-05 supplementary_materials/supplementary_materials.Rmd
Figure and supplement work

--- 2024-01-01 supplementary_materials.Rmd
@@
 ---
-title: "Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\""
+title: "COATi: statistical pairwise alignment of protein coding sequences"
+subtitle: Supplementary Materials
 author: Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright
+mainfont: TeX Gyre Pagella
+bibliography: ../alignpair_letter.bib
 output:
-    pdf_document:
+    "bookdown::pdf_document2":
+        numbered_sections: yes
         fig_caption: yes
         keep_tex: true
-bibliography: alignpair_letter.bib
+        latex_engine: lualatex
+header-includes:
+  - \let\oldthefigure\thefigure
+  - \renewcommand{\thefigure}{S\oldthefigure}
 ---
@@
     # perfect, best, and imperfect alignments
-    source("supplementary_materials/scripts/number_alignments.R")
+    source("scripts/number_alignments.R")
     alns = num_alns(results_file)
@@
     # selection (ka ks)
-    source("supplementary_materials/scripts/kaks.R")
+    source("scripts/kaks.R")
 
@@
 
+# Aligner Commands
+
+We evaluated five different aligners.
+Below are the commands that we used to run them.
+We have abbreviated the commands for clarity, stripping out unimportant arguments.
+Complete workflows can be found in our coati-testing repository on Github.
+
+- COATi: `coati alignpair -m tri-mg ...`
+- ClustalΩ v1.2.4: `clustalo ...`
+- PRANK v.150803: `prank -codon ...`
+- MACSE v2.06: `java -jar macse.jar -fs_lr 10 -stop_lr 10 -prog alignSequences ...`
+- MAFFT v7.505: `mafft --preservecase --globalpair --maxiterate 1000 ...`
+
+COATi can use different alignment models for pairwise alignment. Below are the commands that we used to run different models.
+
+- tri-mg: `coati alignpair -m tri-mg ...`
+- tri-ecm: `coati alignpair -m tri-ecm ...`
+- mar-mg: `coati alignpair -m mar-mg ...`
+- mar-ecm: `coati alignpair -m mar-ecm ...`
+
+# FST Alignment Example
+
+When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST library to generate best alignments by composing the input and output sequences with the COATi FST model.
+While the COATi FST model is too large to display, we can show the result of a composition.
+
+Fig.
+\@ref(fig:aln-example-a) shows a graph depicting the FST that results from composing the COATi FST with the input sequence "CTC" and the output sequence "CTG".
+Every path through this FST represents one possible way to align "CTC" and "CTG", and the sum of all weights along a path is the total weight of the respective alignment.
+Here, the weights of each arc are in negative-log space.
+Note that this graph has been optimized, and weight has been pushed towards the initial state.
+The weight of any specific arc may not be directly mapable to a weight described in the model.
+
+Fig.
+\@ref(fig:aln-example-b) is the best alignment between "CTC" and "CTG", as determined by the shortest path algorithm.
+Figs.
+\@ref(fig:aln-example-a) and \@ref(fig:aln-example-b) were produced by the OpenFST library.
+A bold circle represents a starting node, and a double circle represents a termination node.
+
+```{r aln-example-a, echo = FALSE, fig.align='center', fig.cap="The FST of all possible alignments between `CTC' and `CTG'."}
+knitr::include_graphics("figures/aln-example-graph.pdf")
+```
+
+```{r aln-example-b, echo = FALSE, out.width="75%", fig.align='center', fig.cap="The best alignment of `CTC' and `CTG'."}
+knitr::include_graphics("figures/aln-example-path.pdf")
+```
+
+\newpage
+
 ```{r setup-table-results, echo = FALSE}
@@
 ```{r table-results-1, echo = FALSE, results = 'asis'}
-results = "supplementary_data/tri-mg/results/results_summary.csv"
+results = "data/tri-mg/results/results_summary.csv"
 k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs.
@@
 ```{r table-results-2, echo = FALSE, results = 'asis'}
-results = "supplementary_data/tri-ecm/results/results_summary.csv"
+results = "data/tri-ecm/results/results_summary.csv"
 k = kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs.
@@
 ```{r table-results-3, echo = FALSE, results = 'asis'}
-results = "supplementary_data/mar-mg/results/results_summary.csv"
+results = "data/mar-mg/results/results_summary.csv"
 k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7755 simulated sequence pairs.
@@
 ```{r table-results-4, echo = FALSE, results = 'asis'}
-results = "supplementary_data/mar-ecm/results/results_summary.csv"
+results = "data/mar-ecm/results/results_summary.csv"
 k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7767 simulated sequence pairs.
@@
 ```{r table-results-5, echo = FALSE, results = 'asis'}
-results = "supplementary_data/reverse/results/results_summary.csv"
+results = "data/reverse/results/results_summary.csv"
 k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7798 simulated sequence pairs with gorilla as the reference.
@@
 ```{r table-results-6, echo = FALSE, results = 'asis'}
-source("supplementary_materials/scripts/results_gor_ref.R")
+source("scripts/results_gor_ref.R")
 k = kable(gor_ref_results(), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, MACSE, and codon-triplet-mg with gorila as the reference on 4003 of the 7761 simulated sequence pairs where the gorilla sequence was simulated without early stop codons, incomplete codons, or ambiguous nucleotides.
@@
 Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
-knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
-```
-
-```{r aln-example, echo = FALSE, fig.show='hold', fig.align='center', fig.cap="\\label{fig:aln-example} FST of all possible alignments between `CTC' and `CTG' (top) and the best alignment, determined by the shortest path (bottom).
-Figures generated using the OpenFST library."}
-knitr::include_graphics(path = c("figures/aln-example-graph.pdf", "figures/aln-example.pdf"))
+#knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
 ```
@@
 ```{r pvals, echo = FALSE}
-results = "supplementary_data/tri-mg/results/results_summary.csv"
+results = "data/tri-mg/results/results_summary.csv"
 pval_tri_mg = max(wilcoxon_test(results, "tri-mg"))
-results = "supplementary_data/tri-ecm/results/results_summary.csv"
+results = "data/tri-ecm/results/results_summary.csv"
 pval_tri_ecm = max(wilcoxon_test(results, "tri-ecm"))
-results = "supplementary_data/mar-mg/results/results_summary.csv"
+results = "data/mar-mg/results/results_summary.csv"
 pval_mar_mg = max(wilcoxon_test(results, "mar-mg"))
-results = "supplementary_data/mar-ecm/results/results_summary.csv"
+results = "data/mar-ecm/results/results_summary.csv"
 pval_mar_ecm = max(wilcoxon_test(results, "mar-ecm"))
-results = "supplementary_data/reverse/results/results_summary.csv"
+results = "data/reverse/results/results_summary.csv"
 pval_reverse = max(wilcoxon_test(results, "tri-mg"))
@@
 library(knitr)
-source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg", save = TRUE))
+source("scripts/plot_dseq.R")
+suppressWarnings(plot_dseq_main("data/tri-mg/plot_distance.csv", "tri-mg", save = TRUE))
 knitr::include_graphics(path = "figures/dseq_plots_tri-mg.pdf")
@@
 library(knitr)
-source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm", save = TRUE))
+source("scripts/plot_dseq.R")
+suppressWarnings(plot_dseq_main("data/tri-ecm/plot_distance.csv", "tri-ecm", save = TRUE))
 knitr::include_graphics(path = "figures/dseq_plots_tri-ecm.pdf")
@@
 library(knitr)
-source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg", save = TRUE))
+source("scripts/plot_dseq.R")
+suppressWarnings(plot_dseq_main("data/mar-mg/plot_distance.csv", "mar-mg", save = TRUE))
 knitr::include_graphics(path = "figures/dseq_plots_mar-mg.pdf")
@@
 library(knitr)
-source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm", save = TRUE))
+source("scripts/plot_dseq.R")
+suppressWarnings(plot_dseq_main("data/mar-ecm/plot_distance.csv", "mar-ecm", save = TRUE))
 knitr::include_graphics(path = "figures/dseq_plots_mar-ecm.pdf")
@@
 library(knitr)
-source("supplementary_materials/scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("supplementary_data/reverse/plot_distance.csv", "reverse", save = TRUE))
+source("scripts/plot_dseq.R")
+suppressWarnings(plot_dseq_main("data/reverse/plot_distance.csv", "reverse", save = TRUE))
 knitr::include_graphics(path = "figures/dseq_plots_reverse.pdf")
@@
 
-```{r k2p-dist, echo = FALSE, fig.cap="\\label{fig:k2p} K2P distances.
-Evolutionary distances between reference and retrieved alignments for each method calculated using the Kimura 2-parameter model.
-Notably, COATi performs best while the other tools tend to overestimate the evolutionary distance."}
+```{r k2p-dist, echo = FALSE, fig.align = 'center', out.width="75%", fig.cap="\\label{fig:k2p} Alignments generated by COATi produce accurate evolutionary distances.
+Benchmark alignments were realigned by COATi and four other methods as described in the main text.
+Kimura two-parameter (K2P) distances were estimated from the benchmark alignments and the estimated alignments.
+Each panel is a scatter plot comparing the benchmark disances (Reference) with distances estimated via one of the aligners.
+COATi clearly performs better than other tools, which tend to drastically overestimate some distances."}
 knitr::include_graphics(path = "figures/k2p_distances.pdf")

### 2024-01-05 figures/fig-k2p-empirical.R

@@ -1,0 +1,126 @@
+library(tidyverse)
+library(scales)
+library(here)
+library(patchwork)
+
+PDF_WIDTH <- 7 # 178mm
+PDF_HEIGHT <- PDF_WIDTH/2
+
+#showtext_auto()
+
+cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
+    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+set.seed(17760)
+
+# load data
+aln_data <- read_csv(here("data/aln_data.csv.gz"))
+
+kounts <- aln_data |> count(method, failed=is.na(k2p)) |> filter(failed)
+
+cat("Table of failed sequences\n")
+cat("=========================\n")
+print(kounts)
+
+# identify files that didn't show variation
+k2p_sum <- aln_data |> summarize(
+    coati_na = is.na(k2p[method == "tri-mg"]),
+    all_same = n_distinct(k2p) == 1L,
+    .by = "sequence")
+
+kounts <- k2p_sum |> count(all_same)
+
+cat("Table of trivial alignments\n")
+cat("=========================\n")
+print(kounts)
+
+# remove sequences that COATi failed on
+seq_ids <- k2p_sum |> filter(coati_na == FALSE) |> pull(sequence)
+aln_data <- aln_data |> filter(sequence %in% seq_ids)
+
+
+pal <- cbbPalette[c(2,3,4,6,7)]
+
+#### PLOT 1: K2P density plots
+
+gg1 <- ggplot(aln_data, aes(x = k2p, color=method)) + geom_density(linewidth = 1)
+gg1 <- gg1 + scale_x_continuous(trans = transform_pseudo_log(0.0005, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1))
+gg1 <- gg1 + coord_cartesian(ylim=c(NA, 1.51))
+gg1 <- gg1 + xlab("K2P Distance")
+gg1 <- gg1 + scale_color_manual(values = pal,
+    name="Aligner",  labels = c(
+     "clustalo" = "ClustalΩ", #ClustalΩ
+     "macse" = "MACSE",
+     "mafft" = "MAFFT",
+     "prank" = "PRANK",
+     "tri-mg" = "COATi"
+     )
+    )
+gg1 <- gg1 + guides(color = guide_legend(override.aes=list(fill=pal)))
+gg1 <- gg1 + theme_minimal(9) + theme(
[…]

### 2024-01-05 figures/fig-sel-empirical.R

@@ -1,0 +1,93 @@
+library(here)
+library(tidyverse)
+library(ComplexUpset)
+library(patchwork)
+
+PDF_WIDTH <- 7 # 178mm
+PDF_HEIGHT <- PDF_WIDTH/2
+
+# load data
+aln_data <- read_csv(here("data/aln_data.csv.gz"))
+
+# omega_data <- aln_data |>
+#     mutate(neg_sel = (!is.na(omega) & omega < 1)) |>
+#     filter(neg_sel) |>
+#     (\(x) split(x, x$method))() |>
+#     map(\(x) pull(x, "sequence"))
+
+####
+
+omega_data <- aln_data |>
+    mutate(neg_sel = omega < 1) |>
+    pivot_wider(id_cols = "sequence", names_from = "method", values_from = "neg_sel") |>
+    rename(coati = "tri-mg")
+
+gg1 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
+    wrap = TRUE,
+    n_intersections=10, set_sizes = FALSE, sort_sets = FALSE,
+    name = "Negative Selection Grouping",
+    base_annotations=list(
+        'Intersection size'=intersection_size(
+            mapping=aes(hjust=ifelse(after_stat(y)>3000, 0.5, 0.0)),
+            text=list(
+                vjust=0.5,
+                angle=90,
+                check_overlap=TRUE
+            )
+        )
+    ),
+    labeller = ggplot2::as_labeller(c(
+        coati = "COATi",
+        prank = "PRANK",
+        mafft = "MAFFT",
+        macse = "MACSE",
+        clustalo = "ClustalΩ"
+    ))
+)
+
+
+omega_data <- aln_data |>
+    mutate(neg_sel = omega > 1) |>
+    pivot_wider(id_cols = "sequence", names_from = "method", values_from = "neg_sel") |>
+    rename(coati = "tri-mg")
+
+gg2 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
+    n_intersections=10, set_sizes = FALSE, sort_sets = FALSE,
+    wrap = TRUE,
+    name = "Positive Selection Grouping",
+    base_annotations=list(
+        'Intersection size'=intersection_size(
+            mapping=aes(hjust=ifelse(after_stat(y)>3000, 0.5, 0.0)),
+            text=list(
+                vjust=0.5,
[…]

### 2024-01-05 scripts/aln_data.R

@@ -1,0 +1,117 @@
+library(ape)
+library(stringr)
+library(seqinr)
+
+# sequence,method,CIGAR,K2P,omega
+
+filtered = "tri-mg/data/filtered.csv"
+methods = c("clustalo", "macse", "mafft", "prank", "tri-mg")
+
+files = read.csv(file = filtered, header = FALSE)[,1]
+N = length(files) * length(methods)
+
+results = data.frame(sequence = character(N),
+                     method = character(N),
+                     cigar = character(N),
+                     k2p = numeric(N),
+                     omega = numeric(N))
+
+index = 1
+for(i in seq_along(files)) {
+    gene = files[i]
+    for(j in seq_along(methods)) {
+        method = methods[j]
+        path = paste0("tri-mg/aln/", method, "/", gene)
+        # if the file doesn't exist or the size is 0
+        if(!file.exists(path) | file.info(path)$size == 0) {
+            results[index,] = c(gene, method, NA, NA, NA)
+            index = index + 1
+            next
+        }
+        # cigar
+        cigar = create_cigar(path)
+        # k2p
+        distance = dist.dna(read.dna(path, format = "fasta"), model = "K80")[1]
+        # omega
+        omega = calculate_omega(path)
+        # save data
+        results[index,] = c(gene, method, cigar, distance, omega)
+        index = index + 1
+    }
+}
+
+write.csv(x = results, file = "aln_data.csv", quote = FALSE, row.names = FALSE)
+
+############################## Helper functions ################################
+## omega
+calculate_omega = function(fasta) { #dN/dS
+    if(!file.exists(fasta)) { return(NA) }
+    a = read.alignment(fasta, format = "fasta")
+    if(class(a) != "alignment") {return(NA)}
+    k = kaks(a)
+    if(length(k) == 1) {return(NA)}
+    ka = k$ka[1]
+    ks = k$ks[1]
+    if(ka < 0 || ks < 0) {
+        return(0)
+    } else if(ka == 0 && ks == 0) {
+        return(0)
+    } else if(ks == 0) {
+        return(10)
+    } else {
+        return(ka/ks)
+    }
+}
+
+## CIGAR
+create_cigar = function(fasta) {
+  if(file.exists(fasta)) {
+    seqs = read.fasta(fasta, set.attributes = TRUE, forceDNAtolower = FALSE)
+  } else {
[…]

### 2024-01-05 scripts/k2p_distance_plots.R

@@ -1,0 +1,201 @@
+library(ape)
+library(tidyverse)
+library(patchwork)
+
+models <- c("clustalo", "macse", "mafft", "prank", "tri-mg")
+ref_dir <- "supplementary_data/tri-mg/data/ref_alignments/"
+# models <- c("clustalo", "macse", "mafft", "prank", "mar-mg")
+# ref_dir <- "supplementary_data/mar-mg/data/ref_alignments/"
+alns <- list.files(ref_dir)
+
+distances <- matrix(data = NA, ncol = length(models)+1, nrow = length(alns))
+dir <- "supplementary_data/tri-mg/aln/ref/"
+# dir <- "supplementary_data/mar-mg/aln/ref/"
+
+for(i in seq_along(models)) {
+    for(j in seq_along(alns)) {
+        aln <- paste0(dir, models[i], "/", alns[j])
+        if(!file.exists(aln)) {
+            next
+        } else if(file.size(aln) == 0L) {
+            next
+        } else {
+            distances[j, i] <- dist.dna(read.dna(aln, format = "fasta"), model = "K80")[1]
+        }
+    }
+}
+
+for(i in seq_along(alns)) {
+    distances[i, length(models)+1] <- dist.dna(read.dna(paste0(ref_dir, alns[i]), format = "fasta"), model = "K80")[1]
+}
+
+distances <- as.data.frame(distances)
+colnames(distances) <- c("Clustalo", "MACSE", "MAFFT", "PRANK", "COATi", "Reference")
+
+plot_k2p <- function(distances, m) {
+    ggplot(distances, aes(x = get(m), y = Reference)) +
+        geom_point(alpha = 0.3) +
+        geom_abline(slope = 1, intercept = 0, color  = "red") +
+        xlim(0, 0.4) +
+        ylim(0, 0.4) +
+        coord_fixed(ratio = 1/1) +
+        xlab(m) +
+        theme_minimal()
+}
+
+gg1 = plot_k2p(distances, colnames(distances)[1])
+gg2 = plot_k2p(distances, colnames(distances)[2])
+gg3 = plot_k2p(distances, colnames(distances)[3])
+gg4 = plot_k2p(distances, colnames(distances)[4])
+gg5 = plot_k2p(distances, colnames(distances)[5])
+
+for(m in colnames(distances)[-6]) {
+    gg <- ggplot(distances, aes(x = get(m), y = Reference)) +
+        geom_point(alpha = 0.3) +
+        geom_abline(slope = 1, intercept = 0, color  = "red") +
[…]

### 2024-01-05 supplementary_materials/coati_fst/acceptor.R

@@ -1,0 +1,19 @@
+# convert a sequence in a fasta file into an acceptor
+# Output format is described at http://www.openfst.org/twiki/bin/view/FST/FstQuickTour#CreatingFsts
+ 
+library(stringr)
+library(seqinr)
+
+block_width <- 1
+
+acceptor_main = function(str) {
+	a <- str_split_1(str_to_upper(str),"")
+	n <- seq_along(a)
+	cat(str_c(n, n + 1, a, a, sep="\t"), sep="\n")
+	cat(length(a)+1, sep="\n")
+}
+
+if(!interactive()) {
+	ARGS = commandArgs(trailing=TRUE)
+	acceptor_main(str=ARGS[1])
+}

### 2024-01-05 supplementary_materials/coati_fst/indel.R

@@ -1,0 +1,70 @@
+# construct an indel FST
+
+library(stringr)
+library(Matrix)
+
+# parameters
+deletion <- 0.001
+insertion <- 0.001
+deletion_ext <- 1-1/6
+insertion_ext <- 1-1/6
+
+nucs <- c("A", "C", "G", "T")
+nuc_freqs <- c(0.308,0.185,0.199,0.308)
+
+eps <- "ε"
+
+# Function for creating a transition in an FST
+# Output format is described at http://www.openfst.org/twiki/bin/view/FST/FstQuickTour#CreatingFsts
+arc <- function(src, dest = NULL, ilab = eps, olab = eps, weight = 1.0) {
+    if(weight == 1.0) {
+        weight <- 0.0
+    } else {
+        weight <- -log(weight)
+    }
+
+    if(is.null(dest)) {
+        cat(str_c(src, weight, sep = "\t"), sep = "\n")
+    } else {
+        cat(str_c(src, dest, ilab, olab, weight, sep = "\t"), sep = "\n")
+    }
+}
+
+# build the fst
+# 0 = M
+# 1 = H
+# 2 = I
+# 3 = W
+# 4 = C
+# 5 = D
+# 6 = S
+
+# insertion
+arc(0, 1, weight = insertion)
+arc(0, 3, weight = 1.0 - insertion)
+
+for(i in 1:4) {
+    arc(1, 2, eps, nucs[i], weight = nuc_freqs[i])
+}
+
+arc(2, 1, weight = insertion_ext)
+arc(2, 3, weight = 1.0 - insertion_ext)
+
+# deletion
+arc(3, 4, weight = deletion)
+arc(3, 6, weight = 1.0 - deletion)
+
+for(i in 1:4) {
+    arc(4, 5, nucs[i], eps)
+}
+
+arc(5, 4, weight = deletion_ext)
+arc(5, 6, weight = 1.0 - deletion_ext)
+
+# matches
+for(i in 1:4) {
+    arc(6, 0, nucs[i], nucs[i])
+}
+
+# end
+arc(6)

### 2024-01-05 supplementary_materials/coati_fst/mutation.R

@@ -1,0 +1,128 @@
+# Construct an MG94 FST
+
+library(stringr)
+library(seqinr)
+library(Matrix)
+library(dplyr)
+library(tidyr)
+
+# parameters
+# Yang (1994) Estimating the pattern of nucleotide substitution
+params <- list(
+    time = 0.0133,
+    pi = c(0.308, 0.185, 0.199, 0.308),
+    sigma = c(0.715523, 2.9480112, 0.3246753, 1.1614831, 2.9075202, 0.6430366),
+    omega = 0.2
+)
+
+eps <- "ε"
+
+# Function for creating a transition in an FST
+# Output format is described at http://www.openfst.org/twiki/bin/view/FST/FstQuickTour#CreatingFsts
+arc <- function(src, dest = NULL, ilab = eps, olab = eps, weight = 1.0) {
+    if(weight == 1.0) {
+        weight <- 0.0
+    } else {
+        weight <- -log(weight)
+    }
+
+    if(is.null(dest)) {
+        cat(str_c(src, weight, sep = "\t"), sep = "\n")
+    } else {
+        cat(str_c(src, dest, ilab, olab, weight, sep = "\t"), sep = "\n")
+    }
+}
+
+universal_genetic_code <- function(remove_stops = FALSE) {
+    # genetic code in TCGA order
+    aa    <- "FFLLSSSSYY**CC*WLLLLPPPPHHQQRRRRIIIMTTTTNNKKSSRRVVVVAAAADDEEGGGG"
+    base1 <- "TTTTTTTTTTTTTTTTCCCCCCCCCCCCCCCCAAAAAAAAAAAAAAAAGGGGGGGGGGGGGGGG"
+    base2 <- "TTTTCCCCAAAAGGGGTTTTCCCCAAAAGGGGTTTTCCCCAAAAGGGGTTTTCCCCAAAAGGGG"
+    base3 <- "TCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAGTCAG"
+
+    aa <- str_split_1(aa, "")
+    base1 <- str_split_1(base1, "")
+    base2 <- str_split_1(base2, "")
+    base3 <- str_split_1(base3, "")
+
+    names(aa) <- str_c(base1, base2, base3)
+    
+    if(isTRUE(remove_stops)) {
+        aa <- aa[aa != "*"]
+    }
+
+    # return code in ACGT order
+    aa[order(names(aa))]
+}
+
[…]

### 2024-01-05 supplementary_materials/scripts/distance_pseudo.R

@@ -13,15 +13,15 @@ distance_main = function(metric, coati_model, aligners) {
     }
     
     distances = data.frame(filename = character(), aligner = character(), distance = numeric())
-    ref_alns = list.files(paste0("supplementary_data/", coati_model, "/data/ref_alignments/"))
+    ref_alns = list.files(paste0("data/", coati_model, "/data/ref_alignments/"))
     for(filename in ref_alns) {
-        ref = paste0("supplementary_data/", coati_model, "/data/ref_alignments/", filename)
+        ref = paste0("data/", coati_model, "/data/ref_alignments/", filename)
         # calculate distance between reference for each method
         file_distances = c()
         for(model in aligners) {
-            aln = paste0("supplementary_data/", coati_model, "/aln/ref/", model, "/", filename)
+            aln = paste0("data/", coati_model, "/aln/ref/", model, "/", filename)
             # call metal to calculate distance
-            d = system(paste("supplementary_materials/bin/metal", flag, ref, aln), intern = TRUE, ignore.stderr = TRUE)
+            d = system(paste("bin/metal", flag, ref, aln), intern = TRUE, ignore.stderr = TRUE)
             
             if(length(d) == 0) {
                 break
@@ -38,7 +38,7 @@ distance_main = function(metric, coati_model, aligners) {
     }
     colnames(distances) = c("filename", "aligner", metric)
     write.csv(x = distances,
-              file = paste0("supplementary_data/", coati_model, "/plot_distance.csv"),
+              file = paste0("data/", coati_model, "/plot_distance.csv"),
               col.names = TRUE,
               quote = FALSE,
               row.names = FALSE)

### 2024-01-05 supplementary_materials/scripts/results_gor_ref.R

@@ -6,11 +6,11 @@ library(tidyverse)
 gor_ref_results = function() {
     
     ## Clean and save gor-ref results
-    results_raw = read.csv("supplementary_data/gor-ref/results/results_summary.csv",
+    results_raw = read.csv("data/gor-ref/results/results_summary.csv",
                        header = TRUE)
     alns = results_raw |> filter(aligner == "gor-ref") |> drop_na(score) |> select(ref_name) |> unlist()
     results = results_raw |> filter(ref_name %in% alns)
-    results_file = "supplementary_data/gor-ref/results/results_summary_clean.csv"
+    results_file = "data/gor-ref/results/results_summary_clean.csv"
     write_csv(x = results, file = results_file)
     
     trimg = results |> filter(aligner == "tri-mg")
@@ -33,7 +33,7 @@ gor_ref_results = function() {
     
     ###########################################################################
     # perfect, best, and imperfect alignments
-    source("supplementary_materials/scripts/number_alignments.R")
+    source("scripts/number_alignments.R")
     alns = num_alns(results_file)
     
     # fix row and column names and add to stats table
@@ -43,7 +43,7 @@ gor_ref_results = function() {
     
     ############################################################################
     # selection (ka ks)
-    source("supplementary_materials/scripts/kaks.R")
+    source("scripts/kaks.R")
     
     selection = data.frame("F1.pos" = double(),
                            "F1.neg" = double())
@@ -73,7 +73,7 @@ library(seqinr)
 # count how many alignments didn't pass filtering for coati using gorilla as the reference
 #  and the reasons (i.e., incomplete codons, early stop codons, ambiguous nucleotides)
 filter_counts = function() {
-    results = read.csv("supplementary_data/gor-ref/results/results_summary.csv", header = TRUE)
+    results = read.csv("gor-ref/results/results_summary.csv", header = TRUE)
     filtered = results |> filter(aligner == "gor-ref") |>
         filter(is.na(score)) |> select(ref_name) |> unlist()
     incomplete = 0
@@ -81,7 +81,7 @@ filter_counts = function() {
     ambiguous = 0
     
     for(aln_name in filtered) {
-        aln = read.fasta(file = paste0("supplementary_data/tri-mg/data/no_gaps_ref/",
+        aln = read.fasta(file = paste0("tri-mg/data/no_gaps_ref/",
                                        aln_name),
                          seqonly = TRUE)
     

### 2024-01-11 alignpair_letter.tex
Update data

@@
 
+\subsection*{Empirical Dataset}
+
+Humans and gorillas are two closely related species with very different levels of genome curation.
+The human reference genome has been revised dozens of times and is currently on version GRCh38.p14, while the gorilla reference genome has only been revised a handful of times and is currently on version gorGor4 \citep[cf.\ ENSEMBL database v110;][]{ensembl_hubbard_2002}.
+Additionally significant levels of investment have been made to correctly identify and annotate human genes, while gorilla annotations have received limited support in comparison.
+Together, these reference genomes provide a good opportunity to compare COATi against other aligner as they offer one genome that is high-quality (human) and sister genome that is lower-quality (gorilla).
+
+We used ENSEMBL database v110 \citep{ensembl_hubbard_2002} to create an empirical dataset of protein-coding sequences for both human genes and their gorilla orthologs.
+We first selected human protein-coding genes that belonged to the Consensus Coding Sequence Set, that were located on an autosomal chromosome, and that had a one-to-one gorilla ortholog.
+We selected the canonical isoform for both species and removed any pair in which the total nucleotide length was larger larger than 6,000 nucleotides.
+This resulted in 14,127 sequence pairs and corresponding FASTA files containing CDS sequences.
+Due to the way that canonical isoforms are identified, there is no guarantee that the isoforms are orthologous even though the genes are.
+Therefore, a subset of the sequence pairs in this dataset contain human and gorilla sequences with different exon compositions.
+We have made no attempt to correct these sequence pairs because in our experience genome-wide studies rarely control for such artifacts.
+
 \subsection*{Alignment Strategies}

### 2024-01-11 figures/fig-k2p-empirical.R

@@ -72,8 +72,8 @@ pal <- pal[-5]
 seq_id <- k2p_sum |> filter(all_same == FALSE) |> pull(sequence)
 
 k2p_data <- aln_data |>
-    mutate(k2p_norm = k2p - k2p[method == "tri-mg"], .by = "sequence") |>
-    filter(method != "tri-mg")
+    mutate(k2p_norm = k2p - k2p[method == "coati-tri-mg"], .by = "gene") |>
+    filter(method != "coati-tri-mg")
 
 #k2p_data <- k2p_data |> filter(sequence %in% seq_id)
 

### 2024-01-11 figures/fig-sel-empirical.R

@@ -19,8 +19,8 @@ aln_data <- read_csv(here("data/aln_data.csv.gz"))
 
 omega_data <- aln_data |>
     mutate(neg_sel = omega < 1) |>
-    pivot_wider(id_cols = "sequence", names_from = "method", values_from = "neg_sel") |>
-    rename(coati = "tri-mg")
+    pivot_wider(id_cols = "gene", names_from = "method", values_from = "neg_sel") |>
+    rename(coati = "coati-tri-mg")
 
 gg1 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
     wrap = TRUE,
@@ -48,8 +48,8 @@ gg1 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
 
 omega_data <- aln_data |>
     mutate(neg_sel = omega > 1) |>
-    pivot_wider(id_cols = "sequence", names_from = "method", values_from = "neg_sel") |>
-    rename(coati = "tri-mg")
+    pivot_wider(id_cols = "gene", names_from = "method", values_from = "neg_sel") |>
+    rename(coati = "coati-tri-mg")
 
 gg2 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
     n_intersections=10, set_sizes = FALSE, sort_sets = FALSE,

### 2024-01-12 figures/fig-k2p-empirical.R
Begin updating empirical results

@@ -13,36 +13,15 @@ cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
 set.seed(17760)
 
 # load data
-aln_data <- read_csv(here("data/aln_data.csv.gz"))
-
-kounts <- aln_data |> count(method, failed=is.na(k2p)) |> filter(failed)
-
-cat("Table of failed sequences\n")
-cat("=========================\n")
-print(kounts)
-
-# identify files that didn't show variation
-k2p_sum <- aln_data |> summarize(
-    coati_na = is.na(k2p[method == "tri-mg"]),
-    all_same = n_distinct(k2p) == 1L,
-    .by = "sequence")
-
-kounts <- k2p_sum |> count(all_same)
-
-cat("Table of trivial alignments\n")
-cat("=========================\n")
-print(kounts)
-
-# remove sequences that COATi failed on
-seq_ids <- k2p_sum |> filter(coati_na == FALSE) |> pull(sequence)
-aln_data <- aln_data |> filter(sequence %in% seq_ids)
-
+aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 
 pal <- cbbPalette[c(2,3,4,6,7)]
 
 #### PLOT 1: K2P density plots
+methods <- c("clustalo", "macse", "mafft", "prank", "coati-tri-mg")
+aln_data2 <- aln_data |> filter(method %in% methods)
 
-gg1 <- ggplot(aln_data, aes(x = k2p, color=method)) + geom_density(linewidth = 1)
+gg1 <- ggplot(aln_data2, aes(x = k2p, color=method)) + geom_density(linewidth = 1)
 gg1 <- gg1 + scale_x_continuous(trans = transform_pseudo_log(0.0005, base=10),
     breaks=c(0, 0.001, 0.01, 0.1, 1))
 gg1 <- gg1 + coord_cartesian(ylim=c(NA, 1.51))
@@ -53,7 +32,7 @@ gg1 <- gg1 + scale_color_manual(values = pal,
      "macse" = "MACSE",
      "mafft" = "MAFFT",
      "prank" = "PRANK",
-     "tri-mg" = "COATi"
+     "coati-tri-mg" = "COATi"
      )
     )
 gg1 <- gg1 + guides(color = guide_legend(override.aes=list(fill=pal)))
[…]

### 2024-01-12 figures/fig-sel-empirical.R

@@ -7,7 +7,7 @@ PDF_WIDTH <- 7 # 178mm
 PDF_HEIGHT <- PDF_WIDTH/2
 
 # load data
-aln_data <- read_csv(here("data/aln_data.csv.gz"))
+aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 
 # omega_data <- aln_data |>
 #     mutate(neg_sel = (!is.na(omega) & omega < 1)) |>

### 2024-01-12 tables/tab-gaps.R

@@ -1,0 +1,41 @@
+options(tidyverse.quiet = TRUE,
+    readr.show_col_types = FALSE,
+    conflicts.policy = list(warn = FALSE ))
+
+library(tidyverse)
+library(here)
+
+aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
+
+cigar_to_rle <- function(x) {
+    len <- str_extract_all(x, "\\d+")
+    val <- str_extract_all(x, "[A-Za-z]")
+
+    result <- map2(len, val, function(n, v) {
+        structure(list(lengths = as.integer(n), values = v), class = "rle")
+    })
+    result
+}
+
+tbl_cigar <- function(x) {
+    x <- cigar_to_rle(x)
+    x <- map(x, \(y) as_tibble(unclass(y)))
+    x
+}
+
+tab <- aln_data |> select(gene, method, cigar) |> 
+    mutate(rle = tbl_cigar(cigar)) |>
+    unnest(rle)
+
+k <- tab |> filter(values != "M") |>
+    mutate(lengths = if_else(lengths <= 6, lengths, 7 + ((lengths-1) %% 3))) |>
+    arrange(lengths) |>
+    count(method, lengths, values) |>
+    complete(method, values, lengths = full_seq(lengths, 1), fill = list(n = 0)) |>
+    pivot_wider(names_from = "lengths", values_from = "n")
+
+# print(xtable::xtable(k), floating=FALSE, latex.environments=NULL, booktabs=TRUE)
+
+# tab |> filter(lengths %% 3 > 0 & method == "clustalo")
+
+# tab |> group_by(method, gene) |> summarize(p = sum(lengths[values != "M"])/(sum(lengths) + sum(lengths[values == "M"]))) |> summarize(mean(p, na.rm = TRUE))

### 2024-01-13 alignpair_letter.tex
Update manuscript

@@
 
-\subsection*{Empirical Dataset}
+\subsection*{Empirical Dataset and Alignments}
 
@@
 
-\subsection*{Alignment Strategies}
+\paragraph{Alignment Methods.}
 
-In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ v1.2.4 via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE v2.06, (4) the DNA model of MAFFT v7.505, and (5) the codon model of PRANK v.150803 \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}.
-Together, these five methods allow us to evaluate both different alignment strategies and different software implementations.
+In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE, (4) the DNA model of MAFFT, and (5) the codon model of PRANK \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}.
+COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
+In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATI-rev).
+Together, these six methods allow us to evaluate both different alignment strategies and different software implementations.
+See Supplemental Materials for additional details, including results for COATi's marginal and ECM models.
+
+We aligned our dataset of human-gorilla orthologs using all six alignment methods and calculated multiple biological and technical statistics on each alignment in order to compare how different alignment methods influence biological conclusions.
+%
+Additionally, we checked alignments to ensure that our pipeline did not introduce any artifacts into the results, including unexpected characters, empty columns, and aligned sequences of different lengths.
+We also generated checksums of our sequences to ensure that they were not modified during alignment.
+
+\paragraph{Evolutionary Distances and Gaps.}
+
+Alignment inference impacts the estimation of evolutionary distances.
+To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination.
+K2P corrects for multiple mutations at a site and takes into account differences in transition and transversion rates.
+It also assumes equal nucleotide frequencies and no variations in the rate of substitution across sites.
+While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
+We used the R software package ape \citep{paradis2019ape} to calculate K2P distances.
+%
+Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the lengths and phases of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a gap.
+
+\paragraph{Selection.}
+Alignment inference also impacts the identification of genes that have experienced positive and negative selection.
+To evaluate the impact of aligners on selection identification, we estimated $K_S$ and $K_A$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment.
+Briefly, $K_S$ is the number of substitutions per synonymous site and $K_A$ is the number of substitutions per non-synonymous site between two protein-coding sequences.
+We used the method developed by \cite{ka_ks_li_1993} and independently derived by \cite{Pamilo1993} to estimate these statistics as implemented in the R package seqinr \citep{seqinr}.
+
+Briefly, this method takes two aligned sequences and classifies the sites in each sequence as non-degenerate, two-fold degenerate, or four-fold degenerate based on the standard genetic code.
+A site is non-degenerate if 0/3 possible nucleotide changes to that site are synonymous, two-fold degenerate if 1/3 of the possible changes are synonymous, and four-fold degenerate if 3/3 possible changes are synonymous.
+(The rare three-fold degenerate sites are treated as two-fold degenerate in this method.) First the method calculates the average numbers of sites of each type in the sequence pair ($L_0$, $L_2$, and $L_4$).
+Next, following \cite{Li1985}, it uses Kimura's two-parameter model \citep{kimura1980simple} and the alignment to estimate the numbers of transitions ($A_i$) and transversions ($B_i$) that occurred per each $i$-th type.
+Using these statistics, \cite{ka_ks_li_1993} estimated $K_S$ and $K_A$ as
+\begin{align*}
+K_S &= \frac{L_2 A_2 + L_4 A_4}{L_2 + L_4} + B_4 &
+K_A &= A_0 + \frac{L_0 B_0 + L_2 B_2}{L_0 + L_2}
+\end{align*}
+
+We considered any alignment to be showing evidence of positive selection if $K_A/K_S > 1$ and negative selection if $K_A/K_S < 1$.
+We estimated F\textsubscript{1} scores for both positive and negative selection by comparing estimated alignments to benchmark alignments.
+F\textsubscript{1} is the harmonic mean of precision and recall:
+\[
+F_1 = \frac{2 \text{TP}}{2 \text{TP} + \text{FP} + \text{FN}}
+\]
+where TP is the number of alignments that correctly predicted positive (or negative) selection, FP is the number of alignments that incorrectly predicted positive (or negative) selection, and FN is the number of alignments that incorrectly predicted the absence of positive (or negative) selection.
+F\textsubscript{1} allows us to measure how well aligners produce alignments that correctly identify the presence and absence of positive selection or negative selection.
 
@@
 
+Alignment error also negatively impacts the estimation of evolutionary distances and phylogenetic trees.
+To quantify the impact of aligners on the estimation of evolutionary distances, we compared Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} calculated for the estimated and benchmark alignments.
+
+Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
+
+Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
+To evaluate the impact of aligners on selection identification, we estimated $K_S$ and $K_A$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment.
+
 \paragraph{Alignment Error.}
@@
 
-\paragraph{Selection.}
-Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
-To evaluate the impact of aligners on selection identification, we estimated $K_S$ and $K_A$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment.
-Briefly, $K_S$ is the number of substitutions per synonymous site and $K_A$ is the number of substitutions per non-synonymous site between two protein-coding sequences.
-We used the method developed by \cite{ka_ks_li_1993} and independently derived by \cite{Pamilo1993} to estimate these statistics as implemented in the R package seqinr \citep{seqinr}.
-
-Briefly, this method takes two aligned sequences and classifies the sites in each sequence as non-degenerate, two-fold degenerate, or four-fold degenerate based on the standard genetic code.
-A site is non-degenerate if 0/3 possible nucleotide changes to that site are synonymous, two-fold degenerate if 1/3 of the possible changes are synonymous, and four-fold degenerate if 3/3 possible changes are synonymous.
-(The rare three-fold degenerate sites are treated as two-fold degenerate in this method.) First the method calculates the average numbers of sites of each type in the sequence pair ($L_0$, $L_2$, and $L_4$).
-Next, following \cite{Li1985}, it uses Kimura's two-parameter model \citep{kimura1980simple} and the alignment to estimate the numbers of transitions ($A_i$) and transversions ($B_i$) that occurred per each $i$-th type.
-Using these statistics, \cite{ka_ks_li_1993} estimated $K_S$ and $K_A$ as
-\begin{align*}
-K_S &= \frac{L_2 A_2 + L_4 A_4}{L_2 + L_4} + B_4 &
-K_A &= A_0 + \frac{L_0 B_0 + L_2 B_2}{L_0 + L_2}
-\end{align*}
-
-We considered any alignment to be showing evidence of positive selection if $K_A/K_S > 1$ and negative selection if $K_A/K_S < 1$.
-We estimated F\textsubscript{1} scores for both positive and negative selection by comparing estimated alignments to benchmark alignments.
-F\textsubscript{1} is the harmonic mean of precision and recall:
-\[
-F_1 = \frac{2 \text{TP}}{2 \text{TP} + \text{FP} + \text{FN}}
-\]
-where TP is the number of alignments that correctly predicted positive (or negative) selection, FP is the number of alignments that incorrectly predicted positive (or negative) selection, and FN is the number of alignments that incorrectly predicted the absence of positive (or negative) selection.
-F\textsubscript{1} allows us to measure how well aligners produce alignments that correctly identify the presence and absence of positive selection or negative selection.
-
 % First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate.
@@
 
-\paragraph{Evolutionary Distance.}
-Alignment error also negatively impacts the estimation of evolutionary distances and phylogenetic trees.
-To quantify the impact of aligners on the estimation of evolutionary distances, we compared Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} calculated for the estimated and benchmark alignments.
-This distance corrects for multiple mutations at a site and takes into account differences in transition and transversion rates.
-It also assumes equal nucleotide frequencies and no variations in the rate of substitution across sites.
-While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
-We used the R software package ape \citep{paradis2019ape} to calculate K2P distances.
-
 % e compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura's 2-parameter distance \citep{kimura1980simple}.
@@
 
-\section*{Results and Discussion}
+\section*{Results}
+
+\paragraph{Empirical Data and Alignments.}
+
+We generated 14,127 sequence pairs containing the coding sequences of a human gene and the orthologous gorilla gene.
+22 human sequences contained early stop codons.
+These were not artifacts, but rather UGA codons which encoded for selenocysteine.
+1 human sequence had a length that was not a multiple of 3, and no human sequences contained ambiguous nucleotides.
+Conversely, no gorilla sequences contained early stop codons (i.e.\ proteins containing selenocysteines were improperly annotated).
+20 gorilla sequences had lengths that were not multiples of 3, and 173 sequences contained ambiguous nucleotides.
+
+COATi was not able to align 23 sequence pairs, PRANK was not able to align 24 sequence pairs, and COATi-rev was not able to align 193 sequence pairs.
+MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from our wrapper script.
+
+\begin{figure}[h!]
+    \centering%
+    \includegraphics{figures/fig-k2p-empirical.pdf}
+    \par
+    \caption{COATi.}
+    \label{fig:k2p-empirical}
+\end{figure}
+
+\section*{Discussion}
 % Using 16000 human genes and their gorilla orthologs from the ENSEMBL

### 2024-01-13 supplementary_materials/supplementary_materials.Rmd
Update manuscript

@@
   - \renewcommand{\thefigure}{S\oldthefigure}
+  - \let\oldthetable\thetable
+  - \renewcommand{\thetable}{S\oldthetable}
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, include = FALSE}
+library(knitr)
+library(tidyverse)
+library(kableExtra)
+library(here)
+
+knitr::opts_chunk$set(echo = FALSE)
 ```
 
-```{r echo = FALSE}
-suppressPackageStartupMessages(library(knitr))
-suppressPackageStartupMessages(library(tidyverse))
+```{r}
 
@@
 
-- COATi: `coati alignpair -m tri-mg ...`
-- ClustalΩ v1.2.4: `clustalo ...`
+- COATi: `coati alignpair -m tri-mg ...`; We used COATi two different ways.
+The primary way used human as the reference sequence.
+The secondary way, COATi-rev, used gorilla as the reference, via a wrapper script that reversed the order of sequences before alignment and reversed the order back afterwards.
+- ClustalΩ v1.2.4: `clustalo --seqtype=Protein ...`; We wrapped ClustalΩ with a script that translates DNA sequences into amino acid sequences (including stops) before alignment.
+Any codons that were partial or containing ambiguous characters were translated as "X".
+The script then aligned these translated sequences with ClustalΩ, and then created a DNA alignment that was consistent with the amino-acid alignment.
+- MACSE v2.06: `java -jar macse.jar -prog alignSequences -seq human.fasta -seq_lr gorilla.fasta ...`; We wrapped MACSE with a script that created two temporary fasta files, one containing the human sequence and another containing the gorilla sequence.
+The human sequence was specified as the reliable sequence and the gorilla sequence was specified as the less-reliable sequence.
+MACSE uses "!" to mark gaps that result from frameshifts, and the wrapper script replaced these with "-".
+Additionally, MACSE sometimes produced columns that only contained gaps, and these columns were removed.
 - PRANK v.150803: `prank -codon ...`
-- MACSE v2.06: `java -jar macse.jar -fs_lr 10 -stop_lr 10 -prog alignSequences ...`
-- MAFFT v7.505: `mafft --preservecase --globalpair --maxiterate 1000 ...`
+- MAFFT v7.520: `mafft --preservecase --globalpair --maxiterate 1000 ...`
 
@@
 
-```{r aln-example-a, echo = FALSE, fig.align='center', fig.cap="The FST of all possible alignments between `CTC' and `CTG'."}
+```{r aln-example-a, fig.align='center', fig.cap="The FST of all possible alignments between `CTC' and `CTG'."}
 knitr::include_graphics("figures/aln-example-graph.pdf")
@@
 
-```{r aln-example-b, echo = FALSE, out.width="75%", fig.align='center', fig.cap="The best alignment of `CTC' and `CTG'."}
+```{r aln-example-b, out.width="75%", fig.align='center', fig.cap="The best alignment of `CTC' and `CTG'."}
 knitr::include_graphics("figures/aln-example-path.pdf")
@@
 
-```{r setup-table-results, echo = FALSE}
-suppressMessages(library(kableExtra))
-suppressMessages(library(knitr))
-suppressMessages(library(tidyverse))
+# Gap Patterns
+
+## Lengths
+
+We quantified the lengths of gaps produced by each method across the empirical dataset of human-gorilla protein-coding-sequence pairs (Tab.
+\@ref(tab:gap-table-1)).
+Here the gap type is either "D" for gaps introduced into the gorilla sequence or "I" for gaps introduced in the human sequence.
+"COATi" refers to the full COATi FST model with a MG substitution model (i.e.
+tri-mg).
+"COATi-rev" refers to the same model, but using gorilla as the reference and human as the non-reference sequence.
+Gap lengths of 1--6 nucleotides were binned into their respective columns.
+Gap lengths longer than 6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.
+
+We also quantified the lengths of gaps produced by different COATi models (Tab. \@ref(tab:gap-table-2)).
+
+Note that ClustalΩ gaps with lengths that are not a multiple of 3 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.
+
+```{r gap-table-1, results = 'asis', message = FALSE}
+gap_data <- read_csv(here("supplementary_materials/data/gap-lengths.csv"))
+
+gap_data <- gap_data |> rename("Gap Type" = "values", "Method" = "method",
+  "7+" = "7", "8+" = "8", "9+" = "9")
+
+gap_data1 <- gap_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
+    mutate(Method = case_match(Method,
+        "clustalo" ~ "ClustalΩ",
+        "coati-tri-mg" ~ "COATi",
+        "macse" ~ "MACSE",
+        "mafft" ~ "MAFFT",
+        "prank" ~ "PRANK",
+        "rev-coati-tri-mg" ~ "COATi-rev"
+        )) |>
+    arrange(Method, "Gap Type")
+
+k <- kable(gap_data1, align = "lcrrrrrrrrr", caption="Number of gaps introduced by each method separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position") |>
+  add_header_above(c(" " = 2L, "Gap Lengths" = 9L))
 ```
 
-```{r table-results-1, echo = FALSE, results = 'asis'}
-results = "data/tri-mg/results/results_summary.csv"
-k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs.
-Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 42 empty alignments, calculations are based on 7719 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-```
-```{r table-results-2, echo = FALSE, results = 'asis'}
-results = "data/tri-ecm/results/results_summary.csv"
-k = kable(summary_table(results, "tri-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7761 simulated sequence pairs.
-Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 69 empty alignments, calculations are based on 7692 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-```
-```{r table-results-3, echo = FALSE, results = 'asis'}
-results = "data/mar-mg/results/results_summary.csv"
-k = kable(summary_table(results, "mar-mg"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7755 simulated sequence pairs.
-Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 60 empty alignments, calculations are based on 7695 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-```
-```{r table-results-4, echo = FALSE, results = 'asis'}
-results = "data/mar-ecm/results/results_summary.csv"
-k = kable(summary_table(results, "mar-ecm"), digits = 5, caption = paste("Accuracy of COATi codon-marginal-ecm, PRANK, MAFFT, ClustalOmega, and MACSE on 7767 simulated sequence pairs.
-Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 49 empty alignments, calculations are based on 7718 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-```
-```{r table-results-5, echo = FALSE, results = 'asis'}
-results = "data/reverse/results/results_summary.csv"
-k = kable(summary_table(results, "tri-mg"), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, and MACSE on 7798 simulated sequence pairs with gorilla as the reference.
-Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(footnote(k, general = "PRANK produced 35 empty alignments, calculations are based on 7763 alignments.", footnote_as_chunk = TRUE, general_title = "*"))
-```
-```{r table-results-6, echo = FALSE, results = 'asis'}
-source("scripts/results_gor_ref.R")
-k = kable(gor_ref_results(), digits = 5, caption = paste("Accuracy of COATi codon-triplet-mg, PRANK, MAFFT, ClustalOmega, MACSE, and codon-triplet-mg with gorila as the reference on 4003 of the 7761 simulated sequence pairs where the gorilla sequence was simulated without early stop codons, incomplete codons, or ambiguous nucleotides.
-Perfect alignments have the same score as the true alignment, best alignments have lowest $d_{seq}$, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment."), escape = FALSE) %>% kable_styling(latex_options = "HOLD_position")
-print(k)
+```{r gap-table-2, results = 'asis', message = FALSE}
+gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
+    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
+        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
+    arrange(Method, "Gap Type") |>
+    rename(Model = Method) |>
+    mutate(Model = fct_recode(Model,
+        "TRI-MG" = "coati-tri-mg",
+        "TRI-ECM" = "coati-tri-ecm",
+        "MAR-MG" = "coati-mar-mg",
+        "MAR-ECM" = "coati-mar-ecm",
+        "DNA" = "coati-dna"
+    ))
+
+k <- kable(gap_data2, align = "lcrrrrrrrrr", caption="Number of gaps introduced by different COATi models separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position") |>
+  add_header_above(c(" " = 2L, "Gap Lengths" = 9L))
 ```
 
-\newpage
+## Phases
 
-```{r base-calling-error, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap="\\label{fig:figs} Base calling error FST.
-Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
-#knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
+We quantified the phases of gaps produced by different aligners and COATi models (Tab.
+\@ref(tab:gap-table-3) and \@ref(tab:gap-table-4)).
+Phase 1 gaps begin after the 1st position in a codon in the reference sequence, phase 2 gaps begin after the 2nd position in a codon, and phase 3 begin after the 3rd position in a codon (i.e.
+between codons).
+Phase 3 gaps are also known as phase 0 gaps.
+
+Note that ClustalΩ gaps with phases of 1 and 2 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.
+
+```{r gap-table-3, results = 'asis', message = FALSE}
+gap_data <- read_csv(here("supplementary_materials/data/gap-phases.csv"))
+
+gap_data <- gap_data |> rename("Method" = "method")
+
+gap_data1 <- gap_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
+    mutate(Method = case_match(Method,
+        "clustalo" ~ "ClustalΩ",
+        "coati-tri-mg" ~ "COATi",
+        "macse" ~ "MACSE",
+        "mafft" ~ "MAFFT",
+        "prank" ~ "PRANK",
+        "rev-coati-tri-mg" ~ "COATi-rev"
+        )) |>
+    arrange(Method)
+
+k <- kable(gap_data1, align = "lrrr", caption="Number of gaps introduced by each alignment method separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position") |>
+  add_header_above(c(" " = 1L, "Gap Phases" = 3L))
 ```
 
-```{r pvals, echo = FALSE}
-results = "data/tri-mg/results/results_summary.csv"
-pval_tri_mg = max(wilcoxon_test(results, "tri-mg"))
-results = "data/tri-ecm/results/results_summary.csv"
-pval_tri_ecm = max(wilcoxon_test(results, "tri-ecm"))
-results = "data/mar-mg/results/results_summary.csv"
-pval_mar_mg = max(wilcoxon_test(results, "mar-mg"))
-results = "data/mar-ecm/results/results_summary.csv"
-pval_mar_ecm = max(wilcoxon_test(results, "mar-ecm"))
-results = "data/reverse/results/results_summary.csv"
-pval_reverse = max(wilcoxon_test(results, "tri-mg"))
+```{r gap-table-4, results = 'asis', message = FALSE}
+gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
+    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
+        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
+    arrange(Method) |>
+    rename(Model = Method) |>
+    mutate(Model = fct_recode(Model,
+        "TRI-MG" = "coati-tri-mg",
+        "TRI-ECM" = "coati-tri-ecm",
+        "MAR-MG" = "coati-mar-mg",
+        "MAR-ECM" = "coati-mar-ecm",
+        "DNA" = "coati-dna"
+    ))
+
+k <- kable(gap_data2, align = "lrrr", caption="Number of gaps introduced by different COATi models separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position") |>
+  add_header_above(c(" " = 1L, "Gap Phases" = 3L))
 ```
 
-```{r dseq1, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
-COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
-library(knitr)
-source("scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("data/tri-mg/plot_distance.csv", "tri-mg", save = TRUE))
-knitr::include_graphics(path = "figures/dseq_plots_tri-mg.pdf")
+## Gapiness
+
+We quantified the gapiness of different aligners and COATi models (Tab.
+\@ref(tab:gap-table-5) and \@ref(tab:gap-table-6)).
+Here we define gapiness as the tendency of an aligner to introduce gaps into an alignment, and we measure it by calculating the percentage of nucleotides aligned with a gap.
+Note that this is different than the percent of columns that contain a gap.
+
+```{r gap-table-5, results = 'asis', message = FALSE}
+gap_data <- read_csv(here("supplementary_materials/data/gapiness.csv"))
+
+gap_data <- gap_data |> rename("Method" = "method") |>
+    mutate("Gapiness" = sprintf("%0.2f%%",round(p * 100, 2))) |>
+    select(-p)
+
+gap_data1 <- gap_data |>
+    filter(Method %in% c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
+    mutate(Method = case_match(Method,
+        "clustalo" ~ "ClustalΩ",
+        "coati-tri-mg" ~ "COATi",
+        "macse" ~ "MACSE",
+        "mafft" ~ "MAFFT",
+        "prank" ~ "PRANK",
+        "rev-coati-tri-mg" ~ "COATi-rev"
+        )) |>
+    arrange(Method) |>
+    pivot_wider(names_from = "Method", values_from = "Gapiness")
+
+k <- kable(gap_data1, align = "ccccc", caption="Average gapiness of alignments separated by alignment method.", booktabs = TRUE, linesep = c("\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position")
 ```
 
-```{r dseq2, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
-COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
-library(knitr)
-source("scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("data/tri-ecm/plot_distance.csv", "tri-ecm", save = TRUE))
-knitr::include_graphics(path = "figures/dseq_plots_tri-ecm.pdf")
+```{r gap-table-6, results = 'asis', message = FALSE}
+gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
+    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
+        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
+    arrange(Method) |>
+    rename(Model = Method) |>
+    mutate(Model = fct_recode(Model,
+        "TRI-MG" = "coati-tri-mg",
+        "TRI-ECM" = "coati-tri-ecm",
+        "MAR-MG" = "coati-mar-mg",
+        "MAR-ECM" = "coati-mar-ecm",
+        "DNA" = "coati-dna"
+    ))  |>
+    pivot_wider(names_from = "Model", values_from = "Gapiness")
+
+k <- kable(gap_data2, align = "ccccc", caption="Average gapiness of alignments separated by COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position")
 ```
-
-```{r dseq3, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE.
-COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
-library(knitr)
-source("scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("data/mar-mg/plot_distance.csv", "mar-mg", save = TRUE))
-knitr::include_graphics(path = "figures/dseq_plots_mar-mg.pdf")
-```
-
-```{r dseq4, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE.
-COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
-library(knitr)
-source("scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("data/mar-ecm/plot_distance.csv", "mar-ecm", save = TRUE))
-knitr::include_graphics(path = "figures/dseq_plots_mar-ecm.pdf")
-```
-
-```{r dseq5, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-reverse} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE with gorilla as the reference.
-COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_reverse, digits = 3), "$.")}
-library(knitr)
-source("scripts/plot_dseq.R")
-suppressWarnings(plot_dseq_main("data/reverse/plot_distance.csv", "reverse", save = TRUE))
-knitr::include_graphics(path = "figures/dseq_plots_reverse.pdf")
-```
-
-\newpage
-
-```{r k2p-dist, echo = FALSE, fig.align = 'center', out.width="75%", fig.cap="\\label{fig:k2p} Alignments generated by COATi produce accurate evolutionary distances.
-Benchmark alignments were realigned by COATi and four other methods as described in the main text.
-Kimura two-parameter (K2P) distances were estimated from the benchmark alignments and the estimated alignments.
-Each panel is a scatter plot comparing the benchmark disances (Reference) with distances estimated via one of the aligners.
-COATi clearly performs better than other tools, which tend to drastically overestimate some distances."}
-knitr::include_graphics(path = "figures/k2p_distances.pdf")
-```
-
-<!-- \newpage -->
-
-<!-- # Supplementary Methods -->
-
-<!-- Ks and Ka represent the number of substitutions per synonymous and non-synonymous sites.
-The ratio of nonsynonymous (Ka) to synonymous (Ks) nucleotide substitution rates indicates the selective pressures acting on genes.
-If the ratio is significantly greater than 1, it suggests positive selective pressure, meaning that nonsynonymous substitutions occur more frequently than synonymous substitutions.
-A ratio around 1 can indicate either neutral evolution at the protein level or a mixture of positive and negative selective pressures.
-If the ratio is less than 1, it indicates a pressure to maintain protein sequence, known as purifying selection.
--->
-<!-- Ks and Ka are calculated using the R package seqinr v.4.2-30 [@seqinr]. -->
 

### 2024-01-13 figures/fig-k2p-empirical.R

@@ -15,13 +15,14 @@ set.seed(17760)
 # load data
 aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 
-pal <- cbbPalette[c(2,3,4,6,7)]
+pal <- rev(cbbPalette[c(3,2,4,6,7,8)])
 
 #### PLOT 1: K2P density plots
-methods <- c("clustalo", "macse", "mafft", "prank", "coati-tri-mg")
+methods <- c("clustalo", "macse", "mafft", "prank", "coati-tri-mg", "rev-coati-tri-mg")
 aln_data2 <- aln_data |> filter(method %in% methods)
+aln_data2 <- aln_data2 |> mutate(method = fct_relevel(method, "coati-tri-mg"))
 
-gg1 <- ggplot(aln_data2, aes(x = k2p, color=method)) + geom_density(linewidth = 1)
+gg1 <- ggplot(aln_data2, aes(x = k2p, color =  fct_rev(method))) + geom_density(linewidth = 1)
 gg1 <- gg1 + scale_x_continuous(trans = transform_pseudo_log(0.0005, base=10),
     breaks=c(0, 0.001, 0.01, 0.1, 1))
 gg1 <- gg1 + coord_cartesian(ylim=c(NA, 1.51))
@@ -32,10 +33,11 @@ gg1 <- gg1 + scale_color_manual(values = pal,
      "macse" = "MACSE",
      "mafft" = "MAFFT",
      "prank" = "PRANK",
-     "coati-tri-mg" = "COATi"
+     "coati-tri-mg" = "COATi",
+     "rev-coati-tri-mg" = "COATi-rev"
      )
     )
-gg1 <- gg1 + guides(color = guide_legend(override.aes=list(fill=pal)))
+gg1 <- gg1 + guides(color = guide_legend(override.aes=list(fill=rev(pal)), reverse = TRUE))
 gg1 <- gg1 + theme_minimal(9) + theme(
     legend.position = c(1,1),
     legend.justification = c(1, 1),
@@ -46,7 +48,7 @@ gg1 <- gg1 + theme_minimal(9) + theme(
 
 #### PLOT 2: K2P difference density plots
 
-pal <- pal[-2]
+pal <- pal[-6]
 
 k2p_data <- aln_data2 |>
     mutate(k2p_norm = k2p - k2p[method == "coati-tri-mg"], .by = "gene") |>
@@ -54,7 +56,7 @@ k2p_data <- aln_data2 |>
 
[…]

### 2024-01-13 figures/fig-sel-empirical.R

@@ -4,7 +4,7 @@ library(ComplexUpset)
 library(patchwork)
 
 PDF_WIDTH <- 7 # 178mm
-PDF_HEIGHT <- PDF_WIDTH/2
+PDF_HEIGHT <- PDF_WIDTH*0.6
 
 # load data
 aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
@@ -17,18 +17,19 @@ aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 
 ####
 
+methods <- rev(c("coati-tri-mg", "clustalo", "macse", "mafft", "prank", "rev-coati-tri-mg"))
+
 omega_data <- aln_data |>
     mutate(neg_sel = omega < 1) |>
-    pivot_wider(id_cols = "gene", names_from = "method", values_from = "neg_sel") |>
-    rename(coati = "coati-tri-mg")
+    pivot_wider(id_cols = "gene", names_from = "method", values_from = "neg_sel")
 
-gg1 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
+gg1 <- upset(omega_data, methods,
     wrap = TRUE,
-    n_intersections=10, set_sizes = FALSE, sort_sets = FALSE,
+    n_intersections = 10, set_sizes = FALSE, sort_sets = FALSE,
     name = "Negative Selection Grouping",
     base_annotations=list(
         'Intersection size'=intersection_size(
-            mapping=aes(hjust=ifelse(after_stat(y)>3000, 0.5, 0.0)),
+            mapping=aes(hjust=ifelse(after_stat(y) > 3000, 0.5, 0.0)),
             text=list(
                 vjust=0.5,
                 angle=90,
@@ -37,27 +38,27 @@ gg1 <- upset(omega_data, c("clustalo", "macse", "mafft", "prank", "coati"),
         )
     ),
     labeller = ggplot2::as_labeller(c(
-        coati = "COATi",
+        "coati-tri-mg" = "COATi",
         prank = "PRANK",
         mafft = "MAFFT",
         macse = "MACSE",
-        clustalo = "ClustalΩ"
+        clustalo = "ClustalΩ",
+        "rev-coati-tri-mg" = "COATi-rev"
     ))
 )
 
 
 omega_data <- aln_data |>
     mutate(neg_sel = omega > 1) |>
-    pivot_wider(id_cols = "gene", names_from = "method", values_from = "neg_sel") |>
[…]

### 2024-01-13 tables/tab-gaps.R

@@ -27,6 +27,8 @@ tab <- aln_data |> select(gene, method, cigar) |>
     mutate(rle = tbl_cigar(cigar)) |>
     unnest(rle)
 
+#### LENGTHS ###################################################################
+
 k <- tab |> filter(values != "M") |>
     mutate(lengths = if_else(lengths <= 6, lengths, 7 + ((lengths-1) %% 3))) |>
     arrange(lengths) |>
@@ -34,8 +36,33 @@ k <- tab |> filter(values != "M") |>
     complete(method, values, lengths = full_seq(lengths, 1), fill = list(n = 0)) |>
     pivot_wider(names_from = "lengths", values_from = "n")
 
+write_csv(k, here("tables", "gap-lengths.csv"))
+write_csv(k, here("supplementary_materials", "data", "gap-lengths.csv"))
+
+#### PHASES ####################################################################
+
+k <- tab |> group_by(gene, method) |>
+    mutate(mlen = if_else(values == "I", 0L, lengths),
+        pos = cumsum(mlen) - mlen,
+        phase = ((pos-1) %% 3) + 1) |>
+    ungroup() |>
+    count(method, phase) |>
+    drop_na(phase) |>
+    complete(method, phase = full_seq(phase, 1), fill = list(n = 0)) |>
+    pivot_wider(names_from = "phase", values_from = "n")
+
+write_csv(k, here("tables", "gap-phases.csv"))
+write_csv(k, here("supplementary_materials", "data", "gap-phases.csv"))
+
+#### GAPINESS ##################################################################
+
+k <- tab |> group_by(method, gene) |> 
+    summarize(p = sum(lengths[values != "M"])/(sum(lengths) + sum(lengths[values == "M"]))) |>
+    summarize(p = mean(p, na.rm = TRUE))
+
+write_csv(k, here("tables", "gapiness.csv"))
+write_csv(k, here("supplementary_materials", "data", "gapiness.csv"))
+
 # print(xtable::xtable(k), floating=FALSE, latex.environments=NULL, booktabs=TRUE)
 
-# tab |> filter(lengths %% 3 > 0 & method == "clustalo")
 
-# tab |> group_by(method, gene) |> summarize(p = sum(lengths[values != "M"])/(sum(lengths) + sum(lengths[values == "M"]))) |> summarize(mean(p, na.rm = TRUE))

### 2024-01-16 alignpair_letter.tex
Update supplementary materials.

@@
 
-\section*{Materials and Methods}
+\section*{Methods}
 
@@
 
+% 8263 gapless genes.
+% 5864 gapped genes.
+
 In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
@@
     \par
-    \caption{COATi.}
+    \caption{COATi's alignments produce biologically reasonable evolutionary distances.
+(a) The distribution of K2P distances inferred from alignments generated by each method.
+The averages of each distribution are indicated by vertical lines.
+The averages are COATi = 0.0084, ClustalΩ = 0.0178, MACSE = 0.0149, MAFFT = 0.0154, PRANK = 0.0092, and COATi-rev = 0.0084.
+(b) The distribution of the differences between distances inferred by COATi and other methods.
+The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.}
     \label{fig:k2p-empirical}
 \end{figure}
+
+% Wilcoxon signed rank test
+% coati vs rev-coati: p-value: 0.4212 (two.sided)
+% coati vs prank: p-value: < 2.2e-16 (less)
+% coati vs macse: p-value: < 2.2e-16 (less)
+% coati vs mafft: p-value: < 2.2e-16 (less)
+% coati vs clustalo: p-value: < 2.2e-16 (less)
 

### 2024-01-16 supplementary_materials/supplementary_materials.Rmd
Update supplementary materials.

@@
 
+cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
+    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+set.seed(17761)
+
 knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(dev = "cairo_pdf")
+
+# Alignment stats
+ALN_DATA <- read_csv(here("data", "raw_fasta_aligned_stats.csv.gz"))
+
+# Preprocessed cigar strings from alignment stats
+GAP_DATA <- read_csv(here("supplementary_materials", "data", "gaps.csv.gz"))
 ```
@@
 
-# Gap Patterns
+# Empirical Results
 
-## Lengths
+## Gap Patterns
+
+### Lengths
 
@@
 ```{r gap-table-1, results = 'asis', message = FALSE}
-gap_data <- read_csv(here("supplementary_materials/data/gap-lengths.csv"))
+gap_len_data <- GAP_DATA |> filter(op %in% c("I", "D")) |>
+    mutate(length = if_else(length <= 6, length, 7 + ((length-1) %% 3))) |>
+    arrange(length) |>
+    count(method, length, op) |>
+    complete(method, op, length = full_seq(length, 1), fill = list(n = 0)) |>
+    pivot_wider(names_from = "length", values_from = "n")
 
-gap_data <- gap_data |> rename("Gap Type" = "values", "Method" = "method",
+gap_len_data <- gap_len_data |> rename("Gap Type" = "op", "Method" = "method",
   "7+" = "7", "8+" = "8", "9+" = "9")
 
-gap_data1 <- gap_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
+tab <- gap_len_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
     "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
@@
 
-k <- kable(gap_data1, align = "lcrrrrrrrrr", caption="Number of gaps introduced by each method separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
+k <- kable(tab, align = "lcrrrrrrrrr", caption="Number of gaps introduced by each method separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
 
@@
 ```{r gap-table-2, results = 'asis', message = FALSE}
-gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
+tab <- gap_len_data |> filter(Method %in% c("coati-tri-mg",
     "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
@@
 
-k <- kable(gap_data2, align = "lcrrrrrrrrr", caption="Number of gaps introduced by different COATi models separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
+k <- kable(tab, align = "lcrrrrrrrrr", caption="Number of gaps introduced by different COATi models separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
 
@@
 
-## Phases
+### Phases
 
@@
 ```{r gap-table-3, results = 'asis', message = FALSE}
-gap_data <- read_csv(here("supplementary_materials/data/gap-phases.csv"))
+gap_phase_data <- GAP_DATA |> 
+    mutate(mlen = if_else(op == "I", 0L, length)) |>
+    mutate(pos = cumsum(mlen) - mlen, .by=c(gene, method)) |>
+    mutate(phase = ((pos-1) %% 3) + 1) |>
+    filter(op %in% c("I", "D")) |>
+    count(method, phase) |>
+    drop_na(phase) |>
+    complete(method, phase = full_seq(phase, 1), fill = list(n = 0)) |>
+    pivot_wider(names_from = "phase", values_from = "n")
 
-gap_data <- gap_data |> rename("Method" = "method")
+gap_phase_data <- gap_phase_data |> rename("Method" = "method")
 
-gap_data1 <- gap_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
+tab <- gap_phase_data |> filter(Method %in% c("clustalo", "coati-tri-mg",
     "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
@@
 
-k <- kable(gap_data1, align = "lrrr", caption="Number of gaps introduced by each alignment method separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab, align = "lrrr", caption="Number of gaps introduced by each alignment method separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
 
@@
 ```{r gap-table-4, results = 'asis', message = FALSE}
-gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
+tab <- gap_phase_data |> filter(Method %in% c("coati-tri-mg",
     "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
@@
 
-k <- kable(gap_data2, align = "lrrr", caption="Number of gaps introduced by different COATi models separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab, align = "lrrr", caption="Number of gaps introduced by different COATi models separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
 
@@
 
-## Gapiness
+## Homology Patterns
 
-We quantified the gapiness of different aligners and COATi models (Tab.
-\@ref(tab:gap-table-5) and \@ref(tab:gap-table-6)).
-Here we define gapiness as the tendency of an aligner to introduce gaps into an alignment, and we measure it by calculating the percentage of nucleotides aligned with a gap.
-Note that this is different than the percent of columns that contain a gap.
+We quantified the homology patterns of residues for different aligners and COATi models (Tab.
+\@ref(tab:gap-table-5), Tab.
+\@ref(tab:gap-table-6), and Fig.
+\@ref(fig:gap-figure-1)).
+Here we define the match, mismatch, and gap percentages as the percent of nucleotides aligned against a match, mismatch, and gap respectively.
+Note that this is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.
 
 ```{r gap-table-5, results = 'asis', message = FALSE}
-gap_data <- read_csv(here("supplementary_materials/data/gapiness.csv"))
+gap_count_data <- GAP_DATA |>
+    count(gene, method, op, wt = length) |>
+    drop_na(op) |>
+    pivot_wider(id_cols=c(gene, method),
+        values_from=n, names_from=op,
+        values_fill = 0) |>
+    rename(A = `=`)
 
-gap_data <- gap_data |> rename("Method" = "method") |>
-    mutate("Gapiness" = sprintf("%0.2f%%",round(p * 100, 2))) |>
-    select(-p)
+gap_count_data <- gap_count_data |> mutate(
+    anc_len = A + X + D,
+    dec_len = A + X + I,
+    Pa = (A + A)/(anc_len + dec_len),
+    Px = (X + X)/(anc_len + dec_len),
+    Pg = (D + I)/(anc_len + dec_len),
+)
 
-gap_data1 <- gap_data |>
+gap_frac_data <- gap_count_data |> 
+    summarize(Matches = mean(Pa), Mismatches = mean(Px), Gaps = mean(Pg), .by = method) |>
+    mutate(across(Matches:Gaps,  \(x) scales::percent(x, accuracy = 0.01))) |>
+    rename(Method = method)
+
+tab <- gap_frac_data |>
+    filter(Method %in% c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
+    mutate(Method = case_match(Method,
+        "clustalo" ~ "ClustalΩ",
+        "coati-tri-mg" ~ "COATi",
+        "macse" ~ "MACSE",
+        "mafft" ~ "MAFFT",
+        "prank" ~ "PRANK",
+        "rev-coati-tri-mg" ~ "COATi-rev"
+        )) |>
+    arrange(Method)
+
+k <- kable(tab, align = "rccc", caption="Average homology percentages of alignments separated by alignment method.", booktabs = TRUE, linesep = c("\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position")
+```
+
+```{r gap-table-6, results = 'asis', message = FALSE}
+tab <- gap_frac_data  |> filter(Method %in% c("coati-tri-mg",
+    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
+        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
+    arrange(Method) |>
+    rename(Model = Method) |>
+    mutate(Model = fct_recode(Model,
+        "TRI-MG" = "coati-tri-mg",
+        "TRI-ECM" = "coati-tri-ecm",
+        "MAR-MG" = "coati-mar-mg",
+        "MAR-ECM" = "coati-mar-ecm",
+        "DNA" = "coati-dna"
+    ))
+
+k <- kable(tab, align = "rccc", caption="Average homology percentages of alignments separated by COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"))
+
+k |> kable_styling(latex_options = "HOLD_position")
+```
+
+```{r gap-figure-1, fig.cap="Gap fractions of COATi vs other aligners.
+Each panel is a scatter plot where the x coordinate is the gap fraction of a COATi alignment and y coordinate is the gap fraction of the corresponding alignment from another aligner.", message = FALSE, warning = FALSE}
+
+# Pg
+
+tab <- gap_count_data |>
+    filter(method %in% c(c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg"))) |>
+    pivot_wider(id_cols=gene, names_from=method, values_from=Pg) |>
+    rename(coati = `coati-tri-mg`, `coati-rev` = `rev-coati-tri-mg`) |>
+    relocate(gene, coati)
+
+tab <- tab |> pivot_longer(cols = clustalo:`coati-rev`,
+    names_to="method", values_to="other") |>
+    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
+        "prank", "coati-rev")) |>
+    mutate(method = fct_recode(method,
+        "vs ClustalΩ" = "clustalo",
+        "vs MACSE" = "macse",
+        "vs MAFFT" = "mafft",
+        "vs PRANK" = "prank",
+        "vs COATI-rev" = "coati-rev"
+    ))
+
+gg <- ggplot(tab, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
+gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
+gg <- gg + xlab("COATi") + ylab("Other")
+gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
+gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1), limits = c(0, 1.2))
+gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1))
+
+gg <- gg + theme_grey(9) + theme(
+    panel.grid.minor.x = element_blank(),
+    panel.grid.minor.y = element_blank()
+)
+
+gg
+
+```
+
+\newpage
+
+## Evolutionary Distances
+
+We quantified the evolutionary distances inferred by different aligners and COATi models (Fig. \@ref(fig:k2p-figure-1), Tab. \@ref(tab:k2p-table-1), and Tab. \@ref(tab:k2p-table-2)).
+
+```{r k2p-figure-1, fig.cap="COATi produced shorter evolutionary distances than other aligners.
+Each panel is a scatter plot where the x coordinate is the K2P distance from a COATi alignment and y coordinate is the K2P distance from the corresponding alignment from another aligner.", warning = FALSE}
+k2p_data <- ALN_DATA |>
+    filter(method %in% c(c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg"))) |>
+    pivot_wider(id_cols=gene, names_from=method, values_from=k2p) |>
+    rename(coati = `coati-tri-mg`, `coati-rev` = `rev-coati-tri-mg`) |>
+    relocate(gene, coati)
+
+k2p_data1 <- k2p_data |> pivot_longer(cols = clustalo:`coati-rev`,
+    names_to="method", values_to="other") |>
+    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
+        "prank", "coati-rev")) |>
+    mutate(method = fct_recode(method,
+        "vs ClustalΩ" = "clustalo",
+        "vs MACSE" = "macse",
+        "vs MAFFT" = "mafft",
+        "vs PRANK" = "prank",
+        "vs COATI-rev" = "coati-rev"
+    ))
+
+gg <- ggplot(k2p_data1, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
+gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
+gg <- gg + xlab("COATi") + ylab("Other")
+gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
+gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1))
+gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1))
+
+gg <- gg + theme_grey(9) + theme(
+    panel.grid.minor.x = element_blank(),
+    panel.grid.minor.y = element_blank()
+)
+
+gg
+```
+
+```{r k2p-table-1, results = 'asis', message = FALSE}
+tab <- ALN_DATA |> summarize(m = mean(k2p, na.rm = TRUE), .by = "method") |>
+    rename(Method = method)
+
+tab_data1 <- tab |>
     filter(Method %in% c("clustalo", "coati-tri-mg",
@@
     arrange(Method) |>
-    pivot_wider(names_from = "Method", values_from = "Gapiness")
+    pivot_wider(names_from = "Method", values_from = "m")
 
-k <- kable(gap_data1, align = "ccccc", caption="Average gapiness of alignments separated by alignment method.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab_data1, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each method.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))
 
 k |> kable_styling(latex_options = "HOLD_position")
+
 ```
 
-```{r gap-table-6, results = 'asis', message = FALSE}
-gap_data2 <- gap_data |> filter(Method %in% c("coati-tri-mg",
+```{r k2p-table-2, results = 'asis', message = FALSE}
+tab_data2 <- tab |> filter(Method %in% c("coati-tri-mg",
     "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
@@
     ))  |>
-    pivot_wider(names_from = "Model", values_from = "Gapiness")
+    pivot_wider(names_from = "Model", values_from = "m")
 
-k <- kable(gap_data2, align = "ccccc", caption="Average gapiness of alignments separated by COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab_data2, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))
 

### 2024-01-16 supplementary_materials/supplementary_materials.tex
Update supplementary materials.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{TeX Gyre Pagella}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\let\oldthefigure\thefigure
\renewcommand{\thefigure}{S\oldthefigure}
\let\oldthetable\thetable
\renewcommand{\thetable}{S\oldthetable}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={COATi: statistical pairwise alignment of protein coding sequences},
  pdfauthor={Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{COATi: statistical pairwise alignment of protein coding sequences}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Supplementary Materials}
\author{Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright}
\date{}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{aligner-commands}{%
\section{Aligner Commands}\label{aligner-commands}}

We evaluated five different aligners. Below are the commands that we used to run them. We have abbreviated the commands for clarity, stripping out unimportant arguments. Complete workflows can be found in our coati-testing repository on Github.

\begin{itemize}
\tightlist
\item
  COATi: \texttt{coati\ alignpair\ -m\ tri-mg\ ...}; We used COATi two different ways. The primary way used human as the reference sequence. The secondary way, COATi-rev, used gorilla as the reference, via a wrapper script that reversed the order of sequences before alignment and reversed the order back afterwards.
\item
  ClustalΩ v1.2.4: \texttt{clustalo\ -\/-seqtype=Protein\ ...}; We wrapped ClustalΩ with a script that translates DNA sequences into amino acid sequences (including stops) before alignment. Any codons that were partial or containing ambiguous characters were translated as ``X''. The script then aligned these translated sequences with ClustalΩ, and then created a DNA alignment that was consistent with the amino-acid alignment.
\item
  MACSE v2.06: \texttt{java\ -jar\ macse.jar\ -prog\ alignSequences\ -seq\ human.fasta\ -seq\_lr\ gorilla.fasta\ ...}; We wrapped MACSE with a script that created two temporary fasta files, one containing the human sequence and another containing the gorilla sequence. The human sequence was specified as the reliable sequence and the gorilla sequence was specified as the less-reliable sequence. MACSE uses ``!'' to mark gaps that result from frameshifts, and the wrapper script replaced these with ``-''. Additionally, MACSE sometimes produced columns that only contained gaps, and these columns were removed.
\item
  PRANK v.150803: \texttt{prank\ -codon\ ...}
\item
  MAFFT v7.520: \texttt{mafft\ -\/-preservecase\ -\/-globalpair\ -\/-maxiterate\ 1000\ ...}
\end{itemize}

COATi can use different alignment models for pairwise alignment. Below are the commands that we used to run different models.

\begin{itemize}
\tightlist
\item
  tri-mg: \texttt{coati\ alignpair\ -m\ tri-mg\ ...}
\item
  tri-ecm: \texttt{coati\ alignpair\ -m\ tri-ecm\ ...}
\item
  mar-mg: \texttt{coati\ alignpair\ -m\ mar-mg\ ...}
\item
  mar-ecm: \texttt{coati\ alignpair\ -m\ mar-ecm\ ...}
\end{itemize}

\hypertarget{fst-alignment-example}{%
\section{FST Alignment Example}\label{fst-alignment-example}}

When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST library to generate best alignments by composing the input and output sequences with the COATi FST model. While the COATi FST model is too large to display, we can show the result of a composition.

Fig. \ref{fig:aln-example-a} shows a graph depicting the FST that results from composing the COATi FST with the input sequence ``CTC'' and the output sequence ``CTG''. Every path through this FST represents one possible way to align ``CTC'' and ``CTG'', and the sum of all weights along a path is the total weight of the respective alignment. Here, the weights of each arc are in negative-log space. Note that this graph has been optimized, and weight has been pushed towards the initial state. The weight of any specific arc may not be directly mapable to a weight described in the model.

Fig. \ref{fig:aln-example-b} is the best alignment between ``CTC'' and ``CTG'', as determined by the shortest path algorithm. Figs. \ref{fig:aln-example-a} and \ref{fig:aln-example-b} were produced by the OpenFST library. A bold circle represents a starting node, and a double circle represents a termination node.

\begin{figure}

{\centering \includegraphics{figures/aln-example-graph}

}

\caption{The FST of all possible alignments between `CTC' and `CTG'.}\label{fig:aln-example-a}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{figures/aln-example-path}

}

\caption{The best alignment of `CTC' and `CTG'.}\label{fig:aln-example-b}
\end{figure}

\newpage

\hypertarget{empirical-results}{%
\section{Empirical Results}\label{empirical-results}}

\hypertarget{gap-patterns}{%
\subsection{Gap Patterns}\label{gap-patterns}}

\hypertarget{lengths}{%
\subsubsection{Lengths}\label{lengths}}

We quantified the lengths of gaps produced by each method across the empirical dataset of human-gorilla protein-coding-sequence pairs (Tab. \ref{tab:gap-table-1}). Here the gap type is either ``D'' for gaps introduced into the gorilla sequence or ``I'' for gaps introduced in the human sequence. ``COATi'' refers to the full COATi FST model with a MG substitution model (i.e.~tri-mg). ``COATi-rev'' refers to the same model, but using gorilla as the reference and human as the non-reference sequence. Gap lengths of 1--6 nucleotides were binned into their respective columns. Gap lengths longer than 6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.

We also quantified the lengths of gaps produced by different COATi models (Tab. \ref{tab:gap-table-2}).

Note that ClustalΩ gaps with lengths that are not a multiple of 3 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.

\begin{table}[H]

\caption{\label{tab:gap-table-1}Number of gaps introduced by each method separated by length and type.}
\centering
\begin{tabular}[t]{lcrrrrrrrrr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{9}{c}{Gap Lengths} \\
\cmidrule(l{3pt}r{3pt}){3-11}
Method & Gap Type & 1 & 2 & 3 & 4 & 5 & 6 & 7+ & 8+ & 9+\\
\midrule
COATi & D & 106 & 99 & 651 & 94 & 94 & 226 & 1461 & 1386 & 3240\\
COATi & I & 87 & 73 & 525 & 80 & 97 & 239 & 1399 & 1315 & 3105\\
\addlinespace
COATi-rev & D & 105 & 99 & 634 & 89 & 96 & 217 & 1425 & 1348 & 3169\\
COATi-rev & I & 82 & 79 & 513 & 80 & 97 & 231 & 1360 & 1288 & 3031\\
\addlinespace
ClustalΩ & D & 1 & 0 & 887 & 1 & 1 & 400 & 3 & 7 & 3910\\
ClustalΩ & I & 0 & 0 & 857 & 0 & 1 & 424 & 3 & 2 & 3747\\
\addlinespace
MACSE & D & 396 & 306 & 1237 & 5 & 3 & 657 & 71 & 54 & 4784\\
MACSE & I & 60 & 27 & 1133 & 5 & 9 & 666 & 78 & 97 & 4381\\
\addlinespace
MAFFT & D & 204 & 156 & 714 & 147 & 108 & 276 & 544 & 509 & 2680\\
MAFFT & I & 187 & 154 & 708 & 112 & 92 & 295 & 454 & 424 & 2642\\
\addlinespace
PRANK & D & 0 & 0 & 552 & 0 & 0 & 167 & 0 & 0 & 4812\\
PRANK & I & 0 & 0 & 467 & 0 & 0 & 178 & 0 & 0 & 4686\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:gap-table-2}Number of gaps introduced by different COATi models separated by length and type.}
\centering
\begin{tabular}[t]{lcrrrrrrrrr}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{9}{c}{Gap Lengths} \\
\cmidrule(l{3pt}r{3pt}){3-11}
Model & Gap Type & 1 & 2 & 3 & 4 & 5 & 6 & 7+ & 8+ & 9+\\
\midrule
TRI-MG & D & 106 & 99 & 651 & 94 & 94 & 226 & 1461 & 1386 & 3240\\
TRI-MG & I & 87 & 73 & 525 & 80 & 97 & 239 & 1399 & 1315 & 3105\\
\addlinespace
TRI-ECM & D & 93 & 96 & 665 & 125 & 122 & 257 & 1438 & 1441 & 3203\\
TRI-ECM & I & 86 & 103 & 540 & 105 & 122 & 254 & 1358 & 1370 & 3099\\
\addlinespace
MAR-MG & D & 102 & 106 & 646 & 95 & 89 & 224 & 1465 & 1399 & 3226\\
MAR-MG & I & 84 & 83 & 526 & 80 & 101 & 238 & 1406 & 1313 & 3087\\
\addlinespace
MAR-ECM & D & 110 & 110 & 653 & 107 & 89 & 224 & 1424 & 1389 & 3250\\
MAR-ECM & I & 86 & 89 & 523 & 77 & 99 & 249 & 1408 & 1310 & 3128\\
\addlinespace
DNA & D & 101 & 103 & 646 & 95 & 91 & 224 & 1446 & 1409 & 3206\\
DNA & I & 79 & 79 & 520 & 78 & 99 & 239 & 1369 & 1316 & 3107\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{phases}{%
\subsubsection{Phases}\label{phases}}

We quantified the phases of gaps produced by different aligners and COATi models (Tab. \ref{tab:gap-table-3} and \ref{tab:gap-table-4}). Phase 1 gaps begin after the 1st position in a codon in the reference sequence, phase 2 gaps begin after the 2nd position in a codon, and phase 3 begin after the 3rd position in a codon (i.e.~between codons). Phase 3 gaps are also known as phase 0 gaps.

Note that ClustalΩ gaps with phases of 1 and 2 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.

\begin{table}[H]

\caption{\label{tab:gap-table-3}Number of gaps introduced by each alignment method separated by phase.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Gap Phases} \\
\cmidrule(l{3pt}r{3pt}){2-4}
Method & 1 & 2 & 3\\
\midrule
COATi & 4493 & 3962 & 5822\\
\addlinespace
COATi-rev & 3992 & 3775 & 6176\\
\addlinespace
ClustalΩ & 8 & 6 & 10230\\
\addlinespace
MACSE & 455 & 497 & 13017\\
\addlinespace
MAFFT & 2317 & 2631 & 5458\\
\addlinespace
PRANK & 0 & 0 & 10862\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:gap-table-4}Number of gaps introduced by different COATi models separated by phase.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Gap Phases} \\
\cmidrule(l{3pt}r{3pt}){2-4}
Model & 1 & 2 & 3\\
\midrule
TRI-MG & 4493 & 3962 & 5822\\
\addlinespace
TRI-ECM & 4160 & 4678 & 5639\\
\addlinespace
MAR-MG & 4130 & 3961 & 6179\\
\addlinespace
MAR-ECM & 4071 & 3863 & 6391\\
\addlinespace
DNA & 4228 & 3974 & 6005\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{homology-patterns}{%
\subsection{Homology Patterns}\label{homology-patterns}}

We quantified the homology patterns of residues for different aligners and COATi models (Tab. \ref{tab:gap-table-5}, Tab. \ref{tab:gap-table-6}, and Fig. \ref{fig:gap-figure-1}). Here we define the match, mismatch, and gap percentages as the percent of nucleotides aligned against a match, mismatch, and gap respectively. Note that this is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.

\begin{table}[H]

\caption{\label{tab:gap-table-5}Average homology percentages of alignments separated by alignment method.}
\centering
\begin{tabular}[t]{rccc}
\toprule
Method & Matches & Mismatches & Gaps\\
\midrule
COATi & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
COATi-rev & 95.95\% & 0.79\% & 3.25\%\\
\addlinespace
ClustalΩ & 95.93\% & 1.54\% & 2.52\%\\
\addlinespace
MACSE & 96.13\% & 1.33\% & 2.54\%\\
\addlinespace
MAFFT & 96.12\% & 1.36\% & 2.52\%\\
\addlinespace
PRANK & 95.64\% & 0.84\% & 3.52\%\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:gap-table-6}Average homology percentages of alignments separated by COATi model.}
\centering
\begin{tabular}[t]{rccc}
\toprule
Model & Matches & Mismatches & Gaps\\
\midrule
TRI-MG & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
TRI-ECM & 95.90\% & 0.79\% & 3.30\%\\
\addlinespace
MAR-MG & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
MAR-ECM & 95.94\% & 0.80\% & 3.26\%\\
\addlinespace
DNA & 95.93\% & 0.79\% & 3.27\%\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/gap-figure-1-1.pdf}
\caption{\label{fig:gap-figure-1}Gap fractions of COATi vs other aligners. Each panel is a scatter plot where the x coordinate is the gap fraction of a COATi alignment and y coordinate is the gap fraction of the corresponding alignment from another aligner.}
\end{figure}

\newpage

\hypertarget{evolutionary-distances}{%
\subsection{Evolutionary Distances}\label{evolutionary-distances}}

We quantified the evolutionary distances inferred by different aligners and COATi models (Fig. \ref{fig:k2p-figure-1}, Tab. \ref{tab:k2p-table-1}, and Tab. \ref{tab:k2p-table-2}).

\begin{figure}
\centering
\includegraphics{supplementary_materials_files/figure-latex/k2p-figure-1-1.pdf}
\caption{\label{fig:k2p-figure-1}COATi produced shorter evolutionary distances than other aligners. Each panel is a scatter plot where the x coordinate is the K2P distance from a COATi alignment and y coordinate is the K2P distance from the corresponding alignment from another aligner.}
\end{figure}

\begin{table}[H]

\caption{\label{tab:k2p-table-1}Average K2P distances of alignments for each method.}
\centering
\begin{tabular}[t]{cccccc}
\toprule
COATi & COATi-rev & ClustalΩ & MACSE & MAFFT & PRANK\\
\midrule
0.0084 & 0.0084 & 0.0178 & 0.0149 & 0.0154 & 0.0092\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:k2p-table-2}Average K2P distances of alignments for each COATi model.}
\centering
\begin{tabular}[t]{ccccc}
\toprule
TRI-MG & TRI-ECM & MAR-MG & MAR-ECM & DNA\\
\midrule
0.0084 & 0.0084 & 0.0084 & 0.0084 & 0.0084\\
\bottomrule
\end{tabular}
\end{table}

\end{document}

### 2024-01-16 figures/fig-k2p-empirical.R

@@ -22,6 +22,10 @@ methods <- c("clustalo", "macse", "mafft", "prank", "coati-tri-mg", "rev-coati-t
 aln_data2 <- aln_data |> filter(method %in% methods)
 aln_data2 <- aln_data2 |> mutate(method = fct_relevel(method, "coati-tri-mg"))
 
+mean_k2p <- aln_data2 |>
+    summarize(avg = mean(k2p, na.rm = TRUE), .by = "method") |>
+    arrange(desc(avg))
+
 gg1 <- ggplot(aln_data2, aes(x = k2p, color =  fct_rev(method))) + geom_density(linewidth = 1)
 gg1 <- gg1 + scale_x_continuous(trans = transform_pseudo_log(0.0005, base=10),
     breaks=c(0, 0.001, 0.01, 0.1, 1))
@@ -37,6 +41,10 @@ gg1 <- gg1 + scale_color_manual(values = pal,
      "rev-coati-tri-mg" = "COATi-rev"
      )
     )
+
+gg1 <- gg1 + geom_vline(aes(xintercept = avg, color = method), data = mean_k2p,
+    linewidth = 0.5, linetype = 2, show.legend = FALSE)
+
 gg1 <- gg1 + guides(color = guide_legend(override.aes=list(fill=rev(pal)), reverse = TRUE))
 gg1 <- gg1 + theme_minimal(9) + theme(
     legend.position = c(1,1),

### 2024-01-16 tables/tab-gaps.R

@@ -9,7 +9,7 @@ aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 
 cigar_to_rle <- function(x) {
     len <- str_extract_all(x, "\\d+")
-    val <- str_extract_all(x, "[A-Za-z]")
+    val <- str_extract_all(x, "\\D+")
 
     result <- map2(len, val, function(n, v) {
         structure(list(lengths = as.integer(n), values = v), class = "rle")
@@ -25,43 +25,11 @@ tbl_cigar <- function(x) {
 
 tab <- aln_data |> select(gene, method, cigar) |> 
     mutate(rle = tbl_cigar(cigar)) |>
-    unnest(rle)
+    unnest(rle) |> select(-cigar) |>
+    rename(length = lengths, op = values)
 
-#### LENGTHS ###################################################################
-
-k <- tab |> filter(values != "M") |>
-    mutate(lengths = if_else(lengths <= 6, lengths, 7 + ((lengths-1) %% 3))) |>
-    arrange(lengths) |>
-    count(method, lengths, values) |>
-    complete(method, values, lengths = full_seq(lengths, 1), fill = list(n = 0)) |>
-    pivot_wider(names_from = "lengths", values_from = "n")
-
-write_csv(k, here("tables", "gap-lengths.csv"))
-write_csv(k, here("supplementary_materials", "data", "gap-lengths.csv"))
-
-#### PHASES ####################################################################
-
-k <- tab |> group_by(gene, method) |>
-    mutate(mlen = if_else(values == "I", 0L, lengths),
-        pos = cumsum(mlen) - mlen,
-        phase = ((pos-1) %% 3) + 1) |>
-    ungroup() |>
-    count(method, phase) |>
-    drop_na(phase) |>
-    complete(method, phase = full_seq(phase, 1), fill = list(n = 0)) |>
-    pivot_wider(names_from = "phase", values_from = "n")
-
-write_csv(k, here("tables", "gap-phases.csv"))
-write_csv(k, here("supplementary_materials", "data", "gap-phases.csv"))
-
-#### GAPINESS ##################################################################
-
-k <- tab |> group_by(method, gene) |> 
-    summarize(p = sum(lengths[values != "M"])/(sum(lengths) + sum(lengths[values == "M"]))) |>
-    summarize(p = mean(p, na.rm = TRUE))
-
[…]

### 2024-01-28 alignpair_letter.tex
Add more results and figures

@@
 \usepackage{multicol}
+\usepackage{booktabs}
 
@@
 Additionally significant levels of investment have been made to correctly identify and annotate human genes, while gorilla annotations have received limited support in comparison.
-Together, these reference genomes provide a good opportunity to compare COATi against other aligner as they offer one genome that is high-quality (human) and sister genome that is lower-quality (gorilla).
+Together, these reference genomes provide a good opportunity to compare COATi against other aligners as they offer one genome that is high-quality (human) and sister genome that is lower-quality (gorilla).
 
@@
 Additionally, we checked alignments to ensure that our pipeline did not introduce any artifacts into the results, including unexpected characters, empty columns, and aligned sequences of different lengths.
-We also generated checksums of our sequences to ensure that they were not modified during alignment.
+We also generated checksums of our sequences with gaps removed to ensure that they were not modified during alignment.
 
@@
 K2P corrects for multiple mutations at a site and takes into account differences in transition and transversion rates.
-It also assumes equal nucleotide frequencies and no variations in the rate of substitution across sites.
+It also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites.
 While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
@@
 %
-Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the lengths and phases of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a gap.
+Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the phases and lengths of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a match, mismatch, or gap.
 
@@
 
-\subsection*{Empirical Simulation Algorithm}
-
-% 8263 gapless genes.
-% 5864 gapped genes.
+\subsection*{Simulated Benchmark Alignments}
 
 In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
-Briefly, we downloaded 16000 human genes and their gorilla orthologs from the ENSEMBL database \citep{ensembl_hubbard_2002}.
-After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and aligned the remaining pairs with the five alignment methods.
-Next we separated sequence pairs into two sets: (1) 6048 sequence pairs for which at least one aligner added gaps and (2) 7761 sequence pairs for which all five aligners did not introduce any gaps.
-We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
+We separated our 14,127 empirical sequence pairs into two sets: %
+(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
+(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
 We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments.
 The phases of gaps were preserved as were the spacings of any clusters of gaps.
-Segments of matches that were 99 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
+Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
 (This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.)
@@
 applied to an ungapped alignment of human and gorilla sequences that are 300 nucleotides long.
-First, we modify the CIGAR string to insert flexible lengths for any match segment that is 99 or more nucleotides long while preserving phase.
+First, we modified the CIGAR string to insert flexible lengths for any match segment that is 96 or more nucleotides long while preserving phase.
 This results in a new CIGAR string of
 %
-``101M *M 3D 10M 6I 99M *M'',
+``98M *M 3D 10M 6I 96M *M'',
 %
 where * represents the locations that have flexible length.
-Considering only matches and deletions, this CIGAR string has 213 fixed nucleotides, leaving 87 nucleotides in the human sequence to be allocated to the flexible locations.
-Since there are two locations of flexible length, we draw one random break point uniformly while maintaining phase, producing the final CIGAR string of
+Considering only matches and deletions, this CIGAR string has 207 fixed nucleotides, leaving 93 nucleotides in the human sequence to be allocated to the flexible locations.
+Since there are two locations of flexible length, we drew one random break point uniformly while maintaining phase, producing the final CIGAR string of
 %
-``101M 9M 3D 10M 6I 99M 78M'',
+``98M 12M 3D 10M 6I 96M 81M'',
 %
-which is used to add gaps into the target, ungapped sequence pair.
-To apply deletions, we remove the corresponding nucleotides from the gorilla sequence, and to apply insertions, we add random nucleotides according to their stationary frequency to the gorilla sequence at the corresponding locations.
-The human sequence is left unchanged.
+which was used to add gaps into the target, ungapped sequence pair.
+To apply deletions, we removed the corresponding nucleotides from the gorilla sequence, and to apply insertions, we added nucleotides to the gorilla sequence by sampling codons from their stationary frequency while respecting gap phases.
+The human sequence was left unchanged.
+%
+We generated a benchmark of 8261 sequence pairs with known alignments using this strategy.
 
@@
 
-\subsection*{Alignment Accuracy Measures}
+\subsection*{Measuring Aligner Accuracy}
 
-We used multiple statistics to to quantify the similarity between each alignment in the benchmark and the corresponding output obtained by the different tools.
+We aligned our benchmark alignments using our suite of aligners and quantified the similarity of each alignment that was produced to its respective benchmark.
+As above, we measured Kimura's 2-parameter distance \citep{kimura1980simple} and $K_A/K_S$ \citep{ka_ks_li_1993} for the benchmark and estimated alignments.
+Additionally, we quantified the error of estimated alignments using the $d_{seq}$ metric \citep{metrics_blackburne_whelan_2011}.
 
-Alignment error also negatively impacts the estimation of evolutionary distances and phylogenetic trees.
-To quantify the impact of aligners on the estimation of evolutionary distances, we compared Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} calculated for the estimated and benchmark alignments.
-
-Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
-
-Alignment error negatively impacts the identification of genes that have experienced positive and negative selection.
-To evaluate the impact of aligners on selection identification, we estimated $K_S$ and $K_A$ statistics (also known as $d_s$ and $d_n$) for each estimated alignment and compared them to the benchmark alignment.
-
-\paragraph{Alignment Error.}
-To quantify the error between estimated alignments and the benchmark alignments, we used the alignment error metric $d_{seq}$ \citep{metrics_blackburne_whelan_2011}.
 Intuitively, $d_{seq}$ ranges between zero and one and can be interpreted as the fraction of nucleotides in the sequence pair that are aligned differently between estimated and benchmark alignments.
@@
 
-\paragraph{Perfect, Imperfect, and Best Alignments.}
-We quantified the number of perfect, imperfect and best alignments each aligner produced.
+Using $d_{seq}$ we quantified the number of perfect, imperfect and best alignments each aligner produced.
 We define perfect alignments as alignments with a distance of zero to their benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero.
-Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from impacting our results.
+Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from influencing our results.
 Evolutionary equivalence was determined by scoring alignments using COATi's marginal model.
@@
 
-COATi was not able to align 23 sequence pairs, PRANK was not able to align 24 sequence pairs, and COATi-rev was not able to align 193 sequence pairs.
-MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from our wrapper script.
+As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs.
+PRANK was not able to align 24 sequence pairs.
+MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from its wrapper script.
+Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
+A total of 10296 sequence pairs were aligned equivalently by all five methods, including 8261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1371 that had identical scores.
+
+Compared to other aligners, COATi inferred more indels and aligned more nucleotides against a gap while producing fewer mismatches (Tab.~\ref{tab:gaps-empirical}).
+Additionally, COATi inferred that 60\% indels occurred in phase 1 or phase 2 in our coding sequences.
+MACSE's amino-acid aware nucleotide model inferred 7\%, while MAFFT's non-coding DNA model inferred 48\%.
+PRANK's codon and ClustalΩ's amino-acid models were constrained by their assumptions and inferred no phase 1 or phase 2 gaps.
+Note that ClustalΩ's gaps with phases of 1 and 2 were created by our wrapper script handling DNA sequences with lengths that are not multiples of 3.
+
+\begin{table}[h]
+\caption{\label{tab:gaps-empirical}COATi produced more indels and fewer mismatches compared to other aligners.}
+\centering
+\begin{tabular}[t]{lrrrccc}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Indel Phases\textsuperscript{1}} & \multicolumn{3}{c}{Homology Patterns\textsuperscript{2}}\\
+\cmidrule(l{3pt}r{3pt}){2-4}
+\cmidrule(l{3pt}r{3pt}){5-7}
+Method & \multicolumn{1}{c}{1} & \multicolumn{1}{c}{2} & \multicolumn{1}{c}{3} & Matches & Mismatches & Gaps\\
+\midrule
+COATi & 4493 & 3962 & 5822 & 95.93\% & 0.79\% & 3.28\%\\
+\addlinespace
+ClustalΩ & 8 & 6 & 10230 & 95.93\% & 1.54\% & 2.52\%\\
+\addlinespace
+MACSE & 455 & 497 & 13017 & 96.13\% & 1.33\% & 2.54\%\\
+\addlinespace
+MAFFT & 2317 & 2631 & 5458 & 96.12\% & 1.36\% & 2.52\%\\
+\addlinespace
+PRANK & 0 & 0 & 10862 & 95.64\% & 0.84\% & 3.52\%\\
+\addlinespace
+COATi-rev & 3992 & 3775 & 6176 & 95.95\% & 0.79\% & 3.25\%\\
+\bottomrule
+\addlinespace
+\multicolumn{7}{p{30em}}{\footnotesize\textsuperscript{1}Total number of indels inferred by each aligner separated by phase.}\\
+\multicolumn{7}{p{30em}}{\footnotesize\textsuperscript{2}The percent of nucleotides that were aligned against a match, mismatch, and gap respectively.
+This is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.}
+
+\end{tabular}
+
+% The homology pattern columns contain the percent of nucleotides that were aligned
+% against a match, mismatch, and gap respectively. Note that this is different than the percent of columns that
+% contain a match, mismatch, or gap because match and mismatch columns are counted twice.
+\end{table}
+
+\paragraph{Evolutionary Distances.}
+
+The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable  for human-gorilla gene pairs, with means 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a).
+Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%.
+These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods.
+The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
+%
+COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK.
+Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$.
+Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test).
+Additional analysis of the empirical alignments can be found in Supplementary Methods.
+
+% Wilcoxon signed rank test
+% coati vs rev-coati: p-value: 0.4212 (two.sided)
+% coati vs prank: p-value: < 2.2e-16 (less)
+% coati vs macse: p-value: < 2.2e-16 (less)
+% coati vs mafft: p-value: < 2.2e-16 (less)
+% coati vs clustalo: p-value: < 2.2e-16 (less)
 
@@
 (b) The distribution of the differences between distances inferred by COATi and other methods.
-The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.}
-    \label{fig:k2p-empirical}
+The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.\label{fig:k2p-empirical}}
 \end{figure}
 
-% Wilcoxon signed rank test
-% coati vs rev-coati: p-value: 0.4212 (two.sided)
-% coati vs prank: p-value: < 2.2e-16 (less)
-% coati vs macse: p-value: < 2.2e-16 (less)
-% coati vs mafft: p-value: < 2.2e-16 (less)
-% coati vs clustalo: p-value: < 2.2e-16 (less)
+\paragraph{Selection.}
+
+The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}).
+Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies.
+The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev.
+The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT.
+Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
+%
+In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection.
+ClustalΩ identified 1,228.
+MACSE identified 1,183.
+MAFFT identified 1,340.
+And PRANK identified 1,349.
+
+\begin{figure}[h!]
+    \centering%
+    \includegraphics{figures/fig-sel-empirical.pdf}
+    \par
+    \caption{Aligners varied in which sequence pairs they identified as undergoing positive selection.
+    In this UpSet plot, the bottom panel displays the 16 most frequent intersection patterns among aligners.
+A black circle represents positive selection.
+The most frequent pattern was that no aligner found positive selection while the second most frequent pattern was that all aligners found positive selection.
+Other patterns involved a disagreement between aligners about whether a sequence pair showed evidence of positive or negative selection.
+The top panel displays the number of sequence pairs in each grouping.
+%
+    \label{fig:sel-empirical}}
+\end{figure}
+
+\subsection*{Benchmark Results}
+
+% Software comparison table
+\begin{table}[!ht]
+\definecolor{bestcolor}{RGB}{230,230,230}%
+\centering
+\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\label{table:comp}}
+%
+% Copied from tables/benchmarks.tex
+% footnotes and \addlinespace are manually added
+\begin{tabular}{rccccc}
+  \toprule
+ & COATi & ClustalΩ & MACSE & MAFFT & PRANK \\ 
+  \midrule
+Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
+\addlinespace 
+  Number of Best\textsuperscript{1} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
+  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
+  Number of Imperfect\textsuperscript{3} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
+  \addlinespace 
+  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
+  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
+  \addlinespace
+  F\textsubscript{1} for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
+  F\textsubscript{1} for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
+   \bottomrule
+   \addlinespace
+\multicolumn{6}{p{40em}}{\textsuperscript{1}Best alignments have the lowest $d_{seq}$ (including equivalent alignments)}\\
+\multicolumn{6}{p{40em}}{\textsuperscript{2}Perfect alignments have $d_{seq} = 0$ (including equivalent alignments)}\\
+\multicolumn{6}{p{40em}}{\textsuperscript{3}Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment.}\\
+\end{tabular}
+\end{table}
+
+\clearpage
 
@@
 
-% Software comparison table
-\begin{table}[!ht]
-\centering
-\input{figures/table-comp}
- \vspace{1mm}
- \footnotesize{\textsuperscript{*}PRANK produced 42 empty alignments, calculations are based on 7719 alignments.}
- \caption{COATi generates better alignments than other alignment algorithms.
-Results of COATi, PRANK, MAFFT, Clustal$\Omega$, and MACSE aligning 7761 empirically simulated sequence pairs.
-Best alignments have the lowest $d_{seq}$ (including ties), perfect alignments have the same score as the true alignment, and imperfect alignments have a different score than the true alignment when at least one method found a perfect alignment.}
- \label{table:comp}
-\end{table}
-
 COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
@@
 The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++.
-Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing}.
+COATi is released under an MIT license.
+Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.
 

### 2024-01-28 figures/fig-sel-empirical.R

@@ -3,8 +3,8 @@ library(tidyverse)
 library(ComplexUpset)
 library(patchwork)
 
-PDF_WIDTH <- 7 # 178mm
-PDF_HEIGHT <- PDF_WIDTH*0.6
+PDF_WIDTH <- 3.38 # 7=178mm 3.38=86mm
+PDF_HEIGHT <- PDF_WIDTH
 
 # load data
 aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
@@ -20,20 +20,23 @@ aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 methods <- rev(c("coati-tri-mg", "clustalo", "macse", "mafft", "prank", "rev-coati-tri-mg"))
 
 omega_data <- aln_data |>
-    mutate(neg_sel = omega < 1) |>
-    pivot_wider(id_cols = "gene", names_from = "method", values_from = "neg_sel")
+    filter(!any(is.na(omega)), .by=gene) |>
+    mutate(sel = omega > 1) |>
+    pivot_wider(id_cols = "gene", names_from = "method", values_from = "sel")
 
 gg1 <- upset(omega_data, methods,
-    wrap = TRUE,
-    n_intersections = 10, set_sizes = FALSE, sort_sets = FALSE,
-    name = "Negative Selection Grouping",
+    wrap = FALSE,
+    themes=upset_default_themes(text=element_text(size=9)),
+    n_intersections = 16, set_sizes = FALSE, sort_sets = FALSE,
+    name = "Selection Grouping",
     base_annotations=list(
-        'Intersection size'=intersection_size(
-            mapping=aes(hjust=ifelse(after_stat(y) > 3000, 0.5, 0.0)),
-            text=list(
-                vjust=0.5,
-                angle=90,
-                check_overlap=TRUE
+        'Intersection size'= (
+            intersection_size(
+                mapping=aes(hjust=ifelse(after_stat(y) > 3000, 0.5, 0.0)),
+                text=list(
+                    vjust=0.5,
+                    angle=90,
+                    check_overlap=TRUE )
             )
         )
     ),
@@ -47,13 +50,25 @@ gg1 <- upset(omega_data, methods,
     ))
 )
 
+# base_annotations=list(
+#         'Intersection size (>1500)'=(
+#             intersection_size()
+#             + coord_cartesian(ylim=c(1500, NA))
[…]

### 2024-01-28 scripts/aln_data.R

@@ -1,117 +1,0 @@
-library(ape)
-library(stringr)
-library(seqinr)
-
-# sequence,method,CIGAR,K2P,omega
-
-filtered = "tri-mg/data/filtered.csv"
-methods = c("clustalo", "macse", "mafft", "prank", "tri-mg")
-
-files = read.csv(file = filtered, header = FALSE)[,1]
-N = length(files) * length(methods)
-
-results = data.frame(sequence = character(N),
-                     method = character(N),
-                     cigar = character(N),
-                     k2p = numeric(N),
-                     omega = numeric(N))
-
-index = 1
-for(i in seq_along(files)) {
-    gene = files[i]
-    for(j in seq_along(methods)) {
-        method = methods[j]
-        path = paste0("tri-mg/aln/", method, "/", gene)
-        # if the file doesn't exist or the size is 0
-        if(!file.exists(path) | file.info(path)$size == 0) {
-            results[index,] = c(gene, method, NA, NA, NA)
-            index = index + 1
-            next
-        }
-        # cigar
-        cigar = create_cigar(path)
-        # k2p
-        distance = dist.dna(read.dna(path, format = "fasta"), model = "K80")[1]
-        # omega
-        omega = calculate_omega(path)
-        # save data
-        results[index,] = c(gene, method, cigar, distance, omega)
-        index = index + 1
-    }
-}
-
-write.csv(x = results, file = "aln_data.csv", quote = FALSE, row.names = FALSE)
-
-############################## Helper functions ################################
-## omega
-calculate_omega = function(fasta) { #dN/dS
-    if(!file.exists(fasta)) { return(NA) }
-    a = read.alignment(fasta, format = "fasta")
-    if(class(a) != "alignment") {return(NA)}
-    k = kaks(a)
-    if(length(k) == 1) {return(NA)}
-    ka = k$ka[1]
-    ks = k$ks[1]
-    if(ka < 0 || ks < 0) {
-        return(0)
-    } else if(ka == 0 && ks == 0) {
-        return(0)
-    } else if(ks == 0) {
-        return(10)
-    } else {
-        return(ka/ks)
-    }
-}
-
-## CIGAR
-create_cigar = function(fasta) {
-  if(file.exists(fasta)) {
-    seqs = read.fasta(fasta, set.attributes = TRUE, forceDNAtolower = FALSE)
-  } else {
[…]

### 2024-01-28 tables/tab-benchmarks-print.R

@@ -1,0 +1,56 @@
+library(tidyverse)
+library(here)
+library(xtable)
+
+tab <- read_csv(here("tables", "benchmarks.csv"))
+
+tab <- tab |> mutate(
+        avg_dseq = scales::percent(avg_dseq, accuracy = 0.01),
+        over_k2p = scales::percent(over_k2p, accuracy = 0.1),
+        rmse_k2p = formatC(signif(tab$rmse_k2p, 3), flag = "#", digits = 3),
+        neg_f1 = scales::percent(neg_f1, accuracy = 0.1),
+        pos_f1 = scales::percent(pos_f1, accuracy = 0.1) ) |>
+    relocate(rmse_k2p, .before = over_k2p)
+
+stat_names <- c(
+    "Method" = "method",
+    "Average Alignment Error ($d_{seq}$)" = "avg_dseq",
+    "Number of Best Alignments" = "count_best",
+    "Number of Perfect Alignments" = "count_perfect",
+    "Number of Imperfect Alignments" = "count_imperfect",
+    "RMSE for K2P Distances" = "rmse_k2p",
+    "Overestimated K2P Distances" = "over_k2p",
+    "F\\textsubscript{1} for Positive Selection" = "pos_f1",
+    "F\\textsubscript{1} for Negative Selection" = "neg_f1"
+)
+
+method_levels <- c("coati-tri-mg", "clustalo", "macse",
+    "mafft", "prank")
+
+method_names <- c(
+        "COATi" = "coati-tri-mg",
+        "ClustalΩ" = "clustalo",
+        "MACSE" = "macse",
+        "MAFFT" = "mafft",
+        "PRANK" = "prank"
+)
+
+tab <- tab |> 
+    filter(method %in% method_levels) |>
+    mutate(method = fct_recode(fct_relevel(method, method_levels), !!!method_names)) |>
+    arrange(method) |>
+    rename(!!!stat_names)
+
+x <- t(tab)
+colnames(x) <- x[1,]
+x <- x[-1,]
+x[,1] <- paste0("\\cellcolor{bestcolor}", x[,1])
+
+
+print(xtable::xtable(x, align = c("r", "c", "c", "c", "c", "c")),
+    file = here("tables/benchmarks.tex"),
+    floating = FALSE, 
+    latex.environments = NULL,
+    booktabs = TRUE,
+    sanitize.text.function = \(x) gsub("%", "\\\\%", x)
+)

### 2024-01-28 tables/tab-benchmarks.R

@@ -1,0 +1,130 @@
+options(tidyverse.quiet = TRUE,
+    readr.show_col_types = FALSE,
+    conflicts.policy = list(warn = FALSE ))
+
+library(tidyverse)
+library(here)
+
+simplify_cigar_exp <- function(x) {
+    len <- str_extract_all(x, "\\d+")
+    val <- str_extract_all(x, "\\D+")
+
+    result <- map2_chr(len, val, function(n, v) {
+        if(length(n) == 1L && is.na(n)) {
+            return(NA_character_)
+        }
+        v[v == "=" | v == "X"] <- "M"
+        m <- rep.int(v, as.integer(n))
+        str_flatten(m)
+    })
+    result
+}
+
+calculate_dseq <- function(x, y) {
+    if(is.na(x) || is.na(y)) {
+        return(NA_real_)
+    }
+    if(x == y) {
+        # fast path
+        return(0.0);
+    }
+
+    # calculate homology sets
+    h <- function(z) {
+        z <- str_split_1(z, "")
+        za <- if_else(z == "I", -1, cumsum(z %in% c("M", "D")))
+        zd <- if_else(z == "D", -1, cumsum(z %in% c("M", "I")))
+        ha <- zd[z %in% c("M", "D")]
+        hd <- za[z %in% c("M", "I")]
+        list(ha, hd)
+    }
+
+    hx <- h(x)
+    hy <- h(y)
+
+    # total sequence length
+    tot_len <- length(hx[[1]]) + length(hx[[2]])
+    # calculate hamming distances
+    ham_dist <- sum(hx[[1]] != hy[[1]]) + sum(hx[[2]] != hy[[2]])
+
+    ham_dist/tot_len
+}
+
+bench_data <- read_csv(here("data/benchmark_fasta_aligned_stats.csv.gz"))
+
+# alignment analysis
+aln_data <- bench_data |> select(gene, method, cigar, score) |>
+    mutate(aln = simplify_cigar_exp(cigar)) |>
+    pivot_wider(id_cols = gene, names_from = method, values_from=c(score, aln)) |>
+    pivot_longer(-c(gene, score_benchmark, aln_benchmark),
+        names_to = c(".value", "method"), names_sep = "_") |>
+    relocate(gene, method)
+
+aln_data <- aln_data |> mutate(
[…]

### 2024-01-29 alignpair_letter.tex
Update colorblind colors.
Add new figure projecting the distance between aligners in 2d space

@@
 
+\begin{figure}[h!]
+    \centering%
+    \includegraphics{figures/fig-dseq-benchmarks.pdf}
+    \par
+    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods.
+A principal coordinates analysis of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
+Not only were COATi's alignments on average closest to the benchmark, but other aligners tended to diverge from the benchmark in different directions.
+%
+    \label{fig:dseq-benchmarks}}
+\end{figure}
+
 \clearpage
@@
 % The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
+
+% NOTES:
+% - Benchmark only contains seqeunces with gaps, which will make alignes look more different than they are in practice
+% - Parameters and optimization
+% - simulalation contained a lot of frameshifts that produed stop codons. Probably too many.
+% - redo colors of figures 1 - 3.
+% - define an equivalent alignment again in the benchmark results
 

### 2024-01-29 figures/fig-dseq-benchmarks.R

@@ -1,0 +1,93 @@
+library(tidyverse)
+library(here)
+library(ggrepel)
+
+set.seed(17768)
+
+PDF_WIDTH <- 3.38 # 7=178mm 3.38=86mm
+PDF_HEIGHT <- 0.85*PDF_WIDTH
+
+tab <- read_csv(here("tables", "benchmarks-dseq.csv"))
+methods <- unique(c(tab$method1,tab$method2))
+
+my_methods <- c("benchmark", "clustalo", "macse", "mafft", "prank",
+    "coati-tri-mg")
+
+tab2 <- tab |> filter(method1 %in% my_methods & method2 %in% my_methods)
+methods2 <- unique(c(tab2$method1,tab2$method2))
+
+# Create a PCoA
+m <- matrix(0, length(methods2), length(methods2))
+m[lower.tri(m)] <- tab2$dseq
+colnames(m) <- methods2
+rownames(m) <- methods2
+d <- as.dist(m)
+v <- cmdscale(d, 2)
+colnames(v) <- c("x", "y")
+
+pcoa_tab <- as_tibble(v, rownames = "method")
+
+method_recode <- c(
+        "Benchmark" = "benchmark",
+        "COATi" = "coati-tri-mg",
+        "ClustalΩ" = "clustalo",
+        "MACSE" = "macse",
+        "MAFFT" = "mafft",
+        "PRANK" = "prank",
+        "COATi-rev" = "rev-coati-tri-mg"
+)
+
+pcoa_tab <- pcoa_tab |> mutate(method = fct_recode(method, !!!method_recode))
+
+# Okabe and Ito
+#cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
+#    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+# Paul Tol's Vibrant Set
+# https://personal.sron.nl/~pault/#sec:qualitative
+cbbPalette <- c(
+    "COATi" = "#0077BB",
+    "COATi-rev" = "#33BBEE",
+    "MACSE" = "#009988",
+    "MAFFT" = "#EE7733",
+    "PRANK" = "#CC3311",
+    "ClustalΩ" = "#EE3377",
+    "Benchmark" = "#000000",
+    "missing" = "#BBBBBB")
+
+pal <- cbbPalette
+
+gg1 <- ggplot(pcoa_tab , aes(x = x, y = y, label = method, color =  method)) +
[…]

### 2024-01-29 figures/fig-k2p-empirical.R

@@ -3,24 +3,49 @@ library(scales)
 library(here)
 library(patchwork)
 
+set.seed(17760)
+
 PDF_WIDTH <- 7 # 178mm
 PDF_HEIGHT <- PDF_WIDTH/2
 
-#showtext_auto()
+# Okabe and Ito
+#cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
+#    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+#pal <- rev(cbbPalette[c(3,2,4,6,7,8)])
 
-cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
-    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
-set.seed(17760)
+# Paul Tol's Vibrant Set
+# https://personal.sron.nl/~pault/#sec:qualitative
+cbbPalette <- c(
+    "COATi" = "#0077BB",
+    "COATi-rev" = "#33BBEE",
+    "MACSE" = "#009988",
+    "MAFFT" = "#EE7733",
+    "PRANK" = "#CC3311",
+    "ClustalΩ" = "#EE3377",
+    "Benchmark" = "#000000",
+    "missing" = "#BBBBBB"
+    )
+
+pal <- cbbPalette[c(-7,-8)]
 
 # load data
 aln_data <- read_csv(here("data/raw_fasta_aligned_stats.csv.gz"))
 
-pal <- rev(cbbPalette[c(3,2,4,6,7,8)])
-
 #### PLOT 1: K2P density plots
 methods <- c("clustalo", "macse", "mafft", "prank", "coati-tri-mg", "rev-coati-tri-mg")
+
+method_recode <- c(
+    "COATi" = "coati-tri-mg",
+    "ClustalΩ" = "clustalo",
+    "MACSE" = "macse",
+    "MAFFT" = "mafft",
+    "PRANK" = "prank",
+    "COATi-rev" = "rev-coati-tri-mg"
+)
+
 aln_data2 <- aln_data |> filter(method %in% methods)
 aln_data2 <- aln_data2 |> mutate(method = fct_relevel(method, "coati-tri-mg"))
+aln_data2 <- aln_data2 |> mutate(method = fct_recode(method, !!!method_recode))
 
 mean_k2p <- aln_data2 |>
     summarize(avg = mean(k2p, na.rm = TRUE), .by = "method") |>
[…]

### 2024-01-29 tables/tab-dseq.R

@@ -1,0 +1,82 @@
+library(tidyverse)
+library(here)
+
+simplify_cigar_exp <- function(x) {
+    len <- str_extract_all(x, "\\d+")
+    val <- str_extract_all(x, "\\D+")
+
+    result <- map2_chr(len, val, function(n, v) {
+        if(length(n) == 1L && is.na(n)) {
+            return(NA_character_)
+        }
+        v[v == "=" | v == "X"] <- "M"
+        m <- rep.int(v, as.integer(n))
+        str_flatten(m)
+    })
+    result
+}
+
+calculate_dseq <- function(x, y) {
+    if(is.na(x) || is.na(y)) {
+        return(NA_real_)
+    }
+    if(x == y) {
+        # fast path
+        return(0.0);
+    }
+
+    # calculate homology sets
+    h <- function(z) {
+        z <- str_split_1(z, "")
+        za <- if_else(z == "I", -1, cumsum(z %in% c("M", "D")))
+        zd <- if_else(z == "D", -1, cumsum(z %in% c("M", "I")))
+        ha <- zd[z %in% c("M", "D")]
+        hd <- za[z %in% c("M", "I")]
+        list(ha, hd)
+    }
+
+    hx <- h(x)
+    hy <- h(y)
+
+    # total sequence length
+    tot_len <- length(hx[[1]]) + length(hx[[2]])
+    # calculate hamming distances
+    ham_dist <- sum(hx[[1]] != hy[[1]]) + sum(hx[[2]] != hy[[2]])
+
+    ham_dist/tot_len
+}
+
+bench_data <- read_csv(here("data/benchmark_fasta_aligned_stats.csv.gz"))
+
+methods <- unique(bench_data$method)
+
+dat <- bench_data |> select(gene, method, cigar) |>
+    mutate(aln = simplify_cigar_exp(cigar)) |>
+    select(gene, method, aln) |>
+    arrange(gene, method)
+
+## sanity check
+# dat |> filter(method != methods, .by=gene)
+
+xy <- combn(length(methods), 2)
+x <- xy[1,]
+y <- xy[2,]
+
+tab <- dat |>
+    reframe(method1 = method[x], method2 = method[y],
+        aln1 = aln[x], aln2 = aln[y], .by = gene)
+
+tab <- tab |>
+    mutate(dseq = map2_dbl(aln1, aln2, calculate_dseq, .progress = TRUE))
+
[…]

### 2024-01-31 alignpair_letter.tex
Update discussion.

@@
 
-\subsection*{Simulated Benchmark Alignments}
+\subsection*{Semi-Empirical Benchmark Alignments}
 
@@
 Using $d_{seq}$ we quantified the number of perfect, imperfect and best alignments each aligner produced.
-We define perfect alignments as alignments with a distance of zero to their benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero.
+We define perfect alignments as alignments with a distance of zero to the benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero.
 Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from influencing our results.
@@
 
-\paragraph{Empirical Data and Alignments.}
+\subsection*{Empirical Data and Alignments}
 
@@
 And PRANK identified 1,349.
+Note that we did not explore whether any of these inferences were biologically or statistically significant.
 
@@
 
-\subsection*{Benchmark Results}
+\subsection*{Semi-Empirical Benchmark}
+
+COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set.
+It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
+%
+COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}).
+All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests.
+(P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$).
+Notably, the average alignment error of the second best protocol was four times larger than COATi's.
+We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}).
+COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions.
+Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S5).
+While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.
+
+For evolutionary distances, COATi had the lowest root-mean-square error (0.000694) relative to the benchmark alignments (Tab.~\ref{table:comp}).
+The second lowest aligner, PRANK, had a root-mean-square error that was over 400 times larger than COATi's.
+COATi only overestimated  10\% of its evolutionary distances, while other aligners overestimated 26\%--53\% of their evolutionary distances.
+%
+Additionally, COATi more accurately inferred events of positive and negative selection (Tab.~\ref{table:comp}).
 
@@
 \centering
-\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\label{table:comp}}
+\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\textsuperscript{1}\label{table:comp}}
 %
@@
 \addlinespace 
-  Number of Best\textsuperscript{1} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
+  Number of Best\textsuperscript{2} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
   Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
-  Number of Imperfect\textsuperscript{3} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
+  Number of Imperfect\textsuperscript{2} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
   \addlinespace 
@@
   \addlinespace
-  F\textsubscript{1} for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
-  F\textsubscript{1} for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
+  F\textsubscript{1} Score for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
+  F\textsubscript{1} Score for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
    \bottomrule
    \addlinespace
-\multicolumn{6}{p{40em}}{\textsuperscript{1}Best alignments have the lowest $d_{seq}$ (including equivalent alignments)}\\
-\multicolumn{6}{p{40em}}{\textsuperscript{2}Perfect alignments have $d_{seq} = 0$ (including equivalent alignments)}\\
-\multicolumn{6}{p{40em}}{\textsuperscript{3}Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment.}\\
+\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark was 8,261.}\\
+\multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments).
+Perfect alignments have $d_{seq} = 0$ (including equivalent alignments).
+Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment.
+}\\
 \end{tabular}
@@
     \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods.
-A principal coordinates analysis of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
-Not only were COATi's alignments on average closest to the benchmark, but other aligners tended to diverge from the benchmark in different directions.
+A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
 %
@@
 % - Benchmark only contains seqeunces with gaps, which will make alignes look more different than they are in practice
-% - Parameters and optimization
 % - simulalation contained a lot of frameshifts that produed stop codons. Probably too many.
@@
 
-COATi obtained better results compared to a wide variety of alignment strategies.
-It was significantly more accurate (lower $d_{seq}$) at inferring the empirically simulated alignments compared to other methods; all p-values were less than $1.3 \cdot 10^{-76}$ according to the one-tailed, paired Wilcoxon signed-rank tests (Supplementary Materials Figure 1).
-Notably, the average alignment error of the second best protocol was six times larger than COATi's.
-In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately inferred events of positive and negative selection (Table \ref{table:comp}).
-Furthermore, the estimated evolutionary divergence from the alignments retrieved by COATi was substantially less overestimated than other methods (Table \ref{table:comp}, Supplemental Materials Figure 8).
+We have developed a statistical alignment program using finite-state transducers called COATi.
+On empirical data, COATi produced alignments that provided reasonable biological inferences.
+On semi-empirical, simulated benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid away nucleotide model, MAFFT's DNA model, and PRANK's codon model.
+COATi's advantage over other aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings.
+Different data sets may show different results.
+However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.
 
-Aligning sequences using amino-acid translation via ClustalΩ obtained the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
-These results are not surprising because alignments in amino-acid space only permit phase-3 gaps and are sensitive to frameshift artifacts, which inflate divergence estimates and impact selection detection.
-Conversely, we obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT.
+COATi's primary disadvantage is that its FST model was slower and consumed more resources than other methods.
+However, COATi's marginal model produces results similar to the FST model, was the fastest method tested, and consumed reasonable amounts of memory \citep{garcia2023dissertation}.
+COATi is also not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
+%
+In order to test whether the choice of a reference sequence impact's COATi's results, we aligned our empirical sequence pairs with COATi using gorilla as the reference.
+We found no biologically or statistically significant differences between the results.
+The most noticeable difference was that with gorilla as the reference COATi rejected more sequence pairs due to the presence of ambiguous nucleotides in gorilla sequences.
+
+After COATi, PRANK's codon model was the second most accurate alignment method.
+Like COATi, PRANK also uses statistical models and supports aligning protein-coding sequences using codon models developed for phylogenetics.
+However, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
+In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.
+Similarly, ClustalΩ amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts; however, it was the worst performing protocol that we tested.
+It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
+--- It even performed poorly on its own gap patterns.
+--- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
+
+We obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT.
 These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts.
 If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments.
-And finally, our PRANK protocol used a codon substitution model which should have been an improvement over both amino-acid and nucleotide models; however, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
-In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.
 
-COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
+We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols.
+Aside from the cluster of different COATi methods, we detected no clusters in our analysis.
+Different aligners tended to spread out from one another and in different directions from the benchmark.
+This indicates that each protocol approached alignment inference from a different section of ``aligner space''.
+While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns.
+COATi did well on all patterns.
+
+% COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
 To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference.

### 2024-01-31 supplementary_materials/supplementary_materials.Rmd
Update discussion.

@@
 library(here)
+library(ggrepel)
 
-cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
-    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+#cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
+#    "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
+
+cbbPalette <- c(
+    "COATi" = "#0077BB",
+    "COATi-rev" = "#33BBEE",
+    "MACSE" = "#009988",
+    "MAFFT" = "#EE7733",
+    "PRANK" = "#CC3311",
+    "ClustalΩ" = "#EE3377",
+    "Benchmark" = "#000000",
+    "missing" = "#BBBBBB")
+
 set.seed(17761)
@@
 
-```{r aln-example-a, fig.align='center', fig.cap="The FST of all possible alignments between `CTC' and `CTG'."}
+```{r aln-example-a, fig.align='center', fig.cap="The FST of all possible alignments between \"CTC\" and \"CTG\"."}
 knitr::include_graphics("figures/aln-example-graph.pdf")
@@
 
-```{r aln-example-b, out.width="75%", fig.align='center', fig.cap="The best alignment of `CTC' and `CTG'."}
+```{r aln-example-b, out.width="75%", fig.align='center', fig.cap="The best alignment of \"CTC\" and \"CTG\"."}
 knitr::include_graphics("figures/aln-example-path.pdf")
@@
         "vs PRANK" = "prank",
-        "vs COATI-rev" = "coati-rev"
+        "vs COATi-rev" = "coati-rev"
     ))
+
+pal <- cbbPalette
+names(pal) <- str_c("vs ", names(pal))
 
@@
 gg <- gg + xlab("COATi") + ylab("Other")
-gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
+gg <- gg + scale_color_manual(values = pal)
 gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0003, base=10),
@@
 gg <- gg + xlab("COATi") + ylab("Other")
-gg <- gg + scale_color_manual(values = cbbPalette[c(2,4,6,7,8)])
+gg <- gg + scale_color_manual(values = pal)
 gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
@@
 
+# Benchmark Results
+
+## Alignment Distances
+
+For each sequence pair in the benchmark, we calculated the alignment distance (d~seq~) between the benchmark alignment and the alignments generated by COATi, ClustalΩ, MACSE, MAFFT, and PRANK.
+We also calculated distances between all pairs of aligners.
+Our benchmark contained gap patterns extracted from alignments generated by different aligners.
+Figure \@ref(fig:bench-pcoa-split) contains the results of a metric multidimensional scaling (principle coordinate analysis; PCoA) of the matrix of average distances between aligners, separated by what type of gap patterns was used in the benchmark alignment.
+
+Figure \@ref(fig:bench-pcoa-all) contains a principle coordinate analysis of each aligner, including different COATi models, across the entire benchmark dataset.
+COATi's different models produced similar alignments and cluster together along with the benchmarks.
+
+```{r bench-pcoa-split, message = FALSE, fig.cap="COATi produced accurate alignments regardless of whether the underlying gap pattern was extracted from an alignment generated by another program.
+Each panel is a metric multidimensional scaling of the average distances between aligners."}
+tab_dist <- read_csv(here("supplementary_materials", "data", "benchmarks-dseq-split.csv"))
+
+methods <- unique(c(tab_dist$method1, tab_dist$method2))
+
+tab_xy <- tab_dist |> group_by(origin) |> 
+    group_modify(\(df, k) {
+        m <- matrix(0, length(methods), length(methods))
+        m[lower.tri(m)] <- df$dseq
+        colnames(m) <- methods
+        rownames(m) <- methods
+        d <- as.dist(m)
+        v <- cmdscale(d, 2)
+        colnames(v) <- c("x", "y")
+        as_tibble(v, rownames = "method")
+    }) |> ungroup()
+
+method_recode <- c(
+        "Benchmark" = "benchmark",
+        "COATi" = "coati-tri-mg",
+        "ClustalΩ" = "clustalo",
+        "MACSE" = "macse",
+        "MAFFT" = "mafft",
+        "PRANK" = "prank"
+)
+
+origin_recode <- c(
+        "COATi Gaps" = "coati-tri-mg",
+        "ClustalΩ Gaps" = "clustalo",
+        "MACSE Gaps" = "macse",
+        "MAFFT Gaps" = "mafft",
+        "PRANK Gaps" = "prank"
+)
+
+tab <- tab_xy |> mutate(
+    method = fct_recode(method, !!!method_recode),
+    origin = fct_recode(origin, !!!origin_recode) )
+
+pal <- cbbPalette
+
+gg1 <- ggplot(tab, aes(x = x, y = y, label = method, color =  method)) +
+    geom_point(show.legend = FALSE) +
+    geom_text_repel(show.legend = FALSE, seed = 3)
+
+gg1 <- gg1 + facet_wrap(vars(origin))
+
+gg1 <- gg1 + xlab("Coordinate 1") + ylab("Coordinate 2")
+
+gg1 <- gg1 + scale_color_manual(values = pal)
+
+gg1 <- gg1 + coord_fixed()
+
+gg1 <- gg1 + theme_grey(9)
+
+gg1
+
+```
+
+```{r bench-pcoa-all, message = FALSE, fig.cap="Different COATi models produce accurate alignments that are also similar to one another."}
+tab_dist <- read_csv(here("supplementary_materials", "data", "benchmarks-dseq.csv"))
+
+methods <- unique(c(tab_dist$method1, tab_dist$method2))
+
+m <- matrix(0, length(methods), length(methods))
+m[lower.tri(m)] <- tab_dist $dseq
+colnames(m) <- methods
+rownames(m) <- methods
+d <- as.dist(m)
+v <- cmdscale(d, 2)
+colnames(v) <- c("x", "y")
+tab_xy <- as_tibble(v, rownames = "method")
+
+method_recode <- c(
+    "Benchmark" = "benchmark",
+    "COATi-tri-mg" = "coati-tri-mg",
+    "COATi-tri-ecm" = "coati-tri-ecm",
+    "COATi-mar-mg" = "coati-mar-mg",
+    "COATi-mar-ecm" = "coati-mar-ecm",
+    "COATi-dna" = "coati-dna",
+    "COATi-rev-tri-mg" = "rev-coati-tri-mg",
+    "ClustalΩ" = "clustalo",
+    "MACSE" = "macse",
+    "MAFFT" = "mafft",
+    "PRANK" = "prank"
+)
+
+tab <- tab_xy |> mutate(
+    method = fct_recode(method, !!!method_recode))
+
+pal <- c(
+    "COATi-tri-mg" = "#0077BB",
+    "COATi-tri-ecm" = "#0077BB",
+    "COATi-mar-mg" = "#0077BB",
+    "COATi-mar-ecm" = "#0077BB",
+    "COATi-dna" = "#0077BB",
+    "COATi-rev-tri-mg" = "#33BBEE",
+    "MACSE" = "#009988",
+    "MAFFT" = "#EE7733",
+    "PRANK" = "#CC3311",
+    "ClustalΩ" = "#EE3377",
+    "Benchmark" = "#000000",
+    "missing" = "#BBBBBB")
+
+gg <- ggplot(tab, aes(x = x, y = y, label = method, color =  method)) +
+    geom_point(show.legend = FALSE) +
+    geom_text_repel(show.legend = FALSE, seed = 1, max.overlaps = 20, force = 2)
+
+gg <- gg + xlab("Coordinate 1") + ylab("Coordinate 2")
+gg <- gg + scale_color_manual(values = pal)
+gg <- gg + coord_fixed()
+gg <- gg + theme_grey(9)
+
+gg
+```
+
 <!-- # References -->

### 2024-01-31 supplementary_materials/supplementary_materials.tex
Update discussion.

@@
 
-\caption{The FST of all possible alignments between `CTC' and `CTG'.}\label{fig:aln-example-a}
+\caption{The FST of all possible alignments between "CTC" and "CTG".}\label{fig:aln-example-a}
 \end{figure}
@@
 
-\caption{The best alignment of `CTC' and `CTG'.}\label{fig:aln-example-b}
+\caption{The best alignment of "CTC" and "CTG".}\label{fig:aln-example-b}
 \end{figure}
@@
 
+\hypertarget{benchmark-results}{%
+\section{Benchmark Results}\label{benchmark-results}}
+
+\hypertarget{alignment-distances}{%
+\subsection{Alignment Distances}\label{alignment-distances}}
+
+For each sequence pair in the benchmark, we calculated the alignment distance (d\textsubscript{seq}) between the benchmark alignment and the alignments generated by COATi, ClustalΩ, MACSE, MAFFT, and PRANK.
+We also calculated distances between all pairs of aligners.
+Our benchmark contained gap patterns extracted from alignments generated by different aligners.
+Figure \ref{fig:bench-pcoa-split} contains the results of a metric multidimensional scaling (principle coordinate analysis; PCoA) of the matrix of average distances between aligners, separated by what type of gap patterns was used in the benchmark alignment.
+
+\begin{figure}
+\centering
+\includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-split-1.pdf}
+\caption{\label{fig:bench-pcoa-split}COATi produced accurate alignments regardless of whether the underlying gap pattern was extracted from an alignment generated by another program.
+Each panel is a metric multidimensional scaling of the average distances between aligners.}
+\end{figure}
+
+Figure \ref{fig:bench-pcoa-all} contains a principle coordinate analysis of each aligner, including different COATi models, across the entire benchmark dataset.
+COATi's different models produced similar alignments and cluster together along with the benchmarks.
+
+\begin{figure}
+\centering
+\includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-all-1.pdf}
+\caption{\label{fig:bench-pcoa-all}Different COATi produce accurate alignments that are also similar to one another.}
+\end{figure}
+
 \end{document}

### 2024-01-31 tables/tab-dseq-split.R

@@ -1,0 +1,87 @@
+library(tidyverse)
+library(here)
+
+simplify_cigar_exp <- function(x) {
+    len <- str_extract_all(x, "\\d+")
+    val <- str_extract_all(x, "\\D+")
+
+    result <- map2_chr(len, val, function(n, v) {
+        if(length(n) == 1L && is.na(n)) {
+            return(NA_character_)
+        }
+        v[v == "=" | v == "X"] <- "M"
+        m <- rep.int(v, as.integer(n))
+        str_flatten(m)
+    })
+    result
+}
+
+calculate_dseq <- function(x, y) {
+    if(is.na(x) || is.na(y)) {
+        return(NA_real_)
+    }
+    if(x == y) {
+        # fast path
+        return(0.0);
+    }
+
+    # calculate homology sets
+    h <- function(z) {
+        z <- str_split_1(z, "")
+        za <- if_else(z == "I", -1, cumsum(z %in% c("M", "D")))
+        zd <- if_else(z == "D", -1, cumsum(z %in% c("M", "I")))
+        ha <- zd[z %in% c("M", "D")]
+        hd <- za[z %in% c("M", "I")]
+        list(ha, hd)
+    }
+
+    hx <- h(x)
+    hy <- h(y)
+
+    # total sequence length
+    tot_len <- length(hx[[1]]) + length(hx[[2]])
+    # calculate hamming distances
+    ham_dist <- sum(hx[[1]] != hy[[1]]) + sum(hx[[2]] != hy[[2]])
+
+    ham_dist/tot_len
+}
+
+bench_data <- read_csv(here("data/benchmark_fasta_aligned_stats.csv.gz"))
+
+meta_data <- read_csv(here("data/metadata.csv.gz"))
+
+meta_data <- meta_data |>
+    mutate(gene = sprintf("TEST%06d", 1:n())) |>
+    rename(origin = model)
+
+my_methods <- c("benchmark", "clustalo", "macse", "mafft", "prank",
+    "coati-tri-mg")
+
+cigar_data <- bench_data |> 
+    filter(method %in% my_methods) |>
+    select(gene, method, cigar) |>
+    mutate(aln = simplify_cigar_exp(cigar)) |>
+    select(gene, method, aln) |>
+    arrange(gene, method)
+
+methods <- unique(cigar_data$method)
+
+## sanity check
+# dat |> filter(method != methods, .by=gene)
+
[…]

### 2024-01-31 tables/wilcox-tests.R

@@ -1,0 +1,80 @@
+options(tidyverse.quiet = TRUE,
+    readr.show_col_types = FALSE,
+    conflicts.policy = list(warn = FALSE ))
+
+library(tidyverse)
+library(here)
+
+simplify_cigar_exp <- function(x) {
+    len <- str_extract_all(x, "\\d+")
+    val <- str_extract_all(x, "\\D+")
+
+    result <- map2_chr(len, val, function(n, v) {
+        if(length(n) == 1L && is.na(n)) {
+            return(NA_character_)
+        }
+        v[v == "=" | v == "X"] <- "M"
+        m <- rep.int(v, as.integer(n))
+        str_flatten(m)
+    })
+    result
+}
+
+calculate_dseq <- function(x, y) {
+    if(is.na(x) || is.na(y)) {
+        return(NA_real_)
+    }
+    if(x == y) {
+        # fast path
+        return(0.0);
+    }
+
+    # calculate homology sets
+    h <- function(z) {
+        z <- str_split_1(z, "")
+        za <- if_else(z == "I", -1, cumsum(z %in% c("M", "D")))
+        zd <- if_else(z == "D", -1, cumsum(z %in% c("M", "I")))
+        ha <- zd[z %in% c("M", "D")]
+        hd <- za[z %in% c("M", "I")]
+        list(ha, hd)
+    }
+
+    hx <- h(x)
+    hy <- h(y)
+
+    # total sequence length
+    tot_len <- length(hx[[1]]) + length(hx[[2]])
+    # calculate hamming distances
+    ham_dist <- sum(hx[[1]] != hy[[1]]) + sum(hx[[2]] != hy[[2]])
+
+    ham_dist/tot_len
+}
+
+bench_data <- read_csv(here("data/benchmark_fasta_aligned_stats.csv.gz"))
+
+# alignment analysis
+aln_data <- bench_data |> select(gene, method, cigar, score) |>
+    mutate(aln = simplify_cigar_exp(cigar)) |>
+    pivot_wider(id_cols = gene, names_from = method, values_from=c(score, aln)) |>
+    pivot_longer(-c(gene, score_benchmark, aln_benchmark),
+        names_to = c(".value", "method"), names_sep = "_") |>
+    relocate(gene, method)
+
+aln_data <- aln_data |> mutate(
+    dseq = map2_dbl(aln_benchmark, aln, calculate_dseq)
+)
+
[…]

### 2024-02-05 alignpair_letter.tex
Complete RAC revisions. Version sent to Ziqi and Juan.

@@
 Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
-Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
+Sequence artifacts and errors made during alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
 For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
@@
         Phases refer the the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
-        Phase-1, phase-2, and phase-3 indels are shown in purple, orange, and gray respectively.
-        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the purple indel is type-I (an amino-acid indel only).
-The difference between in-frame and frameshift indels is not displayed.
+        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
+        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only).
+The difference between an in-frame and a frameshift indel is not displayed.
         }
@@
     \caption[FSTs more the generation of sequences]{%
-    Finite state transducers (FSTs) model the generation of an output sequences based on an input sequence.
+    Finite state transducers (FSTs) model the generation of an output sequence based on an input sequence.
     %
@@
     (b) An FST for matching sequences against ambiguous nucleotides (N). 
-This FST is not a true probabilistic FST and cannot be used to simulate output sequences since it is missing a parameter to control how often Ns are added to the output sequence.
-    %
     (c) An FST that results from the composition ($∘$ operation) of the Error FST with the Ambiguity FST.
-As with (b), this composed FST cannot be used to simulate output sequences; however, it does properly weight the ambiguous nucleotide N as representing any other symbol.
     %
@@
 
-Since codon substitution is the first process in COATi's model, the input sequence to the COATi FST must be compatible with a codon-substitution model, i.e.\ be a multiple of three nucleotides and not contain any ambiguous symbols or stop codons.
+COATi FST is not a true probabilistic FST and cannot be used as-is to simulate output sequences based on an input sequence.
+This is because it is missing a parameter to control how often Ns are added to the output sequence.
+This is design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol.
+In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment.
+This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C.
+While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and ignoring any states in between:
+\[
+\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
+\]
+where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence.
+The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.
+
+Since codon substitution is the first process in COATi's model, the input sequence to the COATi FST must be compatible with a codon-substitution model and be a multiple of three nucleotides and not contain any ambiguous symbols or stop codons.
 The phases, reading frames, and amino-acid contexts of alignment columns is determined by the input sequence, and better alignments will be generated if the input is of high quality and free of artifacts.
@@
 %
-(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the ``gap opening parameter'' and $e$ is the ``gap extension'' parameter.
+(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the gap-opening parameter and $e$ is the gap-extension parameter.
 Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
@@
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
-An example of an FST-based alignment can be found in Supplementary Materials Figure 1.
+An example of an FST-based alignment can be found in Supplementary Materials Figure S1.
 
@@
 This treats all gaps equally as the location of the gap is not recorded.
-Finally, $d_{seq}$ is calculated  between an estimated and a benchmark alignment as the average, normalized Hamming distance between the homology sets for each nucleotide in the sequence pair.
-See \cite{metrics_blackburne_whelan_2011} for further details.
+Finally, $d_{seq}$ is calculated  between an estimated and a benchmark alignment as the average, normalized Hamming distance between the homology sets for each nucleotide in the sequence pair \citep{metrics_blackburne_whelan_2011}.
 
@@
 
-Using $d_{seq}$ we quantified the number of perfect, imperfect and best alignments each aligner produced.
+Using $d_{seq}$ we quantified the number of perfect, imperfect, and best alignments each aligner produced.
 We define perfect alignments as alignments with a distance of zero to the benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero.
-Equivalency was included in our definition of perfect alignments to prevent the manner by which aligners break ties between multiple optimal alignments from influencing our results.
+We included equivalency in our definition of perfect alignments to prevent the manner by which aligners break ties from influencing our results.
 Evolutionary equivalence was determined by scoring alignments using COATi's marginal model.
@@
 %
-Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties.
+Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties and equivalent alignments.
 %
@@
 Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
-A total of 10296 sequence pairs were aligned equivalently by all five methods, including 8261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1371 that had identical scores.
+A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.
 
@@
 On empirical data, COATi produced alignments that provided reasonable biological inferences.
-On semi-empirical, simulated benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid away nucleotide model, MAFFT's DNA model, and PRANK's codon model.
-COATi's advantage over other aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings.
+On semi-empirical, simulated benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
+Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps.
+Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results estimate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels.
+This indicates that alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 or phase-2.
+
+In additional to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings.
 Different data sets may show different results.
@@
 COATi's primary disadvantage is that its FST model was slower and consumed more resources than other methods.
-However, COATi's marginal model produces results similar to the FST model, was the fastest method tested, and consumed reasonable amounts of memory \citep{garcia2023dissertation}.
-COATi is also not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
+However, COATi's marginal model produces results similar to the FST model, was the fastest method tested, and consumed reasonable amounts of memory (Supplementary Materials and \citealt{garcia2023dissertation}).
 %
+COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
 In order to test whether the choice of a reference sequence impact's COATi's results, we aligned our empirical sequence pairs with COATi using gorilla as the reference.
 We found no biologically or statistically significant differences between the results.
-The most noticeable difference was that with gorilla as the reference COATi rejected more sequence pairs due to the presence of ambiguous nucleotides in gorilla sequences.
+The most noticeable difference was that, with gorilla as the reference, COATi rejected more sequence pairs due to the presence of ambiguous nucleotides in gorilla sequences.
 
@@
 Like COATi, PRANK also uses statistical models and supports aligning protein-coding sequences using codon models developed for phylogenetics.
-However, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
+Both COATi and PRANK had reasonable distributions of evolutionary distances that lacked large right tails.
+This is likely because COATi and PRANK produced better alignments when tasked with aligning sequence pairs that contained non-orthologous exons.
+However, in contrast with COATi, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
 In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.
-Similarly, ClustalΩ amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts; however, it was the worst performing protocol that we tested.
+
+ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts; however, it was the worst performing protocol that we tested.
 It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
 --- It even performed poorly on its own gap patterns.
---- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
+--- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
 
@@
 We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols.
-Aside from the cluster of different COATi methods, we detected no clusters in our analysis.
+Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S6).
 Different aligners tended to spread out from one another and in different directions from the benchmark.
-This indicates that each protocol approached alignment inference from a different section of ``aligner space''.
+This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''.
 While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns.

### 2024-02-05 figures/fig-fst-base-calling-new.tex
Complete RAC revisions. Version sent to Ziqi and Juan.

@@
 
+\definecolor{TolHighContrastBlue}{RGB}{0,68,136}
+\definecolor{TolVibrantBlue}{RGB}{0,119,187}
+
 % (86 mm) or double (178 mm) column width
@@
   end/.style={fill=none, text=black, inner sep=1pt},
-  block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2em,node distance=4em},
+  block/.style={circle,on grid,fill=TolVibrantBlue,text=white,minimum width=2em,node distance=4em},
   line/.style={draw, -Stealth, thick},

### 2024-02-05 figures/fig-fst-coati.tex
Complete RAC revisions. Version sent to Ziqi and Juan.

@@
 
+\definecolor{TolVibrantBlue}{RGB}{0,119,187}
+
 % (86 mm) or double (178 mm) column width
@@
 
-\draw[white,line width=2pt] ($(w)!0.245!(i)$) -- +(0,4pt);
-\draw[white,line width=2pt] ($(w)!0.385!(i)$) -- +(0,4pt);
-\draw[white,line width=2pt] ($(w)!0.525!(i)$) -- +(0,4pt);
-\draw[white,line width=2pt] ($(w)!0.665!(i)$) -- +(0,4pt);
-
 \coordinate (suba) at ($(w.center)!0.30! 20:(i.center) +(2em,0.5em)$);
@@
 
-\draw[line] (s) to[out=225,in=0] (suba) to[out=180,in=350] (m);
-\draw[line] (s) to[out=240,in=0] (subc) to[out=180,in=335] (m);
-\draw[line] (s) to[out=255,in=0] (subg) to[out=180,in=320] (m);
-\draw[line] (s) to[out=270,in=0] (subt) to[out=180,in=305] (m);
+\draw[white,line width=2pt] (suba -| w) +(0,-2pt) -- +(0,2pt);
+\draw[white,line width=2pt] (subc -| w) +(0,-2pt) -- +(0,2pt);
+\draw[white,line width=2pt] (subg -| w) +(0,-2pt) -- +(0,2pt);
+\draw[white,line width=2pt] (subt -| w) +(0,-2pt) -- +(0,2pt);
+
+\draw[line] (s) to[out=225,in=0] (suba) to ($(suba)+(-20mm,0mm)$) to[out=180,in=330] (m);
+\draw[line] (s) to[out=240,in=0] (subc) to ($(subc)+(-18mm,0mm)$) to[out=180,in=315] (m);
+\draw[line] (s) to[out=255,in=0] (subg) to ($(subg)+(-15mm,0mm)$) to[out=180,in=300] (m);
+\draw[line] (s) to[out=270,in=0] (subt) to ($(subt)+(-10mm,0mm)$) to[out=180,in=285] (m);
 

### 2024-02-09 response_r1.md
Revision 1 - Release Candidate 1

@@
 
-We would like to thank the associate editor and two reviewers for their helpful comments and suggestions. We have addressed all of the concerns brought to our attention as described below.
+Dear Editor,
 
-## Associate Editor
+We are resubmitting our manuscript, "COATi: statistical pairwise alignment of protein coding sequences", for your consideration.
+This manuscript describes a new model for aligning protein coding sequences which produces more accurate alignments, while being robust to common artifacts found in genomic annotations.
+Our approach is implemented in the software package, COATi, which is released via an MIT license and available on GitHub.
+While COATi can generate multiple-sequence alignments, this paper is focused on our pairwise aligner which is has received more attention and validation.
+Additionally, to promote open science and reproducibility all data, procedures, and scripts used to validate COATi and compare it with alternative methods have been uploaded to GitHub.
+
+We would like to thank the associate editor and two reviewers for their helpful comments and suggestions.
+As requested, we have broadened the scope of the manuscript from a software note to a larger research paper about our COATi software.
+Nearly all of the paper has been rewritten, expanded, and reconfigured to provide more details and background about our work.
+Importantly, we now use empirical results from the human-gorilla sequence pairs to initially compare aligners before we show results on semi-empirical benchmarks.
+If certain concepts need additional exposition, we will be happy to make those adjustments.
+
+We have addressed all of the concerns brought to our attention by the associate editor and reviewers as described below.
+
+Sincerely,  
+Reed A. Cartwright, on behalf of all authors
+
+## Associate Editor Comments
 
@@
     We thank the editor and reviews for the positive feedback.
-As requested we have broaden the scope of the manuscript from a software note to a larger research paper about our COATi software.
-Nearly all of the paper has been rewritten, expanded, and reconfigured to provide more details and background about our work.
-If certain concepts need additional exposition, we will be happy to make those adjustments.
+As mentioned above, we have now broadened the scope of the manuscript from a software note to a larger research paper about our COATi software.
+We have rewritten nearly the entire paper as we expanded to scope in response to feedback.
 
@@
 This modeling choice introduces additional directionality into the model.
-If there are two sequences (A and B) the method arbitrarily choses one for the ancestor and one for the descendent.
+If there are two sequences (A and B) the method arbitrarily chooses one for the ancestor and one for the descendant.
 What is not at all clear is whether this directionality is important, i.e., is the same result generated if the model goes A-->B as it would if it went B-->A.
@@
 In particular, we now explain that our input sequence (i.e.
-"the ancestor") is required to have a length that is a multiple of 3 and missing stop codons and ambiguous nucleotides (circa line 84).
+"the ancestor" or "reference") is required to have a length that is a multiple of 3 and missing stop codons and ambiguous nucleotides (circa line 99).
 The output sequence (i.e.
-"the descendant") does not have these restrictions.
+"the descendant" or "non-reference") does not have these restrictions.
 This asymmetry is due to COATi needing one high-quality reference sequence to establish reading-frame contexts.
 
-    Our primary results are presented using an empirically simulated benchmark derived from human-gorilla sequence pairs.
-These simulations were not intended to create gorilla sequences that were compatible with the restrictions that COATi places on input sequences.
-Therefore, we were only able to swap the order of ~50% of the benchmark sequence pairs to measure the impact of directionality on our results.
-The results show a decrease, albeit small, in accuracy across all metrics when the lower quality sequence is used as the reference.
-We now report these results in the main paper and supplementary materials.
+    In order to test the impact of directionality on our results, we have included results when using gorilla as the reference sequence.
+The results are very similar to using human as the reference sequence.
 
@@
 
-## Reviewer 1
+## Reviewer 1 Comments
 
@@
 
-    We now comment on this in the discussion: "COATi is under active development.... We plan on ... implement[ing] new models for aligning long-read sequences of genes against reference genomes."
+    We now comment on this in the discussion.
 
-## Reviewer 2
+## Reviewer 2 Comments
 
@@
 Zhu who graduated the year before.
+
+    We also discovered a bug in COATi's scoring algorithm (used for defining "perfect alignments" in the analysis) as well as minor issues with how we were using MACSE and ClustalΩ.
+We opted to take the time to regenerate our results after fixing these issues.
+Tables and figures barely changed and no conclusions were affected.
 
@@
 
+    To handle indels, the indel FST (Fig 3b) has a state space of four symbols (A, C, G, and T) with a special symbol, epsilon, that represents an empty match.
+An arc that reads nothing (epsilon) and outputs a symbol represents an insertion, while an arc that reads a symbol and outputs nothing (epsilon) represents a deletion.
+
+    We don't specifically model these processes in a single FST as we use the composition algorithm to combine separate FSTs that represent codon substitution, indels, error, and sequencing ambiguity to produce one FST that handles all four processes (Fig 3c).
+
     We hope the updated manuscript offers clarity about these points.
@@
 
-    One of the advantages of using FSTs and the composition algorithm is that it handles this for us. TODO...
+    We are able to easily mix codon models with indels that occur within codons because we are modeling alignments using FSTs.
+The composition algorithm allows us to combine an FST representing a codon substitution model with an FST representing an indel model (Figure 3).
+Composition does produce the complex graphs that you would expect from combining a codon model with an indel model, and we show a short example in Figure S1.
+Composition is a complex but well established algorithm in the FST literature, and we have chosen to refer readers to Mohri et al.
+(2005) if they would like to know more.
 
@@
     We have updated Figure 2 and added a new Figure 3 that we feel more adequately explains the finite-state transducers that we are using and the composition algorithm that allows us to combine them in a reasonable manner.
-We also include in the supplement ....
-TODO
+We also include in the supplement Figures S1 and S2 which show examples of fully composed alignment graph between the codons "CTC" and "CTG" as well as the best path through the graph.
 
@@
 
-    COATi uses the first sequence in the input file as the input/reference sequence, and the second sequence as the output/non-reference sequence.
+    By default, COATi uses the first sequence in the input file as the input/reference sequence and the second sequence as the output/non-reference sequence.
+It also contains flags to customize this behavior.
 This gives users the power to place low-quality sequences as the output sequence when aligning a pair of sequences using COATi FST.
-We now mention this in the manuscript circa line 95.
+We now mention this in the manuscript circa line 108.
 
@@
 
-    We have added details about how we empirically simulated benchmark alignments (circa line 137).
+    We have added details about how we created our semi-empirical benchmark alignments (circa line 227).
 The revision now contains information about our procedure, as well as an example of it being applied to data.
@@
 
-    We used sequinr to calculate Ka and Ks and have expanded our discussion to expain which method that sequinr uses to calculate these stats and how that method works (briefly).
+    We used sequinr to calculate Ka and Ks and have expanded our methods to explain which method that sequinr uses to calculate these stats and how that method works (briefly).
 
@@
 We currently use OpenFST to implement our FST-based algorithms.
-We now mention this in the manuscript (circa line 100).
+We now mention this in the manuscript (circa line 116).
 
@@
 
-    We have quantified each aligner's impact on K2P distance estimation.
-We report these results as the percent of benchmark alignments which aligners produce overestimated distances.
-COATi overestimates only about 11% of benchmarks and is better than the other methods.
-
- 1.
-Comment: *It would also be nice to summarize Ka/Ks and evolutionary distances for empirical human-gorilla aligiments.
+    *It would also be nice to summarize Ka/Ks and evolutionary distances for empirical human-gorilla alignments.
 It strikes me that they already have the data, having used it to get parameters for simulation.*
 
-    TODO...
+    We would like to thank the reviewer for the suggestion to analyze the empirical human-gorilla alignments.
+Using this suggestion, we have expanded our comparison of COATi and the other aligners.
+We now use both empirical human-gorilla sequence pairs and semi-empirical benchmarks.
+From empirical data, we can confirm that aligners that overalign sequences (fewer gaps, more mismatches) do produce higher K2P distances (Table 1 and Figure 4).
+For the benchmarks, aligners that had higher alignment errors had higher root-mean-squared errors for K2P distances and lower F1 scores for selection as well.
 
@@
 
-    TODO...
+    We have adjusted our tables to be less distracting.
 

### 2024-02-09 alignpair_letter.tex
Revision 1 - Release Candidate 1

@@
 Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics.
-COATi implements the Muse and Gaut (\citeyear{muse_gaut_1994}) codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
+COATi implements both the \cite{muse_gaut_1994} codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
 It also lets the user provide a codon substitution matrix.
@@
 This is because it is missing a parameter to control how often Ns are added to the output sequence.
-This is design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol.
+This is a design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol.
 In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment.
@@
 This allows COATi to align lower-quality sequences that may contain artifacts against a high-quality reference sequence.
-We may refer to the output sequence as the ``descendant'', ``non-reference'', or ``query'' sequence.
+We may refer to the output sequence as the ``descendant'' or ``non-reference'' sequence.
 The choice of which sequence is the input sequence and which sequence is the output sequence is left up to the user.
@@
 represent the marginal transition probabilities, where $p \in \{1, 2, 3\}$ is the position of the descendant nucleotide relative to the ancestral reading frame and $I()$ is an indicator function.
-COATi contains marginal models for both Muse and Gaut (\citeyear{muse_gaut_1994}) or the Empirical Codon Model, resulting in the marginal models codon-marginal-mg and codon-marginal-ecm.
+COATi contains marginal models for both \cite{muse_gaut_1994} and the Empirical Codon Model \citep{kosiol_ECM_2007}, resulting in the marginal models codon-marginal-mg and codon-marginal-ecm.
 These models emphasize the position in a codon where the substitution occurs, help restrict the effects of low-quality data in the descendant sequence, and allow more than one substitution per codon.
@@
 
-In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE, (4) the DNA model of MAFFT, and (5) the codon model of PRANK \citep{clustal_omega_sievers_2011,ranwez_macse_2018,katoh2013mafft,prank_loytynoja_2014}.
+In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ \citep{clustal_omega_sievers_2011} via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE \citep{ranwez_macse_2018}, (4) the DNA model of MAFFT \citep{katoh2013mafft}, and (5) the codon model of PRANK \citep{prank_loytynoja_2014}.
 COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
-In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATI-rev).
+In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATi-rev).
 Together, these six methods allow us to evaluate both different alignment strategies and different software implementations.
@@
 
-The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable  for human-gorilla gene pairs, with means 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a).
+The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a).
 Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%.
 These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods.
+As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a).
 The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
@@
 
-In additional to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings.
+In addition to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings.
 Different data sets may show different results.
 However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.
+This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.
 

### 2024-04-16 supplementary_materials/supplementary_materials.Rmd
Add p-dist to supplemental material

@@
 
-```{r}
-
-summary_table = function(results_file, coati_model) {
-    # read results_summary data
-    results = read.csv(results_file)
-
-    # extract dseq values for each aligner
-    coati = results |> filter(aligner == coati_model) |> select(dseq)
-    clustal = results |> filter(aligner == "clustalo") |> select(dseq)
-    macse = results |> filter(aligner == "macse") |> select(dseq)
-    mafft = results |> filter(aligner == "mafft") |> select(dseq)
-    prank = results |> filter(aligner == "prank") |> select(dseq)
-
-    stats_table = data.frame(dseq = c(mean(coati$dseq, na.rm = TRUE),
-                          mean(mafft$dseq, na.rm = TRUE),
-                          mean(prank$dseq, na.rm = TRUE),
-                          mean(macse$dseq, na.rm = TRUE),
-                          mean(clustal$dseq, na.rm = TRUE)), row.names = c(coati_model, "mafft", "prank", "macse", "clustalo"))
-
-    ###########################################################################
-    # perfect, best, and imperfect alignments
-    source("scripts/number_alignments.R")
-    alns = num_alns(results_file)
-
-    # fix row and column names and add to stats table
-    colnames(alns) = c("Perfect alns", "Best alns", "Imperfect alns")
-    alns = alns[c(coati_model, "mafft", "prank", "macse", "clustalo"), ]
-    stats_table = cbind(stats_table, alns)
-
-    ############################################################################
-    # selection (ka ks)
-    source("scripts/kaks.R")
-
-    aligners = c(coati_model, "mafft", "prank", "macse", "clustalo")
-    selection = data.frame("F1.pos" = double(),
-                           "F1.neg" = double())
-
-    # calculate F1 score for positive and negative selection for each aligner
-    for(model in aligners) {
-        model_rows = results[which(results$aligner == model), ]
-        pos = ps_accuracy(model_rows$ref_omega, model_rows$aln_omega)
-        neg = ns_accuracy(model_rows$ref_omega, model_rows$aln_omega)
-
-        selection[nrow(selection) + 1, ] = c(pos, neg)
-    }
-
-    # fix row and column names and add to stats table
-    colnames(selection) = c("F1 pos selection", "F1 neg selection")
-    rownames(selection) = c("coati", "mafft", "prank", "macse", "clustalo")
-    stats_table = cbind(stats_table, selection)
-    if(coati_model == "tri-mg") {
-        coati_model = "Triplet-MG94"
-    } else if(coati_model == "tri-ecm") {
-        coati_model = "Triplet-ECM"
-    } else if(coati_model == "mar-mg") {
-        coati_model = "Marginal-MG94"
-    } else if(coati_model == "mar-ecm") {
-        coati_model = "Marginal-ECM"
-    }
-    rownames(stats_table) = c(coati_model, "MAFFT", "PRANK*", "MACSE", "ClustalOmega")
-
-    # display stats table
-    stats_table
-}
-
-wilcoxon_test = function(results_file, coati_model) {
-    # read results_summary data
-    results = read.csv(results_file)
-
-    # extract dseq values for each aligner
-    coati = results$dseq[which(results$aligner == coati_model)]
-    clustalomega = results$dseq[which(results$aligner == "clustalo")]
-    macse = results$dseq[which(results$aligner == "macse")]
-    mafft = results$dseq[which(results$aligner == "mafft")]
-    prank = results$dseq[which(results$aligner == "prank")]
-
-    # one-side paired wilcoxon rank test
-    wtp = wilcox.test(x = coati, y = prank, paired = TRUE, alternative = "less")
-    wtmf = wilcox.test(x = coati, y = mafft, paired = TRUE, alternative = "less")
-    wtc = wilcox.test(x = coati, y = clustalomega, paired = TRUE, alternative = "less")
-    wtmc = wilcox.test(x = coati, y = macse, paired = TRUE, alternative = "less")
-    w = c(wtp$p.value, wtmf$p.value, wtc$p.value, wtmc$p.value)
-    w
-}
-```
-
 # Aligner Commands
 
-We evaluated five different aligners.
-Below are the commands that we used to run them.
-We have abbreviated the commands for clarity, stripping out unimportant arguments.
-Complete workflows can be found in our coati-testing repository on Github.
+We evaluated five different aligners. Below are the commands that we used to run
+them. We have abbreviated the commands for clarity, stripping out unimportant
+arguments. Complete workflows can be found in our coati-testing repository on
+Github.
 
-- COATi: `coati alignpair -m tri-mg ...`; We used COATi two different ways.
-The primary way used human as the reference sequence.
-The secondary way, COATi-rev, used gorilla as the reference, via a wrapper script that reversed the order of sequences before alignment and reversed the order back afterwards.
-- ClustalΩ v1.2.4: `clustalo --seqtype=Protein ...`; We wrapped ClustalΩ with a script that translates DNA sequences into amino acid sequences (including stops) before alignment.
-Any codons that were partial or containing ambiguous characters were translated as "X".
-The script then aligned these translated sequences with ClustalΩ, and then created a DNA alignment that was consistent with the amino-acid alignment.
-- MACSE v2.06: `java -jar macse.jar -prog alignSequences -seq human.fasta -seq_lr gorilla.fasta ...`; We wrapped MACSE with a script that created two temporary fasta files, one containing the human sequence and another containing the gorilla sequence.
-The human sequence was specified as the reliable sequence and the gorilla sequence was specified as the less-reliable sequence.
-MACSE uses "!" to mark gaps that result from frameshifts, and the wrapper script replaced these with "-".
-Additionally, MACSE sometimes produced columns that only contained gaps, and these columns were removed.
+- COATi: `coati alignpair -m tri-mg ...`; We used COATi two different ways. The
+  primary way used human as the reference sequence. The secondary way,
+  COATi-rev, used gorilla as the reference, via a wrapper script that reversed
+  the order of sequences before alignment and reversed the order back
+  afterwards.
+- ClustalΩ v1.2.4: `clustalo --seqtype=Protein ...`; We wrapped ClustalΩ with a
+  script that translates DNA sequences into amino acid sequences (including
+  stops) before alignment. Any codons that were partial or containing ambiguous
+  characters were translated as "X". The script then aligned these translated
+  sequences with ClustalΩ, and then created a DNA alignment that was consistent
+  with the amino-acid alignment.
+- MACSE v2.06: `java -jar macse.jar -prog alignSequences -seq human.fasta
+  -seq_lr gorilla.fasta ...`; We wrapped MACSE with a script that created two
+  temporary fasta files, one containing the human sequence and another
+  containing the gorilla sequence. The human sequence was specified as the
+  reliable sequence and the gorilla sequence was specified as the less-reliable
+  sequence. MACSE uses "!" to mark gaps that result from frameshifts, and the
+  wrapper script replaced these with "-". Additionally, MACSE sometimes produced
+  columns that only contained gaps, and these columns were removed.
 - PRANK v.150803: `prank -codon ...`
@@
 
-COATi can use different alignment models for pairwise alignment. Below are the commands that we used to run different models.
+COATi can use different alignment models for pairwise alignment. Below are the
+commands that we used to run different models.
 
@@
 
-When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST library to generate best alignments by composing the input and output sequences with the COATi FST model.
-While the COATi FST model is too large to display, we can show the result of a composition.
+When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST
+library to generate best alignments by composing the input and output sequences
+with the COATi FST model. While the COATi FST model is too large to display, we
+can show the result of a composition.
 
-Fig.
-\@ref(fig:aln-example-a) shows a graph depicting the FST that results from composing the COATi FST with the input sequence "CTC" and the output sequence "CTG".
-Every path through this FST represents one possible way to align "CTC" and "CTG", and the sum of all weights along a path is the total weight of the respective alignment.
-Here, the weights of each arc are in negative-log space.
-Note that this graph has been optimized, and weight has been pushed towards the initial state.
-The weight of any specific arc may not be directly mapable to a weight described in the model.
+Fig. \@ref(fig:aln-example-a) shows a graph depicting the FST that results from
+composing the COATi FST with the input sequence "CTC" and the output sequence
+"CTG". Every path through this FST represents one possible way to align "CTC"
+and "CTG", and the sum of all weights along a path is the total weight of the
+respective alignment. Here, the weights of each arc are in negative-log space.
+Note that this graph has been optimized, and weight has been pushed towards the
+initial state. The weight of any specific arc may not be directly mapable to a
+weight described in the model.
 
-Fig.
-\@ref(fig:aln-example-b) is the best alignment between "CTC" and "CTG", as determined by the shortest path algorithm.
-Figs.
-\@ref(fig:aln-example-a) and \@ref(fig:aln-example-b) were produced by the OpenFST library.
-A bold circle represents a starting node, and a double circle represents a termination node.
+Fig. \@ref(fig:aln-example-b) is the best alignment between "CTC" and "CTG", as
+determined by the shortest path algorithm.  Figs. \@ref(fig:aln-example-a) and
+\@ref(fig:aln-example-b) were produced by the OpenFST library. A bold circle
+represents a starting node, and a double circle represents a termination node.
 
@@
 
-We quantified the lengths of gaps produced by each method across the empirical dataset of human-gorilla protein-coding-sequence pairs (Tab.
-\@ref(tab:gap-table-1)).
-Here the gap type is either "D" for gaps introduced into the gorilla sequence or "I" for gaps introduced in the human sequence.
-"COATi" refers to the full COATi FST model with a MG substitution model (i.e.
-tri-mg).
-"COATi-rev" refers to the same model, but using gorilla as the reference and human as the non-reference sequence.
-Gap lengths of 1--6 nucleotides were binned into their respective columns.
-Gap lengths longer than 6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.
+We quantified the lengths of gaps produced by each method across the empirical
+dataset of human-gorilla protein-coding-sequence pairs (Tab. \@ref
+(tab:gap-table-1)). Here the gap type is either "D" for gaps introduced into
+the gorilla sequence or "I" for gaps introduced in the human sequence. "COATi"
+refers to the full COATi FST model with a MG substitution model
+(i.e. tri-mg). "COATi-rev" refers to the same model, but using gorilla as the
+reference and human as the non-reference sequence. Gap lengths of 1--6
+nucleotides were binned into their respective columns. Gap lengths longer than
+6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether
+their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.
 
-We also quantified the lengths of gaps produced by different COATi models (Tab. \@ref(tab:gap-table-2)).
+We also quantified the lengths of gaps produced by different COATi models
+(Tab. \@ref(tab:gap-table-2)).
 
-Note that ClustalΩ gaps with lengths that are not a multiple of 3 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.
+Note that ClustalΩ gaps with lengths that are not a multiple of 3 are created by
+how our wrapper script handles DNA sequences with lengths that are not multiple
+of 3.
 
@@
 
-k <- kable(tab, align = "lcrrrrrrrrr", caption="Number of gaps introduced by each method separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
+k <- kable(tab, align = "lcrrrrrrrrr",
+    caption="Number of gaps introduced by each method separated by length and type.",
+    booktabs = TRUE, linesep = c("", "\\addlinespace"))
 
@@
 
-k <- kable(tab, align = "lcrrrrrrrrr", caption="Number of gaps introduced by different COATi models separated by length and type.", booktabs = TRUE, linesep = c("", "\\addlinespace"))
+k <- kable(tab, align = "lcrrrrrrrrr",
+    caption="Number of gaps introduced by different COATi models separated by length and type.",
+    booktabs = TRUE, linesep = c("", "\\addlinespace"))
 
@@
 
-We quantified the phases of gaps produced by different aligners and COATi models (Tab.
-\@ref(tab:gap-table-3) and \@ref(tab:gap-table-4)).
-Phase 1 gaps begin after the 1st position in a codon in the reference sequence, phase 2 gaps begin after the 2nd position in a codon, and phase 3 begin after the 3rd position in a codon (i.e.
-between codons).
-Phase 3 gaps are also known as phase 0 gaps.
+We quantified the phases of gaps produced by different aligners and COATi
+models (Tab. \@ref(tab:gap-table-3) and \@ref(tab:gap-table-4)). Phase 1 gaps
+begin after the 1st position in a codon in the reference sequence, phase 2 gaps
+begin after the 2nd position in a codon, and phase 3 begin after the 3rd
+position in a codon (i.e. between codons). Phase 3 gaps are also known as phase
+0 gaps.
 
-Note that ClustalΩ gaps with phases of 1 and 2 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.
+Note that ClustalΩ gaps with phases of 1 and 2 are created by how our wrapper
+script handles DNA sequences with lengths that are not multiple of 3.
 
@@
 
-k <- kable(tab, align = "lrrr", caption="Number of gaps introduced by each alignment method separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab, align = "lrrr",
+    caption="Number of gaps introduced by each alignment method separated by phase.",
+    booktabs = TRUE, linesep = c("\\addlinespace"))
 
@@
 
-k <- kable(tab, align = "lrrr", caption="Number of gaps introduced by different COATi models separated by phase.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab, align = "lrrr",
+    caption="Number of gaps introduced by different COATi models separated by phase.",
+    booktabs = TRUE, linesep = c("\\addlinespace"))
 
@@
 
-We quantified the homology patterns of residues for different aligners and COATi models (Tab.
-\@ref(tab:gap-table-5), Tab.
-\@ref(tab:gap-table-6), and Fig.
-\@ref(fig:gap-figure-1)).
-Here we define the match, mismatch, and gap percentages as the percent of nucleotides aligned against a match, mismatch, and gap respectively.
-Note that this is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.
+We quantified the homology patterns of residues for different aligners and COATi
+models (Tab. \@ref(tab:gap-table-5), Tab. \@ref(tab:gap-table-6), and
+Fig. \@ref(fig:gap-figure-1)). Here we define the match, mismatch, and gap
+percentages as the percent of nucleotides aligned against a match, mismatch,
+and gap respectively. Note that this is different than the percent of columns
+that contain a match, mismatch, or gap because match and mismatch columns are
+counted twice.
 
@@
 
-k <- kable(tab, align = "rccc", caption="Average homology percentages of alignments separated by alignment method.", booktabs = TRUE, linesep = c("\\addlinespace"))
+k <- kable(tab, align = "rccc",
+    caption="Average homology percentages of alignments separated by alignment method.",
+    booktabs = TRUE, linesep = c("\\addlinespace"))
 
@@
 gg
+```
 
+\newpage
+
+## Sequence Distances
+
+We quantified the raw sequence distances (p-distance) inferred by different
+aligners and COATi models (Fig. \@ref(fig:p-figure-1),
+Tab. \@ref(tab:p-table-1), and Tab. \@ref(tab:p-table-2)).
+
+```{r p-figure-1, fig.cap="COATi produced shorter sequence distances than other aligners.
+Each panel is a scatter plot where the x coordinate is the p-distance from a COATi alignment and y coordinate is the p-distance from the corresponding alignment from another aligner.", warning = FALSE}
+p_data <- ALN_DATA |>
+    filter(method %in% c(c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg"))) |>
+    pivot_wider(id_cols=gene, names_from=method, values_from=pdist) |>
+    rename(coati = `coati-tri-mg`, `coati-rev` = `rev-coati-tri-mg`) |>
+    relocate(gene, coati)
+
+p_data1 <- p_data |> pivot_longer(cols = clustalo:`coati-rev`,
+    names_to="method", values_to="other") |>
+    mutate(method = fct_relevel(method, "clustalo", "macse", "mafft",
+        "prank", "coati-rev")) |>
+    mutate(method = fct_recode(method,
+        "vs ClustalΩ" = "clustalo",
+        "vs MACSE" = "macse",
+        "vs MAFFT" = "mafft",
+        "vs PRANK" = "prank",
+        "vs COATi-rev" = "coati-rev"
+    ))
+
+pal <- cbbPalette
+names(pal) <- str_c("vs ", names(pal))
+
+gg <- ggplot(p_data1, aes(x = coati, y = other, color = method)) + facet_wrap(vars(method))
+gg <- gg + geom_point(alpha = 0.2, show.legend = FALSE)
+gg <- gg + xlab("COATi") + ylab("Other")
+gg <- gg + scale_color_manual(values = pal)
+gg <- gg + scale_x_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1))
+gg <- gg + scale_y_continuous(trans = scales::transform_pseudo_log(0.0004, base=10),
+    breaks=c(0, 0.001, 0.01, 0.1, 1))
+
+gg <- gg + theme_grey(9) + theme(
+    panel.grid.minor.x = element_blank(),
+    panel.grid.minor.y = element_blank()
+)
+
+gg
+```
+
+```{r p-table-1, results = 'asis', message = FALSE}
+tab <- ALN_DATA |> summarize(m = mean(pdist, na.rm = TRUE), .by = "method") |>
+    rename(Method = method)
+
+tab_data1 <- tab |>
+    filter(Method %in% c("clustalo", "coati-tri-mg",
+    "macse", "mafft", "prank", "rev-coati-tri-mg")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg", "rev-coati-tri-mg"))) |>
+    mutate(Method = case_match(Method,
+        "clustalo" ~ "ClustalΩ",
+        "coati-tri-mg" ~ "COATi",
+        "macse" ~ "MACSE",
+        "mafft" ~ "MAFFT",
+        "prank" ~ "PRANK",
+        "rev-coati-tri-mg" ~ "COATi-rev"
+        )) |>
+    arrange(Method) |>
+    pivot_wider(names_from = "Method", values_from = "m")
+
+k <- kable(tab_data1, digits = 4, align = "ccccc",
+  caption="Average p-distance of alignments for each method.",
+  booktabs = TRUE, linesep = c("\\addlinespace"),
+  format.args = list(nsmall = 4))
+
+k |> kable_styling(latex_options = "HOLD_position")
+```
+
+```{r p-table-2, results = 'asis', message = FALSE}
+tab_data2 <- tab |> filter(Method %in% c("coati-tri-mg",
+    "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna")) |>
+    mutate(Method = fct_relevel(Method, c("coati-tri-mg",
+        "coati-tri-ecm", "coati-mar-mg", "coati-mar-ecm", "coati-dna"))) |>
+    arrange(Method) |>
+    rename(Model = Method) |>
+    mutate(Model = fct_recode(Model,
+        "TRI-MG" = "coati-tri-mg",
+        "TRI-ECM" = "coati-tri-ecm",
+        "MAR-MG" = "coati-mar-mg",
+        "MAR-ECM" = "coati-mar-ecm",
+        "DNA" = "coati-dna"
+    ))  |>
+    pivot_wider(names_from = "Model", values_from = "m")
+
+k <- kable(tab_data2, digits = 4, align = "ccccc",
+    caption="Average p-distance of alignments for each COATi model.",
+    booktabs = TRUE, linesep = c("\\addlinespace"),
+    format.args = list(nsmall = 4))
+
+k |> kable_styling(latex_options = "HOLD_position")
 ```
@@
 
-We quantified the evolutionary distances inferred by different aligners and COATi models (Fig. \@ref(fig:k2p-figure-1), Tab. \@ref(tab:k2p-table-1), and Tab. \@ref(tab:k2p-table-2)).
+We quantified the evolutionary distances inferred by different aligners and
+COATi models (Fig. \@ref(fig:k2p-figure-1), Tab. \@ref(tab:k2p-table-1), and
+Tab. \@ref(tab:k2p-table-2)).
 
@@
 
-k <- kable(tab_data1, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each method.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))
+k <- kable(tab_data1, digits = 4, align = "ccccc",
+  caption="Average K2P-distance of alignments for each method.",
+  booktabs = TRUE, linesep = c("\\addlinespace"),
+  format.args = list(nsmall = 4))
 
 k |> kable_styling(latex_options = "HOLD_position")
-
 ```
@@
 
-k <- kable(tab_data2, digits = 4, align = "ccccc", caption="Average K2P distances of alignments for each COATi model.", booktabs = TRUE, linesep = c("\\addlinespace"), format.args = list(nsmall = 4))
+k <- kable(tab_data2, digits = 4, align = "ccccc",
+    caption="Average K2P-distance of alignments for each COATi model.",
+    booktabs = TRUE, linesep = c("\\addlinespace"),
+    format.args = list(nsmall = 4))
 
@@
 
-For each sequence pair in the benchmark, we calculated the alignment distance (d~seq~) between the benchmark alignment and the alignments generated by COATi, ClustalΩ, MACSE, MAFFT, and PRANK.
-We also calculated distances between all pairs of aligners.
-Our benchmark contained gap patterns extracted from alignments generated by different aligners.
-Figure \@ref(fig:bench-pcoa-split) contains the results of a metric multidimensional scaling (principle coordinate analysis; PCoA) of the matrix of average distances between aligners, separated by what type of gap patterns was used in the benchmark alignment.
+For each sequence pair in the benchmark, we calculated the alignment distance
+(d~seq~) between the benchmark alignment and the alignments generated by COATi,
+ClustalΩ, MACSE, MAFFT, and PRANK. We also calculated distances between all
+pairs of aligners. Our benchmark contained gap patterns extracted from
+alignments generated by different aligners. Figure
+\@ref(fig:bench-pcoa-split) contains the results of a metric multidimensional
+scaling (principle coordinate analysis; PCoA) of the matrix of average
+distances between aligners, separated by what type of gap patterns was used in
+the benchmark alignment.
 
-Figure \@ref(fig:bench-pcoa-all) contains a principle coordinate analysis of each aligner, including different COATi models, across the entire benchmark dataset.
-COATi's different models produced similar alignments and cluster together along with the benchmarks.
+Figure \@ref(fig:bench-pcoa-all) contains a principle coordinate analysis of
+each aligner, including different COATi models, across the entire benchmark
+dataset. COATi's different models produced similar alignments and cluster
+together along with the benchmarks.
 

### 2024-04-16 supplementary_materials/supplementary_materials.tex
Revision 1 - Release Candidate 1
Add p-dist to supplemental material

@@
 }
-\hypertarget{aligner-commands}{%
-\section{Aligner Commands}\label{aligner-commands}}
+\section{Aligner Commands}\label{aligner-commands}
 
-We evaluated five different aligners.
-Below are the commands that we used to run them.
-We have abbreviated the commands for clarity, stripping out unimportant arguments.
-Complete workflows can be found in our coati-testing repository on Github.
+We evaluated five different aligners. Below are the commands that we used to run
+them. We have abbreviated the commands for clarity, stripping out unimportant
+arguments. Complete workflows can be found in our coati-testing repository on
+Github.
 
@@
 \item
-  COATi: \texttt{coati\ alignpair\ -m\ tri-mg\ ...}; We used COATi two different ways.
-The primary way used human as the reference sequence.
-The secondary way, COATi-rev, used gorilla as the reference, via a wrapper script that reversed the order of sequences before alignment and reversed the order back afterwards.
+  COATi: \texttt{coati\ alignpair\ -m\ tri-mg\ ...}; We used COATi two different ways. The
+  primary way used human as the reference sequence. The secondary way,
+  COATi-rev, used gorilla as the reference, via a wrapper script that reversed
+  the order of sequences before alignment and reversed the order back
+  afterwards.
 \item
-  ClustalΩ v1.2.4: \texttt{clustalo\ -\/-seqtype=Protein\ ...}; We wrapped ClustalΩ with a script that translates DNA sequences into amino acid sequences (including stops) before alignment.
-Any codons that were partial or containing ambiguous characters were translated as ``X''.
-The script then aligned these translated sequences with ClustalΩ, and then created a DNA alignment that was consistent with the amino-acid alignment.
+  ClustalΩ v1.2.4: \texttt{clustalo\ -\/-seqtype=Protein\ ...}; We wrapped ClustalΩ with a
+  script that translates DNA sequences into amino acid sequences (including
+  stops) before alignment. Any codons that were partial or containing ambiguous
+  characters were translated as ``X''. The script then aligned these translated
+  sequences with ClustalΩ, and then created a DNA alignment that was consistent
+  with the amino-acid alignment.
 \item
-  MACSE v2.06: \texttt{java\ -jar\ macse.jar\ -prog\ alignSequences\ -seq\ human.fasta\ -seq\_lr\ gorilla.fasta\ ...}; We wrapped MACSE with a script that created two temporary fasta files, one containing the human sequence and another containing the gorilla sequence.
-The human sequence was specified as the reliable sequence and the gorilla sequence was specified as the less-reliable sequence.
-MACSE uses ``!'' to mark gaps that result from frameshifts, and the wrapper script replaced these with ``-''.
-Additionally, MACSE sometimes produced columns that only contained gaps, and these columns were removed.
+  MACSE v2.06: \texttt{java\ -jar\ macse.jar\ -prog\ alignSequences\ -seq\ human.fasta\ \ \ -seq\_lr\ gorilla.fasta\ ...}; We wrapped MACSE with a script that created two
+  temporary fasta files, one containing the human sequence and another
+  containing the gorilla sequence. The human sequence was specified as the
+  reliable sequence and the gorilla sequence was specified as the less-reliable
+  sequence. MACSE uses ``!'' to mark gaps that result from frameshifts, and the
+  wrapper script replaced these with ``-''. Additionally, MACSE sometimes produced
+  columns that only contained gaps, and these columns were removed.
 \item
@@
 
-\hypertarget{fst-alignment-example}{%
-\section{FST Alignment Example}\label{fst-alignment-example}}
+\section{FST Alignment Example}\label{fst-alignment-example}
 
-When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST library to generate best alignments by composing the input and output sequences with the COATi FST model.
-While the COATi FST model is too large to display, we can show the result of a composition.
+When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST
+library to generate best alignments by composing the input and output sequences
+with the COATi FST model. While the COATi FST model is too large to display, we
+can show the result of a composition.
 
-Fig.
-\ref{fig:aln-example-a} shows a graph depicting the FST that results from composing the COATi FST with the input sequence ``CTC'' and the output sequence ``CTG''.
-Every path through this FST represents one possible way to align ``CTC'' and ``CTG'', and the sum of all weights along a path is the total weight of the respective alignment.
-Here, the weights of each arc are in negative-log space.
-Note that this graph has been optimized, and weight has been pushed towards the initial state.
-The weight of any specific arc may not be directly mapable to a weight described in the model.
+Fig. \ref{fig:aln-example-a} shows a graph depicting the FST that results from
+composing the COATi FST with the input sequence ``CTC'' and the output sequence
+``CTG''. Every path through this FST represents one possible way to align ``CTC''
+and ``CTG'', and the sum of all weights along a path is the total weight of the
+respective alignment. Here, the weights of each arc are in negative-log space.
+Note that this graph has been optimized, and weight has been pushed towards the
+initial state. The weight of any specific arc may not be directly mapable to a
+weight described in the model.
 
-Fig.
-\ref{fig:aln-example-b} is the best alignment between ``CTC'' and ``CTG'', as determined by the shortest path algorithm.
-Figs.
-\ref{fig:aln-example-a} and \ref{fig:aln-example-b} were produced by the OpenFST library.
-A bold circle represents a starting node, and a double circle represents a termination node.
+Fig. \ref{fig:aln-example-b} is the best alignment between ``CTC'' and ``CTG'', as
+determined by the shortest path algorithm. Figs. \ref{fig:aln-example-a} and
+\ref{fig:aln-example-b} were produced by the OpenFST library. A bold circle
+represents a starting node, and a double circle represents a termination node.
 
@@
 
-\hypertarget{empirical-results}{%
-\section{Empirical Results}\label{empirical-results}}
+\section{Empirical Results}\label{empirical-results}
 
-\hypertarget{gap-patterns}{%
-\subsection{Gap Patterns}\label{gap-patterns}}
+\subsection{Gap Patterns}\label{gap-patterns}
 
-\hypertarget{lengths}{%
-\subsubsection{Lengths}\label{lengths}}
+\subsubsection{Lengths}\label{lengths}
 
-We quantified the lengths of gaps produced by each method across the empirical dataset of human-gorilla protein-coding-sequence pairs (Tab.
-\ref{tab:gap-table-1}).
-Here the gap type is either ``D'' for gaps introduced into the gorilla sequence or ``I'' for gaps introduced in the human sequence.
-``COATi'' refers to the full COATi FST model with a MG substitution model (i.e.~tri-mg).
-``COATi-rev'' refers to the same model, but using gorilla as the reference and human as the non-reference sequence.
-Gap lengths of 1--6 nucleotides were binned into their respective columns.
-Gap lengths longer than 6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.
+We quantified the lengths of gaps produced by each method across the empirical
+dataset of human-gorilla protein-coding-sequence pairs (Tab. @ref
+(tab:gap-table-1)). Here the gap type is either ``D'' for gaps introduced into
+the gorilla sequence or ``I'' for gaps introduced in the human sequence. ``COATi''
+refers to the full COATi FST model with a MG substitution model
+(i.e.~tri-mg). ``COATi-rev'' refers to the same model, but using gorilla as the
+reference and human as the non-reference sequence. Gap lengths of 1--6
+nucleotides were binned into their respective columns. Gap lengths longer than
+6 nucleotides were binned into columns 7+, 8+, and 9+ depending on whether
+their lengths were 1, 2, or 3 nucleotides longer than a multiple of three.
 
-We also quantified the lengths of gaps produced by different COATi models (Tab. \ref{tab:gap-table-2}).
+We also quantified the lengths of gaps produced by different COATi models
+(Tab. \ref{tab:gap-table-2}).
 
-Note that ClustalΩ gaps with lengths that are not a multiple of 3 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.
+Note that ClustalΩ gaps with lengths that are not a multiple of 3 are created by
+how our wrapper script handles DNA sequences with lengths that are not multiple
+of 3.
 
 \begin{table}[H]
-
+\centering
 \caption{\label{tab:gap-table-1}Number of gaps introduced by each method separated by length and type.}
@@
 \begin{table}[H]
-
+\centering
 \caption{\label{tab:gap-table-2}Number of gaps introduced by different COATi models separated by length and type.}
@@
 
-\hypertarget{phases}{%
-\subsubsection{Phases}\label{phases}}
+\subsubsection{Phases}\label{phases}
 
-We quantified the phases of gaps produced by different aligners and COATi models (Tab.
-\ref{tab:gap-table-3} and \ref{tab:gap-table-4}).
-Phase 1 gaps begin after the 1st position in a codon in the reference sequence, phase 2 gaps begin after the 2nd position in a codon, and phase 3 begin after the 3rd position in a codon (i.e.~between codons).
-Phase 3 gaps are also known as phase 0 gaps.
+We quantified the phases of gaps produced by different aligners and COATi
+models (Tab. \ref{tab:gap-table-3} and \ref{tab:gap-table-4}). Phase 1 gaps
+begin after the 1st position in a codon in the reference sequence, phase 2 gaps
+begin after the 2nd position in a codon, and phase 3 begin after the 3rd
+position in a codon (i.e.~between codons). Phase 3 gaps are also known as phase
+0 gaps.
 
-Note that ClustalΩ gaps with phases of 1 and 2 are created by how our wrapper script handles DNA sequences with lengths that are not multiple of 3.
+Note that ClustalΩ gaps with phases of 1 and 2 are created by how our wrapper
+script handles DNA sequences with lengths that are not multiple of 3.
 
 \begin{table}[H]
-
+\centering
 \caption{\label{tab:gap-table-3}Number of gaps introduced by each alignment method separated by phase.}
@@
 \begin{table}[H]
-
+\centering
 \caption{\label{tab:gap-table-4}Number of gaps introduced by different COATi models separated by phase.}
@@
 
-\hypertarget{homology-patterns}{%
-\subsection{Homology Patterns}\label{homology-patterns}}
+\subsection{Homology Patterns}\label{homology-patterns}
 
-We quantified the homology patterns of residues for different aligners and COATi models (Tab.
-\ref{tab:gap-table-5}, Tab.
-\ref{tab:gap-table-6}, and Fig.
-\ref{fig:gap-figure-1}).
-Here we define the match, mismatch, and gap percentages as the percent of nucleotides aligned against a match, mismatch, and gap respectively.
-Note that this is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.
+We quantified the homology patterns of residues for different aligners and COATi
+models (Tab. \ref{tab:gap-table-5}, Tab. \ref{tab:gap-table-6}, and
+Fig. \ref{fig:gap-figure-1}). Here we define the match, mismatch, and gap
+percentages as the percent of nucleotides aligned against a match, mismatch,
+and gap respectively. Note that this is different than the percent of columns
+that contain a match, mismatch, or gap because match and mismatch columns are
+counted twice.
 
 \begin{table}[H]
-
+\centering
 \caption{\label{tab:gap-table-5}Average homology percentages of alignments separated by alignment method.}
@@
 \begin{table}[H]
-
+\centering
 \caption{\label{tab:gap-table-6}Average homology percentages of alignments separated by COATi model.}
@@
 
-\hypertarget{evolutionary-distances}{%
-\subsection{Evolutionary Distances}\label{evolutionary-distances}}
+\subsection{Sequence Distances}\label{sequence-distances}
 
-We quantified the evolutionary distances inferred by different aligners and COATi models (Fig. \ref{fig:k2p-figure-1}, Tab. \ref{tab:k2p-table-1}, and Tab. \ref{tab:k2p-table-2}).
+We quantified the raw sequence distances (p-distance) inferred by different
+aligners and COATi models (Fig. \ref{fig:p-figure-1},
+Tab. \ref{tab:p-table-1}, and Tab. \ref{tab:p-table-2}).
+
+\begin{figure}
+\centering
+\includegraphics{supplementary_materials_files/figure-latex/p-figure-1-1.pdf}
+\caption{\label{fig:p-figure-1}COATi produced shorter sequence distances than other aligners.
+Each panel is a scatter plot where the x coordinate is the p-distance from a COATi alignment and y coordinate is the p-distance from the corresponding alignment from another aligner.}
+\end{figure}
+
+\begin{table}[H]
+\centering
+\caption{\label{tab:p-table-1}Average p-distance of alignments for each method.}
+\centering
+\begin{tabular}[t]{cccccc}
+\toprule
+COATi & COATi-rev & ClustalΩ & MACSE & MAFFT & PRANK\\
+\midrule
+0.0083 & 0.0083 & 0.0168 & 0.0142 & 0.0147 & 0.0089\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\centering
+\caption{\label{tab:p-table-2}Average p-distance of alignments for each COATi model.}
+\centering
+\begin{tabular}[t]{ccccc}
+\toprule
+TRI-MG & TRI-ECM & MAR-MG & MAR-ECM & DNA\\
+\midrule
+0.0083 & 0.0083 & 0.0083 & 0.0083 & 0.0083\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\newpage
+
+\subsection{Evolutionary Distances}\label{evolutionary-distances}
+
+We quantified the evolutionary distances inferred by different aligners and
+COATi models (Fig. \ref{fig:k2p-figure-1}, Tab. \ref{tab:k2p-table-1}, and
+Tab. \ref{tab:k2p-table-2}).
 
@@
 \begin{table}[H]
-
-\caption{\label{tab:k2p-table-1}Average K2P distances of alignments for each method.}
+\centering
+\caption{\label{tab:k2p-table-1}Average K2P-distance of alignments for each method.}
 \centering
@@
 \begin{table}[H]
-
-\caption{\label{tab:k2p-table-2}Average K2P distances of alignments for each COATi model.}
+\centering
+\caption{\label{tab:k2p-table-2}Average K2P-distance of alignments for each COATi model.}
 \centering
@@
 
-\hypertarget{benchmark-results}{%
-\section{Benchmark Results}\label{benchmark-results}}
+\section{Benchmark Results}\label{benchmark-results}
 
-\hypertarget{alignment-distances}{%
-\subsection{Alignment Distances}\label{alignment-distances}}
+\subsection{Alignment Distances}\label{alignment-distances}
 
-For each sequence pair in the benchmark, we calculated the alignment distance (d\textsubscript{seq}) between the benchmark alignment and the alignments generated by COATi, ClustalΩ, MACSE, MAFFT, and PRANK.
-We also calculated distances between all pairs of aligners.
-Our benchmark contained gap patterns extracted from alignments generated by different aligners.
-Figure \ref{fig:bench-pcoa-split} contains the results of a metric multidimensional scaling (principle coordinate analysis; PCoA) of the matrix of average distances between aligners, separated by what type of gap patterns was used in the benchmark alignment.
+For each sequence pair in the benchmark, we calculated the alignment distance
+(d\textsubscript{seq}) between the benchmark alignment and the alignments generated by COATi,
+ClustalΩ, MACSE, MAFFT, and PRANK. We also calculated distances between all
+pairs of aligners. Our benchmark contained gap patterns extracted from
+alignments generated by different aligners. Figure
+\ref{fig:bench-pcoa-split} contains the results of a metric multidimensional
+scaling (principle coordinate analysis; PCoA) of the matrix of average
+distances between aligners, separated by what type of gap patterns was used in
+the benchmark alignment.
+
+Figure \ref{fig:bench-pcoa-all} contains a principle coordinate analysis of
+each aligner, including different COATi models, across the entire benchmark
+dataset. COATi's different models produced similar alignments and cluster
+together along with the benchmarks.
 
@@
 
-Figure \ref{fig:bench-pcoa-all} contains a principle coordinate analysis of each aligner, including different COATi models, across the entire benchmark dataset.
-COATi's different models produced similar alignments and cluster together along with the benchmarks.
-
 \begin{figure}
@@
 \includegraphics{supplementary_materials_files/figure-latex/bench-pcoa-all-1.pdf}
-\caption{\label{fig:bench-pcoa-all}Different COATi produce accurate alignments that are also similar to one another.}
+\caption{\label{fig:bench-pcoa-all}Different COATi models produce accurate alignments that are also similar to one another.}
 \end{figure}

### 2024-04-17 response_r2.md
Update manuscript

---
title: Response to Reviewers (Second Round)
subtitle: "COATi: statistical pairwise alignment of protein-coding sequences"
author:
	- Juan J. García Mesa
	- Ziqi Zhu
	- Reed A. Cartwright
mainfont: TeX Gyre Pagella
geometry:
	- margin=1in
---

<!-- pandoc --pdf-engine=lualatex -o response_r1.pdf response_r1.md -->

Dear Editor,

We are resubmitting our manuscript, "COATi: statistical pairwise alignment of protein coding sequences", for your consideration. This manuscript describes a new model for aligning protein coding sequences which produces more accurate alignments, while being robust to common artifacts found in genomic annotations. Our approach is implemented in the software package, COATi, which is released via an MIT license and available on GitHub. While COATi can generate multiple-sequence alignments, this paper is focused on our pairwise aligner which is has received more attention and validation. Additionally, to promote open science and reproducibility all data, procedures, and scripts used to validate COATi and compare it with alternative methods have been uploaded to GitHub.

We would like to thank the associate editor and two reviewers for their helpful comments and suggestions. We have addressed all of the concerns brought to our attention by the reviewers as described below.

Sincerely,  
Reed A. Cartwright, on behalf of all authors

## Reviewer 1 Comments

1. Comment: *It is now very clear and easy to read manuscript and a important edition to the alignment literature since it does well relative to the objective it pursues.  It has moved away from statistical alignment formulation in the first submission toward doing well in agreeing with benchmark alignments.  Some discussion of this would be useful.*

    TODO

1. Comment: *And how will the authors turn this into a multiple alignment tool?*

    COATi currently has an early-stage module for estimating pairwise alignments. However, it needs more work (and an additional round of funding) before it meets the high-quality standards exhibited by the pairwise aligner. We have included additional details about COATi MSA in the discussion.

## Review 2 Comments

1. Comment: *This is my second review of the manuscript “COATi: statistical pairwise alignment of protein-coding sequences” by García Mesa et al. and I have to say that it is practically a new manuscript. The first version really excited me, and I felt the science was both sound and very useful to the community, but the paper was difficult to understand. This version is much better; it is well-written, and it confirmed my belief that COATi is an important and useful contribution to the field of sequence alignment.*

    We appreciate your positive words.

1. Comment: *My only somewhat more conceptual comment is about the use of K2P distances. I buy the argument that “… the Kimura’s 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences” (p. 8, line 191) I am a bit mixed about its use. To be honest, I don’t think there are any truly appropriate distances for coding regions. However, p-distances would actually be more intuitive. The authors should consider calculating p-distances and placing them in the supporting information (or even replacing the K2P data). I don’t think this will change conclusions, but it would make things more intuitive for readers. Since I am certain this won’t change conclusions, I don’t want to push it that hard.*

    We calculated p-distances and have included the results in the supplement. The results were equivalent to K2P distances, which we now mention in the manuscript.

2. Comment: *I am not certain why the authors use “phase-3” indels rather than “phase-0”. I find phase-0 more intuitive, and it strikes me that this could be changed easily and would simplify the writing – the authors wouldn’t have to explain that they using phase-3 instead of phase-0!*

    We originally used phase-0 in our terminology. However, in our experience from presenting this research to diverse audiences, phase-3 was easier terminology to explain than phase-0. For example, "a phase-X gap occurs after the X position in a codon" is easier to explain than "a phase-X gap occurs before the X+1 position in a codon". Importantly, by using phase-3, we can succinctly explain the relationship between the frequencies of the gaps as phase-3 > phase-2 > phase-1. We have added additional language to the manuscript to emphasize these points.

3. Comment: *The authors should use “see” rather than c.f. on p. 5, line 82. The abbreviation should be cf., not c.f., and it should be used to mean “compare.” Given the Ranwez et al. references I don’t think the authors of this paper mean “compare.”*

    We do mean "compare" here because COATi's innovation is in similar spirit to MACSE but fundamentally different.

4. Comment: *I would consider adding the fact that their “…COATi results estimate that only 41% indels in protein-coding sequences between humans and gorillas are phase-3 indels” (on p. 15, line 370) to the abstract. This is a nice empirical demonstration that their methods is producing “biologically-reasonable” alignments.*

    Thank you for the suggestion. We have reworked the abstract and included this result.

5. Comment: *I’m a bit confused by the authors’ use of “frameshifts” when discussing the behavior of Clustal Omega (p. 15, line 402). A simple phase-1 or phase-2 indel that is a multiple of 3 is not a frameshift. Are they referring to separate indels that might have an intermediate that has a genuine frameshift? By “genuine frameshift” I mean a case where a presumably inactivating indel mutation that is not a multiple of 3 occurs and persists in a population long enough to have a nearby compensatory mutation that puts the sequence back in frame (i.e., the sum of then indel lengths is a multiple of 3). I think a minor rewrite for clarity would help readers.*

    In this instance, we are referring to an abiological frameshift that produces an abiological amino-acid translation. We have reworked this section for clarify our meaning.

6. Comment: *Overall, I really like this new version of the manuscript!*

    Thank you.

### 2024-04-17 alignpair_letter.tex
Update manuscript

@@
 For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
-While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain these artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
-To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
+More significantly, while indels can occur both within and between codons in natural sequences, most amino-acid- and codon-based aligners assume that indels only occur between codons.
+This mismatch between biology and alignment algorithms produces sub-optimal alignments and errors in downstream analyses.
+
+To address these issues, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
 COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments.
+COATi can infer indels both within and between codons, leading to improved sequences alignments.
+We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with an earlier study in rats and mice.
+We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs via several measures.
+
 \end{abstract}
+
+%While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain these artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
 
@@
 Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}).
-Gaps that occur after the first position or second position of a codon are known as phase-1 and phase-2 gaps, respectively.
-Gaps that occur between codons are known as either phase-3 gaps (this study) or phase-0 gaps \citep[e.g.][]{taylor2004occurrence}.
+Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively.
+Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}.
+Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}.
 While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps.
@@
 
-COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}.
+COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; cf.][]{holmes2001evolutionary}.
 We call this transducer the COATi FST.
@@
 It also lets the user provide a codon substitution matrix.
-A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[c.f.][]{ranwez_macse_2011,ranwez_macse_2018}.
+A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[cf.][]{ranwez_macse_2011,ranwez_macse_2018}.
 This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.
@@
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
-An example of an FST-based alignment can be found in Supplementary Materials Figure S1.
+An example of an FST-based alignment can be found in Supplementary Materials, Figure S1.
 
@@
 To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination.
-K2P corrects for multiple mutations at a site and takes into account differences in transition and transversion rates.
-It also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites.
+K2P distances correct for multiple mutations at a site and takes into account differences in transition and transversion rates.
+K2P also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites.
 While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
-We used the R software package ape \citep{paradis2019ape} to calculate K2P distances.
+We also calculated p-distances \citep{saitou1987neighbor} for each sequence-pair and aligner combination, which are simply the proportions of sites that differ between two sequences.
+We used the R software package ape \citep{paradis2019ape} to calculate the distances.
 %
@@
 
-% The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length.
-The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome.
-Assigning one of the sequences as the reference, to distinguish between insertions and deletions, CIGAR strings can also summarize pairwise alignments by grouping the number of contiguous matches or mismatches `M', deletions `D', and insertions `I'.
-The resulting pattern combines these letters preceded by the number of characters for each section as they appear in the alignment.
-This pattern is introduced by replacing nucleotides with gaps as indicated by deletions on one sequence and randomly introducing residues where the CIGAR strings indicated insertions.
-
-% Several safety checks are in place to ensure the algorithm runs correctly and the result is accurate.
-The assertions are divided into checking lengths and maintaining the reading frame of each section.
-The simulation can fail if the length of the sequences is different or if, without counting insertions, the length of the pattern to be inserted is longer.
-In addition, maintaining the phases of each section in the CIGAR string is important to avoid introducing errors such as frameshifts.
-
-% I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
-% % After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
-% I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
-% Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
-% MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
-% \citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.
-
 \subsection*{Measuring Aligner Accuracy}
@@
 
-% The computation of $d_{seq}$ involves characterizing the gaps present in the alignment.
-Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$.
-The distance between two alignments $A$ and $B$ is the average across all characters of the symmetric difference (or Hamming distance), represented as `$\triangle$' between homology sets over the length of such sets:
-
-% \begin{equation}
-% d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
-% \end{equation}
-
-% \noindent where $c$ is the sum of the sequence lengths.
-
 Using $d_{seq}$ we quantified the number of perfect, imperfect, and best alignments each aligner produced.
@@
 Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.
-
-% \cyan{The first sentence intro is a good idea, but rewrite!}
-% Sequencing artifacts and errors made in alignment can lead to inaccurate results in genomics. This is especially true in studies of positive selection and positively selected genes.
-
-% Ks and Ka are, respectively, the number of substitutions per synonymous site and per non-synonymous site between two protein-coding genes.
-
-% First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate.
-A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous.
-Second, the nucleotide changes between the two sequences are counted and divided as transitional (A$\leftrightarrow$G, C$\leftrightarrow$T) and transversional (\{A, G\}$\leftrightarrow$\{C, T\}).
-Third, the Kimura two-parameter distance \citep{kimura1980simple} is used to estimate the number of transitions and transversions per site type (nondegenerate, twofold degenerate, and fourfold degenerate), which is used as a correction factor for multiple hits.
-Finally, $k_s$ is the estimate of the average transitional rate at twofold and fourfold degenerate sites, and $k_a$ is the estimate of the average transversional rate at nondegenerate and twofold sites.
-In the results, these metrics are reported as the F$_1$ score, which is the harmonic mean of precision (true positives over total positives) and recall (true positives over true positives and false negatives).
-This score ranges between 0 and 1, with a score of 1 representing a perfect result.
-
-% e compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura's 2-parameter distance \citep{kimura1980simple}.
-The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions.
-The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions.
 
@@
 Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test).
-Additional analysis of the empirical alignments can be found in Supplementary Methods.
+Additional analysis of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.
 
@@
 COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions.
-Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S5).
+Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S6).
 While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.
@@
 \section*{Discussion}
-% Using 16000 human genes and their gorilla orthologs from the ENSEMBL
-% database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
-% alignments with empirical gap patterns.
-% We used the data set to evaluate the accuracy of COATi and a suite of popular aligners spanning multiple alignment methods:
-% Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
-% MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
-% \citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}.
-
-% After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and then aligned the remaining pairs with all five methods.
-% At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7719 sequence pairs.
-% We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
-% The accuracy of inferred alignments compared to the benchmarks was measured using \citeauthor{metrics_blackburne_whelan_2011}'s \citeyearpar{metrics_blackburne_whelan_2011} distance metric, $d_{seq}$.
-% Additionally, because different alignments may be evolutionary equivalent, we scored alignments using the codon-marginal-mg model, and any alignment that had the same score as the benchmark alignment was considered equivalent to the benchmark alignment.
-% The accuracy of identifying positive and negative selection was calculated using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
-% \citembe{ka_ks_li_1993} (Supplementary Methods).
-% The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.
-
-% NOTES:
-% - Benchmark only contains seqeunces with gaps, which will make alignes look more different than they are in practice
-% - simulalation contained a lot of frameshifts that produed stop codons. Probably too many.
-% - redo colors of figures 1 - 3.
-% - define an equivalent alignment again in the benchmark results
 
@@
 
-ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts; however, it was the worst performing protocol that we tested.
+ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts.
+Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment.
+ClustalΩ was the worst performing protocol that we tested.
 It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
@@
 We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols.
-Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S6).
+Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7).
 Different aligners tended to spread out from one another and in different directions from the benchmark.
@@
 
-% COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
-To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference.
-Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons or other artifacts in 3758 simulated gorilla sequences.
-While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence.
-Despite this limitation, we analyzed the 4003 alignments and compared the results across methods, including COATi using the human sequence as the reference.
-The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the reference in comparison to the reverse (Supplementary Materials Tab.
-6).
-However, the results for COATi continue to be a significant improvement over other aligners.
-
-COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model.
-While COATi has multiple modes and models, here we have focused on evaluating the utility of COATi FST for estimating pairwise alignments.
-We have shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances in the presence of genomic artifacts.
+COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model, and the present study has shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances and in the presence of genomic artifacts.
+While COATi has multiple models, we have focused on evaluating the utility of COATi's complex finite-state-transducer for estimating pairwise alignments.
 In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
 
-COATi is under active development.
-We plan on extending it to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
-We plan on improving its multiple-sequence alignment and alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
+This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including modules to sample pairwise alignments from alignment space and generate multiple-sequence alignments.
+We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies.
+On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments.
+Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a guide tree.
+
+COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model.
+We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
+We plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
 Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
@@
 \section*{Availability}
+
 The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++.

### 2024-04-18 alignpair_letter.tex
Draft of revision 2

@@
 
-\usepackage[letterpaper,margin=1in]{geometry}
+\usepackage[letterpaper,margin=0.75in]{geometry}
 
@@
 COATi can infer indels both within and between codons, leading to improved sequences alignments.
-We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with an earlier study in rats and mice.
-We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs via several measures.
+We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with previous work in rodents.
+We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs on several measures of alignment quality and accuracy.
 
@@
 To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets.
+We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data.
+We also find that COATi generates more accurate alignments when applied to semi-empirical benchmarks.
 
@@
 On empirical data, COATi produced alignments that provided reasonable biological inferences.
-On semi-empirical, simulated benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
+On semi-empirical benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
 Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps.
-Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results estimate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels.
-This indicates that alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 or phase-2.
+Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels.
+Therefore alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 and phase-2.
 
-In addition to supporting all three phases of gaps, COATi's other advantage over competing aligners in our tests is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligner's default settings.
+In addition to supporting all three phases of gaps, COATi's advantage over the other aligners that we tested is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings.
 Different data sets may show different results.
@@
 This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''.
-While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns.
+While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK did well on ClustalΩ patterns.
 COATi did well on all patterns.
-
-COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model, and the present study has shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances and in the presence of genomic artifacts.
-While COATi has multiple models, we have focused on evaluating the utility of COATi's complex finite-state-transducer for estimating pairwise alignments.
-In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
 
@@
 Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a guide tree.
+
+The present study has shown that COATi offers a biologically significant improvement over other methods.
+COATi produces more accurate alignments of protein-coding sequences and better downstream inferences of sequence divergence, selection, and indel processes.
+While here we have focused on the challenge of aligning pairs of sequences from sister species that may contain genomic artifacts, in other work, we have begun to explore application of COATi to more divergent sequences and evaluate in detail the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
 

### 2024-04-26 response_r2.md
Draft of revision 2
Add additional citations.

@@
 
-    TODO
+    Thank you for the positive words. We have added additional language to the manuscript about COATi's success with benchmark alignments.
 
@@
 
-    COATi currently has an early-stage module for estimating pairwise alignments.
+    COATi currently has an early-stage algorithm for estimating pairwise alignments.
 However, it needs more work (and an additional round of funding) before it meets the high-quality standards exhibited by the pairwise aligner.
@@
 
-    We do mean "compare" here because COATi's innovation is in similar spirit to MACSE but fundamentally different.
+    We have reworked this section and included additional citations to prior work in this area.
 

### 2024-04-26 alignpair_letter.tex
Add additional citations.

@@
 While sophisticated advancements have been made, the challenge of alignment inference has not been fully solved \citep{art_morrison_2015}.
-%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \citep{art_morrison_2015}.
-%
 The alignment of protein-coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \citep[e.g.][]{bininda2005transalign,abascal2010translatorx}.
@@
 While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps.
-Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).
+Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence
+\citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.
 
@@
 
-In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{Lunter2005-hk}.
+In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{hein2001algorithm,holmes2001evolutionary,Lunter2005-hk,demaio2021cumulative}.
 An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies.
 While approaches vary, a statistical aligner for a pair of sequences, $X$ and $Y$, typically finds an alignment, $Aln$, that maximizes the joint probability $P(Aln, X, Y)$ or samples alignments from the posterior $P(Aln | X, Y)$.
-This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{bradley2007transducers}.
+This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{holmes2001evolutionary,redelings2007incorporating,problems_cartwright_2009}.
 Pair-HMMs are computational machines with two output tapes.
@@
 
-COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; cf.][]{holmes2001evolutionary}.
+COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs (Figs.~\ref{fig:base-calling} and \ref{fig:coati-fst}), each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST.
 We call this transducer the COATi FST.
@@
 It also lets the user provide a codon substitution matrix.
-A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[cf.][]{ranwez_macse_2011,ranwez_macse_2018}.
-This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.
+A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons.
+(See \citealt{hein1994algorithm,arvestad1997aligning,pedersen1998comparison,ranwez_macse_2011,ranwez_macse_2018} for earlier approaches.) This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.
 
@@
 Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
-This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions.
+This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions \citep[Cf.][]{holmes2001evolutionary,demaio2021cumulative}.
 %

### 2024-04-26 alignpair_letter_source.tex
Mark revision 2

--- 2024-04-26 alignpair_letter.tex
@@
-% !TEX program=lualatex
+%!TEX program=pdflatex
 \documentclass[12pt,letterpaper]{article}
+\usepackage{ifluatex}
 
@@
 \usepackage{mathtools}
+\ifluatex
   \usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
   \usepackage{fontspec}
-\usepackage{microtype}
-
   \setmainfont{TeX Gyre Pagella}
   \setmathfont{Latin Modern Math}
+\else
+  \usepackage{etex}
+  \usepackage{fixltx2e}
+  \usepackage{morefloats}
+  \usepackage{tgpagella}
+  \usepackage{textgreek}
+  \usepackage[T1]{fontenc}
+\fi
+
+%\usepackage[main=english,greek]{babel}
+
+\usepackage{microtype}
 
@@
 \\
-Juan J. García Mesa\textsuperscript{1,2},
+Juan J. Garc\'{i}a Mesa\textsuperscript{1,2},
 Ziqi Zhu\textsuperscript{1,3},
@@
 Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
-Sequence artifacts and errors made during alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
-For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes.
+Sequencing artifacts and errors made during genome assembly, such as abiological frameshifts and incorrect early stop codons, can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
 More significantly, while indels can occur both within and between codons in natural sequences, most amino-acid- and codon-based aligners assume that indels only occur between codons.
-This mismatch between biology and alignment algorithms produces sub-optimal alignments and errors in downstream analyses.
+This mismatch between biology and alignment algorithms produces suboptimal alignments and errors in downstream analyses.
 
@@
 COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments.
-COATi can infer indels both within and between codons, leading to improved sequences alignments.
-We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with previous work in rodents.
-We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs on several measures of alignment quality and accuracy.
+COATi can infer indels both within and between codons, leading to improved sequence alignments.
+We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas and conclude that 41\% of indels occurred between codons, agreeing with previous work in other species.
+We also applied COATi to semi-empirical benchmark alignments and find that it outperforms several popular alignment programs on several measures of alignment quality and accuracy.
 
@@
 
-Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}).
-Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively.
-Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}.
-Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}.
+Within protein-coding sequences, indels may occur between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}).
+In our nomenclature, gaps that start after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively.
+Here, gaps that occur between codons are phase-3 gaps whereas they may be known as phase-0 gaps in earlier studies \cite[e.g.][]{taylor2004occurrence}.
+Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2, which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}.
 While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps.
-Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence
+Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce suboptimal alignments and inflated estimates of sequence divergence
 \citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.
-
-% Figures are 86 mm or 178 mm wide
-\begin{figure}[h!]
-    \centering%
-    \includegraphics[scale=1]{figures/fig-aln.pdf}
-    \par
-    \caption{
-        Standard algorithms produce suboptimal alignments.
-        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
-        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
-        %
-        Indels in protein-coding sequences can be classified as having one of three difference phases and being one of two different types.
-        Phases refer the the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
-        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
-        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only).
-The difference between an in-frame and a frameshift indel is not displayed.
-        }
-    \label{fig:aln}
-\end{figure}
 
@@
 
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=\textwidth]{figures/fig-fst-base-calling-new.pdf}
-    \caption[FSTs more the generation of sequences]{%
-    Finite state transducers (FSTs) model the generation of an output sequence based on an input sequence.
-    %
-    (a) A graph of a probabilistic FST \citep{cotterell-etal-2014-stochastic} for base-calling errors using a Mealy-machine architecture, where parameter $u$ is the error rate.
-This graph contains two states (S and M) connected by arcs, with labels ``input symbols~:~output symbols~/~weight''.
-Arcs consume symbols from the input sequence and emit symbols to the output sequence.
-Weights describe the probability that an arc is taken given the input symbols.
-Epsilon (ε) is a special symbol denoting that no symbols were either consumed or emitted.
-    %
-    (b) An FST for matching sequences against ambiguous nucleotides (N). 
-    (c) An FST that results from the composition ($∘$ operation) of the Error FST with the Ambiguity FST.
-    %
-    %Let $x$ be a symbol in the input of (a), $y$ be a symbol in the output of (b), and $z$ be a symbol in both the output of (a) and the input of (b).
-An arc in the composed FST (c) has an input symbol $x$, an output symbol $y$, and a weight that is the combination over $z$ of the aggregate weights of all $x$-$z$-$y$ arc pairs.
-    %
-    }\label{fig:base-calling}
-\end{figure}
-
 \subsection*{The COATi FST}
@@
 
-COATi FST is not a true probabilistic FST and cannot be used as-is to simulate output sequences based on an input sequence.
+COATi FST is not a true probabilistic FST \citep{cotterell-etal-2014-stochastic} and cannot be used as-is to simulate output sequences based on an input sequence.
 This is because it is missing a parameter to control how often Ns are added to the output sequence.
@@
 This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C.
-While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and ignoring any states in between:
+While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and eliding any states in between:
 \[
@@
 
-\begin{figure}[h!]
-\includegraphics[width=\textwidth]{figures/fig-fst-coati.pdf}
-\caption{The COATi FST is built from simpler FSTs via composition.
-(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M.
-These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
-The weight of each arc is a conditional probability derived from a codon substitution model.
-See Fig.~\ref{fig:base-calling} for more details about reading this graph.
-%
-(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the gap-opening parameter and $e$ is the gap-extension parameter.
-Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
-This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions \citep[Cf.][]{holmes2001evolutionary,demaio2021cumulative}.
-%
-(c) The COATi FST is derived via composition from the codon substitution, indel, error, and ambiguity FSTs.
-}
-\label{fig:coati-fst}
-\end{figure}
-
 In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}.
@@
 Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment.
-If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm.
+If more than one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm.
 All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}.
@@
 We first selected human protein-coding genes that belonged to the Consensus Coding Sequence Set, that were located on an autosomal chromosome, and that had a one-to-one gorilla ortholog.
-We selected the canonical isoform for both species and removed any pair in which the total nucleotide length was larger larger than 6,000 nucleotides.
+We selected the canonical isoform for both species and removed any pair in which the total nucleotide length was larger than 6,000 nucleotides.
 This resulted in 14,127 sequence pairs and corresponding FASTA files containing CDS sequences.
@@
 
-In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ \citep{clustal_omega_sievers_2011} via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE \citep{ranwez_macse_2018}, (4) the DNA model of MAFFT \citep{katoh2013mafft}, and (5) the codon model of PRANK \citep{prank_loytynoja_2014}.
+In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of Clustal\textOmega{}{} \citep{clustal_omega_sievers_2011} via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE \citep{ranwez_macse_2018}, (4) the DNA model of MAFFT \citep{katoh2013mafft}, and (5) the codon model of PRANK \citep{prank_loytynoja_2014}.
 COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
@@
 
-Alignment inference impacts the estimation of evolutionary distances.
+Alignment inference impacts the estimation of evolutionary distances \citep{redelings2007incorporating}.
 To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination.
-K2P distances correct for multiple mutations at a site and takes into account differences in transition and transversion rates.
+K2P distances correct for multiple mutations at a site and take into account differences in transition and transversion rates.
 K2P also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites.
-While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
+While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straightforward to calculate and provides a quantitative measure of the evolutionary divergence between sequences.
 We also calculated p-distances \citep{saitou1987neighbor} for each sequence-pair and aligner combination, which are simply the proportions of sites that differ between two sequences.
-We used the R software package ape \citep{paradis2019ape} to calculate the distances.
+We used the R \citep{r2024} software package ape \citep{paradis2019ape} to calculate the distances.
 %
@@
 
-In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
+In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels.
 We separated our 14,127 empirical sequence pairs into two sets: %
-(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
+(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, Clustal\textOmega{}, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
 (45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
@@
 Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
-(This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.)
+(This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)
 
@@
 PRANK was not able to align 24 sequence pairs.
-MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from its wrapper script.
+MAFFT, MACSE, and Clustal\textOmega{} were able to align all sequence pairs, although the latter received some help from its wrapper script.
 Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
@@
 MACSE's amino-acid aware nucleotide model inferred 7\%, while MAFFT's non-coding DNA model inferred 48\%.
-PRANK's codon and ClustalΩ's amino-acid models were constrained by their assumptions and inferred no phase 1 or phase 2 gaps.
-Note that ClustalΩ's gaps with phases of 1 and 2 were created by our wrapper script handling DNA sequences with lengths that are not multiples of 3.
+PRANK's codon and Clustal\textOmega{}'s amino-acid models were constrained by their assumptions and inferred no phase 1 or phase 2 gaps.
+Note that Clustal\textOmega{}'s gaps with phases of 1 and 2 were created by our wrapper script handling DNA sequences with lengths that are not multiples of 3.
 
-\begin{table}[h]
+\paragraph{Evolutionary Distances.}
+
+The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a).
+Conversely, the means for Clustal\textOmega{}, MACSE, and MAFFT were much higher, 1.5\%--1.8\%.
+These larger means are due to the fact that the right tails of the distributions for Clustal\textOmega{}, MACSE, and MAFFT were larger than the other methods.
+As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a).
+The differences between the distances inferred by COATi and other methods also shows Clustal\textOmega{}, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
+%
+COATi's alignments produced significantly lower K2P distances than Clustal\textOmega{}, MACSE, MAFFT, and PRANK.
+Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$.
+Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test).
+Additional analyses of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.
+
+% Wilcoxon signed rank test
+% coati vs rev-coati: p-value: 0.4212 (two.sided)
+% coati vs prank: p-value: < 2.2e-16 (less)
+% coati vs macse: p-value: < 2.2e-16 (less)
+% coati vs mafft: p-value: < 2.2e-16 (less)
+% coati vs clustalo: p-value: < 2.2e-16 (less)
+
+\paragraph{Selection.}
+
+The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, Clustal\textOmega{}, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}).
+Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies.
+The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev.
+The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT.
+Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by COATi and COATi-rev only.
+%
+In total, COATi identified 1,367 ($\sim$10\%) sequence pairs as showing evidence of positive selection.
+Clustal\textOmega{} identified 1,228.
+MACSE identified 1,183.
+MAFFT identified 1,340.
+And PRANK identified 1,349.
+Note that we did not explore whether any of these inferences were biologically or statistically significant.
+
+\subsection*{Semi-Empirical Benchmark}
+
+COATi produced better alignments than Clustal\textOmega{}, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set.
+It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
+%
+COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}).
+All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests.
+(P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$).
+Notably, the average alignment error of the second best protocol was four times larger than COATi's.
+We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}).
+COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions.
+Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6).
+While all aligners, except Clustal\textOmega{}, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.
+
+For evolutionary distances, COATi had the lowest root-mean-square error (0.000694) relative to the benchmark alignments (Tab.~\ref{table:comp}).
+The second lowest aligner, PRANK, had a root-mean-square error that was over 400 times larger than COATi's.
+COATi only overestimated  10\% of its evolutionary distances, while other aligners overestimated 26\%--53\% of their evolutionary distances.
+%
+Additionally, COATi more accurately inferred events of positive and negative selection (Tab.~\ref{table:comp}).
+
+\section*{Discussion}
+
+We have developed a statistical alignment program using finite-state transducers called COATi.
+On empirical data, COATi produced alignments that provided reasonable biological inferences.
+On semi-empirical benchmarks, COATi was more accurate than the other aligners/alignment strategies that we tested: Clustal\textOmega{}'s amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
+Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps.
+Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences are phase-3 indels between humans and gorillas.
+Therefore, alignment strategies that only support phase-3 gaps in protein-coding sequences are suboptimally aligning the 59\% of indels that are phase-1 and phase-2.
+
+In addition to supporting all three phases of gaps, COATi's advantage over the other aligners was at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings.
+Different data sets may show different results.
+However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.
+This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.
+
+COATi's primary disadvantage is that its FST model was slower and consumed more resources than other methods.
+However, COATi's marginal model produces results similar to the FST model, was the fastest method tested, and consumed reasonable amounts of memory (Supplementary Materials and \citealt{garcia2023dissertation}).
+%
+COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
+In order to test whether the choice of a reference sequence impact's COATi's results, we aligned our empirical sequence pairs with COATi using gorilla as the reference.
+We found no biologically or statistically significant differences between the results.
+The most noticeable difference was that, with gorilla as the reference, COATi rejected more sequence pairs due to the presence of ambiguous nucleotides in gorilla sequences.
+
+After COATi, PRANK's codon model was the second most accurate alignment method.
+Like COATi, PRANK also uses statistical models and supports aligning protein-coding sequences using codon models developed for phylogenetics.
+Both COATi and PRANK had reasonable distributions of evolutionary distances that lacked large right tails.
+This is likely because COATi and PRANK produced better alignments when tasked with aligning sequence pairs that contained non-orthologous exons.
+However, in contrast with COATi, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
+In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.
+
+Clustal\textOmega{}'s amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts.
+Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment.
+Clustal\textOmega{} was the worst performing protocol that we tested.
+It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
+--- It even performed poorly on its own gap patterns.
+--- Clustal\textOmega{}'s relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
+
+We obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT.
+These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts.
+If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments.
+
+We can look at distances between alignments generated by different aligners to understand the diversity of approaches used by our alignment protocols.
+Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7).
+Different aligners tended to spread out from one another and in different directions from the benchmark.
+This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''.
+While each aligner, except Clustal\textOmega{}, performed well on its own gap patterns, MACSE also did well on Clustal\textOmega{} and MAFFT gap patterns, and PRANK did well on Clustal\textOmega{} patterns.
+COATi did well on all patterns.
+
+This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including a module to sample pairwise alignments from alignment space and a module to generate multiple-sequence alignments.
+We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies.
+On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments.
+Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a user-provided guide tree.
+
+The present study has shown that COATi offers a biologically significant improvement over other methods.
+COATi produces more accurate alignments of protein-coding sequences and better downstream inferences of sequence divergence, selection, and indel processes.
+While here we have focused on the challenge of aligning pairs of sequences from sister species that may contain genomic artifacts, in other work, we have begun to explore application of COATi to more divergent sequences and evaluate in detail the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
+
+COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model.
+We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighting gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
+We also plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
+Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
+
+\section*{Availability}
+
+The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++.
+COATi is released under an MIT license.
+Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.
+
+% \section*{Supplementary information}
+
+\section*{Acknowledgments}
+
+This research was funded by NSF award DBI-1929850.
+%
+The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Oskan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects.
+The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.
+
+\noindent \textit{Conflict of interest:} none declared.
+
+\newpage
+\section*{Figures}
+
+% Figures are 86 mm or 178 mm wide
+\begin{figure}[ht]
+    \centering%
+    %\includegraphics[scale=1]{figures/fig-aln.pdf}
+    \par
+    \caption{
+        Standard algorithms produce suboptimal alignments.
+        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
+        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
+        %
+        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
+        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
+        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
+        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only).
+The difference between an in-frame and a frameshift indel is not displayed.
+        \label{fig:aln}}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    %\includegraphics[width=\textwidth]{figures/fig-fst-base-calling-new.pdf}
+    \par
+    \caption{%
+    Finite state transducers (FSTs) model the generation of an output sequence based on an input sequence.
+    %
+    (a) A graph of a probabilistic FST \citep{cotterell-etal-2014-stochastic} for base-calling errors using a Mealy-machine architecture, where parameter $u$ is the error rate.
+This graph contains two states (S and M) connected by arcs, with labels ``input symbols~:~output symbols~/~weight''.
+Arcs consume symbols from the input sequence and emit symbols to the output sequence.
+Weights describe the probability that an arc is taken given the input symbols.
+Epsilon (\textepsilon) is a special symbol denoting that no symbols were either consumed or emitted.
+    %
+    (b) An FST for matching sequences against ambiguous nucleotides (N). 
+    (c) An FST that results from the composition ($\circ$ operation) of the Error FST with the Ambiguity FST.
+    %
+    %Let $x$ be a symbol in the input of (a), $y$ be a symbol in the output of (b), and $z$ be a symbol in both the output of (a) and the input of (b).
+An arc in the composed FST (c) has an input symbol $x$, an output symbol $y$, and a weight that is the combination over $z$ of the aggregate weights of all $x$-$z$-$y$ arc pairs.
+    %
+    \label{fig:base-calling}}
+\end{figure}
+
+\begin{figure}[ht]
+%\includegraphics[width=\textwidth]{figures/fig-fst-coati.pdf}
+\caption{The COATi FST is built from simpler FSTs via composition.
+(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M.
+These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape.
+The weight of each arc is a conditional probability derived from a codon substitution model.
+See Fig.~\ref{fig:base-calling} for more details about reading this graph.
+%
+(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the gap-opening parameter and $e$ is the gap-extension parameter.
+Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1.
+This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions \citep[Cf.][]{holmes2001evolutionary,demaio2021cumulative}.
+%
+(c) The COATi FST is derived via composition from the codon substitution, indel, error, and ambiguity FSTs.
+\label{fig:coati-fst}}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering%
+    %\includegraphics{figures/fig-k2p-empirical.pdf}
+    \par
+    \caption{COATi's alignments produce biologically reasonable evolutionary distances.
+(a) The distribution of K2P distances inferred from alignments generated by each method.
+The averages of each distribution are indicated by vertical lines.
+The averages are COATi = 0.0084, Clustal\textOmega{} = 0.0178, MACSE = 0.0149, MAFFT = 0.0154, PRANK = 0.0092, and COATi-rev = 0.0084.
+(b) The distribution of the differences between distances inferred by COATi and other methods.
+The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.\label{fig:k2p-empirical}}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering%
+    %\includegraphics{figures/fig-sel-empirical.pdf}
+    \par
+    \caption{Aligners varied in which sequence pairs they identified as undergoing positive selection.
+    In this UpSet plot, the bottom panel displays the 16 most frequent intersection patterns among aligners.
+A black circle represents positive selection.
+The most frequent pattern was that no aligner found positive selection while the second most frequent pattern was that all aligners found positive selection.
+Other patterns involved a disagreement between aligners about whether a sequence pair showed evidence of positive or negative selection.
+The top panel displays the number of sequence pairs in each grouping.\label{fig:sel-empirical}}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering%
+    %\includegraphics{figures/fig-dseq-benchmarks.pdf}
+    \par
+    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods.
+A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.\label{fig:dseq-benchmarks}}
+\end{figure}
+
+\clearpage
+\newpage
+\section*{Tables}
+
+\begin{table}[ht]
 \caption{\label{tab:gaps-empirical}COATi produced more indels and fewer mismatches compared to other aligners.}
@@
 \addlinespace
-ClustalΩ & 8 & 6 & 10230 & 95.93\% & 1.54\% & 2.52\%\\
+Clustal\textOmega{} & 8 & 6 & 10230 & 95.93\% & 1.54\% & 2.52\%\\
 \addlinespace
@@
 
-\paragraph{Evolutionary Distances.}
-
-The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a).
-Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%.
-These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods.
-As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a).
-The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
-%
-COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK.
-Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$.
-Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test).
-Additional analysis of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.
-
-% Wilcoxon signed rank test
-% coati vs rev-coati: p-value: 0.4212 (two.sided)
-% coati vs prank: p-value: < 2.2e-16 (less)
-% coati vs macse: p-value: < 2.2e-16 (less)
-% coati vs mafft: p-value: < 2.2e-16 (less)
-% coati vs clustalo: p-value: < 2.2e-16 (less)
-
-\begin{figure}[h!]
-    \centering%
-    \includegraphics{figures/fig-k2p-empirical.pdf}
-    \par
-    \caption{COATi's alignments produce biologically reasonable evolutionary distances.
-(a) The distribution of K2P distances inferred from alignments generated by each method.
-The averages of each distribution are indicated by vertical lines.
-The averages are COATi = 0.0084, ClustalΩ = 0.0178, MACSE = 0.0149, MAFFT = 0.0154, PRANK = 0.0092, and COATi-rev = 0.0084.
-(b) The distribution of the differences between distances inferred by COATi and other methods.
-The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.\label{fig:k2p-empirical}}
-\end{figure}
-
-\paragraph{Selection.}
-
-The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}).
-Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies.
-The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev.
-The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT.
-Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
-%
-In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection.
-ClustalΩ identified 1,228.
-MACSE identified 1,183.
-MAFFT identified 1,340.
-And PRANK identified 1,349.
-Note that we did not explore whether any of these inferences were biologically or statistically significant.
-
-\begin{figure}[h!]
-    \centering%
-    \includegraphics{figures/fig-sel-empirical.pdf}
-    \par
-    \caption{Aligners varied in which sequence pairs they identified as undergoing positive selection.
-    In this UpSet plot, the bottom panel displays the 16 most frequent intersection patterns among aligners.
-A black circle represents positive selection.
-The most frequent pattern was that no aligner found positive selection while the second most frequent pattern was that all aligners found positive selection.
-Other patterns involved a disagreement between aligners about whether a sequence pair showed evidence of positive or negative selection.
-The top panel displays the number of sequence pairs in each grouping.
-%
-    \label{fig:sel-empirical}}
-\end{figure}
-
-\subsection*{Semi-Empirical Benchmark}
-
-COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set.
-It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
-%
-COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}).
-All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests.
-(P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$).
-Notably, the average alignment error of the second best protocol was four times larger than COATi's.
-We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}).
-COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions.
-Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S6).
-While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.
-
-For evolutionary distances, COATi had the lowest root-mean-square error (0.000694) relative to the benchmark alignments (Tab.~\ref{table:comp}).
-The second lowest aligner, PRANK, had a root-mean-square error that was over 400 times larger than COATi's.
-COATi only overestimated  10\% of its evolutionary distances, while other aligners overestimated 26\%--53\% of their evolutionary distances.
-%
-Additionally, COATi more accurately inferred events of positive and negative selection (Tab.~\ref{table:comp}).
-
 % Software comparison table
-\begin{table}[!ht]
+\begin{table}[ht]
 \definecolor{bestcolor}{RGB}{230,230,230}%
@@
   \toprule
- & COATi & ClustalΩ & MACSE & MAFFT & PRANK \\ 
+ & COATi & Clustal\textOmega{} & MACSE & MAFFT & PRANK \\ 
   \midrule
@@
 
-\begin{figure}[h!]
-    \centering%
-    \includegraphics{figures/fig-dseq-benchmarks.pdf}
-    \par
-    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods.
-A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
-%
-    \label{fig:dseq-benchmarks}}
-\end{figure}
-
 \clearpage
-
-\section*{Discussion}
-
-We have developed a statistical alignment program using finite-state transducers called COATi.
-On empirical data, COATi produced alignments that provided reasonable biological inferences.
-On semi-empirical benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
-Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps.
-Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels.
-Therefore alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 and phase-2.
-
-In addition to supporting all three phases of gaps, COATi's advantage over the other aligners that we tested is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings.
-Different data sets may show different results.
-However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.
-This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.
-
-COATi's primary disadvantage is that its FST model was slower and consumed more resources than other methods.
-However, COATi's marginal model produces results similar to the FST model, was the fastest method tested, and consumed reasonable amounts of memory (Supplementary Materials and \citealt{garcia2023dissertation}).
-%
-COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences.
-In order to test whether the choice of a reference sequence impact's COATi's results, we aligned our empirical sequence pairs with COATi using gorilla as the reference.
-We found no biologically or statistically significant differences between the results.
-The most noticeable difference was that, with gorilla as the reference, COATi rejected more sequence pairs due to the presence of ambiguous nucleotides in gorilla sequences.
-
-After COATi, PRANK's codon model was the second most accurate alignment method.
-Like COATi, PRANK also uses statistical models and supports aligning protein-coding sequences using codon models developed for phylogenetics.
-Both COATi and PRANK had reasonable distributions of evolutionary distances that lacked large right tails.
-This is likely because COATi and PRANK produced better alignments when tasked with aligning sequence pairs that contained non-orthologous exons.
-However, in contrast with COATi, PRANK's codon model only permits phase-3 gaps and is very sensitive to frameshift artifacts.
-In fact, PRANK refuses to align any sequence using a codon model if its length is not a multiple of three.
-
-ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts.
-Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment.
-ClustalΩ was the worst performing protocol that we tested.
-It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.
---- It even performed poorly on its own gap patterns.
---- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
-
-We obtained better results when aligning in nucleotide space using both amino-acid aware MACSE and amino-acid agnostic MAFFT.
-These results indicate that aligning protein-coding sequences in nucleotide space works reasonably well if the sequences come from closely related species, such as humans and gorillas, as it supports all three gap phases and is robust to frameshift artifacts.
-If the sequences were further apart, we predict that amino-acid agnostic approaches would begin to produce unreasonable alignments.
-
-We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols.
-Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7).
-Different aligners tended to spread out from one another and in different directions from the benchmark.
-This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''.
-While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK did well on ClustalΩ patterns.
-COATi did well on all patterns.
-
-This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including modules to sample pairwise alignments from alignment space and generate multiple-sequence alignments.
-We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies.
-On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments.
-Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a guide tree.
-
-The present study has shown that COATi offers a biologically significant improvement over other methods.
-COATi produces more accurate alignments of protein-coding sequences and better downstream inferences of sequence divergence, selection, and indel processes.
-While here we have focused on the challenge of aligning pairs of sequences from sister species that may contain genomic artifacts, in other work, we have begun to explore application of COATi to more divergent sequences and evaluate in detail the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.
-
-COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model.
-We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
-We plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
-Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
-
-\section*{Availability}
-
-The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++.
-COATi is released under an MIT license.
-Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.
-
-% \section*{Supplementary information}
-
-\section*{Acknowledgments}
-
-This research was funded by NSF award DBI-1929850.
-%
-The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Oskan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects.
-The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.
-
-\noindent \textit{Conflict of interest:} none declared.
-
-% \section{References}
+\newpage
 \bibliographystyle{mbe}%
 \setlength{\bibhang}{0pt}
-\bibliography{alignpair_letter.bib}
+%\bibliography{alignpair_letter.bib}
+\begin{thebibliography}{43}
+\providecommand{\natexlab}[1]{#1}
+
+\bibitem[{Abascal et~al.(2010)Abascal, Zardoya, and
+  Telford}]{abascal2010translatorx}
+Abascal F, Zardoya R, and Telford MJ. 2010.
+\newblock {TranslatorX}: multiple alignment of nucleotide sequences guided by
+  amino acid translations.
+\newblock \emph{Nucleic Acids Research} 38:W7--W13.
+
+\bibitem[{Allauzen et~al.(2007)Allauzen, Riley, Schalkwyk, Skut, and
+  Mohri}]{allauzen2007openfst}
+Allauzen C, Riley M, Schalkwyk J, Skut W, and Mohri M. 2007.
+\newblock Openfst: A general and efficient weighted finite-state transducer
+  library.
+\newblock In \emph{International Conference on Implementation and Application
+  of Automata}, pages 11--23. Springer.
+
+\bibitem[{Arvestad(1997)}]{arvestad1997aligning}
+Arvestad L. 1997.
+\newblock Aligning coding dna in the presence of frame-shift errors.
+\newblock In A~Apostolico and J~Hein, editors, \emph{Combinatorial Pattern
+  Matching}, pages 180--190. Berlin, Heidelberg: Springer Berlin Heidelberg.
+
+\bibitem[{Bininda-Emonds(2005)}]{bininda2005transalign}
+Bininda-Emonds O. 2005.
+\newblock {transAlign}: using amino acids to facilitate the multiple alignment
+  of protein-coding {DNA} sequences.
+\newblock \emph{BMC Bioinformatics} 6:1--6.
+
+\bibitem[{Blackburne and Whelan(2011)}]{metrics_blackburne_whelan_2011}
+Blackburne BP and Whelan S. 2011.
+\newblock {Measuring the distance between multiple sequence alignments}.
+\newblock \emph{Bioinformatics} 28:495--502.
+
+\bibitem[{Bradley and Holmes(2007)}]{bradley2007transducers}
+Bradley RK and Holmes I. 2007.
+\newblock Transducers: an emerging probabilistic framework for modeling indels
+  on trees.
+\newblock \emph{Bioinformatics} 23.
+
+\bibitem[{Cartwright(2009)}]{problems_cartwright_2009}
+Cartwright RA. 2009.
+\newblock Problems and solutions for estimating indel rates and length
+  distributions.
+\newblock \emph{Molecular Biology and Evolution} 26:473--480.
+
+\bibitem[{Charif and Lobry(2007)}]{seqinr}
+Charif D and Lobry J. 2007.
+\newblock {SeqinR} 1.0-2: a contributed package to the {R} project for
+  statistical computing devoted to biological sequences retrieval and analysis.
+\newblock In U~Bastolla, M~Porto, H~Roman, and M~Vendruscolo, editors,
+  \emph{Structural approaches to sequence evolution: Molecules, networks,
+  populations}, Biological and Medical Physics, Biomedical Engineering, pages
+  207--232. New York: Springer Verlag.
+
+\bibitem[{Cotterell et~al.(2014)Cotterell, Peng, and
+  Eisner}]{cotterell-etal-2014-stochastic}
+Cotterell R, Peng N, and Eisner J. 2014.
+\newblock Stochastic contextual edit distance and probabilistic {FST}s.
+\newblock In K~Toutanova and H~Wu, editors, \emph{Proceedings of the 52nd
+  Annual Meeting of the Association for Computational Linguistics (Volume 2:
+  Short Papers)}, pages 625--630. Baltimore, Maryland: Association for
+  Computational Linguistics.
+
+\bibitem[{De~Maio(2021)}]{demaio2021cumulative}
+De~Maio N. 2021.
+\newblock The cumulative indel model: fast and accurate statistical
+  evolutionary alignment.
+\newblock \emph{Systematic Biology} 70:236--257.
+
+\bibitem[{Fletcher and Yang(2010)}]{effect_fletcher_2010}
+Fletcher W and Yang Z. 2010.
+\newblock The effect of insertions, deletions, and alignment errors on the
+  branch-site test of positive selection.
+\newblock \emph{Molecular Biology and Evolution} 27:2257--2267.
+
+\bibitem[{Garc\'{i}a~Mesa(2023)}]{garcia2023dissertation}
+Garc\'{i}a~Mesa JJ. 2023.
+\newblock Statistical sequence alignment of protein coding regions.
+\newblock Ph.D. thesis, Arizona State University.
+
+\bibitem[{Hein(1994)}]{hein1994algorithm}
+Hein J. 1994.
+\newblock An algorithm combining dna and protein alignment.
+\newblock \emph{Journal of Theoretical Biology} 167:169--174.
+
+\bibitem[{Hein(2001)}]{hein2001algorithm}
+Hein J. 2001.
+\newblock An algorithm for statistical alignment of sequences related by a
+  binary tree.
+\newblock \emph{Pacific Symposium on Biocomputing} 6:179--190.
+
+\bibitem[{Holmes and Bruno(2001)}]{holmes2001evolutionary}
+Holmes I and Bruno WJ. 2001.
+\newblock Evolutionary {HMMs}: a bayesian approach to multiple alignment.
+\newblock \emph{Bioinformatics} 17:803--820.
+
+\bibitem[{Hubbard et~al.(2002)Hubbard, Barker, Birney, Cameron, Chen, Clark,
+  Cox, Cuff, Curwen, Down et~al.}]{ensembl_hubbard_2002}
+Hubbard T, Barker D, Birney E, Cameron G, Chen Y, Clark L, Cox T, Cuff J,
+  Curwen V, Down T, et~al. 2002.
+\newblock The {Ensembl} genome database project.
+\newblock \emph{Nucleic Acids Research} 30:38--41.
+
+\bibitem[{Hubisz et~al.(2011)Hubisz, Lin, Kellis, and Siepel}]{hubisz2011error}
+Hubisz MJ, Lin MF, Kellis M, and Siepel A. 2011.
+\newblock Error and error mitigation in low-coverage genome assemblies.
+\newblock \emph{{PLoS} One} 6:e17,034.
+
+\bibitem[{Jackman et~al.(2018)Jackman, Coombe, Chu, Warren, Vandervalk, Yeo,
+  Xue, Mohamadi, Bohlmann, Jones et~al.}]{jackman2018tigmint}
+Jackman SD, Coombe L, Chu J, Warren RL, Vandervalk BP, Yeo S, Xue Z, Mohamadi
+  H, Bohlmann J, Jones SJ, et~al. 2018.
+\newblock Tigmint: correcting assembly errors using linked reads from large
+  molecules.
+\newblock \emph{BMC Bioinformatics} 19:1--10.
+
+\bibitem[{Katoh and Standley(2013)}]{katoh2013mafft}
+Katoh K and Standley DM. 2013.
+\newblock {MAFFT} multiple sequence alignment software version 7: improvements
+  in performance and usability.
+\newblock \emph{Molecular Biology and Evolution} 30:772--780.
+
+\bibitem[{Kimura(1980)}]{kimura1980simple}
+Kimura M. 1980.
+\newblock A simple method for estimating evolutionary rates of base
+  substitutions through comparative studies of nucleotide sequences.
+\newblock \emph{Journal of Molecular Evolution} 16:111--120.
+
+\bibitem[{Kosiol et~al.(2007)Kosiol, Holmes, and Goldman}]{kosiol_ECM_2007}
+Kosiol C, Holmes I, and Goldman N. 2007.
+\newblock An empirical codon model for protein sequence evolution.
+\newblock \emph{Molecular Biology and Evolution} 24:1464--1479.
+
+\bibitem[{Li(1993)}]{ka_ks_li_1993}
+Li WH. 1993.
+\newblock Unbiased estimation of the rates of synonymous and nonsynonymous
+  substitution.
+\newblock \emph{Journal of Molecular Evolution} 36:96--99.
+
+\bibitem[{Li et~al.(1985)Li, Wu, and Luo}]{Li1985}
+Li WH, Wu CI, and Luo CC. 1985.
+\newblock {A new method for estimating synonymous and nonsynonymous rates of
+  nucleotide substitution considering the relative likelihood of nucleotide and
+  codon changes.}
+\newblock \emph{Molecular Biology and Evolution} 2:150--174.
+
+\bibitem[{L{\"o}ytynoja(2014)}]{prank_loytynoja_2014}
+L{\"o}ytynoja A. 2014.
+\newblock Phylogeny-aware alignment with {PRANK}.
+\newblock In \emph{Multiple Sequence Alignment Methods}, pages 155--170.
+  Springer.
+
+\bibitem[{Lunter et~al.(2005)Lunter, Drummond, Mikl{\'o}s, and
+  Hein}]{Lunter2005-hk}
+Lunter G, Drummond AJ, Mikl{\'o}s I, and Hein J. 2005.
+\newblock Statistical alignment: Recent progress, new applications, and
+  challenges.
+\newblock In \emph{Statistical Methods in Molecular Evolution}, pages 375--405.
+  New York: Springer-Verlag.
+
+\bibitem[{Mohri et~al.(2005)Mohri, Pereira, and Riley}]{mohri2005weighted}
+Mohri M, Pereira F, and Riley M. 2005.
+\newblock Weighted automata in text and speech processing.
+
+\bibitem[{Morrison(2015)}]{art_morrison_2015}
+Morrison DA. 2015.
+\newblock Is sequence alignment an art or a science?
+\newblock \emph{Systematic Botany} 40:14--26.
+
+\bibitem[{Muse and Gaut(1994)}]{muse_gaut_1994}
+Muse SV and Gaut BS. 1994.
+\newblock A likelihood approach for comparing synonymous and nonsynonymous
+  nucleotide substitution rates, with application to the chloroplast genome.
+\newblock \emph{Molecular Biology and Evolution} 11:715--724.
+
+\bibitem[{Pamilo and Bianchi(1993)}]{Pamilo1993}
+Pamilo P and Bianchi NO. 1993.
+\newblock {Evolution of the Zfx and Zfy genes: rates and interdependence
+  between the genes.}
+\newblock \emph{Molecular Biology and Evolution} 10:271--281.
+
+\bibitem[{Paradis and Schliep(2019)}]{paradis2019ape}
+Paradis E and Schliep K. 2019.
+\newblock ape 5.0: an environment for modern phylogenetics and evolutionary
+  analyses in {R}.
+\newblock \emph{Bioinformatics} 35:526--528.
+
+\bibitem[{Pedersen et~al.(1998)Pedersen, Lyngs{\o}, and
+  Hein}]{pedersen1998comparison}
+Pedersen CNS, Lyngs{\o} R, and Hein J. 1998.
+\newblock Comparison of coding {DNA}.
+\newblock In M~Farach-Colton, editor, \emph{Combinatorial Pattern Matching},
+  pages 153--173. Berlin, Heidelberg: Springer Berlin Heidelberg.
+
+\bibitem[{{R Core Team}(2024)}]{r2024}
+{R Core Team}. 2024.
+\newblock \emph{R: A Language and Environment for Statistical Computing}.
+\newblock R Foundation for Statistical Computing, Vienna, Austria.
+
+\bibitem[{Ranwez et~al.(2018)Ranwez, Douzery, Cambon, Chantret, and
+  Delsuc}]{ranwez_macse_2018}
+Ranwez V, Douzery EJP, Cambon C, Chantret N, and Delsuc F. 2018.
+\newblock {{MACSE} v2: toolkit for the alignment of coding sequences accounting
+  for frameshifts and stop codons}.
+\newblock \emph{Molecular Biology and Evolution} 35:2582--2584.
+
+\bibitem[{Ranwez et~al.(2011)Ranwez, Harispe, Delsuc, and
+  Douzery}]{ranwez_macse_2011}
+Ranwez V, Harispe S, Delsuc F, and Douzery EJ. 2011.
+\newblock {MACSE}: {M}ultiple {A}lignment of {C}oding {SE}quences accounting
+  for frameshifts and stop codons.
+\newblock \emph{{PLoS} One} 6:e22,594.
+
+\bibitem[{Redelings and Suchard(2007)}]{redelings2007incorporating}
+Redelings BD and Suchard MA. 2007.
+\newblock Incorporating indel information into phylogeny estimation for rapidly
+  emerging pathogens.
+\newblock \emph{BMC Evolutionary Biology} 7:1--19.
+
+\bibitem[{Rosenberg(2009)}]{sequence_alignment_rosenberg_2009}
+Rosenberg MS. 2009.
+\newblock \emph{Sequence Alignment: Methods, Models, Concepts, and Strategies}.
+\newblock Univ of California Press.
+
+\bibitem[{Saitou and Nei(1987)}]{saitou1987neighbor}
+Saitou N and Nei M. 1987.
+\newblock The neighbor-joining method: a new method for reconstructing
+  phylogenetic trees.
+\newblock \emph{Molecular Biology and Evolution} 4:406--425.
+
+\bibitem[{Schneider et~al.(2009)Schneider, Souvorov, Sabath, Landan, Gonnet,
+  and Graur}]{estimates_schneider_2009}
+Schneider A, Souvorov A, Sabath N, Landan G, Gonnet GH, and Graur D. 2009.
+\newblock Estimates of positive {Darwinian} selection are inflated by errors in
+  sequencing, annotation, and alignment.
+\newblock \emph{Genome Biology and Evolution} 1:114--118.
+
+\bibitem[{Sievers et~al.(2011)Sievers, Wilm, Dineen, Gibson, Karplus, Li,
+  Lopez, McWilliam, Remmert, S{\"o}ding et~al.}]{clustal_omega_sievers_2011}
+Sievers F, Wilm A, Dineen D, Gibson TJ, Karplus K, Li W, Lopez R, McWilliam H,
+  Remmert M, S{\"o}ding J, et~al. 2011.
+\newblock Fast, scalable generation of high-quality protein multiple sequence
+  alignments using {Clustal Omega}.
+\newblock \emph{Molecular Systems Biology} 7:539.
+
+\bibitem[{Silvestre-Ryan et~al.(2021)Silvestre-Ryan, Wang, Sharma, Lin, Shen,
+  Dider, and Holmes}]{silvestre2021machine}
+Silvestre-Ryan J, Wang Y, Sharma M, Lin S, Shen Y, Dider S, and Holmes I. 2021.
+\newblock {Machine Boss}: rapid prototyping of bioinformatic automata.
+\newblock \emph{Bioinformatics} 37:29--35.
+
+\bibitem[{Taylor et~al.(2004)Taylor, Ponting, and
+  Copley}]{taylor2004occurrence}
+Taylor MS, Ponting CP, and Copley RR. 2004.
+\newblock Occurrence and consequences of coding sequence insertions and
+  deletions in mammalian genomes.
+\newblock \emph{Genome Research} 14:555--566.
+
+\bibitem[{Yoon(2009)}]{yoon_2009_hmm}
+Yoon BJ. 2009.
+\newblock Hidden {Markov} models and their applications in biological sequence
+  analysis.
+\newblock \emph{Current Genomics} 10:402--415.
+
+\bibitem[{Zhu(2022)}]{zhu2022profiling}
+Zhu Z. 2022.
+\newblock Profiling of indel phases in coding regions.
+\newblock Ph.D. thesis, Arizona State University.
+
+\end{thebibliography}
 

### 2024-06-07 alignpair_letter.tex
Draft of final version after acceptence

@@
 \usepackage{url}
+\urlstyle{same}
 \usepackage{graphicx}
@@
 % now start line numbers
-\linenumbers
+% \linenumbers
 
@@
 
-Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}).
-Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively.
-Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}.
-Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}.
+Within protein-coding sequences, indels may occur between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}).
+In our nomenclature, gaps that start after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively.
+Here, gaps that occur between codons are phase-3 gaps whereas they may be known as phase-0 gaps in earlier studies \cite[e.g.][]{taylor2004occurrence}.
+Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2, which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}.
 While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps.
-Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence
+Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce suboptimal alignments and inflated estimates of sequence divergence
 \citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.
@@
         %
-        Indels in protein-coding sequences can be classified as having one of three difference phases and being one of two different types.
-        Phases refer the the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
+        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
+        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
         Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
@@
 Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citep{estimates_schneider_2009, effect_fletcher_2010, hubisz2011error}.
-While genomes for model organisms often get refined over many iterations and contain meticulously curated protein-coding sequences, 
+While genomes for model organisms are often refined over many iterations and contain meticulously curated protein-coding sequences, 
 genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
@@
 
-To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets.
+To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic datasets.
 We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data.
-We also find that COATi generates more accurate alignments when applied to semi-empirical benchmarks.
+We also find that COATi generates more accurate alignments when applied to a semi-empirical benchmark dataset.
 
@@
 
-In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{hein2001algorithm,holmes2001evolutionary,Lunter2005-hk,demaio2021cumulative}.
+In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{hein2001algorithm,holmes2001evolutionary,Lunter2005-hk,Holmes2020,demaio2021cumulative}.
 An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies.
@@
 
-In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels.
+In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels.
 We separated our 14,127 empirical sequence pairs into two sets: %
@@
 (45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
-We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments.
+We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments.
 The phases of gaps were preserved as were the spacings of any clusters of gaps.
 Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
-(This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.)
+(This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)
 
@@
 %
-We generated a benchmark of 8261 sequence pairs with known alignments using this strategy.
+We generated a benchmark dataset of 8261 sequence pairs with known alignments using this strategy.
 
@@
 
-We aligned our benchmark alignments using our suite of aligners and quantified the similarity of each alignment that was produced to its respective benchmark.
-As above, we measured Kimura's 2-parameter distance \citep{kimura1980simple} and $K_A/K_S$ \citep{ka_ks_li_1993} for the benchmark and estimated alignments.
+We aligned our benchmark alignments using our suite of aligners and quantified the similarity of each estimated alignment to its respective benchmark.
+As above, we measured Kimura's 2-parameter distance \citep{kimura1980simple} and $K_A/K_S$ \citep{ka_ks_li_1993} for the benchmark alignments and estimated alignments.
 Additionally, we quantified the error of estimated alignments using the $d_{seq}$ metric \citep{metrics_blackburne_whelan_2011}.
@@
 Evolutionary equivalence was determined by scoring alignments using COATi's marginal model.
-Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark was greater than zero.
+Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark alignment was greater than zero.
 %
@@
 Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test).
-Additional analysis of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.
+Additional analyses of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.
 
@@
 The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT.
-Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
+Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by COATi and COATi-rev only.
 %
@@
 
-COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set.
+COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark dataset.
 It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
@@
 Notably, the average alignment error of the second best protocol was four times larger than COATi's.
-We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}).
-COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions.
-Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S6).
+We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}).
+COATi was the closest aligner to the benchmark dataset, and other aligners tended to diverge from the benchmark dataset in different directions.
+Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6).
 While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.
@@
 \centering
-\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\textsuperscript{1}\label{table:comp}}
+\caption{COATi generates better alignments than other alignment algorithms on a semi-empirical benchmark dataset.\textsuperscript{1}\label{table:comp}}
 %
@@
    \addlinespace
-\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark was 8,261.}\\
+\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark dataset was 8,261.}\\
 \multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments).
@@
     \par
-    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods.
+    \caption{COATi's alignments were closer to the semi-empirical benchmark dataset than other methods.
 A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
@@
 On empirical data, COATi produced alignments that provided reasonable biological inferences.
-On semi-empirical benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
+On a semi-empirical benchmark dataset, COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
 Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps.
-Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels.
-Therefore alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 and phase-2.
+Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences are phase-3 indels between humans and gorillas.
+Therefore, alignment strategies that only support phase-3 gaps in protein-coding sequences are suboptimally aligning the 59\% of indels that are phase-1 and phase-2.
 
-In addition to supporting all three phases of gaps, COATi's advantage over the other aligners that we tested is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings.
+In addition to supporting all three phases of gaps, COATi's advantage over the other aligners was at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings.
 Different data sets may show different results.
@@
 
-We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols.
+We can look at distances between alignments generated by different aligners to understand the diversity of approaches used by our alignment protocols.
 Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7).
-Different aligners tended to spread out from one another and in different directions from the benchmark.
+Different aligners tended to spread out from one another and in different directions from the benchmark dataset.
 This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''.
@@
 
-This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including modules to sample pairwise alignments from alignment space and generate multiple-sequence alignments.
+This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including a module to sample pairwise alignments from alignment space and a module to generate multiple-sequence alignments.
 We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies.
 On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments.
-Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a guide tree.
+Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a user-provided guide tree.
 
@@
 COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model.
-We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
-We plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
+We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighting gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
+We also plan on improving its alignment sampling capabilities, as well as implementing new models for aligning long-read sequences of genes against reference genomes.
 Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
@@
 
-The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++.
+The source code for COATi, along with documentation, is freely available on GitHub: \nolinkurl{https://github.com/CartwrightLab/coati} (DOI: \nolinkurl{10.5281/zenodo.11499800}) and is implemented in C++.
 COATi is released under an MIT license.
-Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.
+Additional information, code, and workflows to replicate the analysis can be found on GitHub: \nolinkurl{https://github.com/jgarciamesa/coati-testing} (DOI: \nolinkurl{10.5281/zenodo.11515933}) and \nolinkurl{https://github.com/jgarciamesa/alignpair_letter} (DOI: \nolinkurl{10.5281/zenodo.11512411}).
 
@@
 %
-The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Oskan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects.
+The authors would like to thank Profs.\ Marco Mangone, Ted Pavlic, Banu Ozkan, Jay Taylor, and Jeremy Wideman for their helpful support on two separate PhD dissertation projects.
 The authors would also like to thank the associate editor and two reviewers for their helpful comments on earlier versions of this manuscript.

### 2024-06-07 alignpair_letter_source.tex
Draft of final version after acceptence

@@
 % now start line numbers
-\linenumbers
+%\linenumbers
 
@@
 (45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs.
-We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments.
+We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments.
 The phases of gaps were preserved as were the spacings of any clusters of gaps.
@@
 
-COATi produced better alignments than Clustal\textOmega{}, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set.
+COATi produced better alignments than Clustal\textOmega{}, MACSE, MAFFT, and PRANK when evaluated by our benchmark dataset.
 It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
@@
 We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}).
-COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions.
+COATi was the closest aligner to the benchmark dataset, and other aligners tended to diverge from the benchmark dataset in different directions.
 Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6).
@@
 On empirical data, COATi produced alignments that provided reasonable biological inferences.
-On semi-empirical benchmarks, COATi was more accurate than the other aligners/alignment strategies that we tested: Clustal\textOmega{}'s amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
+On a semi-empirical benchmark dataset, COATi was more accurate than the other aligners/alignment strategies that we tested: Clustal\textOmega{}'s amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model.
 Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps.
@@
 Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7).
-Different aligners tended to spread out from one another and in different directions from the benchmark.
+Different aligners tended to spread out from one another and in different directions from the benchmark dataset.
 This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''.
@@
 We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighting gap openings to reflect known selection on indel phases \citep{zhu2022profiling}.
-We also plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes.
+We also plan on improving its alignment sampling capabilities, as well as implementing new models for aligning long-read sequences of genes against reference genomes.
 Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
@@
     \par
-    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods.
+    \caption{COATi's alignments were closer to the semi-empirical benchmark dataset than other methods.
 A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.\label{fig:dseq-benchmarks}}
@@
 %\bibliography{alignpair_letter.bib}
-\begin{thebibliography}{43}
+\begin{thebibliography}{44}
 \providecommand{\natexlab}[1]{#1}
@@
 Abascal F, Zardoya R, and Telford MJ. 2010.
-\newblock {TranslatorX}: multiple alignment of nucleotide sequences guided by
+\newblock {TranslatorX}: Multiple alignment of nucleotide sequences guided by
   amino acid translations.
@@
 Allauzen C, Riley M, Schalkwyk J, Skut W, and Mohri M. 2007.
-\newblock Openfst: A general and efficient weighted finite-state transducer
+\newblock {OpenFst}: A general and efficient weighted finite-state transducer
   library.
@@
 Arvestad L. 1997.
-\newblock Aligning coding dna in the presence of frame-shift errors.
+\newblock Aligning coding {DNA} in the presence of frame-shift errors.
 \newblock In A~Apostolico and J~Hein, editors, \emph{Combinatorial Pattern
@@
 Bininda-Emonds O. 2005.
-\newblock {transAlign}: using amino acids to facilitate the multiple alignment
+\newblock {transAlign}: Using amino acids to facilitate the multiple alignment
   of protein-coding {DNA} sequences.
@@
 Blackburne BP and Whelan S. 2011.
-\newblock {Measuring the distance between multiple sequence alignments}.
+\newblock Measuring the distance between multiple sequence alignments.
 \newblock \emph{Bioinformatics} 28:495--502.
@@
 Bradley RK and Holmes I. 2007.
-\newblock Transducers: an emerging probabilistic framework for modeling indels
+\newblock Transducers: An emerging probabilistic framework for modeling indels
   on trees.
@@
 Hein J. 1994.
-\newblock An algorithm combining dna and protein alignment.
+\newblock An algorithm combining {DNA} and protein alignment.
 \newblock \emph{Journal of Theoretical Biology} 167:169--174.
@@
 
+\bibitem[{Holmes(2020)}]{Holmes2020}
+Holmes I. 2020.
+\newblock A model of indel evolution by finite-state, continuous-time machines.
+\newblock \emph{Genetics} 216:1187--1204.
+
 \bibitem[{Holmes and Bruno(2001)}]{holmes2001evolutionary}
 Holmes I and Bruno WJ. 2001.
-\newblock Evolutionary {HMMs}: a bayesian approach to multiple alignment.
+\newblock Evolutionary {HMMs}: A {Bayesian} approach to multiple alignment.
 \newblock \emph{Bioinformatics} 17:803--820.
@@
   H, Bohlmann J, Jones SJ, et~al. 2018.
-\newblock Tigmint: correcting assembly errors using linked reads from large
+\newblock Tigmint: Correcting assembly errors using linked reads from large
   molecules.
@@
 Katoh K and Standley DM. 2013.
-\newblock {MAFFT} multiple sequence alignment software version 7: improvements
+\newblock {MAFFT} multiple sequence alignment software version 7: Improvements
   in performance and usability.
@@
 \newblock Weighted automata in text and speech processing.
+\newblock arXiv https://arxiv.org/abs/cs/0503077.
 
@@
 Pamilo P and Bianchi NO. 1993.
-\newblock {Evolution of the Zfx and Zfy genes: rates and interdependence
+\newblock {Evolution of the {Zfx} and {Zfy} genes: rates and interdependence
   between the genes.}
@@
 Paradis E and Schliep K. 2019.
-\newblock ape 5.0: an environment for modern phylogenetics and evolutionary
+\newblock {ape} 5.0: an environment for modern phylogenetics and evolutionary
   analyses in {R}.
@@
 Ranwez V, Douzery EJP, Cambon C, Chantret N, and Delsuc F. 2018.
-\newblock {{MACSE} v2: toolkit for the alignment of coding sequences accounting
+\newblock {{MACSE} v2: Toolkit for the alignment of coding sequences accounting
   for frameshifts and stop codons}.
@@
 Saitou N and Nei M. 1987.
-\newblock The neighbor-joining method: a new method for reconstructing
+\newblock The neighbor-joining method: A new method for reconstructing
   phylogenetic trees.
@@
 Silvestre-Ryan J, Wang Y, Sharma M, Lin S, Shen Y, Dider S, and Holmes I. 2021.
-\newblock {Machine Boss}: rapid prototyping of bioinformatic automata.
+\newblock {Machine Boss}: Rapid prototyping of bioinformatic automata.
 \newblock \emph{Bioinformatics} 37:29--35.

### 2024-06-10 alignpair_letter.tex
Accepted Version

@@
         Standard algorithms produce suboptimal alignments.
-        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
+        (a) shows the true alignment of an ancestral sequence (A) and a descendant sequence (D).
         (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
@@
 In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment.
-This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C.
+This allows COATi FST to assign the same weight to gaps at the beginning and the end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C.
 While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and eliding any states in between:
@@
 We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments.
-The phases of gaps were preserved as were the spacings of any clusters of gaps.
+We preserved the phases of gaps and the spacings of any clusters of gaps.
 Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair.
@@
 MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from its wrapper script.
-Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
+Additionally, MACSE produced alignments with columns that contained only gaps, which were removed by its wrapper script.
 A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.
@@
 In addition to supporting all three phases of gaps, COATi's advantage over the other aligners was at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings.
-Different data sets may show different results.
+Different datasets may show different results.
 However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}.
@@
 --- It even performed poorly on its own gap patterns.
---- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
+--- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our dataset and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
 

### 2024-06-10 alignpair_letter_source.tex
Accepted Version

@@
 MAFFT, MACSE, and Clustal\textOmega{} were able to align all sequence pairs, although the latter received some help from its wrapper script.
-Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
+Additionally, MACSE produced alignments with columns that contained only gaps, which were removed by its wrapper script.
 A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.
@@
 --- It even performed poorly on its own gap patterns.
---- Clustal\textOmega{}'s relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
+--- Clustal\textOmega{}'s relatively poor performance likely derives from its default assumptions being a poor fit for our dataset and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.
 
